```{r}
library(dunn.test)
library(tidyverse)
library(ggpubr)
plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#1331a7",
                             "Intermediate" = "steelblue","Not found" = 'darkgrey')
```


Read in the early/late table and Liu junction counts

```{r data_import}
master_table <- fread(here::here("data/master_table.csv"))
master_table <- master_table %>%
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "" ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_sh == "" ~ tdp43_sensitivity_be2,
                                tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                                tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late',
                                tdp43_sensitivity_sh == 'Intermediate'~  'Intermediate')) %>% 
    filter(!is.na(new_cate))
liu = fread("~/cluster/vyplab/alb_projects/data/liu_facs_neurons/leaf_clustered/liuFacs_perind.counts.gz")
liu = liu %>% select(-matches('unsorted')) %>% 
    melt(id.vars = c("chrom")) %>% 
    separate(value, into =c("numerator","denominator"),convert = TRUE) %>% 
    mutate(psi = numerator / denominator)

liu = liu %>% 
    separate(chrom,into = c("chr","start","end",'cluster'),sep = ":") %>% 
    mutate(strand = str_sub(cluster,-1,-1))

liu = liu  %>% 
    mutate(paste_into_igv_junction = paste0(chr,":",start,"-",end))

```



Now explore expression of 

1. Early/Late

How often is an early/late event more expressed in TDP-43 negative events than not?

```{r early_late_sig_up}


test_diff = liu %>% 
    filter(paste_into_igv_junction %in% master_table$paste_into_igv_junction) %>% 
    left_join(master_table %>% select(new_cate,paste_into_igv_junction,gene,cryptic)) %>% 
    mutate(condition = ifelse(grepl('positive',variable),"TDP-43 positive","TDP-43 negative")) %>% 
    filter(!is.na(new_cate)) %>% 
    group_by(paste_into_igv_junction) %>% 
    nest() |> 
    mutate(wicox_test_result = map(data, ~ broom::tidy(wilcox.test(psi ~ condition, data = .x))),
           mean_positive = map_dbl(data, ~ mean(.x$psi[.x$condition == "TDP-43 positive"], na.rm = TRUE)),
           mean_negative = map_dbl(data, ~ mean(.x$psi[.x$condition == "TDP-43 negative"], na.rm = TRUE)),
           ratio = log2(mean_negative / mean_positive)) |> 
    unnest() %>% 
    mutate(adj_pvalue = p.adjust(p.value))

these_ones = test_diff %>% filter(adj_pvalue < 0.1 & ratio > 0) %>% 
    pull(paste_into_igv_junction)


contingency_table <- master_table %>%
    select(paste_into_igv_junction, new_cate) %>%
    filter(!is.na(new_cate)) %>%
    mutate(sig_diff_pos_neg = paste_into_igv_junction %in% these_ones) %>%
    group_by(new_cate, sig_diff_pos_neg) %>%
    summarise(n = n(), .groups = "drop") %>%
    pivot_wider(names_from = sig_diff_pos_neg, values_from = n, values_fill = 0) %>% 
    column_to_rownames("new_cate") 

pairwise_results = rstatix::pairwise_fisher_test(contingency_table) 

comparisons <- list(
    c("Early", "Intermediate"),
    c("Early", "Late"),
    c("Intermediate", "Late")
)
    

contingency_table %>%
    rownames_to_column('new_cate') %>%
    melt() %>%
    group_by(new_cate) %>%
    mutate(prop = value / sum(value)) %>% 
    ggplot(aes(x = new_cate, y = value, fill = variable)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(
        aes(label = scales::percent(prop, accuracy = 1)),
        position = position_stack(0.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'Sig. up in TDP-43 negative') +
    scale_fill_manual(values = c("#5BC0EB", "#FDE74C")) + 
    theme(legend.position = 'bottom') +
    ggsignif::geom_signif(
        comparisons = comparisons,
        annotations = scales::pvalue(pairwise_results$p.adj, accuracy = 0.001),
        y_position = c(340, 320, 300),  # Adjusted y-positions above the bars
        textsize = 3,
        tip_length = 0
    )
```

Of those events which are, which are more highly expressed?

```{r early_late_expression}

plot_data = liu %>% 
    mutate(selective = ifelse(paste_into_igv_junction %in% these_ones,'Upregulated in TDP-43 negative vs pos.', 
                              'N.S.' )) %>% 
    filter(paste_into_igv_junction %in% master_table$paste_into_igv_junction) %>% 
    left_join(master_table %>% select(new_cate,paste_into_igv_junction,gene)) %>% 
    mutate(condition = ifelse(grepl('positive',variable),"TDP-43 positive","TDP-43 negative")) %>% 
    unique() %>% 
    filter(!is.na(new_cate)) %>% 
    mutate(psi = ifelse(is.nan(psi),0,psi))

plot_data %>% 
    ggplot(aes(x = new_cate, y = psi,fill = new_cate)) + 
    geom_boxplot() +
    theme_classic() + 
    scale_fill_manual(values = plot_sensitivity_colors) + 
    stat_compare_means(hide.ns = TRUE,
                       textsize = 3,
                       tip.length = 0) +
    theme(legend.position = "none") +
    facet_wrap(~selective+condition) + 
    ylab("PSI")

dunn_results <- dunn.test(subset(plot_data,condition == "TDP-43 negative" & selective != "N.S.")$psi,
                          subset(plot_data,condition == "TDP-43 negative" & selective != "N.S.")$new_cate,
                          kw = TRUE, label = TRUE)

plot_data %>% 
    filter(selective != "N.S.") %>% 
    ggplot(aes(x = new_cate, y = psi,fill = new_cate)) + 
    geom_boxplot() +
    theme_classic() + 
    scale_fill_manual(values = plot_sensitivity_colors) + 
    theme(legend.position = "none") +
    facet_wrap(~condition) +
    ggsignif::geom_signif(
        data = subset(plot_data,condition == "TDP-43 negative"),
        comparisons = comparisons,
        annotations = scales::pvalue(dunn_results$P.adjusted, accuracy = 0.001),
        y_position = c(1, 1.1, 1.2),  # Adjusted y-positions above the bars
        textsize = 3,
        tip_length = 0
    ) + 
    ylab("PSI")

    

```




2. NMD sensitive vs Not

Are NMD-sensitive vs not more or less likely to be significantly up in the TDP-43 negative samples? No - although we didn't expect them to be

```{r chx_up}


    
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))


test_diff_chx = liu %>% 
    filter(paste_into_igv_junction %in% big_data_chx$paste_into_igv_junction) %>% 
    left_join(big_data_chx %>% select(paste_into_igv_junction,color_gene_name,gene_name)) %>% 
    unique() %>% 
    mutate(psi = ifelse(is.nan(psi),0,psi)) %>% 
    mutate(condition = ifelse(grepl('positive',variable),"TDP-43 positive","TDP-43 negative")) %>% 
    group_by(paste_into_igv_junction) %>% 
    nest() |> 
    mutate(wicox_test_result = map(data, ~ broom::tidy(wilcox.test(psi ~ condition, data = .x))),
           mean_positive = map_dbl(data, ~ mean(.x$psi[.x$condition == "TDP-43 positive"], na.rm = TRUE)),
           mean_negative = map_dbl(data, ~ mean(.x$psi[.x$condition == "TDP-43 negative"], na.rm = TRUE)),
           ratio = log2(mean_negative / mean_positive)) |> 
    unnest() %>% 
    mutate(adj_pvalue = p.adjust(p.value))

these_ones_chx = test_diff %>% filter(adj_pvalue < 0.1 & ratio > 0) %>% 
    pull(paste_into_igv_junction)


contingency_table_chx <- big_data_chx %>%
    distinct(paste_into_igv_junction, color_gene_name) %>%
    mutate(sig_diff_pos_neg = paste_into_igv_junction %in% these_ones_chx) %>%
    group_by(color_gene_name, sig_diff_pos_neg) %>%
    summarise(n = n(), .groups = "drop") %>%
    pivot_wider(names_from = sig_diff_pos_neg, values_from = n, values_fill = 0) %>% 
    column_to_rownames("color_gene_name") 

pairwise_results_chx = rstatix::pairwise_fisher_test(contingency_table_chx) 

comparisons_nmd <- list(
    c("NMD rescued", "non-NMD rescued"))
    

contingency_table_chx %>%
    rownames_to_column('color_gene_name') %>%
    melt() %>%
    group_by(color_gene_name) %>%
    mutate(prop = value / sum(value)) %>% 
    ggplot(aes(x = color_gene_name, y = value, fill = variable)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(
        aes(label = scales::percent(prop, accuracy = 1)),
        position = position_stack(0.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'Sig. up in TDP-43 negative') +
    scale_fill_manual(values = c("#5BC0EB", "#FDE74C")) + 
    theme(legend.position = 'bottom') +
    ggsignif::geom_signif(
        comparisons = comparisons_nmd,
        annotations = scales::pvalue(pairwise_results_chx$p.adj, accuracy = 0.001),
        y_position = c(450),  # Adjusted y-positions above the bars
        textsize = 3,
        tip_length = 0
    )

```



Are NMD-sensitive vs not more expressed? 

```{r chx_expression,echo=FALSE}

plot_data_chx = liu %>% 
    filter(paste_into_igv_junction %in% big_data_chx$paste_into_igv_junction) %>% 
    mutate(selective = ifelse(paste_into_igv_junction %in% these_ones_chx,'Upregulated in TDP-43 negative vs pos.', 
                          'N.S.' )) %>% 
    left_join(big_data_chx %>% select(paste_into_igv_junction,color_gene_name)) %>% 
    unique() %>% 
    mutate(psi = ifelse(is.nan(psi),0,psi)) %>% 
    mutate(condition = ifelse(grepl('positive',variable),"TDP-43 positive","TDP-43 negative"))

plot_data_chx %>% 
    ggplot(aes(x = color_gene_name, y = psi,fill = color_gene_name)) + 
    geom_boxplot() +
    theme_classic() + 
    scale_fill_manual(values = c("#3288BD","#D53E4F")) + 
    stat_compare_means(hide.ns = TRUE,
                       textsize = 3,
                       tip.length = 0) +
    theme(legend.position = "none") +
    facet_wrap(~selective+condition) + 
    ylab("PSI") +
    xlab(element_blank())

```

