---
title: "TDP-43 cryptic exon biology main figures"
author: "MZ"
date: "2024-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
###load packages/libraries/scripts
library(tidyverse)
library(ggrepel)
library(patchwork)
library(ggalluvial)

source(here::here('trial_nmd_plots.R'))

#check you're in right folder
here::here()
```

## Figure 1 - Dose-response curves in SH-SY5Y and SK-N-BE(2) 
```{r, echo = FALSE, warning=FALSE, message=FALSE}
#1A and 1B
levels_sy5y_qpcr <- read.csv(here::here('data',"unc13a_b_tdp43_curves_nature_paper.csv"))
levels_dz_qpcr <- read.csv(here::here('data','qpcr_dz.csv'))
#1C and 1D
diff_expr_sh <- read.csv(here::here('data', 'diff_expr_sh.csv'))
diff_expr_sk <- read.csv(here::here('data', 'diff_expr_sk.csv'))
#1E, 1F, and 1G
master_table <- read.csv(here::here("data/master_table.csv"))

#####
print("1A and 1B")
a <- levels_sy5y_qpcr %>%
  ggplot(aes(x = as.character(Treatment), y = Value)) +
  geom_point() +
  geom_boxplot() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  theme_classic() +
  theme(text = element_text(size = 10))
c <- levels_dz_qpcr %>%
  filter(experiment == "new_new") %>%
  ggplot(aes(x = as.character(sample), y = 100*value)) +
  geom_point() +
  geom_boxplot() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  #facet_wrap(facets = vars(experiment)) +
  theme_classic() +
  theme(text = element_text(size = 10))
a+c

print("1C and 1D")
plot_de <- diff_expr_sh %>%
    filter(symbol != "RNA5-8SN2") %>% 
  mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
  mutate(alpha_gene_name = ifelse(padj < 0.05, "sig", "non_sig")) %>%
  mutate(gene_label = case_when(symbol %in% c("UNC13A", "STMN2") & alpha_gene_name == "sig" ~ symbol, T ~ ""))
plot_de_meta <- as.data.frame(table(plot_de$color_gene_name, plot_de$alpha_gene_name, plot_de$source))
e <- plot_de %>%
  ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
  geom_point(aes(color = color_gene_name, alpha = alpha_gene_name),
             size = 0.2,
             position = position_jitter(width = 0, height = 0.4),
             show.legend = F) +
  geom_text(data = plot_de_meta[plot_de_meta$Var2 == "sig",], 
            aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
            size = 5,
            show.legend = F) +
  geom_vline(xintercept = 0) +
  geom_text_repel(aes(label = gene_label), max.overlaps = Inf) +
  coord_flip() +
  scale_color_brewer(palette = "Set1") +
  scale_alpha_manual(values = c(0.2, 1)) +
  xlim(c(-10,10)) +
  xlab('Gene expression Log2 FC compared to controls') +
  ylab('TDP-43 KD level') +
  scale_y_discrete(labels = c("+", "++", "+++", "++++", "+++++")) +
  theme_classic() +
  theme(text = element_text(size = 10))
plot_de_dz <- diff_expr_sk %>%
    filter(symbol != "RNA5-8SN2") %>% 
  mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
  mutate(alpha_gene_name = ifelse(padj < 0.05, "sig", "non_sig")) %>%
  mutate(gene_label = case_when(symbol %in% c("UNC13A", "STMN2") & alpha_gene_name == "sig" ~ symbol, T ~ ""))
plot_de_dz_meta <- as.data.frame(table(plot_de_dz$color_gene_name, plot_de_dz$alpha_gene_name, plot_de_dz$source))
f <- plot_de_dz %>%
  ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
  geom_point(aes(color = color_gene_name, alpha = alpha_gene_name),
             size = 0.2,
             position = position_jitter(width = 0, height = 0.4),
             show.legend = F) +
  geom_text(data = plot_de_dz_meta[plot_de_dz_meta$Var2 == "sig",], 
            aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
            size = 5,
            show.legend = F) +
  geom_vline(xintercept = 0) +
  geom_text_repel(aes(label = gene_label), max.overlaps = Inf) +
  coord_flip() +
  scale_color_brewer(palette = "Set1") +
  scale_alpha_manual(values = c(0.2, 1)) +
  xlim(c(-10,10)) +
  xlab('Log2 Fold Change compared to untreated') +
  ylab('Doxycycline concentration (ug/mL)') +
  scale_y_discrete(labels = c("+", "++", "+++", "++++", "+++++", "++++++", "+++++++")) +
  theme_classic() +
  theme(text = element_text(size = 10)) 
e + f

print("1E and 1F")
list_gene = c("STMN2", "UNC13A", "HDGFL2", "AARS1")
big_data_filtereds_early_late_sh <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_sh) %>% 
    select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_sh,max_psi_sh)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox025)) %>% 
    distinct(gene, .keep_all = T) %>% 
    filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:12)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox0125", "dox0187", "dox021", "dox025", "dox075"))) %>%
  mutate(label_junction = case_when(name == "dox025" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_sh$tdp43_sensitivity_sh)
table(big_data_filtereds_early_late_sh$max_psi_sh)
GG <- big_data_filtereds_early_late_sh %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_sh), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) + #, "#984EA3", "#000000")) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = c("-", "+", "++", "+++", "++++", "+++++")) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_sh) + 
  theme_classic()
big_data_filtereds_early_late_be2 <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_be2) %>% 
    select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox002,dox003,dox004,dox005,dox0075,dox01,dox1,
             #dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_be2,max_psi_be2)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox01)) %>% 
    distinct(gene, .keep_all = T) %>% 
  filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:14)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox002","dox003","dox004","dox005","dox0075","dox01","dox1"))) %>%
  mutate(label_junction = case_when(name == "dox01" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_be2$tdp43_sensitivity_be2)
table(big_data_filtereds_early_late_be2$max_psi_be2)
HH <- big_data_filtereds_early_late_be2 %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_be2), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) + #, "#984EA3", "#000000")) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = c("-", "+", "++", "+++", "++++", "+++++", "++++++", "+++++++")) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_be2) + 
  theme_classic()
GG + HH

print("1G")  
summary <- master_table %>%
  filter(cryptic == "yes" & 
             tdp43_sensitivity_sh != "" & tdp43_sensitivity_be2 != "") %>% 
  group_by(tdp43_sensitivity_sh, tdp43_sensitivity_be2) %>%
  summarise(count = n())%>%
  ungroup()
ggplot(summary,
       aes(y = count, axis1 = tdp43_sensitivity_sh, axis2 = tdp43_sensitivity_be2)) +
  geom_alluvium(aes(fill = tdp43_sensitivity_sh), width = 1/12, knot.pos = 0) +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  guides(fill = "none") +
  geom_stratum(alpha = .25, width = 1/8) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  theme_void()
```

A. Dose-response curves for SH-SY5Y
B. Dose-response curves for SK-N-BE(2)
C. Differential expression in SH-SY5Y
D. Differential expression in SK-N-BE(2)
E. Spaghetti plot in SH-SY5Y
F. Spaghetti plot in SK-N-BE(2)
G. alluvial plot

## Figure 2 - Features
```{r}

```

A. Conservation score for CE in SH-SY5Y
B. Conservation score for non-CE in SH-SY5Y
C. Kmers enrichment analysis in SH-SY5Y
D. Density UG motifs by early-late in SH-SY5Y
E. Density UG motifs by top-bottom in SH-SY5Y
F. Distance iCLIP TDP43 binding site SH-SY5Y
G. RNA maps based on figure F

## Figure 3 - NMD
```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("3A")

print("3B")
gene_list_slope <- c("STMN2", "UNC13A", "AARS1", "HDGFL2", "CYFIP2", "SYNE1")
input_list <- c("Control_Control",
                "Cycloheximide_Control", 
                "Control_TDP43KD", 
                "Cycloheximide_TDP43KD")
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))
a <- slope_plot_nmd(big_data_chx, 
               gene_list_slope) #### NMD rescued and unaffected

print("3C")

```

A. 
B.
C.
D.

## Figure 3 - compendium / in vivo
```{r, echo=FALSE, warning=FALSE, message=FALSE}
##load targeted seq table and compute CE ratio / rename samples in agreement to UNC13A paper
maxc <- read.csv("~/Desktop/ce_bio/count_table_of_aligned_remove_200bp_plus_product_reads.csv") %>%
  mutate(fraction_cryptic = n_cryptic / n_counts_for_gene) %>%
  mutate(sample_name_ari = case_when(sample_name == "P64_11" ~ "HC1",
                                     sample_name == "P17_07" ~ "HC2",
                                     sample_name == "P47_11" ~ "HC3",
                                     sample_name == "P35_07" ~ "HC4",
                                     sample_name == "P56_13" ~ "FTD1",
                                     sample_name == "P45_15" ~ "FTD2",
                                     sample_name == "P63_05" ~ "FTD3",
                                     sample_name == "P28_07" ~ "FTD4",
                                     sample_name == "P40_04" ~ "FTD5",
                                     sample_name == "P07_15" ~ "FTD6",
                                     sample_name == "P16_09" ~ "FTD7",
                                     sample_name == "P86_08" ~ "FTD8",
                                     sample_name == "P13_13" ~ "FTD9",
                                     sample_name == "P11_07" ~ "FTE10",
                                     sample_name == "ddH2O" ~ "HO2O")) #%>%
  #pivot_wider(names_from = sample_name_ari, values_from = ratio_cryptic)
table(maxc$sample_name_ari)

##use only TDP43path-specific genes (use list from AL) - this is exluding more than 2/3 of genes!!
al <- read.csv("~/Desktop/ce_bio/expression_by_pathology_updated.csv") %>%
  dplyr::filter(fraction_path > 0.01 & fraction_not_path < 0.005) %>% ####need to carefully check this with AL
  distinct(gene, .keep_all = T) ##check when binding with annotated validation table - should be ok since left_join
names(al)[4] <- "gene_name"
maxc_al <- maxc %>%
  left_join(al, by = "gene_name") %>%
  dplyr::filter(!is.na(paste_into_igv_junction))
table(maxc_al$sample_name_ari)
table(maxc_al$gene_name)

##check igv files for these genes
maxc_type <- maxc_al %>%
  select(-type) %>%
  mutate(gene = gene_name) %>%
  filter(gene_name != "SETD5") %>%
  left_join(master_table, by = "paste_into_igv_junction") %>%
  filter(!is.na(tdp43_sensitivity_sh)) %>% 
  mutate(group = ifelse(grepl("FT", sample_name_ari), "FTD",
                        ifelse(grepl("HC", sample_name_ari), "HC",
                               "H2O"))) %>% 
  group_by(gene_name) %>%
  filter(sum(n_cryptic) > 0) %>%
  ungroup() %>% 
    filter(group != "H2O")

maxc_type %>%
  filter(tdp43_sensitivity_sh != "") %>%
  filter(gene_name != "SETD5") %>%
  #filter(detected == "Yes") %>%
  filter(group != "H2O") %>%
  ggplot(aes(x = tdp43_sensitivity_sh)) +
  geom_bar(aes(fill = tdp43_sensitivity_sh)) +
  facet_wrap(facets = vars(group), ncol = 2) +
  scale_fill_brewer(palette = "Set1") +
  xlab("") +
  ylab("Count of cryptic exons detected by targeted panel") +
  labs(fill = "Category") +
  theme_bw()

maxc_earli <- maxc_type %>%
  filter(tdp43_sensitivity_sh == "Earli") %>%
  filter(gene_name != "SETD5") %>% 
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") + #position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Earli") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_early <- maxc_type %>%
  filter(tdp43_sensitivity_sh == "Early") %>%
  filter(gene_name != "SETD5") %>% 
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") +#position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Early") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_intermediate <- maxc_type %>%
  filter(tdp43_sensitivity_sh == "Intermediate") %>%
  filter(gene_name != "SETD5") %>%
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") +#position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Greens") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Intermediate") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_late <- maxc_type %>%
  filter(tdp43_sensitivity_sh == "Late") %>%
  filter(gene_name != "SETD5") %>%
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") +#position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Late") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_early / maxc_intermediate / maxc_late #/ maxc_na

maxc_type %>%
  filter(group != "H2O") %>%
    filter(tdp43_sensitivity_sh != "") %>% 
  ggplot(aes(x = sample_name_ari, y = gene_name, fill = fraction_cryptic)) +
  geom_tile() +
  #scale_fill_viridis_d() +
  scale_fill_gradient(low = "white", high = "black") +
  facet_grid(rows = vars(tdp43_sensitivity_sh), cols = vars(group), 
             scales = "free", space = "free") +
  ylab("") + xlab("Sample name") + labs(fill = "CE fraction") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))
```
A.
B.
C. Detection of CEs from in vivo samples
D.
