---
title: "TDP-43 cryptic exon biology main figures"
author: "MZ"
date: "2024-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
###load packages/libraries/scripts
library(tidyverse)
library(ggrepel)
library(patchwork)
library(ggalluvial)
library(ggsignif)
library(ggseqlogo)
library(Biostrings)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ComplexHeatmap)
library(RColorBrewer)

source(here::here('trial_nmd_plots.R'))

#check you're in right folder
here::here()

#plot parameters
plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#1331a7", "Intermediate" = "steelblue")
plot_psi_colors <- c("Top" = "#388f78", "Bottom" = "#de9e4a", "Intermediate" = "steelblue")
plot_text_size <- 10
```

## Figure 1 - Dose-response curves in SH-SY5Y and SK-N-BE(2) 
```{r, echo = FALSE, warning=FALSE, message=FALSE}
#1A and 1B
levels_sy5y_qpcr <- read.csv(here::here('data',"unc13a_b_tdp43_curves_nature_paper.csv"))
levels_dz_qpcr <- read.csv(here::here('data','qpcr_dz.csv'))
#1C and 1D
diff_expr_sh <- read.csv(here::here('data', 'diff_expr_sh.csv'))
diff_expr_sk <- read.csv(here::here('data', 'diff_expr_sk.csv'))
#1E, 1F, and 1G
master_table <- read.csv(here::here("data/master_table.csv"))

#####
print("1A and 1B")
a <- levels_sy5y_qpcr %>%
  ggplot(aes(x = as.character(Treatment), y = Value)) +
  geom_point() +
  geom_boxplot() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
c <- levels_dz_qpcr %>%
  filter(experiment == "new_new") %>%
  ggplot(aes(x = as.character(sample), y = 100*value)) +
  geom_point() +
  geom_boxplot() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  #facet_wrap(facets = vars(experiment)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
a+c

print("1C and 1D")
plot_de <- diff_expr_sh %>%
    filter(symbol != "RNA5-8SN2") %>% 
  mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
  mutate(alpha_gene_name = ifelse(padj < 0.05, "sig", "non_sig")) %>%
  mutate(gene_label = case_when(symbol %in% c("UNC13A", "STMN2") & alpha_gene_name == "sig" ~ symbol, T ~ ""))
plot_de_meta <- as.data.frame(table(plot_de$color_gene_name, plot_de$alpha_gene_name, plot_de$source))
e <- plot_de %>%
  ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
  geom_point(aes(color = color_gene_name, alpha = alpha_gene_name),
             size = 0.2,
             position = position_jitter(width = 0, height = 0.4),
             show.legend = F) +
  geom_text(data = plot_de_meta[plot_de_meta$Var2 == "sig",], 
            aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
            size = 5,
            show.legend = F) +
  geom_vline(xintercept = 0) +
  geom_text_repel(aes(label = gene_label), max.overlaps = Inf) +
  coord_flip() +
  scale_color_brewer(palette = "Set1") +
  scale_alpha_manual(values = c(0.2, 1)) +
  xlim(c(-10,10)) +
  xlab('Gene expression Log2 FC compared to controls') +
  ylab('TDP-43 KD level') +
  scale_y_discrete(labels = c("+", "++", "+++", "++++", "+++++")) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
plot_de_dz <- diff_expr_sk %>%
    filter(symbol != "RNA5-8SN2") %>% 
  mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
  mutate(alpha_gene_name = ifelse(padj < 0.05, "sig", "non_sig")) %>%
  mutate(gene_label = case_when(symbol %in% c("UNC13A", "STMN2") & alpha_gene_name == "sig" ~ symbol, T ~ ""))
plot_de_dz_meta <- as.data.frame(table(plot_de_dz$color_gene_name, plot_de_dz$alpha_gene_name, plot_de_dz$source))
f <- plot_de_dz %>%
  ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
  geom_point(aes(color = color_gene_name, alpha = alpha_gene_name),
             size = 0.2,
             position = position_jitter(width = 0, height = 0.4),
             show.legend = F) +
  geom_text(data = plot_de_dz_meta[plot_de_dz_meta$Var2 == "sig",], 
            aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
            size = 5,
            show.legend = F) +
  geom_vline(xintercept = 0) +
  geom_text_repel(aes(label = gene_label), max.overlaps = Inf) +
  coord_flip() +
  scale_color_brewer(palette = "Set1") +
  scale_alpha_manual(values = c(0.2, 1)) +
  xlim(c(-10,10)) +
  xlab('Log2 Fold Change compared to untreated') +
  ylab('Doxycycline concentration (ug/mL)') +
  scale_y_discrete(labels = c("+", "++", "+++", "++++", "+++++", "++++++", "+++++++")) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size)) 
e + f

print("1E and 1F")
list_gene = c("STMN2", "UNC13A", "HDGFL2", "AARS1")
big_data_filtereds_early_late_sh <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_sh) %>% 
    select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_sh,max_psi_sh)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox025)) %>% 
    distinct(gene, .keep_all = T) %>% 
    filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:12)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox0125", "dox0187", "dox021", "dox025", "dox075"))) %>%
  mutate(label_junction = case_when(name == "dox025" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_sh$tdp43_sensitivity_sh)
table(big_data_filtereds_early_late_sh$max_psi_sh)
GG <- big_data_filtereds_early_late_sh %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_sh), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = plot_sensitivity_colors) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = c("-", "+", "++", "+++", "++++", "+++++")) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_sh) + 
  theme_classic()
big_data_filtereds_early_late_be2 <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_be2) %>% 
    select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox002,dox003,dox004,dox005,dox0075,dox01,dox1,
             #dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_be2,max_psi_be2)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox01)) %>% 
    distinct(gene, .keep_all = T) %>% 
  filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:14)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox002","dox003","dox004","dox005","dox0075","dox01","dox1"))) %>%
  mutate(label_junction = case_when(name == "dox01" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_be2$tdp43_sensitivity_be2)
table(big_data_filtereds_early_late_be2$max_psi_be2)
HH <- big_data_filtereds_early_late_be2 %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_be2), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = plot_sensitivity_colors) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = c("-", "+", "++", "+++", "++++", "+++++", "++++++", "+++++++")) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_be2) + 
  theme_classic()
GG + HH

print("1G")  
summary <- master_table %>%
  filter(cryptic == "yes" & 
             tdp43_sensitivity_sh != "" & tdp43_sensitivity_be2 != "") %>% 
  group_by(tdp43_sensitivity_sh, tdp43_sensitivity_be2) %>%
  summarise(count = n())%>%
  ungroup()
ggplot(summary,
       aes(y = count, axis1 = tdp43_sensitivity_sh, axis2 = tdp43_sensitivity_be2)) +
  geom_alluvium(aes(fill = tdp43_sensitivity_sh), width = 1/12, knot.pos = 0) +
  guides(fill = "none") +
  geom_stratum(alpha = .5, width = 1/8) +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  theme_void()
```

A. Dose-response curves for SH-SY5Y
B. Dose-response curves for SK-N-BE(2)
C. Differential expression in SH-SY5Y
D. Differential expression in SK-N-BE(2)
E. Spaghetti plot in SH-SY5Y
F. Spaghetti plot in SK-N-BE(2)
G. alluvial plot

## Figure 2 - Features
```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("2A and 2B")

get_signif_stars <- function(p_value) {
  if (p_value < 0.001) {
    return("***")
  } else if (p_value < 0.01) {
    return("**")
  } else if (p_value < 0.05) {
    return("*")
  } else {
    return("NS")
  }
}
  
get_significance <- function(df, value, group, category_col, category) {
  test_formula <- as.formula(paste(value, "~", group))
  p_test <- kruskal.test(test_formula, data = df %>% filter(!!sym(category_col) == category))
  #p1_annotation <- paste("p =", format(p1_test$p.value, digits = 3), get_signif_stars(p1_test$p.value))
  return(get_signif_stars(p_test$p.value))
}

cons_score_df <- master_table %>% 
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>% 
  subset(!is.na(phastCons_alt) & !is.na(phastCons_unchanged)) %>%
  pivot_longer(cols = c(phastCons_alt, phastCons_unchanged), names_to = "category", values_to = "cons_score") 

cons_score_df_cryptic <- cons_score_df %>% filter(cryptic == "yes") 
cons_score_df_not_cryptic <- cons_score_df %>% filter(cryptic == "no") 
a <- cons_score_df_cryptic %>% 
  ggplot(aes(x = category, y = cons_score, fill = tdp43_sensitivity_sh)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Novel splice site", "Unchanged splice site")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_cryptic,
                                    "cons_score", "tdp43_sensitivity_sh", "category", "phastCons_alt"), 
                   get_significance(cons_score_df_cryptic,
                                    "cons_score", "tdp43_sensitivity_sh", "category", "phastCons_unchanged")) 
  ) +
  #facet_wrap(facets = vars(cryptic)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

b <- cons_score_df_not_cryptic %>% 
  ggplot(aes(x = category, y = cons_score, fill = tdp43_sensitivity_sh)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Alternatively used splice site", "Unchanged splice site")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "tdp43_sensitivity_sh", "category", "phastCons_alt"), 
                   get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "tdp43_sensitivity_sh", "category", "phastCons_unchanged"))
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
a + b
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#C. Kmers enrichment analysis in SH-SY5Y
print("2C")

kmer_size <- 6
yg_motifs <- c('GTGTGT', 'TGTGTG', 'CGTGTG', 'TGCGTG', 'TGTGCG', 'GTGTGC')
ya_motifs <- c('ATGTGT', 'GTATGT', 'GTGTAT', 'TGTGTA', 'TGTATG', 'TGCATG')
aa_motifs <- c('GTGTGA', 'AATGAA', 'GAATGA', 'TGAATG', 'ATGAAT', 'GTGAAT', 'GAATGT', 'TTGAAT')
all_motifs <- c(yg_motifs, ya_motifs, aa_motifs)

count_hexamers <- function(row) {
  if (row[["junc_cat"]] == "exon_skip" || is.na(row[["paste_into_igv_junction_annotated"]])) {
    return(NULL)
  }
  
  coords <- strsplit(row[["paste_into_igv_junction_annotated"]], "[:-]")[[1]]
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = coords[1], start = as.numeric(coords[2]), end = as.numeric(coords[3]))
  if (row[["strand"]] == "-") {
    seq <- reverseComplement(seq)
  }
  counts <- sapply(all_motifs, function(motif) countPattern(motif, seq, max.mismatch = 0))
  names(counts) <- all_motifs
  
  return(counts)
}
master_table_filt <- master_table %>% 
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>% filter(junc_cat != "exon_skip")
kmer_table <- apply(master_table_filt, 1, count_hexamers)
all_counts_df <- as.data.frame(t(kmer_table))

all_counts_df$type <- master_table_filt$tdp43_sensitivity_sh
all_counts_df$cryptic <- master_table_filt$cryptic

summary_counts <- all_counts_df %>%
  group_by(type, cryptic) %>%
  summarise(across(all_motifs, sum))

means <- apply(summary_counts[, -c(1,2)], 1, mean)
sds <- apply(summary_counts[, -c(1,2)], 1, sd)
z_scores <- sweep(summary_counts[, -c(1,2)], MARGIN = 1, STATS = means, FUN = "-") / sds
heatmap_data <- as.matrix(z_scores)

sort_kmers_by_zscore <- function(kmers, z_scores) {
  kmers[order(z_scores[kmers], decreasing = TRUE)]
}

sorted_yg <- sort_kmers_by_zscore(yg_motifs, heatmap_data[1,])
sorted_ya <- sort_kmers_by_zscore(ya_motifs, heatmap_data[1,])
sorted_aa <- sort_kmers_by_zscore(aa_motifs, heatmap_data[1,])

sorted_all_kmers <- c(sorted_yg, sorted_ya, sorted_aa)
categories <- c(rep("YG", 6), rep("YA", 6), rep("AA", 8))

annotation <- HeatmapAnnotation(Motif = factor(categories, levels = c("YG", "YA", "AA")),
                                show_legend = c(F, F, F),
                                col = list(Motif = c(YG = "purple", YA = "orange", AA = "darkgreen")),
                                which = "col")

heatmap_data <- as.matrix(z_scores)
rownames(heatmap_data) <- paste(summary_counts$type, ifelse(summary_counts$cryptic == "yes", "Cryptic", "Not cryptic"))
c <- Heatmap(heatmap_data[c("Early Cryptic","Early Not cryptic","Late Cryptic","Late Not cryptic") , sorted_all_kmers],
        cluster_columns = F,
        cluster_rows = F,
        show_row_names = T,
        show_column_names = T,
        show_row_dend = F,
        bottom_annotation = annotation,
        col = colorRampPalette(brewer.pal(3, "YlGnBu"))(100),
        show_heatmap_legend = FALSE)
c
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#D. Density UG motifs by early-late in SH-SY5Y
print("2D and 2E")
sequences <- sapply(seq_along(master_table$paste_into_igv_junction_annotated), function(i) {
  coord <- master_table$paste_into_igv_junction_annotated[i]
  chr <- sub("^(.*):.*$", "\\1", coord)
  range <- as.numeric(unlist(strsplit(sub(".*:(.*)$", "\\1", coord), "-")))
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = chr, start = range[1], end = range[2])

  if (master_table$strand[i] == "-") {
    seq <- reverseComplement(seq)
  }
  
  return(seq)
})

master_table$longest_ug_region <- Biostrings::vcountPattern('TGTGTG', DNAStringSet(sequences), max.mismatch = 1)

master_table <- master_table %>% 
  mutate(
    start = as.numeric(sub(".*:(\\d+)-.*", "\\1", paste_into_igv_junction_annotated)),
    end = as.numeric(sub(".*-(\\d+)", "\\1", paste_into_igv_junction_annotated)),
    intron_len = end - start,
    ug_density = longest_ug_region / intron_len
  ) %>%
  select(-start, -end)

d <- master_table %>% 
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>% 
  ggplot(aes(x = cryptic, y = ug_density, fill = tdp43_sensitivity_sh)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(tdp43_sensitivity_sh %in% c("Early", "Late")),
                                    "ug_density", "tdp43_sensitivity_sh", "cryptic", "no"), 
                   get_significance(master_table %>% filter(tdp43_sensitivity_sh %in% c("Early", "Late")),
                                    "ug_density", "tdp43_sensitivity_sh", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

#E. Density UG motifs by top-bottom in SH-SY5Y
master_table$max_psi_sh <- factor(master_table$max_psi_sh, levels=c("Top", "Bottom"))
e <- master_table %>% 
  filter(max_psi_sh %in% c("Top", "Bottom")) %>%
  ggplot(aes(x = cryptic, y = ug_density, fill = max_psi_sh)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) + 
  scale_fill_manual(values = plot_psi_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(max_psi_sh %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_sh", "cryptic", "no"), 
                   get_significance(master_table %>% filter(max_psi_sh %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_sh", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
d + e

#F. Distance iCLIP TDP43 binding site SH-SY5Y
print("2F")
df_iclip <- read.csv(here::here('data', 'all_events.closest_tdp_bind.bed'), sep="\t")
df_iclip$tdp_len <- df_iclip$end - df_iclip$start + 1

f <- df_iclip %>% 
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>%
  filter(abs(dist) <= 1000) %>%
  ggplot(aes(x = dist, fill = tdp43_sensitivity_sh)) +
  geom_density(alpha = 0.5, adjust = 1) +  
  facet_wrap(~cryptic, 
             ncol = 2,
             labeller = labeller(cryptic=c("yes" = "Cryptic", "no" = "Not cryptic"))) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  labs(x = 'Distance to the nearest TDP-binding site', y = 'Density') +  
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

f

#G. RNA maps based on figure F
early_donors <- readDNAStringSet("data/maxent/donors_early_sh.fa")
late_donors <- readDNAStringSet("data/maxent/donors_late_sh.fa")

g1 <- ggplot() + 
  geom_logo(data = as.character(early_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g2 <- ggplot() + 
  geom_logo(data = as.character(late_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g <- g1 / g2
g
```

A. Conservation score for CE in SH-SY5Y
B. Conservation score for non-CE in SH-SY5Y
C. Kmers enrichment analysis in SH-SY5Y
D. Density UG motifs by early-late in SH-SY5Y
E. Density UG motifs by top-bottom in SH-SY5Y
F. Distance iCLIP TDP43 binding site SH-SY5Y
G. RNA maps based on figure F

## Figure 3 - NMD
```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("3A")

print("3B")


print("3C")
gene_list_slope <- c("STMN2", "UNC13A", "AARS1", "HDGFL2", "CYFIP2", "SYNE1")
input_list <- c("Control_Control",
                "Cycloheximide_Control", 
                "Control_TDP43KD", 
                "Cycloheximide_TDP43KD")
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))
a <- slope_plot_nmd(big_data_chx, 
               gene_list_slope) #### NMD rescued and unaffected

print("3D")

```

A. Categories - Overlap splicing changes?
C. Slope plot in SH-SY5Y
D. Differential expression of NMD sensitive and not
E. Differential expression of NMD sensitive/not but more detailed
F. TDP-43 dependent loss of dark cryptics
G. 4-SU

## Figure 4 - compendium / in vivo


A.
B.
C. Detection of CEs from in vivo samples
D.
