---
title: "TDP-43 cryptic exon biology main figures"
author: "MZ"
date: "2024-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
###load packages/libraries/scripts
#BiocManager::install("ComplexHeatmap")
library(data.table)
library(tidyverse)
library(ggrepel)
library(patchwork)
library(ggalluvial)
library(ggsignif)
library(ggseqlogo)
library(Biostrings)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ComplexHeatmap)
library(RColorBrewer)
library(ggstats)

source(here::here('trial_nmd_plots.R'))

#check you're in right folder
here::here()

#plot parameters
plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#1331a7", "Intermediate" = "steelblue")
plot_psi_colors <- c("Top" = "#388f78", "Bottom" = "#de9e4a", "Intermediate" = "steelblue")
plot_text_size <- 10
```

## Figure 1 - Dose-response curves in SH-SY5Y and SK-N-BE(2) 
```{r, echo = FALSE, warning=FALSE, message=FALSE}
#1A and 1B
levels_sy5y_qpcr <- read.csv(here::here('data',"unc13a_b_tdp43_curves_nature_paper.csv"))
levels_dz_qpcr <- read.csv(here::here('data','qpcr_dz.csv'))
#1C and 1D
diff_expr_sh <- read.csv(here::here('data', 'diff_expr_sh.csv'))
diff_expr_sk <- read.csv(here::here('data', 'diff_expr_sk.csv'))
#1E, 1F, and 1G
master_table <- read.csv(here::here("data/master_table.csv"))

#####
print("1A and 1B - Dose-response curves for SH-SY5Y and SK-N-BE(2)")
a <- levels_sy5y_qpcr %>%
  ggplot(aes(x = as.character(Treatment), y = Value)) +
  geom_point() +
  geom_boxplot() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

sy5y_qpcr = levels_sy5y_qpcr %>%
    group_by(Treatment) %>% 
    summarize(avg_f = mean(Value)) %>% 
    ungroup() %>% 
    dplyr::slice(-1) %>% pull(avg_f)

c <- levels_dz_qpcr %>%
  filter(experiment == "new_new") %>%
  ggplot(aes(x = as.character(sample), y = 100*value)) +
  geom_point() +
  geom_boxplot() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  #facet_wrap(facets = vars(experiment)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

be2_qpcr = levels_dz_qpcr %>%
    filter(experiment == "new_new") %>% 
    group_by(sample) %>% 
    summarize(avg_f = mean(value) *100) %>% 
    ungroup() %>% 
    dplyr::slice(-1) %>% pull(avg_f)
a+c
```




```{r de_by_levels}
print("1C and 1D - Differential expression in SH-SY5Y and SK-N-BE(2)")

fold_cut = 0.5

plot_de <- diff_expr_sh %>%
    filter(symbol != "RNA5-8SN2") %>% 
    mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
    mutate(alpha_gene_name = ifelse(padj < 0.05 & abs(log2FoldChange) >= fold_cut, "sig", "non_sig")) %>%
    mutate(gene_label = case_when(symbol %in% c("UNC13A", "STMN2") & alpha_gene_name == "sig" ~ symbol, T ~ ""))

plot_de_meta <- as.data.frame(table(plot_de$color_gene_name, plot_de$alpha_gene_name, plot_de$source))


e <- plot_de %>%
    ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
    geom_point(aes(color = color_gene_name, 
                   alpha = alpha_gene_name, 
                   shape = alpha_gene_name, 
                   fill = alpha_gene_name),
               size = 0.2,
               position = position_jitter(width = 0, height = 0.4),
               show.legend = F) +
    geom_text(data = plot_de_meta[plot_de_meta$Var2 == "sig",], 
              aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
              size = 5,
              show.legend = F) +
    geom_vline(xintercept = 0) +
    geom_text_repel(aes(label = gene_label), max.overlaps = Inf) +
    coord_flip() +
    scale_color_brewer(palette = "Set1") +
    scale_alpha_manual(values = c(0.2, 1)) +
    scale_shape_manual(values = c(21, 21)) +  # Use shape 21 for both
    scale_fill_manual(values = c("grey", "black")) +  # Grey for non-sig, black for sig
    xlim(c(-10, 10)) +
    xlab('Gene expression Log2 FC compared to controls') +
    ylab('TDP-43 KD level') +
    scale_y_discrete(labels = paste0(c("+", "++", "+++", "++++", "+++++"),'\n',round(sy5y_qpcr,2),rep(" %",5))) +
    theme_classic() +
    theme(text = element_text(size = plot_text_size)) +
    ggtitle("SH-SY5Y")

plot_de_dz <- diff_expr_sk %>%
    filter(symbol != "RNA5-8SN2") %>% 
    mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
    mutate(alpha_gene_name = ifelse(padj < 0.05 & abs(log2FoldChange) >= fold_cut, "sig", "non_sig")) %>%
  mutate(gene_label = case_when(symbol %in% c("UNC13A", "STMN2") & alpha_gene_name == "sig" ~ symbol, T ~ "")) 
plot_de_dz_meta <- as.data.frame(table(plot_de_dz$color_gene_name, plot_de_dz$alpha_gene_name, plot_de_dz$source))

f <- plot_de_dz %>%
    ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
    geom_point(aes(color = color_gene_name, 
                   alpha = alpha_gene_name, 
                   shape = alpha_gene_name, 
                   fill = alpha_gene_name),
               size = 0.2,
               position = position_jitter(width = 0, height = 0.4),
               show.legend = F) +
    geom_text(data = plot_de_dz_meta[plot_de_dz_meta$Var2 == "sig",], 
              aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
              size = 5,
              show.legend = F) +
    geom_vline(xintercept = 0) +
    geom_text_repel(aes(label = gene_label), max.overlaps = Inf) +
    coord_flip() +
    scale_color_brewer(palette = "Set1") +
    scale_alpha_manual(values = c(0.2, 1)) +
    scale_shape_manual(values = c(21, 21)) +  # Use shape 21 for both
    scale_fill_manual(values = c("grey", "black")) +  # Grey for non-sig, black for sig
    xlim(c(-10, 10)) +
    xlab('Gene expression Log2 FC compared to controls') +
    ylab('TDP-43 KD level') +
    scale_y_discrete(labels = paste0(c("+", "++", "+++", "++++", "+++++","++++++","+++++++"),'\n',round(be2_qpcr,2),rep(" %",7))) +
    theme_classic() +
    theme(text = element_text(size = plot_text_size)) +
    ggtitle("SK-N-BE(2)")

e + f
```


```{r spaghetti}
print("1E and 1F - Spaghetti plot in SH-SY5Y and SK-N-BE(2)")
list_gene = c("STMN2", "UNC13A", "HDGFL2", "AARS1")
big_data_filtereds_early_late_sh <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_sh) %>% 
    select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_sh,max_psi_sh)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox025)) %>% 
    distinct(gene, .keep_all = T) %>% 
    filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:12)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox0125", "dox0187", "dox021", "dox025", "dox075"))) %>%
  mutate(label_junction = case_when(name == "dox025" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_sh$tdp43_sensitivity_sh)
table(big_data_filtereds_early_late_sh$max_psi_sh)
GG <- big_data_filtereds_early_late_sh %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_sh), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = plot_sensitivity_colors) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = c("-", "+", "++", "+++", "++++", "+++++")) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_sh) + 
  theme_classic()
big_data_filtereds_early_late_be2 <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_be2) %>% 
    select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox002,dox003,dox004,dox005,dox0075,dox01,dox1,
             #dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_be2,max_psi_be2)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox01)) %>% 
    distinct(gene, .keep_all = T) %>% 
  filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:14)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox002","dox003","dox004","dox005","dox0075","dox01","dox1"))) %>%
  mutate(label_junction = case_when(name == "dox01" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_be2$tdp43_sensitivity_be2)
table(big_data_filtereds_early_late_be2$max_psi_be2)
HH <- big_data_filtereds_early_late_be2 %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_be2), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = plot_sensitivity_colors) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = c("-", "+", "++", "+++", "++++", "+++++", "++++++", "+++++++")) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_be2) + 
  theme_classic()
GG + HH

print("1G - Alluvial plot")  
s <- master_table %>%
  filter(cryptic == "yes" & 
             tdp43_sensitivity_sh != "" & tdp43_sensitivity_be2 != "") %>% 
  group_by(tdp43_sensitivity_sh, tdp43_sensitivity_be2) %>%
  count(name = 'count') %>% 
  ungroup()

ggplot(s,
       aes(y = count, axis1 = tdp43_sensitivity_sh, axis2 = tdp43_sensitivity_be2)) +
  geom_alluvium(aes(fill = tdp43_sensitivity_sh), width = 1/12, knot.pos = 0) +
  guides(fill = "none") +
  geom_stratum(alpha = .5, width = 1/8) +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  theme_void()
```

A. Dose-response curves for SH-SY5Y
B. Dose-response curves for SK-N-BE(2)
C. Differential expression in SH-SY5Y
D. Differential expression in SK-N-BE(2)
E. Spaghetti plot in SH-SY5Y
F. Spaghetti plot in SK-N-BE(2)
G. Alluvial plot




## Figure 2 - Features
```{r, echo=FALSE, warning=FALSE, message=FALSE}
master_table <- master_table %>%
  mutate(new_cate = case_when(tdp43_sensitivity_be2 == "" ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_sh == "" ~ tdp43_sensitivity_be2,
                            tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                            tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late'))

get_signif_stars <- function(p_value) {
  if (p_value < 0.001) {
    return("***")
  } else if (p_value < 0.01) {
    return("**")
  } else if (p_value < 0.05) {
    return("*")
  } else {
    return("NS")
  }
}

get_significance <- function(df, value, group, category_col, category) {
  test_formula <- as.formula(paste(value, "~", group))
  p_test <- kruskal.test(test_formula, data = df %>% filter(!!sym(category_col) == category))
  #p1_annotation <- paste("p =", format(p1_test$p.value, digits = 3), get_signif_stars(p1_test$p.value))
  return(get_signif_stars(p_test$p.value))
}

print("2A - Splice site strength")
df_5ss_strength <- read.csv(here::here('data', 'maxent/5ss_strength_combined.csv'))
df_3ss_strength <- read.csv(here::here('data', 'maxent/3ss_strength_combined.csv'))
df_ss_strength <- rbind(
  df_5ss_strength  %>%
    mutate(site = "5' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, new_cate, site, ID_exon),
  df_3ss_strength  %>%
    mutate(site = "3' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, new_cate, site, ID_exon)
) %>%
  filter(new_cate %in% c("Early", "Late"))
h <- df_ss_strength %>%
  ggplot(aes(x=site, y = ss_ratio, color = new_cate)) +
  geom_boxplot() +
  geom_abline(intercept = 1, slope = 0, color = "gray", linetype = "dashed") +
  # geom_signif(
  #  y_position = c(max(df_ss_strength$ss_ratio) * 1.05,max(df_ss_strength$ss_ratio) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
  #  annotation = c(get_significance(df_ss_strength,
  #                                  "ss_ratio", "new_cate", "site", "3' splice site"),
  #                 get_significance(df_ss_strength,
  #                                  "ss_ratio", "new_cate", "site", "5' splice site")), tip_length = 0.02
  # ) +
  scale_color_manual(values = plot_sensitivity_colors) +
  #ylim(-4.1, 2.5) +
  labs(y = "MaxEnt score cryptic site vs annotated site", x = "") +
  theme_minimal() +
  theme(legend.position = "top") +
  guides(color = "none")
h

print("2B - Conservation score for CE and non-ce in combined categories")
cons_score_df <- master_table %>%
  filter(new_cate %in% c("Early", "Late")) %>%
  subset(!is.na(phastCons_alt) & !is.na(phastCons_unchanged)) %>%
  pivot_longer(cols = c(phastCons_alt, phastCons_unchanged), names_to = "category", values_to = "cons_score")

cons_score_df_cryptic <- cons_score_df %>% filter(cryptic == "yes")
cons_score_df_not_cryptic <- cons_score_df %>% filter(cryptic == "no")
a <- cons_score_df_cryptic %>%
  ggplot(aes(x = category, y = cons_score, fill = new_cate)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Novel splice site", "Unchanged splice site")) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_alt"),
                   get_significance(cons_score_df_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_unchanged"))
  ) +
  #facet_wrap(facets = vars(cryptic)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

b <- cons_score_df_not_cryptic %>%
  ggplot(aes(x = category, y = cons_score, fill = new_cate)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Alternatively used splice site", "Unchanged splice site")) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_alt"),
                   get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_unchanged"))
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
a + b
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#C. Kmers enrichment analysis in SH-SY5Y
print("2C - Kmers enrichment analysis in SH-SY5Y")

kmer_size <- 6
yg_motifs <- c('TGTGTG', 'GTGTGT', 'GTGTGC', 'CGTGTG', 'TGCGTG', 'TGTGCG')
ya_motifs <- c('ATGTGT', 'TGTGTA', 'TGCATG', 'GTGTAT', 'TGTATG', 'GTATGT')
aa_motifs <- c('AATGAA', 'TTGAAT', 'ATGAAT', 'GTGTGA', 'TGAATG', 'GAATGA', 'GTGAAT', 'GAATGT')
all_motifs <- c(yg_motifs, ya_motifs, aa_motifs)
categories <- c(rep("YG", 6), rep("YA", 6), rep("AA", 8))

count_hexamers <- function(row) {
  if (row[["junc_cat"]] == "exon_skip" || is.na(row[["paste_into_igv_junction_annotated"]])) {
    return(NULL)
  }

  coords <- strsplit(row[["paste_into_igv_junction_annotated"]], "[:-]")[[1]]
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = coords[1], start = as.numeric(coords[2]), end = as.numeric(coords[3]))
  if (row[["strand"]] == "-") {
    seq <- reverseComplement(seq)
  }
  counts <- sapply(all_motifs, function(motif) countPattern(motif, seq, max.mismatch = 0))
  names(counts) <- all_motifs

  return(counts)
}

master_table_filt <- master_table %>%
  filter(new_cate %in% c("Early", "Late")) %>%
  filter(junc_cat != "exon_skip") %>%
  distinct(paste_into_igv_junction_annotated, .keep_all = T)
kmer_table <- apply(master_table_filt, 1, count_hexamers)
all_counts_df <- as.data.frame(t(kmer_table))

all_counts_df$type <- master_table_filt$new_cate
all_counts_df$cryptic <- master_table_filt$cryptic

summary_counts <- all_counts_df %>%
  group_by(type, cryptic) %>%
  summarise(across(all_motifs, sum))

means <- apply(summary_counts[, -c(1,2)], 1, mean)
sds <- apply(summary_counts[, -c(1,2)], 1, sd)
z_scores <- sweep(summary_counts[, -c(1,2)], MARGIN = 1, STATS = means, FUN = "-") / sds
heatmap_data <- as.matrix(z_scores)
rownames(heatmap_data) <- paste(summary_counts$type, ifelse(summary_counts$cryptic == "yes", "Cryptic", "Not cryptic"))

sort_kmers_by_zscore <- function(kmers, z_scores) {
  kmers[order(z_scores[kmers], decreasing = TRUE)]
}

sort_categories_by_zscore <- function(kmers, z_scores) {
  categories[order(z_scores[kmers], decreasing = TRUE)]
}

sorted_all_kmers <- sort_kmers_by_zscore(all_motifs, heatmap_data["Early Cryptic",]+heatmap_data["Late Cryptic",])
sorted_categories <- sort_categories_by_zscore(all_motifs, heatmap_data["Early Cryptic",]+heatmap_data["Late Cryptic",])

annotation <- HeatmapAnnotation(Motif = factor(sorted_categories, levels = c("YG", "YA", "AA")),
                                show_legend = c(F, F, F),
                                col = list(Motif = c(YG = "palegreen3", YA = "turquoise3", AA = "tan1")),
                                which = "col")
c <- Heatmap(heatmap_data[c("Early Cryptic","Early Not cryptic","Late Cryptic","Late Not cryptic") , sorted_all_kmers],
        cluster_columns = F,
        cluster_rows = F,
        show_row_names = T,
        show_column_names = T,
        show_row_dend = F,
        bottom_annotation = annotation,
        col = colorRampPalette(brewer.pal(12, "RdPu"))(100),
        show_heatmap_legend = FALSE)
c
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#D. Density UG motifs by early-late in SH-SY5Y
print("2D and 2E - Density UG motifs by early-late and top-bottom in SH-SY5Y")
sequences <- sapply(seq_along(master_table$paste_into_igv_junction_annotated), function(i) {
  coord <- master_table$paste_into_igv_junction_annotated[i]
  chr <- sub("^(.*):.*$", "\\1", coord)
  range <- as.numeric(unlist(strsplit(sub(".*:(.*)$", "\\1", coord), "-")))
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = chr, start = range[1], end = range[2])

  if (master_table$strand[i] == "-") {
    seq <- reverseComplement(seq)
  }

  return(seq)
})

master_table$longest_ug_region <- Biostrings::vcountPattern('TGTGTG', DNAStringSet(sequences), max.mismatch = 1)

master_table <- master_table %>%
  mutate(
    start = as.numeric(sub(".*:(\\d+)-.*", "\\1", paste_into_igv_junction_annotated)),
    end = as.numeric(sub(".*-(\\d+)", "\\1", paste_into_igv_junction_annotated)),
    intron_len = end - start,
    ug_density = longest_ug_region / intron_len
  ) %>%
  select(-start, -end)

<<<<<<< Updated upstream
d <- master_table %>%
  filter(new_cate %in% c("Early", "Late")) %>%
=======
d <- master_table %>% 
    filter(cryptic == 'yes') %>% 
  # filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>% 
>>>>>>> Stashed changes
  ggplot(aes(x = cryptic, y = ug_density, fill = new_cate)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(new_cate %in% c("Early", "Late")),
                                    "ug_density", "new_cate", "cryptic", "no"),
                   get_significance(master_table %>% filter(new_cate %in% c("Early", "Late")),
                                    "ug_density", "new_cate", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

#E. Density UG motifs by top-bottom in SH-SY5Y
master_table$max_psi_sh <- factor(master_table$max_psi_sh, levels=c("Top", "Bottom"))
e <- master_table %>%
  filter(max_psi_sh %in% c("Top", "Bottom")) %>%
  ggplot(aes(x = cryptic, y = ug_density, fill = max_psi_sh)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) +
  scale_fill_manual(values = plot_psi_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(max_psi_sh %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_sh", "cryptic", "no"),
                   get_significance(master_table %>% filter(max_psi_sh %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_sh", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
d + e
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#F. Distance iCLIP TDP43 binding site SH-SY5Y
print("2F - Distance iCLIP TDP43 binding site SH-SY5Y")
df_iclip <- read.csv(here::here('data', 'all_events.closest_tdp_bind.bed'), sep="\t")
df_iclip$tdp_len <- df_iclip$end - df_iclip$start + 1

f <- df_iclip %>%
  filter(new_cate %in% c("Early", "Late")) %>%
  filter(abs(dist) <= 1000) %>%
  ggplot(aes(x = dist, fill = new_cate)) +
  geom_density(alpha = 0.5, adjust = 1) +
  facet_wrap(~cryptic,
             ncol = 2,
             labeller = labeller(cryptic=c("yes" = "Cryptic", "no" = "Not cryptic"))) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  labs(x = 'Distance to the nearest TDP-binding site', y = 'Density') +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

f

print("2G - RNA maps based on figure A")
#G. RNA maps based on figure F
early_donors <- readDNAStringSet("data/maxent/donors_early_combined.fa")
late_donors <- readDNAStringSet("data/maxent/donors_late_combined.fa")

g1 <- ggplot() +
  geom_logo(data = as.character(early_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g2 <- ggplot() +
  geom_logo(data = as.character(late_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g <- g1 / g2
g

```

A. Splice site strength
B. Conservation score for CE and non-CE
C. Kmers enrichment analysis
D. Density UG motifs by early-late
E. Density UG motifs by top-bottom
F. Distance iCLIP TDP43 binding site
G. RNA maps based on figure F

## Figure 3 - NMD

A. Categories - Overlap splicing changes?
B. Late/early is not driven by NMD sensitivity
C. Slope plot in SH-SY5Y
D. Differential expression of NMD sensitive and not
E. Differential expression of NMD sensitive/not but more detailed
F. TDP-43 dependent loss of dark cryptics
G. 4-SU

```{r LateEarlySushi, echo=FALSE, warning=FALSE, message=FALSE}
# reading in the master table again for debug, can remove later
master_table <- read.csv(here::here("data/master_table.csv"))
master_table <- master_table %>%
  mutate(new_cate = case_when(tdp43_sensitivity_be2 == "" ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_sh == "" ~ tdp43_sensitivity_be2,
                            tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                            tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late'))
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))


sh_nmd_sens = big_data_chx %>% 
    distinct(paste_into_igv_junction,color_gene_name) %>% 
    left_join(master_table) %>% 
    filter(!is.na(tdp43_sensitivity_sh)) %>% 
    filter(tdp43_sensitivity_sh != "") %>% 
    filter(junction_for_clustering == 'yes') %>% 
    group_by(color_gene_name,tdp43_sensitivity_sh) %>% 
    reframe(n = n_distinct(paste_into_igv_junction)) 

ggplot(sh_nmd_sens) +
    aes(x = tdp43_sensitivity_sh, fill = color_gene_name, weight = n,by = as.factor(tdp43_sensitivity_sh)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'Cycloheximide treatment') +
    scale_fill_manual(values = c("#3288BD","#D53E4F")) + 
    theme(legend.position = 'bottom') 

sh_nmd_sens %>% 
    pivot_wider(names_from = 'color_gene_name',values_from = 'n') %>% 
    column_to_rownames("tdp43_sensitivity_sh") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()

```

```{r LateEarlyBE2}
sk_nmd_sens = big_data_chx %>% 
    distinct(paste_into_igv_junction,color_gene_name) %>% 
    left_join(master_table) %>% 
    filter(!is.na(tdp43_sensitivity_be2)) %>% 
    filter(tdp43_sensitivity_be2 != "") %>% 
    group_by(color_gene_name,tdp43_sensitivity_be2) %>% 
    reframe(n = n_distinct(paste_into_igv_junction)) 

ggplot(sk_nmd_sens) +
    aes(x = tdp43_sensitivity_be2, fill = color_gene_name, weight = n, by = as.factor(tdp43_sensitivity_be2)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'Cycloheximide treatment') +
    scale_fill_manual(values = c("#3288BD","#D53E4F")) + 
    theme(legend.position = 'bottom')

sk_nmd_sens %>% 
    pivot_wider(names_from = 'color_gene_name',values_from = 'n') %>% 
    column_to_rownames("tdp43_sensitivity_be2") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()
```

```{r NMDSensNewCate}

new_cate_nmd_sens = big_data_chx %>% 
    distinct(paste_into_igv_junction,color_gene_name) %>% 
    left_join(master_table) %>% 
    filter(cryptic == 'yes') %>% 
    filter(!is.na(new_cate)) %>% 
    filter(new_cate != "") %>% 
    group_by(color_gene_name,new_cate) %>% 
    reframe(n = n_distinct(paste_into_igv_junction)) 

ggplot(new_cate_nmd_sens) +
    aes(x = new_cate, fill = color_gene_name, weight = n, by = as.factor(new_cate)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'Cycloheximide treatment') +
    scale_fill_manual(values = c("#3288BD","#D53E4F")) + 
    theme(legend.position = 'bottom')

new_cate_nmd_sens %>% 
    pivot_wider(names_from = 'new_cate',values_from = 'n') %>% 
    column_to_rownames("color_gene_name") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()

```




```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("3A")

print("3B")


print("3C - Slope plot in SH-SY5Y")
gene_list_slope <- c("STMN2", "UNC13A", "AARS1", "HDGFL2", "CYFIP2", "SYNE1")
input_list <- c("Control_Control",
                "Cycloheximide_Control", 
                "Control_TDP43KD", 
                "Cycloheximide_TDP43KD")
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))
a <- slope_plot_nmd(big_data_chx, 
               gene_list_slope) #### NMD rescued and unaffected
```

### Are NMD sensitive events up/downregulated?

```{r nmd_up_down_sh}
simple_chx = big_data_chx %>% 
    distinct(gene_name,color_gene_name) %>% 
    group_by(gene_name) %>%
    mutate(status = ifelse(any(color_gene_name == 'NMD rescued'), "NMD rescued", color_gene_name)) %>% 
    ungroup() %>% 
    distinct(gene_name,status)


diff_plt_sh = diff_expr_sh %>% 
    left_join(simple_chx,by = c("symbol" = "gene_name")) %>% 
    filter(symbol != "RNA5-8SN2") %>% 
    mutate(fc_direction = ifelse(log2FoldChange > 0, "up","down")) %>%
    mutate(sig_de = ifelse(padj < 0.05 & abs(log2FoldChange) >= fold_cut, "sig", "non_sig")) %>% 
    filter(!is.na(sig_de)) %>% 
    mutate(sig_de = glue::glue("{sig_de}\n{fc_direction}")) %>% 
    distinct(symbol,status,fc_direction,sig_de,source) %>% 
    mutate(status = ifelse(is.na(status),'non-splicing',status)) %>% 
    mutate(status = ifelse(status == '','non-splicing',status)) 
    
sh_d_gene = diff_plt_sh %>% 
    group_by(status,source) %>% 
    summarize(denom = n_distinct(symbol))

diff_plt_sh %>% 
    group_by(status,fc_direction,sig_de,source) %>% 
    summarize(n_genes = n_distinct(symbol)) %>% 
    ungroup() %>% 
    left_join(sh_d_gene) %>% 
    mutate(prop = n_genes/denom) %>% 
    ggplot(aes(x = status,
               y = prop,
               fill = sig_de)) + 
    geom_col(position = position_dodge(preserve = "single")) + 
    facet_wrap(~source,scales = 'free_x') +
    # scale_fill_manual(values = c("#3288BD","#D53E4F","darkgrey")) +
    theme_classic() + 
    ylab("Proportion of genes") + 
    theme(legend.position = 'bottom')  
    

```


<<<<<<< Updated upstream
```{r}
parse_coords <- function(coord_str) {
  parts <- str_match(coord_str, "(chr\\w+):(\\d+)-(\\d+)")
  list(chrom = parts[, 2], start = as.numeric(parts[, 3]), end = as.numeric(parts[, 4]))
}

merge_junctions <- function(df, distance_threshold = 2000) {
  df <- df %>% arrange(paste_into_igv_junction)

  result <- list()
  i <- 1
  while (i <= nrow(df)) {
    current <- parse_coords(df$paste_into_igv_junction[i])

    if (i < nrow(df)) {
      next_junction <- parse_coords(df$paste_into_igv_junction[i + 1])

      # Check if chromosomes are the same and distance is less than threshold
      if (current$chrom == next_junction$chrom &&
          (next_junction$start - current$end) < distance_threshold &&
          (next_junction$start - current$end) > 0) {

        # Merge the junctions
        merged_coord <- sprintf("%s:%d-%d", current$chrom, current$end, next_junction$start)
        result[[length(result) + 1]] <- c(
          list(paste_into_igv_junction = merged_coord),
          new_cate = df$new_cate[i],
          cryptic = df$cryptic[i],
          gene = df$gene[i]
        )

        i <- i + 2  # Skip the next row as it's been merged
        next
      }
    }

    # If not merged, don't add to the result
    i <- i + 1
  }

  # Convert result to data frame
  result_df <- do.call(rbind, lapply(result, data.frame))

  # Remove rows with same start or end coordinates
  result_df <- result_df %>%
    mutate(
      start = sapply(paste_into_igv_junction, function(x) parse_coords(x)$start),
      end = sapply(paste_into_igv_junction, function(x) parse_coords(x)$end)
    ) %>%
    group_by(start, end) %>%
    filter(n() == 1) %>%
  ungroup()
}

exons_combine <- merge_junctions(master_table) %>%
  mutate(
    symmetric = (end-start+1) %% 3==0
  )

exons1 <- read.table("data/noDox-dox0075.cryptic_exons.bed",col.names = c("chrom","start","end","name","score","strand"))
exons1 <- exons1 %>%
  tidyr::separate(name, into = c("gene", "paste_into_igv_junction"), sep = "\\|", remove = FALSE)

exons1 <- merge(exons1,
                master_table %>%
                  select(paste_into_igv_junction, new_cate, cryptic),
                by = "paste_into_igv_junction", all.x=T) %>%
  distinct(paste_into_igv_junction, .keep_all = T) %>%
  mutate(
    symmetric = (end-start) %% 3==0
  )

exons2 <- read.table("data/all_the_ce_length_manual_filtered.bed.txt", fill = TRUE, header = T)
exons2 <- exons2 %>%
  mutate(
    symmetric = (end-start) %% 3==0
  )
exons2$paste_into_igv_junction <- paste0(exons2$chr, ":", exons2$start + 1, "-", exons2$end)
master_table <- master_table %>%
    mutate(
      start = sapply(paste_into_igv_junction, function(x) parse_coords(x)$start),
      end = sapply(paste_into_igv_junction, function(x) parse_coords(x)$end)
    )

exons2 <- merge(exons2,
                master_table %>%
                  select(gene, paste_into_igv_junction, new_cate, cryptic),
                by.x="gene_name",by.y = "gene", all.x=F) %>%
  distinct(gene_name, symmetric, .keep_all = T)
exons2 <- exons2[!duplicated(exons2$gene_name) & !duplicated(exons2$gene_name, fromLast = TRUE), ]

exons2 <- exons2 %>% select(gene_name, paste_into_igv_junction.x, new_cate, cryptic, symmetric)
colnames(exons2) <- c("gene", "paste_into_igv_junction", "new_cate", "cryptic", "symmetric")

result_mrg <- rbind(
  exons_combine %>% select(gene, paste_into_igv_junction, new_cate, cryptic, symmetric),
  exons1 %>% select(gene, paste_into_igv_junction, new_cate, cryptic, symmetric),
  exons2 %>% select(gene, paste_into_igv_junction, new_cate, cryptic, symmetric)) %>%
  na.omit() %>%
  distinct()

junction_counts <- result_mrg %>%
  group_by(new_cate, cryptic, symmetric) %>%
  summarise(count = n()) %>%
  ungroup()

ggplot(junction_counts %>%
         filter(new_cate != ""), aes(x = new_cate, y = count, fill = symmetric)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~cryptic)+
  labs(
    #title = "Number of Symmetric and Non-Symmetric Exons by Type",
    x = "",
    y = "Number of exons",
    fill = "Symmetric"
  ) +
  scale_fill_manual(values = c("TRUE" = "darkslategray", "FALSE" = "tan2")) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size), axis.text.x = element_text(angle = 45, hjust = 1))

junction_counts %>%
  filter(cryptic=="yes") %>%
   filter(new_cate!="Intermediate") %>%
    pivot_wider(names_from = symmetric, values_from = count, values_fill = 0L) %>%
  select(-new_cate, -cryptic) %>%
    as.matrix() %>%
    chisq.test()

```

```{r}
data_chx <- big_data_chx %>%
    distinct(paste_into_igv_junction,color_gene_name) %>%
    left_join(master_table) %>%
    filter(!is.na(new_cate)) %>%
    filter(new_cate != "")

counts_chx_symmetric <- merge(result_mrg,
      data_chx) %>%
  group_by(new_cate, symmetric, color_gene_name) %>%
  summarise(count = n()) %>%
  ungroup()

ggplot(counts_chx_symmetric %>%
         filter(new_cate != ""), aes(x = color_gene_name, y = count, fill = symmetric)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~new_cate)+
  labs(
    #title = "Number of Symmetric and Non-Symmetric Exons by Type",
    x = "",
    y = "Number of exons",
    fill = "Symmetric"
  ) +
  scale_fill_manual(values = c("TRUE" = "darkslategray", "FALSE" = "tan2")) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size), axis.text.x = element_text(angle = 45, hjust = 1))

```
=======
```{r de_late_early_be2}

diff_plt_be2 = diff_expr_sk %>% 
    left_join(unique(master_table[junction_for_clustering == 'yes',.(gene,last_exon,cryptic,tdp43_sensitivity_be2)]),by = c("symbol" = 'gene')) %>% 
    filter(symbol != "RNA5-8SN2") %>% 
    mutate(fc_direction = ifelse(log2FoldChange > 0, "up","down")) %>%
    mutate(sig_de = ifelse(padj < 0.05 & abs(log2FoldChange) >= fold_cut, "sig", "non_sig")) %>% 
    filter(!is.na(sig_de)) %>% 
    mutate(sig_de = glue::glue("{sig_de}\n{fc_direction}")) %>% 
    distinct(symbol,last_exon,cryptic,tdp43_sensitivity_be2,fc_direction,sig_de,source) %>% 
    mutate(tdp43_sensitivity_be2 = ifelse(is.na(tdp43_sensitivity_be2),'non-splicing',tdp43_sensitivity_be2)) %>% 
    mutate(tdp43_sensitivity_be2 = ifelse(tdp43_sensitivity_be2 == '','non-splicing',tdp43_sensitivity_be2)) 
    
be2_d_gene = diff_plt_be2 %>% 
    group_by(source,tdp43_sensitivity_be2) %>% 
    summarize(denom = n_distinct(symbol))

diff_plt_be2 %>% 
    group_by(tdp43_sensitivity_be2,fc_direction,sig_de,source) %>% 
    summarize(n_genes = n_distinct(symbol)) %>% 
    ungroup() %>% 
    left_join(be2_d_gene) %>% 
    mutate(prop = n_genes/denom) %>% 
    ggplot(aes(x = sig_de,
               y = prop,
               fill = tdp43_sensitivity_be2)) + 
    geom_col(position = position_dodge(preserve = "single")) + 
    facet_wrap(~source,scales = 'free_x')


```




>>>>>>> Stashed changes


## Figure 4 - compendium / in vivo

```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("4A")
##load targeted seq table and compute CE ratio / rename samples in agreement to UNC13A paper
maxc <- read.csv(here::here('data',"count_table_of_aligned_remove_200bp_plus_product_reads.csv")) %>%
  mutate(fraction_cryptic = n_cryptic / n_counts_for_gene) %>%
  mutate(sample_name_ari = case_when(sample_name == "P64_11" ~ "HC1",
                                     sample_name == "P17_07" ~ "HC2",
                                     sample_name == "P47_11" ~ "HC3",
                                     sample_name == "P35_07" ~ "HC4",
                                     sample_name == "P56_13" ~ "FTD1",
                                     sample_name == "P45_15" ~ "FTD2",
                                     sample_name == "P63_05" ~ "FTD3",
                                     sample_name == "P28_07" ~ "FTD4",
                                     sample_name == "P40_04" ~ "FTD5",
                                     sample_name == "P07_15" ~ "FTD6",
                                     sample_name == "P16_09" ~ "FTD7",
                                     sample_name == "P86_08" ~ "FTD8",
                                     sample_name == "P13_13" ~ "FTD9",
                                     sample_name == "P11_07" ~ "FTE10",
                                     sample_name == "ddH2O" ~ "HO2O")) #%>%
  #pivot_wider(names_from = sample_name_ari, values_from = ratio_cryptic)
table(maxc$sample_name_ari)

##use only TDP43path-specific genes (use list from AL) - this is exluding more than 2/3 of genes!!
al <- read.csv(here::here('data',"expression_by_pathology_updated.csv")) %>%
  dplyr::filter(fraction_path > 0.01 & fraction_not_path < 0.005) %>% ####need to carefully check this with AL
  distinct(gene, .keep_all = T) ##check when binding with annotated validation table - should be ok since left_join
names(al)[4] <- "gene_name"
maxc_al <- maxc %>%
  left_join(al, by = "gene_name") %>%
  dplyr::filter(!is.na(paste_into_igv_junction))
table(maxc_al$sample_name_ari)
table(maxc_al$gene_name)

##check igv files for these genes
master_table = master_table %>% 
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "" ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_sh == "" ~ tdp43_sensitivity_be2,
                            tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_be2 == 'Early' | tdp43_sensitivity_sh == 'Early' ~ 'Early',
                            tdp43_sensitivity_be2 == 'Late' | tdp43_sensitivity_sh == 'Late' ~ 'Late'))  

maxc_type <- maxc_al %>%
  select(-type) %>%
  mutate(gene = gene_name) %>%
  filter(gene_name != "SETD5") %>%
  left_join(master_table, by = "paste_into_igv_junction") %>%
  filter(!is.na(new_cate)) %>% 
  mutate(group = ifelse(grepl("FT", sample_name_ari), "FTD",
                        ifelse(grepl("HC", sample_name_ari), "HC",
                               "H2O"))) %>% 
  group_by(gene_name) %>%
  filter(sum(n_cryptic) > 0) %>%
  ungroup() %>% 
    filter(group != "H2O")

maxc_type %>%
  filter(new_cate != "") %>%
  filter(gene_name != "SETD5") %>%
  #filter(detected == "Yes") %>%
  filter(group != "H2O") %>%
  ggplot(aes(x = new_cate)) +
  geom_bar(aes(fill = new_cate)) +
  facet_wrap(facets = vars(group), ncol = 2) +
  scale_fill_brewer(palette = "Set1") +
  xlab("") +
  ylab("Count of cryptic exons detected by targeted panel") +
  labs(fill = "Category") +
  theme_bw()

maxc_earli <- maxc_type %>%
  filter(tdp43_sensitivity_sh == "Earli") %>%
  filter(gene_name != "SETD5") %>% 
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") + #position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Earli") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_early <- maxc_type %>%
  filter(new_cate == "Early") %>%
  filter(gene_name != "SETD5") %>% 
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") +#position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Early") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_intermediate <- maxc_type %>%
  filter(new_cate == "Intermediate") %>%
  filter(gene_name != "SETD5") %>%
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") +#position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Greens") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Intermediate") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_late <- maxc_type %>%
  filter(new_cate == "Late") %>%
  filter(gene_name != "SETD5") %>%
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") +#position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Late") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_early / maxc_intermediate / maxc_late #/ maxc_na

maxc_type
```

A.

B.

C. Detection of CEs from in vivo samples

D.
```{r readSC junc, echo=FALSE, warning=FALSE, message=FALSE}
# Based on analyses by Jack Humphrey

# reading in the master table again for debug, can remove later
master_table <- read.csv(here::here("data/master_table.csv"))

# read in the annotation for cells
anno_df = fread("data/anno_df_parse.csv")
# read in the pasre junctions and filter for the cryptic events in the master table
junc_df <- fread("data/all_parse_junctions_per_cell.tsv.gz")

junc_cryptic_master = junc_df |> 
    filter(junction %in% master_table$paste_into_igv_junction) |> 
    left_join((master_table %>% distinct(gene,paste_into_igv_junction,
                                         cryptic,
                                         tdp43_sensitivity_sh,
                                         tdp43_sensitivity_be2,
                                         longest_ug_region,
                                         dist_nearest_ug_region)),by = c("junction" = "paste_into_igv_junction")) |> 
    unique() |> 
    filter(cryptic == 'yes') 

# add in the cell barcode information and filter for cells with information
junc_cryptic_master <- junc_cryptic_master  %>%
    left_join(anno_df, by = c("barcode" = "donor_barcode") ) %>% 
    filter(!is.na(cluster))


```


```{r proportionExpressed, echo=FALSE, warning=FALSE, message=FALSE}
quick_bind = master_table %>% 
    select(gene,paste_into_igv_junction,junc_cat,tdp43_sensitivity_sh,tdp43_sensitivity_be2) %>% 
    mutate(gene = glue::glue("{gene}_{junc_cat}")) %>% 
    mutate(gene = make.unique(gene))

cluster_size_df <- 
    anno_df %>% group_by(cluster_label) %>% tally(name = "total")

cryptic_prop_cluster <- junc_cryptic_master %>%
    left_join(cluster_size_df) %>% 
    filter(!is.na(cluster)) %>%
    group_by(junction,cluster_label) %>% 
    reframe(n = n_distinct(barcode),gene,tdp43_sensitivity_sh,tdp43_sensitivity_be2) %>% 
    unique() %>% 
    left_join(cluster_size_df, by = "cluster_label") %>% 
    mutate(prop = n / total) 




```


```{r proprotionDetected,  echo=FALSE, warning=FALSE, message=FALSE}

fish_sh = quick_bind %>%
    filter(paste_into_igv_junction %in% s) %>% 
    mutate(overlap = paste_into_igv_junction %in% junc_cryptic_master$junction) %>% 
    select(tdp43_sensitivity_sh,overlap) %>% 
    filter(tdp43_sensitivity_sh %in% c("Early","Late")) %>% 
    table() 

fish_sh = quick_bind %>%
    mutate(overlap = paste_into_igv_junction %in% junc_cryptic_master$junction) %>% 
    select(tdp43_sensitivity_sh,overlap) %>% 
    filter(tdp43_sensitivity_sh %in% c("Early","Late")) %>% 
    table() 


ft = rstatix::fisher_test(fish_sh)

fish_sh %>% 
    melt() %>% 
    pivot_wider(names_from = overlap, values_from = value) %>%
    mutate( prop = `TRUE` / (`TRUE` + `FALSE`) ) %>%
    mutate(prop_label = paste0(`TRUE`, "/", `FALSE`) ) %>%
    ggplot(aes(x = tdp43_sensitivity_sh, y = prop)) + geom_col() +
    geom_text(aes(label = prop_label), nudge_y =  0.01) +
    labs(y = "Proportion detected",x = element_blank()) +
    geom_signif(
        comparisons = list(c("Early", "Late")),
        annotations = paste("p =", round(ft$p, digits = 3)),
        tip_length = 0.03
    ) +
    theme_classic()



fish_sk = quick_bind %>%
    mutate(overlap = paste_into_igv_junction %in% junc_cryptic_master$junction) %>% 
    select(tdp43_sensitivity_be2,overlap) %>% 
    filter(tdp43_sensitivity_be2 %in% c("Early","Late")) %>% 
    table() 

ft_sk = rstatix::fisher_test(fish_sk)

fish_sk %>% 
    melt() %>% 
    pivot_wider(names_from = overlap, values_from = value) %>%
    mutate( prop = `TRUE` / (`TRUE` + `FALSE`) ) %>%
    mutate(prop_label = paste0(`TRUE`, "/", `FALSE`) ) %>%
    ggplot(aes(x = tdp43_sensitivity_be2, y = prop)) + 
    geom_col() +
    geom_text(aes(label = prop_label), nudge_y =  0.01) +
    labs(y = "Proportion detected",x = element_blank()) +
    geom_signif(
        comparisons = list(c("Early", "Late")),
        annotations = paste("p =", round(ft_sk$p, digits = 3)),
        tip_length = 0.03
    ) +
    theme_classic()

```

