---
title: "TDP-43 cryptic exon biology main figures"
author: "MZ;AM;ALB"
date: "2024-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
###load packages/libraries/scripts
#BiocManager::install("ComplexHeatmap")
library(zoo)
library(gridExtra)
library(data.table)
library(tidyverse)
library(ggrepel)
library(patchwork)
library(ggalluvial)
library(ggsignif)
library(ggseqlogo)
library(Biostrings)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ComplexHeatmap)
library(RColorBrewer)
library(scales)
library(ggstats)
library(ggpubr)
library(pheatmap)
library(pdftools)
library(cowplot)
library(ggplotify)
source(here::here('scripts','odds_ratio_plots.R'))
source(here::here('scripts','trial_nmd_plots.R'))
source(here::here('scripts',"create_coverage_vectors.R"))

give.n <- function(x){
return(c(y = median(x)*1.05, label = length(x)))
}


#plot parameters
plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#173dd3",
                             "Intermediate" = "steelblue","Not found" = 'darkgrey',"Ambiguous" = 'darkred')
plot_psi_colors <- c("Top" = "#388f78", "Bottom" = "#de9e4a", "Intermediate" = "steelblue")
plot_inframe_colors <- c("In-frame" = "darkslategray", "Out-frame" = "tan2")
plot_text_size <- 10
```


## Cryptic compendium and recurrent cryptics

```{r overall dataset}
compendium_events = fread('data/compendium_cryptics.csv')

compendium_events = compendium_events %>% 
    mutate(plot_comparison = case_when(comparison == "Ferguson-HeLA" ~ 'HeLa',
                                       comparison == "Appocher-SKNBE2" ~ 'SK-N-BE(2) - 1',
                                       comparison == 'Brown-i3Cortical' ~ "i3 Cortical - 1",
                                       comparison == "Brown-SKNBE2" ~ 'SK-N-BE(2) - 2',
                                       comparison == 'Brown-SHSY5Y' ~ "SH-SY5Y - 2",
                                       comparison == 'Melamed-SHSY5Y'  ~ "SH-SY5Y - 1",
                                       comparison == 'Seddighi-i3Cortical' ~ "i3 Cortical - 2",
                                       comparison == 'Humphrey-i3Cortical' ~ "i3 Cortical - 3",
                                       comparison == 'Vargas-i3LMN' ~ "i3 Lower motor neuron",
                                       comparison == 'Klim-MotorNeuron' ~ "iPSC Motor neuron"))
c_tbl = compendium_events %>% 
    distinct(paste_into_igv_junction,plot_comparison) %>% 
    mutate(found = TRUE) %>% 
    pivot_wider(names_from = 'plot_comparison',values_from = 'found',values_fill = FALSE) %>% 
    tibble::column_to_rownames('paste_into_igv_junction') %>% as.matrix()

```

```{r}

# Calculate the absolute overlap counts
overlap_matrix <- matrix(0, ncol=ncol(c_tbl), nrow=ncol(c_tbl))
colnames(overlap_matrix) <- colnames(c_tbl)
rownames(overlap_matrix) <- colnames(c_tbl)

for(i in 1:ncol(c_tbl)) {
    for(j in 1:ncol(c_tbl)) {
        # Count intersection (number of TRUE values in both)
        overlap_matrix[i,j] <- sum(c_tbl[,i] & c_tbl[,j])
    }
}

# Create annotation data frame for the column sums
annotation_df <- data.frame(TotalJunctions = colSums(c_tbl))
rownames(annotation_df) <- colnames(overlap_matrix)

# Create annotation colors (optional)
# Scale the colors based on the values in TotalJunctions
annotation_colors <- list(
  TotalJunctions = colorRampPalette(c("white", "darkblue"))(100)
)
norm_matrix <- overlap_matrix / colSums(c_tbl)
hc <- hclust(dist(norm_matrix), method = "complete")
matrix_order <- hc$order

# Reorder the matrix rows and columns using the same order
ordered_matrix <- norm_matrix[matrix_order, matrix_order]

ordered_display_mat <- matrix(paste0(round(ordered_matrix * 100, 0), "%"), 
                              nrow = nrow(ordered_matrix))

annotation_df = annotation_df %>% mutate(plot_name = rownames(annotation_df)) %>% 
    mutate(TotalJunctions = paste(plot_name,TotalJunctions,sep = ' - ')) 
# Reorder annotation if it exists
if(exists("annotation_df")) {
    ordered_annotation_df <- annotation_df[matrix_order, , drop = FALSE]
} else {
    # Create a simple annotation by cell type if none exists
    # This is just an example - modify according to your actual data
    cell_types <- c("Cancer", "Neuronal", "Neuronal", "Cancer", "Cancer", 
                    "Neuronal", "Neuronal", "Cancer", "Neuronal", "Neuronal")
    ordered_annotation_df <- data.frame(Cell_Type = cell_types[matrix_order])
    annotation_colors <- list(Cell_Type = c(Cancer = "blue", Neuronal = "green"))
}
om2 = ordered_matrix %>% as.data.frame() %>% rownames_to_column('plot_name') %>% left_join(annotation_df) %>% column_to_rownames('TotalJunctions') %>% select(-plot_name)
# Create the heatmap with the ordered matrix
pheatmap(om2 * 100,
         color = colorRampPalette(c("white", "orange", "red"))(100),
         cluster_rows = FALSE,  # We've already clustered
         cluster_cols = FALSE,  # We've already clustered
          display_numbers = ordered_display_mat,
         main = "Cryptic Splice Junction Overlap (%)",
         fontsize = 12,
         fontsize_number = 12,
         annotation_names_row = TRUE)

```


```{r appearnace of most frequent}
compendium_events %>% 
    mutate(gene_name_new = ifelse(grepl("^ENS", gene_name), NA_character_, gene_name)) %>% 
    group_by(paste_into_igv_junction) %>% 
    fill(gene_name_new) %>% 
    ungroup() %>% 
    filter(n_significant_data >= 6) %>%
    distinct(gene_name_new, plot_comparison) %>% 
    mutate(gene_name_new = ifelse(gene_name_new == 'RP1-283E3.8', 'CDK11A', gene_name_new)) %>% 
    mutate(gene_name_new = ifelse(gene_name_new == 'RP4-583P15.15', 'ZGPAT', gene_name_new)) %>% 
    distinct(gene_name_new, plot_comparison) %>% 
    group_by(gene_name_new) %>% 
    mutate(n_datasets = n_distinct(plot_comparison)) %>% 
    ungroup() %>% 
    mutate(gene_name_new = fct_reorder(gene_name_new, n_datasets)) %>% 
    mutate(present = 1) %>% 
    ggplot(aes(y = gene_name_new, x = plot_comparison,fill = present)) + 
    geom_tile(color = "black", fill = "grey", linewidth = 0.3) +
        scale_y_discrete(drop = FALSE) + # Shows all genes even if not in every comparison
    theme_minimal() +
    theme(
        panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_rect(fill = "white", color = NA),
        axis.text.y = element_text(face = "italic")
    ) +
    labs(x = element_blank(), y = element_blank()) +
    theme(axis.text = element_text(size = 12)) 

 
```


## Cryptic compendium expression in post-mortem


```{r overlap_biomarker_indataset}
summary_post = fread("data/summary_metrics_post_mortem.csv")


not_in_the_cord = summary_post %>%
    filter(fpr_value <=0.1 & tpr_value >= 0.1) %>%
    filter(value_tdp_path >=5) %>%
    filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction) %>%
    distinct(group,gene_name) %>%
    mutate(observed = TRUE) %>%
    pivot_wider(names_from = 'group',
                values_from = 'observed',
                values_fill = FALSE) %>%
    unique() %>%
    mutate(total_count = rowSums(across(RiMOD:`Motor Cortex`))) %>%
    filter(Cord == FALSE & total_count == 3)

marker_euler = 
    summary_post %>% 
    filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction) %>% 
    filter(fpr_value <=0.1 & tpr_value >= 0.1) %>% 
    filter(value_tdp_path >=5) %>% 
    distinct(group,gene_name) %>% 
    mutate(observed = TRUE) %>% 
    pivot_wider(names_from = 'group',
                values_from = 'observed',
                values_fill = FALSE) %>% 
    unique() %>% 
    # mutate(total_count = rowSums(across(RiMOD:`Motor Cortex`)))
    select(-gene_name) 

custom_labels <- c("RiMOD\nFTLD\nFrontal Cortex",
                   "NYGC FTLD\nFrontalTemporal Cortex",
                   "NYGC ALS\nSpinal Cord",
                    "NYGC ALS\nMotor Cortex")

colnames(marker_euler) = custom_labels
gg_euler = marker_euler %>% 
    eulerr::euler(shape = "ellipse") %>% 
    plot(
     fills = c("#EF7F7F", "#ACD7E5", "#4A5240", "#5F4BB6"),
     labels = list(col = "white", cex = 1.1),  # Try labels as a list with color
     quantities = list(fontsize = 20, col = "white"),  # White numbers with larger font
     # Text styling
     text = list(fontsize = 10, col = "white"),        # White labels with larger font
     # Additional styling options
     edges = list(lwd = 4)             # White borders, thicker
)

gg_euler = as.ggplot(gg_euler)

#get the genes in the overlap section
genes_everywhere = summary_post %>% 
    filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction) %>% 
    filter(fpr_value <=0.1 & tpr_value >= 0.1) %>% 
    filter(value_tdp_path >=5) %>% 
    distinct(group,paste_into_igv_junction) %>% 
    group_by(paste_into_igv_junction) %>% 
    add_count() %>% 
    filter(n == 4) %>% 
    ungroup() %>% 
    left_join(summary_post %>% distinct(paste_into_igv_junction,gene_name,frac_ob_tdp_path)) %>% 
    group_by(gene_name) %>% 
    summarise(avg_d = mean(frac_ob_tdp_path)) %>% 
    arrange(-avg_d)

#get the ordering of the genes not in the cord
genes_not_in_cord = summary_post %>% 
    filter(gene_name %in% not_in_the_cord$gene_name) %>% 
    filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction) %>% 
    filter(fpr_value <=0.1 & tpr_value >= 0.1) %>% 
    filter(value_tdp_path >=5) %>% 
    distinct(group,paste_into_igv_junction) %>% 
    group_by(paste_into_igv_junction) %>% 
    add_count() %>% 
    filter(n == 3) %>% 
    ungroup() %>% 
    left_join(summary_post %>% distinct(paste_into_igv_junction,gene_name,frac_ob_tdp_path)) %>% 
    group_by(gene_name) %>% 
    summarise(avg_d = mean(frac_ob_tdp_path)) %>% 
    arrange(-avg_d) %>% 
    mutate(gene_name = ifelse(gene_name == 'CORO7-PAM16,CORO7','CORO7',gene_name))

paste0(genes_everywhere$gene_name,collapse = "\n")

paste0(genes_not_in_cord$gene_name,collapse = "\n")




gg_euler +
    # Expand the plot limits to add more space on the left
    xlim(-0.5, 1.5) +  # Extend left side from 0 to -0.5
    ylim(0, 1) +     # Keep vertical limits the same
    # Add arrow pointing to center (now with more space)
    annotate("segment", 
             x = 1, y = 0.6,          # Start point (now in the expanded left space)
             xend = 0.665, yend = 0.45,     # End point (center of euler plot)
             arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             color = "black", 
             size = 1.2) +
    # Add text box annotation in the expanded space
    annotate("label", 
             x = 1.12, y = 0.5,          # Position in the new left space
             label = paste0(genes_not_in_cord$gene_name,collapse = "\n"), 
             fill = "white", 
             color = "black",
             size = 6,
             fontface = "bold.italic",
             hjust = 0.5) +
        # Add text box annotation in the expanded space
    annotate("label", 
             x = -0.2, y = 0.5,          # Position in the new left space
             label = paste0(genes_everywhere$gene_name,collapse = "\n"), 
             fill = "white", 
             color = "black",
             size = 6,
             fontface = "bold.italic",
             hjust = 0.5) +
    theme_void() +
    theme(plot.background = element_rect(fill = "white", color = NA),
          plot.margin = margin(60, 20, 20, 60, "pt"))  # Add extra margin on left (60pt)
    
```




```{r fpr_compendium}
summary_post = fread("data/summary_metrics_post_mortem.csv")

fpr_across_ds_plot_table = summary_post %>% 
    left_join(compendium_events %>% 
    distinct(n_significant_data,paste_into_igv_junction)) %>% 
        mutate(bind_n = case_when(n_significant_data <= 3 ~ "1-3",
                              n_significant_data >= 4 & n_significant_data <= 6 ~ "4-6",
                              TRUE ~ "7-10")) %>% 
    filter(!is.na(n_significant_data)) %>% 
    mutate(plot_name = case_when(paste_into_igv_junction == "chr12:131992230-132005077" ~ 'EP400',
                                 paste_into_igv_junction == 'chr10:3099352-3099557' ~ 'PFKP',
                                 # paste_into_igv_junction == 'chr1:1709640-1719358' ~ "CDK11A",
                                 # paste_into_igv_junction == 'chr2:3458515-3460263' ~ 'TRAPPC12',
                                 paste_into_igv_junction == 'chr6:16585907-16589446' ~ 'ATXN1')) %>% 
    unique() %>% 
    # group_by(paste_into_igv_junction) %>% 
    # mutate(plot_name = ifelse(value == max(value),plot_name, NA_character_)) %>% 
    ungroup() 

fpr_across_ds_plot_table %>% 
    ggplot(aes(x = bind_n,
               y = fpr_value,
               fill = group)) + 
    geom_boxplot(outliers = FALSE) +
    ylab("False positive rate TDP-43 proteinopathy vs not") +
    xlab("N datasets cryptic detected") +
    theme_bw() +
    scale_fill_manual(values = c("#4A5240","#ACD7E5","#5F4BB6","#EF7F7F"),
                      name = "") +
    scale_color_manual(values = c("#4A5240","#ACD7E5","#5F4BB6","#EF7F7F"),
                  name = "") +
    theme(legend.position = 'bottom') +
    geom_point(pch = 21, data = fpr_across_ds_plot_table %>% filter(!is.na(plot_name)),position = position_dodge2(width = 0.75)) +
    geom_text_repel(aes(label = plot_name,color = group),min.segment.length = 0,position = position_dodge2(width = 0.9)) +
    geom_hline(yintercept = 0.1, linetype = 'dotted') +
    scale_y_continuous(
        labels = percent_format(),
    ) +
    theme(axis.text = element_text(size = 12))
    


```



## NMD

```{r makewidechx}
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))
big_data_chx = big_data_chx %>% 
    separate(paste_into_igv_junction,remove = FALSE,into = c("chr",'start','end'),
             convert = TRUE)


wide_chx = big_data_chx |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_chx = Control_Control < 0.05 &  Control_TDP43KD > 0.1) |> 
    mutate(chx_cryptic = Control_Control < 0.05 &  Cycloheximide_TDP43KD > 0.1) |> 
    mutate(chx_effect = case_when((Cycloheximide_TDP43KD - Control_TDP43KD) > 0.1 ~ "NMD rescued",
                                  Control_TDP43KD > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_chx %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id))


chx_dark = wide_chx |> 
    filter(normal_cryptic_chx == FALSE & chx_cryptic == TRUE & chx_effect == "NMD rescued" & Cycloheximide_Control < 0.1) |> 
    pull(paste_into_igv_junction) |> unique()

wide_chx = wide_chx %>% 
    mutate(chx_effect = ifelse(paste_into_igv_junction %in% chx_dark, "Dark cryptic",chx_effect))
```

```{r makewideupf1}
big_data_upf1 <- read.csv(here::here('data','nmd_or_not_upf1.csv'))
big_data_upf1 = big_data_upf1 %>% 
    separate(paste_into_igv_junction,remove = FALSE,into = c("chr",'start','end'),
             convert = TRUE)


wide_upf1 = big_data_upf1 |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_upf1 = ctrl_ctrl < 0.05 &  TDP43_ctrl > 0.1) |> 
    mutate(upf1_cryptic = ctrl_ctrl < 0.05 &  TDP43_UPF1 > 0.1) |> 
    mutate(upf1_effect = case_when((TDP43_UPF1 - TDP43_ctrl) > 0.1 ~ "NMD rescued",
                                  TDP43_ctrl > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_upf1 %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id))


upf1_dark = wide_upf1 |> 
    filter(normal_cryptic_upf1 == FALSE & upf1_cryptic == TRUE & upf1_effect == "NMD rescued" & ctrl_UPF1 < 0.1) |> 
    pull(paste_into_igv_junction) |> unique()

wide_upf1 = wide_upf1 %>% 
    mutate(upf1_effect = ifelse(paste_into_igv_junction %in% upf1_dark, "Dark cryptic",upf1_effect))
```

## NMDi all slope plots

```{r slopeplot_all, echo=FALSE, warning=FALSE, message=FALSE}
gene_list_slope = c("STMN2", 
                    "KCNQ2", 
                    "ATG4B", 
                    "HDGFL2", 
                    "KCNQ2", 
                    "ACTL6B", 
                    "GPSM2", 
                    "AKT3", 
                    "HDGFL2",
                    "PKN1",
                    "UNC13A",
                    'CYFIP2',
                    "SYNE1")



input_list = c("Control_Control",
                "Cycloheximide_Control", 
                "Control_TDP43KD", 
                "Cycloheximide_TDP43KD")
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))

big_data_chx$alpha_gene_name = NULL
big_data_chx$label_junction = NULL

data_summary_full = fread('data/cryptic_detection_summary_long_version.csv')

true_dark_selective = data_summary_full %>% 
    filter(true_dark == TRUE) %>%
    filter(fpr_value <= 0.1 & value_tdp_path >=5) %>%
    pull(paste_into_igv_junction) %>% 
    unique()

chx_slope_plot_table <- slope_plot_nmd(big_data_chx, 
               gene_list_slope,
               input_list = c("Control_Control",
                "Cycloheximide_Control", 
                "Control_TDP43KD", 
                "Cycloheximide_TDP43KD"),
               nmd_label = 'CHX treatment',
                true_dark_junctions = true_dark_selective) #### NMD rescued and unaffected

big_data_upf1 <- read.csv(here::here('data','nmd_or_not_upf1.csv'))
big_data_upf1$alpha_gene_name = NULL
big_data_upf1$label_junction = NULL

upf1_slope_plot_table <- slope_plot_nmd(big_data_upf1, 
               gene_list_slope,
               input_list = c("ctrl_ctrl",
                "ctrl_UPF1", 
                "TDP43_ctrl", 
                "TDP43_UPF1"),nmd_label = 'UPF1 KD',
               true_dark_junctions = true_dark_selective) #### NMD rescued and unaffected


psi_all_groups = arrow::read_parquet("data/nuc_cyto/fractionation_psi_all_groups.parquet")

psi_all_groups = psi_all_groups   %>%  dplyr::rename(.id = comparison)

psi_all_groups_cyto = psi_all_groups %>% 
    filter(!grepl("na",lsv_id)) %>% 
    filter(grepl("_cytosol",.id))  

smg1_slope_plot_table_cyto <- slope_plot_nmd(psi_all_groups_cyto, 
                                             gene_list_slope,
                                             input_list =  c("nt_ctrl_cytosol",
                                                                          "nt_nmd_cytosol",
                                                                          "dox_ctrl_cytosol",
                                                                          "dox_nmd_cytosol"),
                                             nmd_label = 'SMG1i\ncytosolic fraction',
                                             true_dark_junctions = true_dark_selective)
pseudo_psi = psi_all_groups %>% filter(.id %in% c("nt_ctrl","nt_nmd","dox_ctrl","dox_nmd"))

pseudo_bulk_plot_slope = slope_plot_nmd(psi_all_groups,
                                        gene_list = gene_list_slope,
                                        input_list =  c("nt_ctrl",
                                                        "nt_nmd",
                                                        "dox_ctrl",
                                                        "dox_nmd"),
                                        nmd_label = 'SMG1i',
                                        true_dark_junctions = true_dark_selective)


ggarrange(chx_slope_plot_table$plot,upf1_slope_plot_table$plot,pseudo_bulk_plot_slope$plot,nrow = 1)

```


```{r makewidesmg1}
big_data_smg1 = pseudo_bulk_plot_slope$filtered_table %>% 
    separate(paste_into_igv_junction,remove = FALSE,into = c("chr",'start','end'),
             convert = TRUE)


wide_smg1 = big_data_smg1 |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_smg1 = nt_ctrl < 0.05 &  dox_ctrl > 0.1) |> 
    mutate(smg1_cryptic = nt_ctrl < 0.05 &  dox_nmd > 0.1) |> 
    mutate(smg1_effect = case_when((dox_nmd - dox_ctrl) > 0.1 ~ "NMD rescued",
                                  dox_ctrl > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_smg1 %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id))


smg1_dark = wide_smg1 |> 
    filter(normal_cryptic_smg1 == FALSE & smg1_cryptic == TRUE & smg1_effect == "NMD rescued" & nt_nmd < 0.1) |> 
    pull(paste_into_igv_junction) |> unique()

wide_smg1 = wide_smg1 %>% 
    mutate(smg1_effect = ifelse(paste_into_igv_junction %in% smg1_dark, "Dark cryptic",smg1_effect))
```

### Are NMD sensitive events up/downregulated?

```{r nmd_up_down_sh_fractionation}
nuc_kd = fread("data/nuc_cyto/deseq2_results_nuc_kd.csv") %>% 
        mutate(ensgene =  gsub("\\..*","",gene_id))

cyto_kd = fread("data/nuc_cyto/deseq2_results_cyto_kd.csv") %>% 
            mutate(ensgene =  gsub("\\..*","",gene_id))

pseudo_kd = fread("data/nuc_cyto/deseq2_results_nuc_kd.csv") %>% 
            mutate(ensgene =  gsub("\\..*","",gene_id))


nuc_kd = nuc_kd %>% 
    left_join(wide_smg1,by = c("ensgene")) %>% 
    filter(!is.na(smg1_effect))  %>% 
    group_by(ensgene) %>% 
    mutate(max_psi = max(dox_nmd)) %>% 
    ungroup() %>% 
    mutate(bin_max = cut_interval(max_psi,n = 2)) %>% 
    mutate(fraction = 'nuclear RNA')


cyto_kd = cyto_kd %>% 
    left_join(wide_smg1,by = c("ensgene")) %>% 
    filter(!is.na(smg1_effect))  %>% 
    group_by(ensgene) %>% 
    mutate(max_psi = max(dox_nmd)) %>% 
    ungroup() %>% 
    mutate(bin_max = cut_interval(max_psi,n = 2)) %>% 
    mutate(fraction = 'cytoplasmic RNA')

facet_labels <- c("Low expressed (PSI < 46%)", "High expressed (PSI > 46%)")
names(facet_labels) <- unique(nuc_kd$bin_max)  

nuc_kd %>% 
    rbind(cyto_kd) %>% 
    group_by(smg1_effect,bin_max) %>% 
    mutate(n = n_distinct(paste_into_igv_junction)) %>% 
    mutate(x = glue::glue('{smg1_effect}\nn = {n}')) %>% 
    ggplot(aes(x = x,
               y = log2FoldChange,
               fill = fraction)) + 
    geom_hline(yintercept = 0, linetype = 'dotted') +
    geom_boxplot(position = position_dodge(preserve = "single")) +
    theme_classic() +
    labs(fill = 'Subcellular fraction') +
    xlab(element_blank()) +
    scale_fill_manual(values = c("lightblue", "moccasin")) + 
    ylab("Log2 Fold Change TDP-43 KD") +
    facet_wrap(~bin_max, labeller = labeller(bin_max = facet_labels),scales = 'free_x') +
    theme(text = element_text(size = 18)) +
    stat_compare_means(hide.ns = TRUE, 
                       method = "wilcox.test",
                       label = "p.signif",  
                       p.adjust.method = "BH",
                       size = 10) + 
    theme(legend.position = 'bottom')


```

```{r nmd_up_down_sh_chx}

chx_rna_full = fread(here::here('data',"chx_all_experiments_deseq2.csv"))

chx_nmd_effect = chx_rna_full %>% 
    left_join(wide_chx) %>% 
    filter(!is.na(chx_effect)) %>% 
    filter(experiment == "Control_TDP43KD|Control_Control") %>% 
    group_by(ensgene) %>% 
    mutate(max_psi = max(Cycloheximide_TDP43KD)) %>% 
    ungroup() %>% 
    mutate(bin_max = cut_interval(max_psi,n = 2)) %>% 
    ggplot(aes(x = bin_max,
               y = log2fold_change,fill = chx_effect)) + 
    geom_hline(yintercept = 0,linetype = 'dotted') +
    geom_boxplot(position = position_dodge(preserve = "single")) +
    stat_summary(fun.data = give.n,
                 geom = "text",
                 fun.y = median,
                 position = position_dodge(width = 0.75),color = 'white',) +
    theme_classic() +
    xlab(element_blank()) +
    scale_fill_manual(values = c("#0D1821","#3288BD","#D53E4F","#BFCC94")) + 
    ylab("Log2 Fold Change TDP-43 KD") +
    labs(fill = 'NMD sensitivity') +
    theme(text  = element_text(size = 18)) +
    theme(legend.position = "bottom")


    
```


```{r nmd_up_down_upf1}

upf1_rna_full = fread(here::here('data',"upf1_all_gene_expression.csv"))



upf1_nmd_effect = upf1_rna_full %>% 
    left_join(wide_upf1) %>% 
    filter(!is.na(upf1_effect)) %>% 
    filter(experiment == "ctrlctrl_ctrlTDP43") %>% 
    group_by(ensgene) %>% 
    mutate(max_psi = max(TDP43_UPF1)) %>% 
    ungroup() %>% 
    mutate(bin_max = cut_interval(max_psi,n = 2)) %>% 
    ggplot(aes(x = bin_max,
               y = log2fold_change,fill = upf1_effect)) + 
    geom_hline(yintercept = 0,linetype = 'dotted') +
    geom_boxplot(position = position_dodge(preserve = "single")) +
    stat_summary(fun.data = give.n,
                 geom = "text",
                 fun.y = median,
                 position = position_dodge(width = 0.75),color = 'white') +
    theme_classic() +
    xlab(element_blank()) +
    scale_fill_manual(values = c("#0D1821","#3288BD","#D53E4F","#BFCC94")) + 
    ylab("Log2 Fold Change TDP-43 KD") +
    theme(legend.position = 'none') +
    theme(text  = element_text(size = 18)) 

```



# Localization of cryptics at baseline

```{r}
sj_star_counts = arrow::read_parquet('data/nuc_cyto/star_splice_junctioncounts.parquet')

sj_star_counts %>% 
    filter(paste_into_igv_junction %in% pseudo_bulk_plot_slope$filtered_table$paste_into_igv_junction) %>% 
    # left_join(librar_sizes) %>% 
    # mutate(rpkm = (uniquely_mapping / `Uniquely mapped`) * 10^6) %>% 
    mutate(rpkm = (uniquely_mapping / exonic_reads) * 10^6) %>% 
    left_join(pseudo_bulk_plot_slope$filtered_table %>% distinct(paste_into_igv_junction,color_gene_name))  %>% 
    filter(tdp_kd == 'dox') %>% 
    group_by(paste_into_igv_junction,nmdi,compartment) %>% 
    summarise(mean_rpkm = mean(rpkm),color_gene_name) %>% 
    unique() %>% 
    pivot_wider(names_from = c('compartment','nmdi'),
                values_from = 'mean_rpkm',
                values_fill = 0) %>% 
    ungroup() %>% 
    mutate(log2rpkm_baseline = log2(nucleus_ctrl / cytosol_ctrl)) %>% 
    mutate(log2rpkm_nmd = log2(nucleus_nmd / cytosol_nmd)) %>% 
    mutate(log2rpkm_nuc_change_nmd = log2(nucleus_nmd / nucleus_ctrl)) %>% 
    mutate(log2rpkm_cyto_change_nmd = log2(cytosol_nmd / cytosol_ctrl)) %>% 
    mutate(change_in_local = (log2rpkm_nmd - log2rpkm_baseline)) %>% 
    mutate(delta_log2_nc = (log2rpkm_cyto_change_nmd - log2rpkm_nuc_change_nmd)) %>% 
    ggplot(aes(x = log2rpkm_baseline,colour = color_gene_name)) + 
    geom_rect(aes(xmin = -Inf, xmax = 0, ymin = -Inf, ymax = 1), fill = "lightblue", alpha = 0.2, inherit.aes = FALSE) +  # Bottom-left
    geom_rect(aes(xmin = 0, xmax = Inf, ymin = -Inf, ymax = 1), fill = "moccasin", alpha = 0.2, inherit.aes = FALSE)     +# Top-right
    stat_ecdf(size = 2) +
    xlab('Nuclear / Cytoplasmic Ratio RPKM') +
    scale_color_manual(values = c("#3288BD","#D53E4F")) +
    theme_classic() +
    labs(color = "SMG1i pseudo-bulk") +
    theme(axis.text  = element_text(size = 18),
          axis.title = element_text(size = 18),
          legend.text = element_text(size = 18)) +
    theme(legend.position = 'bottom')
```

# localization of cryptic isoform compare to localization of parent gene - Expression in KD 


```{r}


cryptics =(pseudo_bulk_plot_slope$filtered_table %>% 
                                             ungroup() %>% 
                                             filter(mean_psi_per_lsv_junction >= 0.10))$paste_into_igv_junction

cryptic_rpkm = sj_star_counts %>% 
    filter(paste_into_igv_junction %in% cryptics) %>% 
    # left_join(librar_sizes) %>% 
    # mutate(rpkm = (uniquely_mapping / `Uniquely mapped`) * 10^6) %>% 
    mutate(rpkm = (uniquely_mapping / exonic_reads) * 10^6) %>% 
    left_join(pseudo_bulk_plot_slope$filtered_table %>% distinct(paste_into_igv_junction,color_gene_name,gene_name))  %>% 
    filter(tdp_kd == 'dox') %>% 
    group_by(paste_into_igv_junction,nmdi,compartment) %>% 
    summarise(mean_rpkm = mean(rpkm),color_gene_name,gene_name) %>% 
    unique() %>% 
    pivot_wider(names_from = c('compartment','nmdi'),
                values_from = 'mean_rpkm',
                values_fill = 0) %>% 
    ungroup() %>% 
    left_join(pseudo_bulk_plot_slope$filtered_table %>% distinct(paste_into_igv_junction,lsv_id)) %>% 
    ungroup()

these_lsv = pseudo_bulk_plot_slope$filtered_table %>% 
        ungroup() %>% 
        pull(lsv_id)

ann_junctions = psi_all_groups %>% 
    filter(de_novo_junctions == 0) %>% 
    filter(lsv_id %in% these_lsv) %>% 
    ungroup() %>% 
    distinct(lsv_id,paste_into_igv_junction) %>% 
    dplyr::rename(annotated_junction = paste_into_igv_junction)
    
annotated_rpkm = sj_star_counts %>%
    filter(paste_into_igv_junction %in% ann_junctions$annotated_junction) %>%
    # left_join(librar_sizes) %>%
    # mutate(rpkm = (uniquely_mapping / `Uniquely mapped`) * 10^6) %>%
    mutate(rpkm = (uniquely_mapping / exonic_reads) * 10^6) %>%
    filter(tdp_kd == 'dox') %>%
    left_join(ann_junctions,by = c('paste_into_igv_junction' = "annotated_junction")) %>%
    group_by(lsv_id,nmdi,compartment) %>%
    summarise(mean_rpkm = mean(rpkm),lsv_id) %>%
    unique() %>%
    pivot_wider(names_from = c('compartment','nmdi'),
                values_from = 'mean_rpkm',
                values_fill = 0,
                names_prefix = 'annotated_')

baseline_rpkm_plot = cryptic_rpkm %>% 
    left_join(annotated_rpkm) %>% 
    mutate(log2_annotated = log2((annotated_nucleus_ctrl)/ (annotated_cytosol_ctrl))) %>% 
    mutate(log2rpkm_baseline = log2((nucleus_ctrl)/ (cytosol_ctrl))) %>% 
    mutate(log2_cryptic_nuclear_bias = log2rpkm_baseline - log2_annotated) %>% 
    group_by(gene_name) %>% 
    mutate(color_gene_name = ifelse(any(color_gene_name == 'NMD rescued'),'NMD rescued', color_gene_name)) %>% 
    mutate(plot_name = ifelse(log2rpkm_baseline == max(log2rpkm_baseline) &gene_name %in% gene_list_slope ,gene_name,NA_character_)) %>% 
    ggplot(aes(x = log2_annotated, y = log2rpkm_baseline)) +
    geom_rect(aes(xmin = -Inf, xmax = 0, ymin = -Inf, ymax = 0), fill = "lightblue", alpha = 0.2, inherit.aes = FALSE) +  # Bottom-left
    # geom_rect(aes(xmin = 0, xmax = Inf, ymin = -Inf, ymax = 0), fill = "lightgreen", alpha = 0.2, inherit.aes = FALSE)  +# Bottom-right
    # geom_rect(aes(xmin = -Inf, xmax = 0, ymin = 0, ymax = Inf), fill = "mistyrose", alpha = 0.2, inherit.aes = FALSE)   +# Top-left
    geom_rect(aes(xmin = 0, xmax = Inf, ymin = 0, ymax = Inf), fill = "moccasin", alpha = 0.2, inherit.aes = FALSE)     +# Top-right
    geom_point() +
    geom_smooth(method = 'lm',colour = 'black') +
    stat_cor(  method = "spearman",cor.coef.name = c("rho"),p.accuracy = 0.001,size = 6) +
    facet_wrap(~color_gene_name) +
    theme_classic() +
    geom_hline(yintercept = 0,linetype = 'dotted') +
    geom_vline(xintercept = 0,linetype = 'dotted') +
    ylab('Log2 Nuclear / Cytoplasmic cryptic junction RPKM') +
    xlab('Log2 Nuclear / Cytoplasmic annotated junction RPKM') +
    geom_text_repel(aes(label = plot_name),bg.colour = 'white') +
    theme(text = element_text(size = 18))




    
```

# localization of cryptic isoform after NMDi 

```{r}
nmdi_rpkm_plot = cryptic_rpkm %>% 
    left_join(annotated_rpkm) %>% 
    mutate(log2_annotated = log2((annotated_nucleus_nmd)/ (annotated_cytosol_ctrl))) %>% 
    mutate(log2rpkm_nmd = log2((nucleus_nmd)/ (cytosol_nmd))) %>% 
    group_by(gene_name) %>% 
    mutate(color_gene_name = ifelse(any(color_gene_name == 'NMD rescued'),'NMD rescued', color_gene_name)) %>% 
    mutate(plot_name = ifelse(log2rpkm_nmd == max(log2rpkm_nmd) &gene_name %in% gene_list_slope ,gene_name,NA_character_)) %>% 
    ggplot(aes(x = log2_annotated, y = log2rpkm_nmd)) +
    geom_rect(aes(xmin = -Inf, xmax = 0, ymin = -Inf, ymax = 0), fill = "lightblue", alpha = 0.2, inherit.aes = FALSE) +  # Bottom-left
    # geom_rect(aes(xmin = 0, xmax = Inf, ymin = -Inf, ymax = 0), fill = "lightgreen", alpha = 0.2, inherit.aes = FALSE)  +# Bottom-right
    # geom_rect(aes(xmin = -Inf, xmax = 0, ymin = 0, ymax = Inf), fill = "mistyrose", alpha = 0.2, inherit.aes = FALSE)   +# Top-left
    geom_rect(aes(xmin = 0, xmax = Inf, ymin = 0, ymax = Inf), fill = "moccasin", alpha = 0.2, inherit.aes = FALSE)     +# Top-right
    geom_point() +
    geom_smooth(method = 'lm',color = 'black') +
    stat_cor(  method = "spearman",cor.coef.name = c("rho"),p.accuracy = 0.001,size = 6) +
    facet_wrap(~color_gene_name) +
    theme_classic() +
    geom_hline(yintercept = 0,linetype = 'dotted') +
    geom_vline(xintercept = 0,linetype = 'dotted') +
    ylab('Log2 Nuclear / Cytoplasmic cryptic junction RPKM\nNMDi') +
    xlab('Log2 Nuclear / Cytoplasmic annotated junction RPKM\nNMDi') +
    geom_text_repel(aes(label = plot_name),bg.colour = 'white') +
    theme(text = element_text(size = 18))

ggarrange(baseline_rpkm_plot,nmdi_rpkm_plot)  


```


# Sensitivity/Detection rate of cryptics in tissue by NMD sensitivity -?

```{r}

nygc = fread("~/Desktop/nygc_postmortem_summary.csv")
rimod = fread("~/Desktop/rimod_post.csv")
summary_post = rbind(nygc,rimod)

chx_summary_postmortem = generate_summary_table(summary_post,
                       wide_chx,
                       category_column = 'color_gene_name')
fisher_chx = run_fisher_test(chx_summary_postmortem)

chx_detection_plots = make_final_plot(chx_summary_postmortem,fisher_chx)

upf1_summary_postmortem = generate_summary_table(summary_post,
                       wide_upf1 )
fisher_upf1= run_fisher_test(upf1_summary_postmortem)

upf1_detection_plots = make_final_plot(upf1_summary_postmortem,fisher_upf1)


smg1_summary_postmortem = generate_summary_table(summary_post,
                       wide_smg1 %>% filter(smg1_cryptic != 'Dark cryptic'))
fisher_smg1 = run_fisher_test(smg1_summary_postmortem)

smg1_detection_plots = make_final_plot(smg1_summary_postmortem,fisher_smg1)



# Create the individual plots with titles
# Top row: remove x-axis labels and ticks, turn off legends
chx_detection <- chx_detection_plots$detection_plot + 
    ggtitle('CHX Detection Rate') + 
    theme(text = element_text(size = 18),
          plot.title = element_text(size = 15)) +
    guides(x = guide_axis(n.dodge = 2)) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

upf1_detection <- upf1_detection_plots$detection_plot + 
    ggtitle('UPF1 KD Detection Rate') +
    theme(text = element_text(size = 18),
          plot.title = element_text(size = 15)) +
    guides(x = guide_axis(n.dodge = 2)) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

smg1_detection <- smg1_detection_plots$detection_plot + 
    ggtitle('SMG1i Detection Rate') +
    theme(text = element_text(size = 18),
          plot.title = element_text(size = 15)) +
    guides(x = guide_axis(n.dodge = 2)) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

# Bottom row: keep x-axis labels and ticks, use dodge position for overlapping labels
chx_selectivity <- chx_detection_plots$selectivity_plot + 
    ggtitle('CHX Selectivity Rate') +
    theme(text = element_text(size = 18),
          plot.title = element_text(size = 15)) +
    guides(x = guide_axis(n.dodge = 2)) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

upf1_selectivity <- upf1_detection_plots$selectivity_plot + 
    ggtitle('UPF1 KD Selectivity Rate') +
    theme(text = element_text(size = 18),
          plot.title = element_text(size = 15)) +
    guides(x = guide_axis(n.dodge = 2)) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))


smg1_selectivity <- smg1_detection_plots$selectivity_plot + 
    ggtitle('SMG1i Selectivity Rate') +
    theme(text = element_text(size = 18),
          plot.title = element_text(size = 15)) +
    guides(x = guide_axis(n.dodge = 2)) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))



arranged_plots_detect <- grid.arrange(
    chx_detection, upf1_detection, smg1_detection,
    ncol = 3, nrow = 1
)
arranged_plots_sel <- grid.arrange(
    chx_selectivity, upf1_selectivity, smg1_selectivity,
    ncol = 3, nrow = 1
)

```



# Sensitivity/Detection rate of cryptics in tissue by localization?

```{r sensitivity_by_localization}

my_table = cryptic_rpkm %>% 
    left_join(annotated_rpkm) %>% 
    mutate(log2rpkm_nmd = log2((nucleus_nmd  + 0.001)/ (cytosol_nmd  + 0.001))) %>% 
    mutate(log2rpkm_baseline = log2((nucleus_ctrl + 0.001)/ (cytosol_ctrl +0.001))) %>% 
    mutate(log2rpkm_cytoshift = log2((cytosol_nmd + 0.001)/ (cytosol_ctrl +0.001))) %>% 
    mutate(baseline_localization = case_when(log2rpkm_baseline >= 0 ~ "cytoplasmic",
                                             log2rpkm_baseline <= 0 ~ "nuclear")) %>% 
    mutate(nmd_localization = case_when(log2rpkm_nmd >= 0 ~ "cytoplasmic",
                                         log2rpkm_nmd <= 0 ~ "nuclear")) %>% 
    mutate(nmd_cytoshift = case_when(log2rpkm_cytoshift >= 0.5 ~ "cytoplasmic shift NMDi",
                                         log2rpkm_cytoshift < 0.5 ~ "unshifted")) 


nmd_localization_effect = generate_summary_table(summary_post,
                       my_table,
                       category_column = 'baseline_localization') 

fisher_nmd_localization_effect = run_fisher_test(nmd_localization_effect,category_column = 'baseline_localization')

nmd_localization_effect_plots = make_final_plot(nmd_localization_effect,
                                                fisher_nmd_localization_effect,
                                                category_column = 'baseline_localization',
                                        color_choices = c("lightblue","moccasin"))
print(nmd_localization_effect_plots$detection_plot + ggtitle('Localization at baseline'))
print(nmd_localization_effect_plots$selectivity_plot+ ggtitle('Localization at baseline'))

#nmd localization
nmd_localization_effect = generate_summary_table(summary_post,
                       my_table,
                       category_column = 'nmd_localization') 

fisher_nmd_localization_effect = run_fisher_test(nmd_localization_effect,category_column = 'nmd_localization')

nmd_localization_effect_plots = make_final_plot(nmd_localization_effect,
                                                fisher_nmd_localization_effect,
                                                category_column = 'nmd_localization',
                                        color_choices = c("lightblue","moccasin"))
print(nmd_localization_effect_plots$detection_plot + ggtitle('Localization under NMDi'))
print(nmd_localization_effect_plots$selectivity_plot+ ggtitle('Localization under NMDi'))

#effective of NMD inhibition on localization
nmd_localization_effect = generate_summary_table(summary_post,
                       my_table,
                       category_column = 'nmd_cytoshift') 

fisher_nmd_localization_effect = run_fisher_test(nmd_localization_effect,category_column = 'nmd_cytoshift')

nmd_localization_effect_plots = make_final_plot(nmd_localization_effect,
                                                fisher_nmd_localization_effect,
                                                category_column = 'nmd_cytoshift',
                                        color_choices = c("lightblue","moccasin"))
print(nmd_localization_effect_plots$detection_plot + ggtitle('NMDi effect on localization'))
print(nmd_localization_effect_plots$selectivity_plot+ ggtitle('NMDi effect on localization'))

``` 




## Dose-response curves in SH-SY5Y and SK-N-BE(2) 

## Figure 2 - NMD

### Are NMD sensitive events up/downregulated?
```{r nmd_up_down_chx}

chx_rna_full = fread(here::here('data',"chx_all_experiments_deseq2.csv"))
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))

wide_chx = big_data_chx |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_chx = Control_Control < 0.05 &  Control_TDP43KD > 0.1) |> 
    mutate(chx_cryptic = Control_Control < 0.05 &  Cycloheximide_TDP43KD > 0.1) |> 
    mutate(chx_effect = case_when((Cycloheximide_TDP43KD - Control_TDP43KD) > 0.05 ~ "NMD rescued",
                                  Control_TDP43KD > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_chx %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id))

chx_dark = wide_chx |> 
    filter(normal_cryptic_chx == FALSE & chx_cryptic == TRUE & chx_effect == "NMD rescued") |> 
    pull(paste_into_igv_junction) |> unique()

wide_chx = wide_chx %>% 
    mutate(chx_effect = ifelse(paste_into_igv_junction %in% chx_dark, "Dark cryptic",chx_effect))

chx_rna_full %>% 
    left_join(wide_chx) %>% 
    filter(!is.na(chx_effect)) %>% 
    filter(experiment == "Control_TDP43KD|Control_Control") %>% 
    group_by(ensgene) %>% 
    mutate(max_psi = max(Cycloheximide_TDP43KD)) %>% 
    ungroup() %>% 
    mutate(bin_max = cut_interval(max_psi,n = 2)) %>% 
    ggplot(aes(x = bin_max,
               y = log2fold_change,fill = chx_effect)) + 
    geom_hline(yintercept = 0,linetype = 'dotted') +
    geom_boxplot(position = position_dodge(preserve = "single")) +
    # stat_summary(fun.data = give.n, 
    #              geom = "text",
    #              fun.y = median,
    #              position = position_dodge(width = 0.75)) +
    theme_classic() +
    xlab("Max PSI NMDi + TDP-43 KD") +
    scale_fill_manual(values = c("#0D1821","#3288BD","#D53E4F","#BFCC94")) + 
    ylab("Log2 Fold Change TDP-43 KD SH-SY5Y")

   
```



# Figure 3 - Dose-response curves in SH-SY5Y and SK-N-BE(2)


```{r, echo = FALSE, warning=FALSE, message=FALSE}
master_table <- read.csv(here::here("data/master_table.csv"))
master_table <- master_table %>%
    mutate(tdp43_sensitivity_be2 = ifelse(tdp43_sensitivity_be2 == "","Not found",tdp43_sensitivity_be2)) %>% 
    mutate(tdp43_sensitivity_sh = ifelse(tdp43_sensitivity_sh == "","Not found",tdp43_sensitivity_sh)) %>% 
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "Not found" ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_sh == "Not found" ~ tdp43_sensitivity_be2,
                                tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                                tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late',
                                tdp43_sensitivity_sh == 'Intermediate'~  'Intermediate')) %>% 
        mutate(new_cate = ifelse(is.na(new_cate),"Ambiguous",new_cate)) 

#1A and 1B
levels_sy5y_qpcr <- read.csv(here::here('data',"qPCR_RNA.csv"))
levels_sy5y_qpcr = levels_sy5y_qpcr %>%
    filter(Gene == 'TDP-43')
levels_dz_qpcr <- read.csv(here::here('data','qpcr_dz.csv'))
levels_dz_qpcr = levels_dz_qpcr %>%
  filter(experiment == "new_new")
#1C and 1D
diff_expr_sh <- read.csv(here::here('data', 'diff_expr_sh.csv'))
diff_expr_sk <- read.csv(here::here('data', 'diff_expr_sk.csv'))
#1E, 1F, and 1G
master_table <- read.csv(here::here("data/master_table.csv"))

#####
print("1A - Dose-response curves for SH-SY5Y and SK-N-BE(2)")
a = levels_sy5y_qpcr %>%
    filter(Gene == 'TDP-43') %>% 
    mutate(source = c(rep(unique(diff_expr_sh$source),each = 3),rep(0,3))) %>% 
  ggplot(aes(x = as.character(source), y = Value)) +
    geom_violin() +
  geom_jitter() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  theme_classic() +
  theme(text = element_text(size = 18))
c <- levels_dz_qpcr %>%
  filter(experiment == "new_new") %>%
  ggplot(aes(x = as.character(sample), y = 100*value)) +
    geom_violin() +
  geom_jitter() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  #facet_wrap(facets = vars(experiment)) +
  theme_classic() +
  theme(text = element_text(size = 18))
a+c
#ggsave("~/Desktop/curves_qpcr.svg", width = 10, height = 6, dpi = 300)

```

```{r de_by_levels}

avg_tdp_level_sh = levels_sy5y_qpcr %>% 
    group_by(Treatment) %>% 
    summarize(m = mean(Value)) %>% pull(m) %>% head(5)

print("1B - Differential expression in SH-SY5Y and SK-N-BE(2)")
plot_de <- diff_expr_sh %>%
    filter(symbol != "RNA5-8SN2") %>% 
  mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
  mutate(alpha_gene_name = ifelse(padj < 0.05, "sig", "non_sig")) 
plot_de_stmn_unc = plot_de %>% filter(symbol %in% c("STMN2","UNC13A"))
plot_de_meta <- as.data.frame(table(plot_de$color_gene_name, plot_de$alpha_gene_name, plot_de$source))
e <- plot_de %>%
    ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
    geom_point(aes(color = color_gene_name, 
                   alpha = alpha_gene_name, 
                   shape = alpha_gene_name, 
                   fill = alpha_gene_name),
               size = 0.2,
               position = position_jitter(width = 0, height = 0.4),
               show.legend = F) +
    geom_text(data = plot_de_meta[plot_de_meta$Var2 == "sig",], 
              aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
              size = 5,
              show.legend = F) +
    geom_vline(xintercept = 0) +
    coord_flip() +
    scale_color_manual(values = c(down = "#2B4162",up = "#721817",STMN2 = "#E31A1C",UNC13A ="#FB9A99"),breaks = c("STMN2","UNC13A")) +
    scale_alpha_manual(values = c(0.2, 1)) +
    scale_shape_manual(values = c(21, 21)) +  # Use shape 21 for both
    scale_fill_manual(values = c("grey", "black")) +  # Grey for non-sig, black for sig
    xlim(c(-10, 10)) +
    xlab('Gene expression Log2 FC compared to controls') +
    ylab('TDP-43 knockdown level') +
    scale_y_discrete(labels = paste0(c("+", "++", "+++", "++++", "+++++"),'\n',round(avg_tdp_level_sh,0),rep(" %",5))) +
    theme_classic() +
    theme(text = element_text(size = plot_text_size), plot.title = element_text(hjust = 0.5)) +
    ggtitle("SH-SY5Y") + 
    geom_point(inherit.aes = FALSE,
              aes(x = log2FoldChange, y = as.factor(source),color = symbol),
              data = plot_de_stmn_unc) + 
        labs(color = "") +
      theme(text = element_text(size = 18))


avg_be2_qpcr = levels_dz_qpcr %>%
  filter(experiment == "new_new") %>% 
    group_by(sample) %>% 
    summarize(m = mean(value)) %>% pull(m) %>% tail(7) * 100

plot_de_dz <- diff_expr_sk %>%
    filter(symbol != "RNA5-8SN2") %>% 
  mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
  mutate(alpha_gene_name = ifelse(padj < 0.05, "sig", "non_sig"))

plot_de_stmn_unc_dz = plot_de_dz %>% filter(symbol %in% c("STMN2","UNC13A"))

plot_de_dz_meta <- as.data.frame(table(plot_de_dz$color_gene_name, plot_de_dz$alpha_gene_name, plot_de_dz$source))
f <- plot_de_dz %>%
    ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
    geom_point(aes(color = color_gene_name, 
                   alpha = alpha_gene_name, 
                   shape = alpha_gene_name, 
                   fill = alpha_gene_name),
               size = 0.2,
               position = position_jitter(width = 0, height = 0.4),
               show.legend = F) +
    geom_text(data = plot_de_dz_meta[plot_de_dz_meta$Var2 == "sig",], 
              aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
              size = 5,
              show.legend = F) +
    geom_vline(xintercept = 0) +
    coord_flip() +
    scale_color_manual(values = c(down = "#2B4162",up = "#721817",STMN2 = "#E31A1C",UNC13A ="#FB9A99"),breaks = c("STMN2","UNC13A")) +
    scale_alpha_manual(values = c(0.2, 1)) +
    scale_shape_manual(values = c(21, 21)) +  # Use shape 21 for both
    scale_fill_manual(values = c("grey", "black")) +  # Grey for non-sig, black for sig
    xlim(c(-10, 10)) +
    xlab('Gene expression Log2 FC compared to controls') +
    ylab('TDP-43 knockdown level') +
    scale_y_discrete(labels = paste0(c("+", "++", "+++", "++++", "+++++","++++++","+++++++"),'\n',round(avg_be2_qpcr,0),rep(" %",7))) +
    theme_classic() +
    theme(text = element_text(size = plot_text_size), plot.title = element_text(hjust = 0.5)) +
    ggtitle("SK-N-BE(2)") +
    geom_point(inherit.aes = FALSE,
          aes(x = log2FoldChange, y = as.factor(source),color = symbol),
          data = plot_de_stmn_unc_dz) +
    labs(color = "") +
      theme(text = element_text(size = 18))

ggarrange(e, f, common.legend = TRUE, legend = 'bottom')
#ggsave("~/Desktop/de_curves.png", width = 8, height = 4, dpi = 300)

```

```{r late_early_spaghetti_plot}
print("1C - Spaghetti plot in SH-SY5Y and SK-N-BE(2)")
list_gene = c("STMN2", "UNC13A", "HDGFL2", "AARS1","RANBP1",'PHF2')

avg_tdp_level_sh = levels_sy5y_qpcr %>% 
    group_by(Treatment) %>% 
    summarize(m = mean(Value)) %>% pull(m) %>% head(5)
avg_tdp_level_sh = c(100,avg_tdp_level_sh)


big_data_filtereds_early_late_sh <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_sh) %>% 
    dplyr::select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_sh,max_psi_sh)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox025)) %>% 
    distinct(gene, .keep_all = T) %>% 
    filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:12)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox0125", "dox0187", "dox021", "dox025", "dox075"))) %>%
  mutate(label_junction = case_when(name == "dox025" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_sh$tdp43_sensitivity_sh)
table(big_data_filtereds_early_late_sh$max_psi_sh)
GG <- big_data_filtereds_early_late_sh %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_sh), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = plot_sensitivity_colors) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = #c("-", "+", "++", "+++", "++++", "+++++")
                   paste0(round(avg_tdp_level_sh,0),rep("%",5))) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_sh) + 
  theme_classic()

avg_be2_qpcr = levels_dz_qpcr %>%
  filter(experiment == "new_new") %>% 
    group_by(sample) %>% 
    summarize(m = mean(value)) %>% pull(m) %>% tail(7) * 100
avg_be2_qpcr = c(100,avg_be2_qpcr)

big_data_filtereds_early_late_be2 <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_be2) %>% 
    dplyr::select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox002,dox003,dox004,dox005,dox0075,dox01,dox1,
             #dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_be2,max_psi_be2)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox01)) %>% 
    distinct(gene, .keep_all = T) %>% 
  filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:14)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox002","dox003","dox004","dox005","dox0075","dox01","dox1"))) %>%
  mutate(label_junction = case_when(name == "dox01" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_be2$tdp43_sensitivity_be2)
table(big_data_filtereds_early_late_be2$max_psi_be2)

HH = big_data_filtereds_early_late_be2 %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_be2), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = plot_sensitivity_colors) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = #c("-", "+", "++", "+++", "++++", "+++++", "++++++", "+++++++")
                   paste0(round(avg_be2_qpcr,0),rep("%",7))) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_be2) + 
  theme_classic()
GG + HH
#ggsave("~/Desktop/earlylate.svg", width = 18, height = 6, dpi = 300)
```

```{r alluvial_early_late_new}
# Create the alluvial plot
print("1D - alluvial plot")
master_table <- read.csv(here::here("data/master_table.csv"))
master_table <- master_table %>%
    mutate(tdp43_sensitivity_be2 = ifelse(tdp43_sensitivity_be2 == "","Not found",tdp43_sensitivity_be2)) %>% 
    mutate(tdp43_sensitivity_sh = ifelse(tdp43_sensitivity_sh == "","Not found",tdp43_sensitivity_sh)) %>% 
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "Not found" ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_sh == "Not found" ~ tdp43_sensitivity_be2,
                                tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                                tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late',
                                tdp43_sensitivity_sh == 'Intermediate'~  'Intermediate')) %>% 
        mutate(new_cate = ifelse(is.na(new_cate),"Ambiguous",new_cate)) 

tmp = master_table %>% 
    filter(tdp43_sensitivity_be2 != "Not found") %>% 
    filter(tdp43_sensitivity_sh != "Not found")
df_long <- to_lodes_form(tmp, key = "axis", axes = c("tdp43_sensitivity_sh", "tdp43_sensitivity_be2"))

df_long %>% 
    filter(junction_for_clustering == 'yes') %>% 
    mutate(axis = ifelse(axis == "tdp43_sensitivity_sh", "SH-SY5Y", "SK-N-BE(2)")) %>% 
    ggplot(
       aes(x = axis,
           stratum = stratum, 
           alluvium = alluvium,
           fill = stratum, 
           label = stratum)) +
    scale_fill_manual(values = plot_sensitivity_colors) +
    geom_flow(stat = "alluvium", 
              lode.guidance = "frontback") +
    geom_stratum() +
    theme(legend.position = "bottom") +
    facet_wrap(~cryptic,
               labeller = as_labeller(
               c(no = "Alternative splicing", yes = "Cryptic splicing"))) + 
    theme_classic() +
    geom_text(stat = "stratum",
              aes(label = after_stat(count))) +
    labs(fill = element_blank(),
         x = element_blank(),
         y = 'N junctions') 
#ggsave("~/Desktop/alluvial.svg", width = 6, height = 4, dpi = 300)
```

```{r scatterplot}
master_table <- read.csv(here::here("data/master_table.csv"))

master_table <- master_table %>%
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "" ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_sh == "" ~ tdp43_sensitivity_be2,
                                tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                                tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late',
                                tdp43_sensitivity_sh == 'Intermediate'~  'Intermediate')) %>% 
    mutate(new_cate = ifelse(is.na(new_cate),"Ambiguous",new_cate))


print("1E - Concordance between SH and SK scatterplot")
#scatterplot from upgrade
master_table %>% 
    ggplot(aes(x = dox075, y = dox1)) +
    geom_point() +
    geom_smooth(method = 'lm', se = F) +
    stat_cor() +
    xlab("Max PSI in SH-SY5Y") +
    ylab("Max PSI in SK-N-BE(2)") +
    theme_classic()
#ggsave("~/Desktop/scatter.svg", width = 6, height = 4, dpi = 300)

```

```{r makewidechx_1}
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))
big_data_chx = big_data_chx %>% 
    separate(paste_into_igv_junction,remove = FALSE,into = c("chr",'start','end'),
             convert = TRUE)


wide_chx = big_data_chx |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_chx = Control_Control < 0.05 &  Control_TDP43KD > 0.1) |> 
    mutate(chx_cryptic = Control_Control < 0.05 &  Cycloheximide_TDP43KD > 0.1) |> 
    mutate(chx_effect = case_when((Cycloheximide_TDP43KD - Control_TDP43KD) > 0.05 ~ "NMD rescued",
                                  Control_TDP43KD > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_chx %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id))


chx_dark = wide_chx |> 
    filter(normal_cryptic_chx == FALSE & chx_cryptic == TRUE & chx_effect == "NMD rescued") |> 
    pull(paste_into_igv_junction) |> unique()

wide_chx = wide_chx %>% 
    mutate(chx_effect = ifelse(paste_into_igv_junction %in% chx_dark, "Dark cryptic",chx_effect))
```

```{r makewideupf1_1}
big_data_upf1 <- read.csv(here::here('data','nmd_or_not_upf1.csv'))
big_data_upf1 = big_data_upf1 %>% 
    separate(paste_into_igv_junction,remove = FALSE,into = c("chr",'start','end'),
             convert = TRUE)

wide_upf1 = big_data_upf1 |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_upf1 = ctrl_ctrl < 0.05 &  TDP43_ctrl > 0.1) |> 
    mutate(upf1_cryptic = ctrl_ctrl < 0.05 &  TDP43_UPF1 > 0.1) |> 
    mutate(upf1_effect = case_when((TDP43_UPF1 - TDP43_ctrl) > 0.05 ~ "NMD rescued",
                                  TDP43_ctrl > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_upf1 %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id)) 

upf1_dark = wide_upf1 |> 
    filter(normal_cryptic_upf1 == FALSE & upf1_cryptic == TRUE & upf1_effect == "NMD rescued") |> 
    pull(paste_into_igv_junction) |> unique()

wide_upf1 = wide_upf1 %>% 
    mutate(upf1_effect = ifelse(paste_into_igv_junction %in% upf1_dark, "Dark cryptic",upf1_effect))
```

```{r EarlyLateSushi}
sh_nmd_sens = wide_chx %>% 
    distinct(paste_into_igv_junction,chx_effect) %>% 
    left_join(master_table) %>% 
    filter(!is.na(tdp43_sensitivity_sh)) %>% 
    filter(tdp43_sensitivity_sh != "") %>% 
    filter(junction_for_clustering == 'yes') %>% 
    mutate(chx_effect = ifelse(chx_effect == "Dark cryptic","NMD rescued",chx_effect)) %>% 
    group_by(chx_effect,tdp43_sensitivity_sh) %>% 
    reframe(n = n_distinct(paste_into_igv_junction))

ggplot(sh_nmd_sens) +
    aes(x = tdp43_sensitivity_sh, fill = chx_effect, weight = n,by = as.factor(tdp43_sensitivity_sh)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'CHX treatment') +
    scale_fill_manual(values = c("#3288BD","#D53E4F","#BFCC94")) + 
    theme(legend.position = 'bottom', legend.text.position = 'bottom') 
#ggsave("~/Desktop/sh_nmd_sens.svg", width = 5, height = 4, dpi = 300)

sh_nmd_sens %>% 
    filter(chx_effect != "PSI > 95% at baseline") %>% 
    pivot_wider(names_from = 'chx_effect',values_from = 'n') %>% 
    column_to_rownames("tdp43_sensitivity_sh") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()
```

```{r curves nmd}
print("3F - TDP-43 dependent loss of dark cryptics genes")
####read in data!!
big_delta_de <- read.csv(here::here('data','big_delta_de.csv'))

for_plot <- big_delta_de[,c(4,8,10,11)] %>%
  dplyr::filter(symbol == "STMN2" | symbol == "AARS1" | symbol == "HDGFL2" | symbol == "UNC13A" |
                  symbol == "SYNE1" | symbol == "CYFIP2")
for_plot_0 <- for_plot %>% 
  distinct(symbol, .keep_all = T) %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
for_plot <- rbind(for_plot_0,for_plot) %>% 
  mutate(symbol = factor(symbol, levels = c("AARS1", "HDGFL2", "SYNE1", "CYFIP2", "UNC13A", "STMN2")))

big_delta_de_tdp <- big_delta_de[,c(4,8,10,11)] %>% 
  dplyr::filter(symbol == "TARDBP")
for_plot_0_tdp <- big_delta_de_tdp[1,] %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
big_delta_de_tdp <- rbind(for_plot_0_tdp,big_delta_de_tdp)

for_plot$source <- as.factor(for_plot$source)
a <- for_plot %>%   
  ggplot() +
  geom_col(aes(x = as.character(source), y = log2FoldChange, fill = symbol), data = big_delta_de_tdp, alpha = 0.3, show.legend = F) +
  geom_point(aes(x = source, y = log2FoldChange, color = symbol, group = symbol), show.legend = F) +
  geom_line(aes(x = source, y = log2FoldChange, color = symbol, group = symbol), show.legend = F) +
  scale_fill_manual(values = "black") +
  scale_color_manual(values = c("#A6CEE3", "#1F78B4", "#FF7F00", "#FDBF6F", "#FB9A99", "#E31A1C"))+ # "#B2DF8A" "#33A02C" "#CAB2D6" "#6A3D9A")) +
  scale_y_continuous(limits = c(-6.2,0.6)) +
  labs(x = "TDP-43 knockdown level", y = "Log2 Fold Change compared to control", color = "", fill = "Gene") +
  scale_x_discrete(labels = paste0(round(avg_tdp_level_sh,0),rep(" %",6))) +
  theme_classic() #+
# theme(axis.text.x = element_text(angle = 90))

big_delta_de_dz <- read.csv(here::here('data','big_delta_de_dz.csv'))

big_delta_de_dz_tdp <- big_delta_de_dz[,c(4,8,10,11)] %>% 
  dplyr::filter(symbol == "TARDBP")
for_plot_dz_0_tdp <- big_delta_de_dz_tdp[1,] %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
big_delta_de_dz_tdp <- rbind(for_plot_dz_0_tdp,big_delta_de_dz_tdp)

for_plot_dz <- big_delta_de_dz[,c(4,8,10,11)] %>%
  dplyr::filter(symbol == "STMN2" | symbol == "AARS1" | symbol == "HDGFL2" | symbol == "UNC13A" |
                  symbol == "SYNE1" | symbol == "CYFIP2")
for_plot_dz_0 <- for_plot_dz %>% 
  distinct(symbol, .keep_all = T) %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
for_plot_dz <- rbind(for_plot_dz_0,for_plot_dz) %>% 
  mutate(symbol = factor(symbol, levels = c("AARS1", "HDGFL2", "SYNE1", "CYFIP2", "UNC13A", "STMN2")))

for_plot_dz$source <- as.factor(for_plot_dz$source)
b <- for_plot_dz %>%   
  ggplot() +
  geom_col(aes(x = as.character(source), y = log2FoldChange, fill = symbol), data = big_delta_de_dz_tdp, alpha = 0.3) +
  geom_point(aes(x = source, y = log2FoldChange, color = symbol, group = symbol)) +
  geom_line(aes(x = source, y = log2FoldChange, color = symbol, group = symbol)) +
  scale_fill_manual(values = "black") +
  scale_color_manual(values = c("#A6CEE3", "#1F78B4", "#FF7F00", "#FDBF6F", "#FB9A99", "#E31A1C"))+ # "#B2DF8A" "#33A02C" "#CAB2D6" "#6A3D9A")) +
  scale_y_continuous(limits = c(-5.2,0.5)) +
  labs(x = "TDP43 knockdown level", y = "Log2 Fold Change compared to control", color = "", fill = "Gene") +
  scale_x_discrete(labels = paste0(round(avg_be2_qpcr,0),rep(" %",8))) +
  theme_classic()

a+b
#ggsave("~/Desktop/caio.png", width = 8, height = 4, dpi = 300)
```

```{r new_cate_creation}
master_table %>% 
    filter(new_cate != "Ambiguous") %>% 
    select(paste_into_igv_junction,cryptic,tdp43_sensitivity_sh,tdp43_sensitivity_be2,new_cate) %>% 
    mutate(cell_contribution = case_when(tdp43_sensitivity_sh == "Not found" ~ "SK-N-BE(2)",
                                         tdp43_sensitivity_be2 == "Not found" ~ "SH-SY5Y",
                                         TRUE ~ "Both cell lines")) %>% 
    group_by(cryptic,new_cate) %>% 
    count(cell_contribution) %>% 
    ggplot(aes(x = new_cate,
               y = n,
               fill = cell_contribution)) + 
    geom_col() + 
    facet_wrap(~cryptic,
               labeller = as_labeller(
               c(no = "Alternative splicing", yes = "Cryptic splicing"))) + 
    theme_classic() +
    labs(fill = element_blank(),
         x = element_blank(),
         y = 'N junctions') +
    theme(legend.position = "bottom") +
    scale_fill_manual(values = c("#7DDE92","#DB504A","#D7CDCC"))
    
master_table %>% 
    filter(new_cate != "Ambiguous") %>% 
    select(paste_into_igv_junction, cryptic, tdp43_sensitivity_sh, tdp43_sensitivity_be2, new_cate) %>% 
    mutate(cell_contribution = case_when(
        tdp43_sensitivity_sh == "Not found" ~ "SK-N-BE(2)",
        tdp43_sensitivity_be2 == "Not found" ~ "SH-SY5Y",
        TRUE ~ "Both cell lines"
    )) %>% 
    group_by(cryptic, new_cate, cell_contribution) %>% 
    summarise(n = n_distinct(paste_into_igv_junction)) %>% 
    mutate(percentage = n / sum(n))  %>% 
    ggplot(aes(x = "", y = percentage, fill = cell_contribution)) + 
    geom_bar(stat = "identity", width = 1) + 
    coord_polar(theta = "y") + 
facet_grid(new_cate ~ cryptic, labeller = as_labeller(
    c(no = "Alternative splicing", yes = "Cryptic splicing",
      Early = "Early", Intermediate = "Intermediate", Late = "Late")
)) + 
    theme_classic() +
    labs(fill = element_blank(), 
         x = element_blank(), 
         y = element_blank()) +
    theme(legend.position = "bottom") +
    scale_fill_manual(values = c("#7DDE92","#DB504A","#D7CDCC")) +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank())

```





## Figure 4 - Features
```{r, echo=FALSE, warning=FALSE, message=FALSE}
get_signif_stars <- function(p_value) {
  if (p_value < 0.001) {
    return("***")
  } else if (p_value < 0.01) {
    return("**")
  } else if (p_value < 0.05) {
    return("*")
  } else {
    return("NS")
  }
}

```

```{r splice_site_strength}
get_significance <- function(df, value, group, category_col, category) {
  test_formula <- as.formula(paste(value, "~", group))
  p_test <- kruskal.test(test_formula, data = df %>% filter(!!sym(category_col) == category))
  #p1_annotation <- paste("p =", format(p1_test$p.value, digits = 3), get_signif_stars(p1_test$p.value))
  return(get_signif_stars(p_test$p.value))
}

print("2A - Splice site strength")
df_5ss_strength <- read.csv(here::here('data', 'maxent/5ss_strength_combined.csv'))
df_3ss_strength <- read.csv(here::here('data', 'maxent/3ss_strength_combined.csv'))
df_ss_strength <- rbind(
  df_5ss_strength  %>%
    mutate(site = "5' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, new_cate, site, ID_exon),
  df_3ss_strength  %>%
    mutate(site = "3' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, new_cate, site, ID_exon)
) %>%
  filter(new_cate %in% c("Early", "Late"))

h <- df_ss_strength %>%
  ggplot(aes(x=site, y = log2(ss_ratio), fill = new_cate)) +
  geom_boxplot() +
  geom_abline(intercept = 0, slope = 0, color = "black", linetype = "dashed") +
  # geom_signif(
  #  y_position = c(max(df_ss_strength$ss_ratio) * 1.05,max(df_ss_strength$ss_ratio) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
  #  annotation = c(get_significance(df_ss_strength,
  #                                  "ss_ratio", "new_cate", "site", "3' splice site"),
  #                 get_significance(df_ss_strength,
  #                                  "ss_ratio", "new_cate", "site", "5' splice site")), tip_length = 0.02
  # ) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  #ylim(-4.1, 2.5) +
  labs(y = "Log2 Ratio MaxEnt score cryptic site vs annotated site", x = "") +
  theme_minimal() +
  theme(legend.position = "top") +
  guides(fill = "none")
h
```

```{r conservation_cell_only}
print("2B - Conservation score for CE and non-ce in combined categories")
cons_score_df <- master_table %>%
  filter(new_cate %in% c("Early", "Late")) %>%
  subset(!is.na(phastCons_alt) & !is.na(phastCons_unchanged)) %>%
  pivot_longer(cols = c(phastCons_alt, phastCons_unchanged), names_to = "category", values_to = "cons_score")

cons_score_df_cryptic <- cons_score_df %>% filter(cryptic == "yes")
cons_score_df_not_cryptic <- cons_score_df %>% filter(cryptic == "no")

a <- cons_score_df_cryptic %>%
  ggplot(aes(x = category, y = cons_score, fill = new_cate)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Novel splice site", "Unchanged splice site")) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_alt"),
                   get_significance(cons_score_df_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_unchanged"))
  ) +
  #facet_wrap(facets = vars(cryptic)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

b <- cons_score_df_not_cryptic %>%
  ggplot(aes(x = category, y = cons_score, fill = new_cate)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Alternatively used splice site", "Unchanged splice site")) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_alt"),
                   get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_unchanged"))
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
a + b
```

```{r exon_in_frame}
result_mrg = fread('data/result_exon_table_merged.csv')
tmp_gr = result_mrg %>% 
    separate(exon_coord,into = c("chr","start","end")) %>% 
    makeGRangesFromDataFrame(,keep.extra.columns = TRUE)

seq_tmp <- getSeq(BSgenome.Hsapiens.UCSC.hg38,tmp_gr)


result_mrg = result_mrg %>% 
    mutate(width = seq_tmp %>% width()) %>% 
    mutate(div_3 = ifelse(width %% 3==0,"In-frame","Out-frame")) 


exon_counts <- result_mrg %>%
    filter(last_exon == FALSE) %>% 
  group_by(new_cate, cryptic, div_3) %>%
  summarise(count = n_distinct(exon_coord)) %>%
  ungroup()

frame_early <- exon_counts %>%
  #filter(cryptic=="yes") %>%
    filter(new_cate == "Early") %>%
    pivot_wider(names_from = div_3, values_from = count, values_fill = 0L) %>%
    select(-new_cate) %>%
    column_to_rownames('cryptic') %>% 
    chisq.test() %>% broom::tidy()

frame_late <- exon_counts %>%
    filter(new_cate == "Late") %>%
    pivot_wider(names_from = div_3, values_from = count, values_fill = 0L) %>%
    select(-new_cate) %>%
    column_to_rownames('cryptic') %>% 
    chisq.test() %>% broom::tidy()

exon_counts %>% 
    filter(new_cate %in% c("Early","Late")) %>%
    ggplot(aes(x = cryptic, y = count, fill = div_3)) +
  geom_bar(stat = "identity", position = "dodge") +
        facet_wrap(~new_cate) +
  geom_signif(data = data.frame(new_cate = c("Early","Late"), div_3=c('In-frame', 'Out-frame')),
              aes(y_position=c(61, 61), xmin=c(0.8, 0.8), xmax=c(2.2, 2.2),
              annotations=c(get_signif_stars(frame_early$p.value), 
                            get_signif_stars(frame_late$p.value))), tip_length=0.02, manual = T) +

  labs(
    #title = "Number of Symmetric and Non-Symmetric Exons by Type",
    x = "",
    y = "Number of exons",
    fill = ""
  ) +
  scale_x_discrete(labels=c("Alternative splicing", "Cryptic splicing"))+
  scale_fill_manual(
    values = plot_inframe_colors,
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size), axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
all_exons <- read.csv("data/inclusion_rates.csv")
all_exons <- all_exons %>%
  mutate(
    div_3 = ifelse((exon_stop-exon_start) %% 3==0,"In-frame","Out-frame")
  )

all_exons <- all_exons %>%
  filter(!is.na(incl_rate))%>%
  mutate(incl_rate_group = cut(incl_rate, 
                                breaks = c(0.01, 0.25, 0.85, 0.99, Inf), 
                                labels = c("1-25%", "25-85%", "85-99%", ">99%")))
                                #labels = c("<0.85", "0.85-0.995", ">0.995")))

plot_cont_table <- na.omit(all_exons) %>%
  group_by(incl_rate_group, div_3) %>%
  summarise(count = n(), .groups = 'drop')

cont_table <- plot_cont_table %>%
    pivot_wider(names_from = div_3, values_from = count)

group_pairs <- list(
  c(1,2), 
  c(1,3),  
  c(1,4)   
)

p_values <- sapply(group_pairs, function(pair) {
  chisq.test(as.matrix(cont_table %>% select(`In-frame`, `Out-frame`))[pair,])$p.value
})

ggplot(plot_cont_table, aes(x = incl_rate_group, y = count, fill = div_3)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "",
       x = "Inclusion rate",
       y = "Number of exons",
       fill = "") +
  theme_classic() +
  scale_fill_manual(
    values = plot_inframe_colors,
  ) +
  geom_signif(comparisons = list(c("1-25%", "25-85%"), 
                                 c("1-25%", "85-99%"), 
                                 c("1-25%", ">99%")),
              annotations = sprintf("p = %.1e", p_values),
              y_position = c(max(plot_cont_table$count)*1.1, 
                             max(plot_cont_table$count)*1.2, 
                             max(plot_cont_table$count)*1.3))
```

```{r kmer sh-sy5y, echo=FALSE, warning=FALSE, message=FALSE}
#C. Kmers enrichment analysis in SH-SY5Y
print("2C - Kmers enrichment analysis in SH-SY5Y")

kmer_size <- 6
yg_motifs <- c('TGTGTG', 'GTGTGT', 'GTGTGC', 'CGTGTG', 'TGCGTG', 'TGTGCG')
ya_motifs <- c('ATGTGT', 'TGTGTA', 'TGCATG', 'GTGTAT', 'TGTATG', 'GTATGT')
aa_motifs <- c('AATGAA', 'TTGAAT', 'ATGAAT', 'GTGTGA', 'TGAATG', 'GAATGA', 'GTGAAT', 'GAATGT')
all_motifs <- c(yg_motifs, ya_motifs, aa_motifs)
categories <- c(rep("YG", 6), rep("YA", 6), rep("AA", 8))
flank_size = 250
flank_region = master_table %>% 
      filter(new_cate %in% c("Early", "Late")) %>%
    mutate(new_max = case_when(max_psi_be2 == max_psi_sh ~ max_psi_sh,
                               max_psi_be2 == "" ~ max_psi_sh,
                               max_psi_sh == "" ~ max_psi_be2,
                               max_psi_be2 == 'Medium' ~ max_psi_sh,
                               max_psi_sh == 'Medium' ~ max_psi_be2)) %>% 
    filter(junc_cat != "exon_skip") %>% 
    filter(cryptic == 'yes') %>% 
    separate(paste_into_igv_junction,into = c("seqnames","og_start","og_end"),convert = TRUE,remove = FALSE) %>% 
    mutate(start = case_when(junc_cat == "acceptor" & strand == "+" ~ og_end - flank_size,
                             junc_cat == "donor" & strand == "+" ~ og_start - flank_size,
                             junc_cat == "acceptor" & strand == "-" ~ og_start - flank_size,
                             junc_cat == "donor" & strand == "-" ~ og_end - flank_size)) %>% 
    mutate(end = case_when(junc_cat == "acceptor" & strand == "+" ~ og_end + flank_size,
                           junc_cat == "donor" & strand == "+" ~ og_start + flank_size,
                           junc_cat == "acceptor" & strand == "-" ~ og_start + flank_size,
                           junc_cat == "donor" & strand == "-" ~ og_end + flank_size)) %>% 
    mutate(name = snakecase::to_snake_case(paste(gene,junc_cat,paste_into_igv_junction,cryptic,new_cate,sep  = "_"))) %>% 
    unique() %>% 
    makeGRangesFromDataFrame(,keep.extra.columns = TRUE)

count_hexamers_granges <- function(granges_obj, motifs = all_motifs, genome = BSgenome.Hsapiens.UCSC.hg38) {
    # Input validation
    if (!is(granges_obj, "GRanges")) {
        stop("Input must be a GRanges object")
    }
    
    if (length(granges_obj) == 0) {
        warning("Empty GRanges object provided")
        return(NULL)
    }
    
    # Extract sequences for all ranges at once (vectorized operation)
    sequences <- getSeq(genome, granges_obj)
    
    # Handle strand information if present
    if ("strand" %in% names(mcols(granges_obj)) || any(strand(granges_obj) != "*")) {
        # Use strand information from GRanges object
        minus_strand <- as.logical(strand(granges_obj) == "-")
        sequences[minus_strand] <- reverseComplement(sequences[minus_strand])
    }
    
    # Count motifs for each sequence (vectorized)
    count_matrix <- sapply(motifs, function(motif) {
        vcountPattern(motif, sequences, max.mismatch = 0)
    })
    
    # Set row names and column names
    rownames(count_matrix) <- names(granges_obj)
    colnames(count_matrix) <- motifs
    
    return(count_matrix)
}

kmer_counts <- count_hexamers_granges(flank_region)

# Convert back to data frame for existing analysis
kmer_df <- as.data.frame(kmer_counts)
kmer_df$type <- mcols(flank_region)$new_cate
kmer_df$cryptic <- mcols(flank_region)$cryptic


summary_counts <- kmer_df %>%
  group_by(type, cryptic) %>%
  summarise(across(all_motifs, sum))

means <- apply(summary_counts[, -c(1,2)], 1, mean)
sds <- apply(summary_counts[, -c(1,2)], 1, sd)
z_scores <- sweep(summary_counts[, -c(1,2)], MARGIN = 1, STATS = means, FUN = "-") / sds
heatmap_data <- as.matrix(z_scores)
rownames(heatmap_data) <- paste(summary_counts$type, ifelse(summary_counts$cryptic == "yes", "Cryptic", "Not cryptic"))

sort_kmers_by_zscore <- function(kmers, z_scores) {
  kmers[order(z_scores[kmers], decreasing = TRUE)]
}

sort_categories_by_zscore <- function(kmers, z_scores) {
  categories[order(z_scores[kmers], decreasing = TRUE)]
}

sorted_all_kmers <- sort_kmers_by_zscore(all_motifs, heatmap_data["Early Cryptic",]+heatmap_data["Late Cryptic",])
sorted_categories <- sort_categories_by_zscore(all_motifs, heatmap_data["Early Cryptic",]+heatmap_data["Late Cryptic",])

annotation <- HeatmapAnnotation(Motif = factor(sorted_categories, levels = c("YG", "YA", "AA")),
                                #show_legend = c(F, F, F),
                                col = list(Motif = c(YG = "#35590F", YA = "#4E72F7", AA = "#B22724")),
                                which = "col")
c <- Heatmap(heatmap_data[c("Early Cryptic","Late Cryptic") , sorted_all_kmers],
        cluster_columns = F,
        cluster_rows = F,
        show_row_names = T,
        show_column_names = T,
        show_row_dend = F,
        bottom_annotation = annotation)
        # col = colorRampPalette(brewer.pal(12, "RdPu"))(100),
        # show_heatmap_legend = FALSE)
c

```

```{r kmer barplot}

```



```{r ug_dense sh-y5y, echo=FALSE, warning=FALSE, message=FALSE}
#D. Density UG motifs by early-late in SH-SY5Y
print("2D and 2E - Density UG motifs by early-late and top-bottom in SH-SY5Y")
sequences <- sapply(seq_along(master_table$paste_into_igv_junction_annotated), function(i) {
  coord <- master_table$paste_into_igv_junction_annotated[i]
  chr <- sub("^(.*):.*$", "\\1", coord)
  range <- as.numeric(unlist(strsplit(sub(".*:(.*)$", "\\1", coord), "-")))
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = chr, start = range[1], end = range[2])

  if (master_table$strand[i] == "-") {
    seq <- reverseComplement(seq)
  }

  return(seq)
})

master_table$longest_ug_region <- Biostrings::vcountPattern('TGTGTG', DNAStringSet(sequences), max.mismatch = 1)

master_table <- master_table %>%
  mutate(
    start = as.numeric(sub(".*:(\\d+)-.*", "\\1", paste_into_igv_junction_annotated)),
    end = as.numeric(sub(".*-(\\d+)", "\\1", paste_into_igv_junction_annotated)),
    intron_len = end - start,
    ug_density = longest_ug_region / intron_len
  ) %>%
  select(-start, -end)

d <- master_table %>%
  filter(new_cate %in% c("Early", "Late")) %>%
  ggplot(aes(x = cryptic, y = ug_density, fill = new_cate)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(new_cate %in% c("Early", "Late")),
                                    "ug_density", "new_cate", "cryptic", "no"),
                   get_significance(master_table %>% filter(new_cate %in% c("Early", "Late")),
                                    "ug_density", "new_cate", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

#E. Density UG motifs by top-bottom in SH-SY5Y
master_table$max_psi_sh <- factor(master_table$max_psi_sh, levels=c("Top", "Bottom"))
e <- master_table %>%
  filter(max_psi_sh %in% c("Top", "Bottom")) %>%
  ggplot(aes(x = cryptic, y = ug_density, fill = max_psi_sh)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) +
  scale_fill_manual(values = plot_psi_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(max_psi_sh %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_sh", "cryptic", "no"),
                   get_significance(master_table %>% filter(max_psi_sh %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_sh", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
d + e
```

```{r postar3_coverage_fraction_early and late}
tdp_3 = rtracklayer::import('data/postar3_tardbp_reduced.bed')

coverage_numeric_tdp_flank <- create_coverage_vectors(flank_region, tdp_3)
coverage_matrix_tdp<- do.call(rbind, coverage_numeric_tdp_flank) %>%
    as_tibble() %>%
    mutate(name = flank_region$name,
           ID_exon = flank_region$ID_exon) %>%
    dplyr::relocate(name,ID_exon)

coverage_matrix_tdp %>%
    mutate(cryptic = ifelse(grepl("_yes",name),"cryptic","alternative")) %>%
    mutate(type = ifelse(grepl("acceptor",name),"acceptor","donor")) %>%
    mutate(timing = gsub(".*_","",name)) %>%
    reshape2::melt(id.vars = c("name","cryptic","timing","type","ID_exon")) %>%
    mutate(pos = parse_number(as.character(variable))) %>%
    left_join(data.frame(name = flank_region$name, strand = strand(flank_region))) %>%
    mutate(pos = ifelse(strand == "-", max(pos) - pos + 1, pos)) %>%
    group_by(pos,timing,type,cryptic) %>%
    summarise(cov_val = mean(value)) %>%
    mutate(pos = pos - (flank_size)) %>%
    ungroup() %>%
    group_by(timing,type,cryptic) %>%
    filter(timing %in% c("early","late")) %>%
    filter(cryptic == 'cryptic') %>%
    mutate(
        smoothed_weight = rollmean(
            cov_val,
            k = 5,
            fill = NA,
            align = "center"
        )) %>%
    ungroup() %>%
    # mutate(type = fct_relevel(type,"donor")) %>%
    ggplot(aes(x = pos, y = smoothed_weight,color = timing)) +
    geom_line(size = 1.6) +
    facet_wrap(~type) +
    theme_bw() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    ylab("Fraction of junctions bound by TDP-43 Postar3") +
    geom_vline(xintercept = 0,linetype = 'dotted') +
    xlab("Distance from cryptic splice site (bp)") +
    scale_color_manual(values = c("#de53b0" ,   "#173dd3")) +
    theme(text = element_text(size = 18))

```




```{r, echo=FALSE, warning=FALSE, message=FALSE}
#F. Distance iCLIP TDP43 binding site SH-SY5Y
exact_splice_site <- resize(flank_region, width = 1, fix = "center")
overlaps <- distanceToNearest(exact_splice_site, tdp_3)

print("2F - Distance iCLIP TDP43 binding site SH-SY5Y")
df_iclip <- read.csv(here::here('data', 'all_events.closest_tdp_bind.bed'), sep="\t")

f1 <- df_iclip %>%
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>%
  filter(junc_cat != "exon_skip") %>%
  filter(abs(dist) <= 1000) %>%
  #filter(cryptic!="yes")%>%
  ggplot(aes(x = dist, fill = cryptic)) +
  geom_density(alpha = 0.5, adjust = 0.1) +
  coord_cartesian(xlim=c(-150,150))+
  scale_fill_manual(values = c("seagreen2", "darkorchid")) +
  guides(fill = "none") +
  labs(x = 'Distance to the nearest TDP-binding site', y = 'Density') +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

f1

overlaps_exact = overlaps %>% 
    as.data.frame()

overlaps_exact$gene = exact_splice_site$gene
overlaps_exact$junc_cat = exact_splice_site$junc_cat
overlaps_exact$new_cate = exact_splice_site$new_cate

overlaps_exact %>% 
    ggplot(aes(x = distance, color = new_cate)) + 
    geom_density()
```

```{r binding_sites_heatmap}
# Calculate total counts for each TDP43 sensitivity group
total_counts <- df_iclip %>%
  group_by(tdp43_sensitivity_sh, cryptic) %>%
  summarize(total = n())

# Calculate percentages
heatmap_data <- df_iclip %>%
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>%
  filter(abs(dist) <= 250) %>%
  group_by(tdp43_sensitivity_sh, cryptic) %>%
  summarize(count = n(), .groups = "drop_last") %>%
  left_join(total_counts, by = c("tdp43_sensitivity_sh", "cryptic")) %>%
  mutate(percentage = round(count / total * 100, 0)) %>%
  select(-count, -total) %>%
  pivot_wider(
    names_from = cryptic, 
    values_from = percentage, 
    values_fill = 0
  ) %>%
  column_to_rownames(var = "tdp43_sensitivity_sh")

heatmap_plot <- heatmap_data %>%
  rownames_to_column(var = "tdp43_sensitivity_sh") %>%
  pivot_longer(cols = c("yes", "no"), names_to = "Cryptic", values_to = "Percentage")
heatmap_plot$tdp43_sensitivity_sh <- factor(heatmap_plot$tdp43_sensitivity_sh, levels=c("Late","Early"))
ggplot(heatmap_plot, aes(x = Cryptic, y = tdp43_sensitivity_sh, fill = Percentage)) +
  geom_tile() +
  geom_text(aes(label = paste0(Percentage, "%")), vjust = 0.5, hjust = 0.5) +
  scale_fill_gradient(low="saddlebrown", high="palegreen1") +
  labs(title = "", 
       y = "TDP43 sensitivity", 
       x = "Cryptic") +
  scale_x_discrete(labels=c("Alternative splicing", "Cryptic splicing"))+
  guides(fill="none") +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
```

```{r rnamaps_figure}
#G. RNA maps 
early_donors <- readDNAStringSet("data/maxent/donors_early_combined.fa")
late_donors <- readDNAStringSet("data/maxent/donors_late_combined.fa")

g1 <- ggplot() +
  geom_logo(data = as.character(early_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g2 <- ggplot() +
  geom_logo(data = as.character(late_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g <- g1 / g2
g
```

## Detection by early late post-mortem

```{r fpr_early_late_max_min}

nygc = fread("~/Desktop/nygc_postmortem_summary.csv")
rimod = fread("~/Desktop/rimod_post.csv")
summary_post = rbind(nygc,rimod)
cryptic_early_late = master_table %>% 
    filter(cryptic == 'yes') %>% 
    filter(new_cate %in% c("Early","Late"))

early_late_post = generate_summary_table(summary_post,
                       cryptic_early_late,
                            category_column = 'new_cate')
fisher_timing = run_fisher_test(early_late_post,
                                category_column = 'new_cate')

early_late_plots = make_final_plot(early_late_post,
                                      fisher_timing,
                                      category_column = 'new_cate',
                                      color_choices = c("#de53b0", "#173dd3"))
early_late_plots$detection_plot + 
          ggtitle('Early/Late Detection Rate') + 
    theme(text = element_text(size = 18)) +
        guides(x = guide_axis(n.dodge = 2)) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))


early_late_plots$selectivity_plot + ggtitle('Early/Late Selectivity Rate') +
    theme(text = element_text(size = 18)) +
        theme(text = element_text(size = 18)) +
        guides(x = guide_axis(n.dodge = 2)) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))


```



```{r}
full_summary_table = fread("/Users/annaleigh/Desktop/long_version.csv")
full_summary_table %>% 
    filter(chx_effect == 'Dark cryptic') %>% 
    filter(datasets_detected == 'CHX -SHY5Y') %>% 
    filter(fpr_value < 0.1 & value_tdp_path >=5)

```




```{r detection_rate_early_late_max_min}


master_table %>% 
    mutate(new_max = case_when(max_psi_be2 == max_psi_sh ~ max_psi_sh,
                               max_psi_be2 == "" ~ max_psi_sh,
                               max_psi_sh == "" ~ max_psi_be2,
                               max_psi_be2 == 'Medium' ~ max_psi_sh,
                               max_psi_sh == 'Medium' ~ max_psi_be2)) %>% 
    filter(!is.na(new_max)) %>% 
    left_join(postmortem_summary) %>% 
    filter(new_cate %in% c("Early","Late")) %>% 
    filter(new_max %in% c("Top","Bottom")) %>% 
    filter(fpr_value <= 0.1) %>% 
    filter(!is.na(group)) %>% 
    ggplot(aes(x = new_max, y =frac_ob_tdp_path,fill = new_cate)) + 
    geom_boxplot() +
    # facet_wrap(~group,scales = 'free')  +
    stat_summary(fun.data = give.n, geom = "text",
                                  fun.y = median,
                                  position = position_dodge(width = 0.75),color = 'white') +
  stat_compare_means(label = "p.signif",size = 15,label.y.npc = 'middle') +
    theme_classic() +
    scale_fill_manual(values = c("#de53b0","#173dd3")) +
    # scale_fill_manual(values = plot_psi_colors) +
    theme(text = element_text(size = 18)) +
    ylab("Fraction samples observed") +
    ggtitle('FPR <= 0.1')

```



## Does NMD sensitivity determine early/lateness?

```{r NMDSensNewCate_chx}

new_cate_nmd_sens = wide_chx %>% 
    distinct(paste_into_igv_junction,chx_effect) %>% 
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    filter(new_cate != "") %>% 
    filter(junction_for_clustering == 'yes') %>% 
    mutate(chx_effect = ifelse(chx_effect == "Dark cryptic","NMD rescued",chx_effect)) %>% 
    group_by(chx_effect,new_cate) %>% 
    reframe(n = n_distinct(paste_into_igv_junction))

ggplot(new_cate_nmd_sens) +
    aes(x = new_cate, fill = chx_effect, weight = n, by = as.factor(new_cate)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'Cycloheximide treatment') +
    scale_fill_manual(values = c("#3288BD","#D53E4F","#BFCC94")) + 
    theme(legend.position = 'bottom') +
    theme(text = element_text(size = 18))

new_cate_nmd_sens %>% 
    filter(chx_effect != "PSI > 95% at baseline") %>% 
    pivot_wider(names_from = 'chx_effect',values_from = 'n') %>% 
    column_to_rownames("new_cate") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()

```

```{r NMDSensNewCate_upf1}

new_cate_nmd_sens_upf1 = wide_chx %>% 
    distinct(paste_into_igv_junction,chx_effect) %>% 
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    filter(new_cate != "") %>% 
    filter(junction_for_clustering == 'yes') %>% 
    mutate(chx_effect = ifelse(chx_effect == "Dark cryptic","NMD rescued",chx_effect)) %>% 
    group_by(chx_effect,new_cate) %>% 
    reframe(n = n_distinct(paste_into_igv_junction))

ggplot(new_cate_nmd_sens) +
    aes(x = new_cate, fill = chx_effect, weight = n, by = as.factor(new_cate)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'Cycloheximide treatment') +
    scale_fill_manual(values = c("#3288BD","#D53E4F","#BFCC94")) + 
    theme(legend.position = 'bottom') 

new_cate_nmd_sens %>% 
    filter(chx_effect != "PSI > 95% at baseline") %>% 
    pivot_wider(names_from = 'chx_effect',values_from = 'n') %>% 
    column_to_rownames("new_cate") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()

```





```{r }




chx_g_e = chx_slope_plot_table$filtered_table %>% 
    ungroup() %>%  
    distinct(color_gene_name,gene_name) %>% 
    filter(gene_name %in% master_table$gene) %>% group_by(gene_name) %>%
    # Create a new column indicating if any row in the group has "NMD rescued"
    mutate(nmd_status = if_else(any(color_gene_name == "NMD rescued"), 
                                "NMD rescued", 
                                "non-NMD rescued")) %>%
    # Remove the grouping
    ungroup() %>% 
    select(-color_gene_name) %>% 
    unique()
smgi_g_e = pseudo_bulk_plot_slope$filtered_table %>% 
    ungroup() %>%  
    distinct(color_gene_name,gene_name) %>% 
    filter(gene_name %in% master_table$gene) %>% group_by(gene_name) %>%
    # Create a new column indicating if any row in the group has "NMD rescued"
    mutate(nmd_status = if_else(any(color_gene_name == "NMD rescued"), 
                                "NMD rescued", 
                                "non-NMD rescued")) %>%
    # Remove the grouping
    ungroup() %>% 
    select(-color_gene_name) %>% 
    unique()
chx_g_e %>% 
    left_join(smgi_g_e,by = c('gene_name')) %>% 
    filter(nmd_status.x != nmd_status.y)
####read in data!!
big_delta_de <- read.csv(here::here('data','big_delta_de.csv'))

for_plot <- big_delta_de[,c(4,8,10,11)] %>%
  # dplyr::filter(symbol == "STMN2" | symbol == "AARS1" | symbol == "HDGFL2" | symbol == "UNC13A" |
  #                 symbol == "SYNE1" | symbol == "CYFIP2")
    filter(symbol %in% master_table$gene)
for_plot_0 <- for_plot %>% 
  distinct(symbol, .keep_all = T) %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
for_plot <- rbind(for_plot_0,for_plot) %>% 
  mutate(symbol = factor(symbol, levels = c("AARS1", "HDGFL2", "SYNE1", "CYFIP2", "UNC13A", "STMN2")))

big_delta_de_tdp <- big_delta_de[,c(4,8,10,11)] %>% 
  dplyr::filter(symbol == "TARDBP")
for_plot_0_tdp <- big_delta_de_tdp[1,] %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
big_delta_de_tdp <- rbind(for_plot_0_tdp,big_delta_de_tdp)

for_plot$source <- as.factor(for_plot$source)
for_plot %>%   
    left_join(chx_g_e,by = c("symbol" = 'gene_name')) %>% 
    left_join(master_table %>% distinct(gene,new_cate,cryptic),by = c("symbol" = 'gene')) %>% 
    filter(new_cate %in% c("Early","Late")) %>% 
    filter(!is.na(nmd_status)) %>% 
    filter(cryptic == 'yes' & nmd_status == 'NMD rescued' & log2FoldChange > 1)
    ggplot(aes(x = source, y= log2FoldChange,fill = new_cate)) + 
    facet_grid(vars(nmd_status),vars(cryptic)) + 
    geom_boxplot()
  ggplot() +
  # geom_col(aes(x = as.character(source), y = log2FoldChange, fill = symbol), 
  #          data = big_delta_de_tdp, alpha = 0.3, show.legend = F) +
  geom_point(aes(x = as.character(source), y = log2FoldChange, color = symbol, group = symbol), ) +
  geom_line(aes(x = as.character(source), y = log2FoldChange, color = symbol, group = symbol), show.legend = F) +
  scale_fill_manual(values = "black") +
    facet_wrap(~nmd_status)
  # scale_color_manual(values = c("#A6CEE3", "#1F78B4", "#FF7F00", "#FDBF6F", "#FB9A99", "#E31A1C"))+ # "#B2DF8A" "#33A02C" "#CAB2D6" "#6A3D9A")) +
  scale_y_continuous(limits = c(-6.2,0.6)) +
  labs(x = "Doxycycline concentration", y = "Log2 Fold Change compared to control", color = "", fill = "Gene") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))
a
#ggsave("~/Desktop/curves_nmd.png", width = 6, height = 4)



big_delta_de_dz <- read.csv(here::here('data','big_delta_de_dz.csv'))


big_delta_de_dz_tdp <- big_delta_de_dz[,c(4,8,10,11)] %>% 
  dplyr::filter(symbol == "TARDBP")
for_plot_dz_0_tdp <- big_delta_de_dz_tdp[1,] %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
big_delta_de_dz_tdp <- rbind(for_plot_dz_0_tdp,big_delta_de_dz_tdp)

for_plot_dz <- big_delta_de_dz[,c(4,8,10,11)] %>%
  dplyr::filter(symbol == "STMN2" | symbol == "AARS1" | symbol == "HDGFL2" | symbol == "UNC13A" |
                  symbol == "SYNE1" | symbol == "CYFIP2")
for_plot_dz_0 <- for_plot_dz %>% 
  distinct(symbol, .keep_all = T) %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
for_plot_dz <- rbind(for_plot_dz_0,for_plot_dz) %>% 
  mutate(symbol = factor(symbol, levels = c("AARS1", "HDGFL2", "SYNE1", "CYFIP2", "UNC13A", "STMN2")))

for_plot_dz$source <- as.factor(for_plot_dz$source)
b <- for_plot_dz %>%   
  ggplot() +
  geom_col(aes(x = as.character(source), y = log2FoldChange, fill = symbol), data = big_delta_de_dz_tdp, alpha = 0.3) +
  geom_point(aes(x = source, y = log2FoldChange, color = symbol, group = symbol)) +
  geom_line(aes(x = source, y = log2FoldChange, color = symbol, group = symbol)) +
  scale_fill_manual(values = "black") +
  # scale_color_manual(values = c("#A6CEE3", "#1F78B4", "#FF7F00", "#FDBF6F", "#FB9A99", "#E31A1C"))+ # "#B2DF8A" "#33A02C" "#CAB2D6" "#6A3D9A")) +
  scale_y_continuous(limits = c(-5.2,0.5)) +
  labs(x = "Doxycycline concentration", y = "Log2 Fold Change compared to control", color = "", fill = "Gene") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))
b
#ggsave("~/Desktop/curves_nmd_dz.png", width = 6, height = 4)

#library(patchwork)

a+b

#ggsave("~/Desktop/caio.png", caio, width = 8, height = 4)




print("3G")

```



## Figure 4 - compendium / in vivo

```{r}

compendium_events = fread('data/compendium_cryptics.csv')

bind_table = master_table %>% 
    distinct(paste_into_igv_junction,gene,strand) %>% 
    dplyr::rename(gene_name = gene) %>% 
    rbind((compendium_events %>% distinct(gene_name,paste_into_igv_junction,strand))) %>% 
    dplyr::rename(junctions_coords = paste_into_igv_junction) %>% 
    unique() %>% 
    group_by(junctions_coords) %>% 
    mutate(gene_name = paste0(gene_name, collapse = ",")) %>% 
    ungroup() %>% 
    select(-strand) %>% 
    unique()


```




```{r tpr nmd sens}
facet_labels <- c(

tpr_value_rimod = "  RiMOD\nFTLD Frontal Cortex",

tpr_value_ftd_cortex = " NYGC\nFTLD Frontal\nTemporal Cortex",

tpr_value_als_cortex = "NYGC\nALS Motor Cortex",

tpr_value_cord = "NYGC\nALS Spinal Cord"

)

chx_insensitive = 
    chx_slope_plot_table$filtered_table %>% 
    filter(color_gene_name == 'NMD rescued') %>% 
    pull(paste_into_igv_junction) %>% unique()

smg1_insensitive = smg1_slope_plot_table_cyto$filtered_table %>% 
    filter(color_gene_name == 'non-NMD rescued') %>% 
    pull(paste_into_igv_junction) %>% unique()

upf1_insensitive = upf1_slope_plot_table$filtered_table %>% 
    filter(color_gene_name == 'NMD rescued') %>% 
    pull(paste_into_igv_junction) %>% unique()

combined_effect = c(chx_insensitive,smg1_insensitive,upf1_insensitive)

combine_events(chx_slope_plot_table$filtered_table,smg1_slope_plot_table_cyto$filtered_table,
               upf1_slope_plot_table$filtered_table,
               column = 'paste_into_igv_junction')


rimod_detection %>% 
    filter(fpr_value < 0.1) %>% 
    left_join(smg1_slope_plot_table_cyto$filtered_table %>% distinct(paste_into_igv_junction,color_gene_name)) %>% 
    filter(!is.na(color_gene_name)) %>% 
    ggplot(aes(x = color_gene_name, y= auc)) + 
    geom_boxplot() +
            stat_summary(fun.data = give.n, geom = "text",
                     fun.y = median,
                     position = position_dodge(width = 0.75)) +
    ggpubr::stat_compare_means(
    method = "wilcox.test",  # Explicitly specify method if needed
    formula = value ~ chx_rescue         # Explicit formula to avoid ambiguity
  )

tpr_across_ds_plot_table %>% 
    filter(tpr_value_cord >= 0.1) %>% 
    mutate(nmd_stable = ifelse(paste_into_igv_junction %in% chx_insensitive,"NMD-stable","other")) %>% 
    ggplot(aes(x = nmd_stable, y = tpr_value_cord)) + 
    geom_boxplot() +
        stat_summary(fun.data = give.n, geom = "text",
                     fun.y = median,
                     position = position_dodge(width = 0.75)) +
    ggpubr::stat_compare_means(
    method = "wilcox.test",  # Explicitly specify method if needed
    formula = value ~ chx_rescue         # Explicit formula to avoid ambiguity
  )

tpr_across_ds_plot_table %>% 
    filter(tpr_value_rimod >= 0.1) %>% 
    mutate(nmd_stable = ifelse(paste_into_igv_junction %in% chx_insensitive,"NMD-stable","other")) %>% 
    ggplot(aes(x = nmd_stable, y = tpr_value_rimod)) + 
    geom_boxplot() +
        stat_summary(fun.data = give.n, geom = "text",
                     fun.y = median,
                     position = position_dodge(width = 0.75)) +
    ggpubr::stat_compare_means(
    method = "wilcox.test",  # Explicitly specify method if needed
    formula = value ~ chx_rescue         # Explicit formula to avoid ambiguity
  )

tpr_across_ds_plot_table %>% 
    filter(tpr_value_ftd_cortex >= 0.1) %>% 
    mutate(nmd_stable = ifelse(paste_into_igv_junction %in% chx_insensitive,"NMD-stable","other")) %>% 
    ggplot(aes(x = nmd_stable, y = tpr_value_ftd_cortex)) + 
    geom_boxplot() +
        stat_summary(fun.data = give.n, geom = "text",
                     fun.y = median,
                     position = position_dodge(width = 0.75)) +
    ggpubr::stat_compare_means(
    method = "wilcox.test",  # Explicitly specify method if needed
    formula = value ~ chx_rescue         # Explicit formula to avoid ambiguity
  )

```



```{r targeted_panel, echo=FALSE, warning=FALSE, message=FALSE}
print("4A")
##load targeted seq table and compute CE ratio / rename samples in agreement to UNC13A paper
maxc <- read.csv(here::here('data',"count_table_of_aligned_remove_200bp_plus_product_reads.csv")) %>%
  mutate(fraction_cryptic = n_cryptic / n_counts_for_gene) %>%
  mutate(sample_name_ari = case_when(sample_name == "P64_11" ~ "HC1",
                                     sample_name == "P17_07" ~ "HC2",
                                     sample_name == "P47_11" ~ "HC3",
                                     sample_name == "P35_07" ~ "HC4",
                                     sample_name == "P56_13" ~ "FTD1",
                                     sample_name == "P45_15" ~ "FTD2",
                                     sample_name == "P63_05" ~ "FTD3",
                                     sample_name == "P28_07" ~ "FTD4",
                                     sample_name == "P40_04" ~ "FTD5",
                                     sample_name == "P07_15" ~ "FTD6",
                                     sample_name == "P16_09" ~ "FTD7",
                                     sample_name == "P86_08" ~ "FTD8",
                                     sample_name == "P13_13" ~ "FTD9",
                                     sample_name == "P11_07" ~ "FTE10",
                                     sample_name == "ddH2O" ~ "HO2O")) #%>%
  #pivot_wider(names_from = sample_name_ari, values_from = ratio_cryptic)
table(maxc$sample_name_ari)

##use only TDP43path-specific genes (use list from AL) - this is exluding more than 2/3 of genes!!
al <- read.csv(here::here('data',"expression_by_pathology_updated.csv")) %>%
  dplyr::filter(fraction_path > 0.01 & fraction_not_path < 0.005) %>% ####need to carefully check this with AL
  distinct(gene, .keep_all = T) ##check when binding with annotated validation table - should be ok since left_join
names(al)[4] <- "gene_name"
maxc_al <- maxc %>%
  left_join(al, by = "gene_name") %>%
  dplyr::filter(!is.na(paste_into_igv_junction))
table(maxc_al$sample_name_ari)
table(maxc_al$gene_name)

##check igv files for these genes
maxc_type <- maxc_al %>%
  select(-type) %>%
  mutate(gene = gene_name) %>%
  filter(gene_name != "SETD5") %>%
  left_join(master_table, by = "paste_into_igv_junction") %>%
  filter(!is.na(tdp43_sensitivity_sh)) %>% 
  mutate(group = ifelse(grepl("FT", sample_name_ari), "FTD",
                        ifelse(grepl("HC", sample_name_ari), "HC",
                               "H2O"))) %>% 
  group_by(gene_name) %>%
  filter(sum(n_cryptic) > 0) %>%
  ungroup() %>% 
    filter(group != "H2O")

maxc_type %>%
  filter(!is.na(new_cate)) %>%
  filter(gene_name != "SETD5") %>%
  #filter(n_cryptic>0) %>% 
  #filter(detected == "Yes") %>%
  filter(group != "H2O") %>%
  ggplot(aes(x = new_cate, y = n_cryptic)) +
  geom_col(aes(fill = new_cate)) +
  facet_wrap(facets = vars(group), ncol = 2) +
  scale_fill_brewer(palette = "Set1") +
  xlab("") +
  ylab("Count of cryptic exons detected by targeted panel") +
  labs(fill = "Category") +
  theme_bw()
ggsave("~/Desktop/grouped_panel.png", width = 8, height = 4)

maxc_earli <- maxc_type %>%
  filter(new_cate == "Early") %>%
  filter(gene_name != "SETD5") %>% 
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") + #position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Earli") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_early <- maxc_type %>%
  filter(new_cate == "Early") %>%
  filter(gene_name != "SETD5") %>% 
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") +#position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Early") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_intermediate <- maxc_type %>%
  filter(new_cate == "Intermediate") %>%
  filter(gene_name != "SETD5") %>%
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") +#position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Greens") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Intermediate") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_late <- maxc_type %>%
  filter(new_cate == "Late") %>%
  filter(gene_name != "SETD5") %>%
  ggplot(aes(x = sample_name_ari, y = fraction_cryptic)) +
  geom_bar(aes(fill = gene_name), stat = "identity", position = "stack", color="black") +#position = position_dodge(width = 0.8)) +
  #facet_grid(facets = vars(gene_name)) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(limits = c(0,0.3)) +
  labs(x = "", y = "", fill = "Late") +
  #scale_fill_manual(values = randomcoloR::distinctColorPalette(k = 30)) +
  theme_bw()
maxc_early / maxc_intermediate / maxc_late #/ maxc_na


```

```{r}
data = maxc_type %>% 
    distinct(sample_name_ari,new_cate,pc_cryptic,gene_name,group)  %>% 
    filter(!is.na(new_cate)) %>% 
    arrange(desc(group),sample_name_ari,new_cate)
# Define new_cate as an ordered factor
data$new_cate <- factor(data$new_cate, levels = c("Early", "Intermediate", "Late"))


heatmap_matrix = data %>% 
    select(-group,-new_cate) %>% 
    pivot_wider(names_from = sample_name_ari, values_from = pc_cryptic) %>% 
  column_to_rownames("gene_name") %>% 
    as.matrix()
# Row annotation with an ordered factor
row_ha <- rowAnnotation(
  new_cate = data$new_cate[match(rownames(heatmap_matrix), data$gene_name)],
  col = list(new_cate = c("Early" = "#de53b0", "Intermediate" = "steelblue", "Late" = "#1331a7"))
)

# Column annotation (sample group)
column_ha <- HeatmapAnnotation(
  group = data$group[match(colnames(heatmap_matrix), data$sample_name_ari)], 
  col = list(group = c("FTD" = "red", "HC" = "blue"))
)
ro <- order(factor(data$new_cate[match(rownames(heatmap_matrix), data$gene_name)], 
                          levels = c("Early", "Intermediate", "Late")))
Heatmap(
  heatmap_matrix,
  name = "pc_cryptic",
  top_annotation = column_ha,
  right_annotation = row_ha,
  row_order = ro,  # Use row_order to order rows by `new_cate`
  cluster_columns = FALSE,  # Keep manual ordering of columns
  cluster_rows = FALSE,  # Keep manual ordering of rows
  show_column_names = TRUE,
  show_row_names = TRUE
)

```





```{r}
summary_post %>% 
    filter(fpr_value <=0.1 & value_tdp_path >=5) %>% 
    select(group,paste_into_igv_junction,frac_ob_tdp_path,group,gene_name) %>% 
    pivot_wider(names_from = 'group', values_from = 'frac_ob_tdp_path') %>% 
    filter(!is.na(`Motor Cortex`) & !is.na(Cord)) %>% 
    mutate(motor_rank = rank(`Motor Cortex`)) %>% 
    mutate(cord_rank = rank(Cord)) %>% 
    ggplot(aes(x = motor_rank, y = cord_rank)) + 
    geom_point() + 
    geom_text_repel(aes(label = gene_name)) + 
    geom_abline() + 
    stat_cor(method = 'spearman')

deseq2_normalized_salmon %>% 
    filter(symbol %in% c("FIRRE","EPB41L4A","KCNIP1")) %>% 
    filter(variable %in% nygc_metadata$sample) %>% 
    left_join(nygc_metadata %>% select(sample,group,tdp_proteinopathy,
                                       rna_tissue_source_simplified),by = c("variable" = 'sample')) %>% 
    filter(!is.na(group)) %>% 
    ggplot(aes(x = group, y = value,fill = group)) +
    geom_boxplot() +
    facet_wrap(~symbol, scales = 'free' )
```

```{r}
# make figure - gives error so commenitng out for now
#row1 <- cowplot::plot_grid(ss_strength_plot, cons_score_plot, nrow=1,label_size = 16, labels=c("A","B"), rel_widths = c(1,2))
#row2 <- cowplot::plot_grid(symmetry, grid.grabExpr(draw(kmer_heatmap)), nrow=1,label_size = 16, labels=c("C","D"), rel_widths = c(1,2))
#donors_logo <- magick::image_read_pdf(here::here('main_fig_plus_files',"donors_logo.pdf"), density = 300)
#row3 <- cowplot::plot_grid(ug_repeats, g, nrow=1,label_size = 16, labels=c("E","F"), rel_widths = c(1,2))
#row4 <- cowplot::plot_grid(f1, sites_heatmap, nrow=1,label_size = 16, labels=c("G","H", "I"), rel_widths = c(2,1))

#fig2 <- cowplot::plot_grid( row1, row2, row3, row4, nrow=4,  label_size = 16, labels=c("","","",""))
```
