---
title: "TDP-43 cryptic exon biology main figures"
author: "MZ;AM;ALB"
date: "2024-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
###load packages/libraries/scripts
#BiocManager::install("ComplexHeatmap")
library(data.table)
library(tidyverse)
library(ggrepel)
library(patchwork)
library(ggalluvial)
library(ggsignif)
library(ggseqlogo)
library(Biostrings)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ComplexHeatmap)
library(RColorBrewer)
library(scales)
library(ggstats)
library(ggpubr)
library(pdftools)

source(here::here('scripts','trial_nmd_plots.R'))
give.n <- function(x){
return(c(y = median(x)*1.05, label = length(x)))
}


#check you're in right folder
here::here()

#plot parameters
plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#173dd3",
                             "Intermediate" = "steelblue","Not found" = 'darkgrey',"Ambiguous" = 'darkred')
plot_psi_colors <- c("Top" = "#388f78", "Bottom" = "#de9e4a", "Intermediate" = "steelblue")
plot_inframe_colors <- c("In-frame" = "darkslategray", "Out-frame" = "tan2")
plot_text_size <- 18
```

## Figure 1 - tissue

## Figure 2 - NMD

### Are NMD sensitive events up/downregulated?
```{r nmd_up_down_sh_chx}

chx_rna_full = fread(here::here('data',"chx_all_experiments_deseq2.csv"))
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))

wide_chx = big_data_chx |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_chx = Control_Control < 0.05 &  Control_TDP43KD > 0.1) |> 
    mutate(chx_cryptic = Control_Control < 0.05 &  Cycloheximide_TDP43KD > 0.1) |> 
    mutate(chx_effect = case_when((Cycloheximide_TDP43KD - Control_TDP43KD) > 0.05 ~ "NMD rescued",
                                  Control_TDP43KD > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_chx %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id))

chx_dark = wide_chx |> 
    filter(normal_cryptic_chx == FALSE & chx_cryptic == TRUE & chx_effect == "NMD rescued") |> 
    pull(paste_into_igv_junction) |> unique()

wide_chx = wide_chx %>% 
    mutate(chx_effect = ifelse(paste_into_igv_junction %in% chx_dark, "Dark cryptic",chx_effect))

chx_rna_full %>% 
    left_join(wide_chx) %>% 
    filter(!is.na(chx_effect)) %>% 
    filter(experiment == "Control_TDP43KD|Control_Control") %>% 
    group_by(ensgene) %>% 
    mutate(max_psi = max(Cycloheximide_TDP43KD)) %>% 
    ungroup() %>% 
    mutate(bin_max = cut_interval(max_psi,n = 2)) %>% 
    ggplot(aes(x = bin_max,
               y = log2fold_change,fill = chx_effect)) + 
    geom_hline(yintercept = 0,linetype = 'dotted') +
    geom_boxplot(position = position_dodge(preserve = "single")) +
    # stat_summary(fun.data = give.n, 
    #              geom = "text",
    #              fun.y = median,
    #              position = position_dodge(width = 0.75)) +
    theme_classic() +
    xlab("Max PSI NMDi + TDP-43 KD") +
    scale_fill_manual(values = c("#0D1821","#3288BD","#D53E4F","#BFCC94")) + 
    ylab("Log2 Fold Change TDP-43 KD SH-SY5Y")

   
```



## Figure 3 - Dose-response curves in SH-SY5Y and SK-N-BE(2) 

A. Dose-response curves for SH-SY5Y and SK-N-BE(2)
B. Differential expression in SH-SY5Y and in SK-N-BE(2)
C. Spaghetti plot in SH-SY5Y and in SK-N-BE(2)
D. Alluvial plot
E. Concordance between SH and SK (scatterplot) 
F. NMD early late
G. expression nmd genes in curves

```{r, echo = FALSE, warning=FALSE, message=FALSE}
#1A and 1B
levels_sy5y_qpcr <- read.csv(here::here('data',"qPCR_RNA.csv"))
levels_sy5y_qpcr = levels_sy5y_qpcr %>%
    filter(Gene == 'TDP-43')
levels_dz_qpcr <- read.csv(here::here('data','qpcr_dz.csv'))
levels_dz_qpcr = levels_dz_qpcr %>%
  filter(experiment == "new_new")
#1C and 1D
diff_expr_sh <- read.csv(here::here('data', 'diff_expr_sh.csv'))
diff_expr_sk <- read.csv(here::here('data', 'diff_expr_sk.csv'))
#1E, 1F, and 1G
master_table <- read.csv(here::here("data/master_table.csv"))

#####
print("1A - Dose-response curves for SH-SY5Y and SK-N-BE(2)")
a = levels_sy5y_qpcr %>%
    filter(Gene == 'TDP-43') %>% 
    mutate(source = c(rep(unique(diff_expr_sh$source),each = 3),rep(0,3))) %>% 
  ggplot(aes(x = as.character(source), y = Value)) +
    geom_violin() +
  geom_jitter() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
c <- levels_dz_qpcr %>%
  filter(experiment == "new_new") %>%
  ggplot(aes(x = as.character(sample), y = 100*value)) +
    geom_violin() +
  geom_jitter() +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 levels (% of untreated)') +
  #facet_wrap(facets = vars(experiment)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
a+c
#ggsave("~/Desktop/curves_qpcr.png", width = 10, height = 6, dpi = 300)

```

```{r de_by_levels}

avg_tdp_level_sh = levels_sy5y_qpcr %>% 
    group_by(Treatment) %>% 
    summarize(m = mean(Value)) %>% pull(m) %>% head(5)

print("1B - Differential expression in SH-SY5Y and SK-N-BE(2)")
plot_de <- diff_expr_sh %>%
    filter(symbol != "RNA5-8SN2") %>% 
  mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
  mutate(alpha_gene_name = ifelse(padj < 0.05, "sig", "non_sig")) 
plot_de_stmn_unc = plot_de %>% filter(symbol %in% c("STMN2","UNC13A"))
plot_de_meta <- as.data.frame(table(plot_de$color_gene_name, plot_de$alpha_gene_name, plot_de$source))
e <- plot_de %>%
    ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
    geom_point(aes(color = color_gene_name, 
                   alpha = alpha_gene_name, 
                   shape = alpha_gene_name, 
                   fill = alpha_gene_name),
               size = 0.2,
               position = position_jitter(width = 0, height = 0.4),
               show.legend = F) +
    geom_text(data = plot_de_meta[plot_de_meta$Var2 == "sig",], 
              aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
              size = 5,
              show.legend = F) +
    geom_vline(xintercept = 0) +
    coord_flip() +
    scale_color_manual(values = c(down = "#2B4162",up = "#721817",STMN2 = "#E31A1C",UNC13A ="#FB9A99"),breaks = c("STMN2","UNC13A")) +
    scale_alpha_manual(values = c(0.2, 1)) +
    scale_shape_manual(values = c(21, 21)) +  # Use shape 21 for both
    scale_fill_manual(values = c("grey", "black")) +  # Grey for non-sig, black for sig
    xlim(c(-10, 10)) +
    xlab('Gene expression Log2 FC compared to controls') +
    ylab('TDP-43 KD level') +
    scale_y_discrete(labels = paste0(c("+", "++", "+++", "++++", "+++++"),'\n',round(avg_tdp_level_sh,0),rep(" %",5))) +
    theme_classic() +
    theme(text = element_text(size = plot_text_size), plot.title = element_text(hjust = 0.5)) +
    ggtitle("SH-SY5Y") + 
    geom_point(inherit.aes = FALSE,
              aes(x = log2FoldChange, y = as.factor(source),color = symbol),
              data = plot_de_stmn_unc) + 
        labs(color = "") 


avg_be2_qpcr = levels_dz_qpcr %>%
  filter(experiment == "new_new") %>% 
    group_by(sample) %>% 
    summarize(m = mean(value)) %>% pull(m) %>% tail(7) * 100

plot_de_dz <- diff_expr_sk %>%
    filter(symbol != "RNA5-8SN2") %>% 
  mutate(color_gene_name = ifelse(log2FoldChange > 0, "up","down")) %>%
  mutate(alpha_gene_name = ifelse(padj < 0.05, "sig", "non_sig"))

plot_de_stmn_unc_dz = plot_de_dz %>% filter(symbol %in% c("STMN2","UNC13A"))

plot_de_dz_meta <- as.data.frame(table(plot_de_dz$color_gene_name, plot_de_dz$alpha_gene_name, plot_de_dz$source))
f <- plot_de_dz %>%
    ggplot(aes(x = log2FoldChange, y = as.factor(source))) +
    geom_point(aes(color = color_gene_name, 
                   alpha = alpha_gene_name, 
                   shape = alpha_gene_name, 
                   fill = alpha_gene_name),
               size = 0.2,
               position = position_jitter(width = 0, height = 0.4),
               show.legend = F) +
    geom_text(data = plot_de_dz_meta[plot_de_dz_meta$Var2 == "sig",], 
              aes(x = ifelse(Var1 == "down", -8, 8), y = Var3, label = Freq, color = Var1),
              size = 5,
              show.legend = F) +
    geom_vline(xintercept = 0) +
    coord_flip() +
    scale_color_manual(values = c(down = "#2B4162",up = "#721817",STMN2 = "#E31A1C",UNC13A ="#FB9A99"),breaks = c("STMN2","UNC13A")) +
    scale_alpha_manual(values = c(0.2, 1)) +
    scale_shape_manual(values = c(21, 21)) +  # Use shape 21 for both
    scale_fill_manual(values = c("grey", "black")) +  # Grey for non-sig, black for sig
    xlim(c(-10, 10)) +
    xlab('Gene expression Log2 FC compared to controls') +
    ylab('TDP-43 KD level') +
    scale_y_discrete(labels = paste0(c("+", "++", "+++", "++++", "+++++","++++++","+++++++"),'\n',round(avg_be2_qpcr,0),rep(" %",7))) +
    theme_classic() +
    theme(text = element_text(size = plot_text_size), plot.title = element_text(hjust = 0.5)) +
    ggtitle("SK-N-BE(2)") +
    geom_point(inherit.aes = FALSE,
          aes(x = log2FoldChange, y = as.factor(source),color = symbol),
          data = plot_de_stmn_unc_dz) +
    labs(color = "") 

ggarrange(e, f, common.legend = TRUE, legend = 'bottom')
#ggsave("~/Desktop/de_curves.png", width = 8, height = 4, dpi = 300)

```

```{r late_early_spaghetti_plot}
print("1C - Spaghetti plot in SH-SY5Y and SK-N-BE(2)")
list_gene = c("STMN2", "UNC13A", "HDGFL2", "AARS1","RANBP1",'PHF2')

avg_tdp_level_sh = levels_sy5y_qpcr %>% 
    group_by(Treatment) %>% 
    summarize(m = mean(Value)) %>% pull(m) %>% head(5)
avg_tdp_level_sh = c(100,avg_tdp_level_sh)


big_data_filtereds_early_late_sh <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_sh) %>% 
    dplyr::select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_sh,max_psi_sh)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox025)) %>% 
    distinct(gene, .keep_all = T) %>% 
    filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:12)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox0125", "dox0187", "dox021", "dox025", "dox075"))) %>%
  mutate(label_junction = case_when(name == "dox025" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_sh$tdp43_sensitivity_sh)
table(big_data_filtereds_early_late_sh$max_psi_sh)
GG <- big_data_filtereds_early_late_sh %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_sh), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = plot_sensitivity_colors) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = #c("-", "+", "++", "+++", "++++", "+++++")
                   paste0(round(avg_tdp_level_sh,0),rep("%",5))) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_sh) + 
  theme_classic()

avg_be2_qpcr = levels_dz_qpcr %>%
  filter(experiment == "new_new") %>% 
    group_by(sample) %>% 
    summarize(m = mean(value)) %>% pull(m) %>% tail(7) * 100
avg_be2_qpcr = c(100,avg_be2_qpcr)

big_data_filtereds_early_late_be2 <- master_table %>% #[,c(2,1,3:9,20:24)] %>%
    mutate(dox0 = dox0_be2) %>% 
    dplyr::select(c(gene,paste_into_igv_junction,strand,junc_cat,cryptic,junction_for_clustering,
             dox0,dox002,dox003,dox004,dox005,dox0075,dox01,dox1,
             #dox0,dox0125,dox0187,dox021,dox025,dox075,
             tdp43_sensitivity_be2,max_psi_be2)) %>% 
    group_by(gene) %>% 
    arrange(desc(dox01)) %>% 
    distinct(gene, .keep_all = T) %>% 
  filter(cryptic == "yes" & #junction_for_clustering == "yes" & 
             !is.na(dox0)) %>%
  pivot_longer(cols = c(7:14)) %>% 
  mutate(name = factor(name, levels = c("dox0","dox002","dox003","dox004","dox005","dox0075","dox01","dox1"))) %>%
  mutate(label_junction = case_when(name == "dox01" & gene %in% list_gene ~ gene, T ~ ""))
table(big_data_filtereds_early_late_be2$tdp43_sensitivity_be2)
table(big_data_filtereds_early_late_be2$max_psi_be2)
HH <- big_data_filtereds_early_late_be2 %>%
  ggplot(mapping = aes(x = name, y = value, group = paste_into_igv_junction)) +
  geom_line(aes(color = tdp43_sensitivity_be2), show.legend = F) +
  geom_text_repel(aes(label = label_junction), point.padding = 0.3,
                  nudge_y = 0.2, min.segment.length = 0.5, box.padding  = 2, max.overlaps = Inf, size=4, show.legend = F) +
  #geom_hline(yintercept = 0.2, linetype = "dotted") +
  scale_color_manual(values = plot_sensitivity_colors) +
  #scale_alpha_manual(values = c(0.1,1)) +
  xlab("TDP-43 knockdown level") +
  ylab("PSI") +
  scale_x_discrete(labels = #c("-", "+", "++", "+++", "++++", "+++++", "++++++", "+++++++")
                   paste0(round(avg_be2_qpcr,0),rep("%",7))) +
  scale_y_continuous(breaks=seq(0,1,0.2)) +
  facet_wrap(~tdp43_sensitivity_be2) + 
  theme_classic()
GG + HH
#ggsave("~/Desktop/earlylate.png", width = 18, height = 6, dpi = 300)
```

```{r alluvial_early_late_new}
# Create the alluvial plot
print("1D - alluvial plot")
master_table <- read.csv(here::here("data/master_table.csv"))
master_table <- master_table %>%
    mutate(tdp43_sensitivity_be2 = ifelse(tdp43_sensitivity_be2 == "","Not found",tdp43_sensitivity_be2)) %>% 
    mutate(tdp43_sensitivity_sh = ifelse(tdp43_sensitivity_sh == "","Not found",tdp43_sensitivity_sh)) %>% 
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "Not found" ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_sh == "Not found" ~ tdp43_sensitivity_be2,
                                tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                                tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late',
                                tdp43_sensitivity_sh == 'Intermediate'~  'Intermediate')) %>% 
        mutate(new_cate = ifelse(is.na(new_cate),"Ambiguous",new_cate)) 




df_long <- to_lodes_form(master_table, key = "axis", axes = c("tdp43_sensitivity_sh", "tdp43_sensitivity_be2")) %>% 
    mutate(axis = ifelse(axis == "tdp43_sensitivity_sh", "SH-SY5Y", "SK-N-BE(2)"))


df_long %>% 
    filter(junction_for_clustering == 'yes') %>% 
    ggplot(
       aes(x = axis,
           stratum = stratum, 
           alluvium = alluvium,
           fill = stratum, 
           label = stratum)) +
    scale_fill_manual(values = plot_sensitivity_colors) +
    geom_flow(stat = "alluvium", 
              lode.guidance = "frontback") +
    geom_stratum() +
    theme(legend.position = "bottom") +
    facet_wrap(~cryptic,
               labeller = as_labeller(
               c(no = "Alternative splicing", yes = "Cryptic splicing"))) + 
    theme_classic() +
    geom_text(stat = "stratum",
              aes(label = after_stat(count))) +
    labs(fill = element_blank(),
         x = element_blank(),
         y = 'N junctions') 
#ggsave("~/Desktop/alluvial.png", width = 6, height = 4, dpi = 300)
```

```{r scatterplot}
master_table <- read.csv(here::here("data/master_table.csv"))

master_table <- master_table %>%
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "" ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_sh == "" ~ tdp43_sensitivity_be2,
                                tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                                tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late',
                                tdp43_sensitivity_sh == 'Intermediate'~  'Intermediate')) %>% 
    mutate(new_cate = ifelse(is.na(new_cate),"Ambiguous",new_cate))


print("1E - Concordance between SH and SK scatterplot")
#scatterplot from upgrade
master_table %>% 
    ggplot(aes(x = dox075, y = dox1)) +
    geom_point() +
    geom_smooth(method = 'lm', se = F) +
    stat_cor() +
    xlab("Max PSI in SH-SY5Y") +
    ylab("Max PSI in SK-N-BE(2)") +
    theme_classic()
#ggsave("~/Desktop/scatter.png", width = 6, height = 4, dpi = 300)

```

```{r makewidechx}
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))
big_data_chx = big_data_chx %>% 
    separate(paste_into_igv_junction,remove = FALSE,into = c("chr",'start','end'),
             convert = TRUE)


wide_chx = big_data_chx |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_chx = Control_Control < 0.05 &  Control_TDP43KD > 0.1) |> 
    mutate(chx_cryptic = Control_Control < 0.05 &  Cycloheximide_TDP43KD > 0.1) |> 
    mutate(chx_effect = case_when((Cycloheximide_TDP43KD - Control_TDP43KD) > 0.05 ~ "NMD rescued",
                                  Control_TDP43KD > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_chx %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id))


chx_dark = wide_chx |> 
    filter(normal_cryptic_chx == FALSE & chx_cryptic == TRUE & chx_effect == "NMD rescued") |> 
    pull(paste_into_igv_junction) |> unique()

wide_chx = wide_chx %>% 
    mutate(chx_effect = ifelse(paste_into_igv_junction %in% chx_dark, "Dark cryptic",chx_effect))
```

```{r makewideupf1}
big_data_upf1 <- read.csv(here::here('data','nmd_or_not_upf1.csv'))
big_data_upf1 = big_data_upf1 %>% 
    separate(paste_into_igv_junction,remove = FALSE,into = c("chr",'start','end'),
             convert = TRUE)


wide_upf1 = big_data_upf1 |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_upf1 = ctrl_ctrl < 0.05 &  TDP43_ctrl > 0.1) |> 
    mutate(upf1_cryptic = ctrl_ctrl < 0.05 &  TDP43_UPF1 > 0.1) |> 
    mutate(upf1_effect = case_when((TDP43_UPF1 - TDP43_ctrl) > 0.05 ~ "NMD rescued",
                                  TDP43_ctrl > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_upf1 %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id)) 


upf1_dark = wide_upf1 |> 
    filter(normal_cryptic_upf1 == FALSE & upf1_cryptic == TRUE & upf1_effect == "NMD rescued") |> 
    pull(paste_into_igv_junction) |> unique()

wide_upf1 = wide_upf1 %>% 
    mutate(upf1_effect = ifelse(paste_into_igv_junction %in% upf1_dark, "Dark cryptic",upf1_effect))
```

```{r EarlyLateSushi}
sh_nmd_sens = wide_chx %>% 
    distinct(paste_into_igv_junction,chx_effect) %>% 
    left_join(master_table) %>% 
    filter(!is.na(tdp43_sensitivity_sh)) %>% 
    filter(tdp43_sensitivity_sh != "") %>% 
    filter(junction_for_clustering == 'yes') %>% 
    mutate(chx_effect = ifelse(chx_effect == "Dark cryptic","NMD rescued",chx_effect)) %>% 
    group_by(chx_effect,tdp43_sensitivity_sh) %>% 
    reframe(n = n_distinct(paste_into_igv_junction))

ggplot(sh_nmd_sens) +
    aes(x = tdp43_sensitivity_sh, fill = chx_effect, weight = n,by = as.factor(tdp43_sensitivity_sh)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'CHX treatment') +
    scale_fill_manual(values = c("#3288BD","#D53E4F","#BFCC94")) + 
    theme(legend.position = 'bottom', legend.text.position = 'bottom') 
ggsave("~/Desktop/sh_nmd_sens.png", width = 5, height = 4, dpi = 300)

sh_nmd_sens %>% 
    filter(chx_effect != "PSI > 95% at baseline") %>% 
    pivot_wider(names_from = 'chx_effect',values_from = 'n') %>% 
    column_to_rownames("tdp43_sensitivity_sh") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()

```

```{r curves nmd}
print("3F - TDP-43 dependent loss of dark cryptics genes")
####read in data!!
big_delta_de <- read.csv(here::here('data','big_delta_de.csv'))

for_plot <- big_delta_de[,c(4,8,10,11)] %>%
  dplyr::filter(symbol == "STMN2" | symbol == "AARS1" | symbol == "HDGFL2" | symbol == "UNC13A" |
                  symbol == "SYNE1" | symbol == "CYFIP2")
for_plot_0 <- for_plot %>% 
  distinct(symbol, .keep_all = T) %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
for_plot <- rbind(for_plot_0,for_plot) %>% 
  mutate(symbol = factor(symbol, levels = c("AARS1", "HDGFL2", "SYNE1", "CYFIP2", "UNC13A", "STMN2")))

big_delta_de_tdp <- big_delta_de[,c(4,8,10,11)] %>% 
  dplyr::filter(symbol == "TARDBP")
for_plot_0_tdp <- big_delta_de_tdp[1,] %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
big_delta_de_tdp <- rbind(for_plot_0_tdp,big_delta_de_tdp)

for_plot$source <- as.factor(for_plot$source)
a <- for_plot %>%   
  ggplot() +
  geom_col(aes(x = as.character(source), y = log2FoldChange, fill = symbol), data = big_delta_de_tdp, alpha = 0.3, show.legend = F) +
  geom_point(aes(x = source, y = log2FoldChange, color = symbol, group = symbol), show.legend = F) +
  geom_line(aes(x = source, y = log2FoldChange, color = symbol, group = symbol), show.legend = F) +
  scale_fill_manual(values = "black") +
  scale_color_manual(values = c("#A6CEE3", "#1F78B4", "#FF7F00", "#FDBF6F", "#FB9A99", "#E31A1C"))+ # "#B2DF8A" "#33A02C" "#CAB2D6" "#6A3D9A")) +
  scale_y_continuous(limits = c(-6.2,0.6)) +
  labs(x = "Doxycycline concentration", y = "Log2 Fold Change compared to control", color = "", fill = "Gene") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))

#ggsave("~/Desktop/curves_nmd.png", width = 6, height = 4)

big_delta_de_dz <- read.csv(here::here('data','big_delta_de_dz.csv'))


big_delta_de_dz_tdp <- big_delta_de_dz[,c(4,8,10,11)] %>% 
  dplyr::filter(symbol == "TARDBP")
for_plot_dz_0_tdp <- big_delta_de_dz_tdp[1,] %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
big_delta_de_dz_tdp <- rbind(for_plot_dz_0_tdp,big_delta_de_dz_tdp)

for_plot_dz <- big_delta_de_dz[,c(4,8,10,11)] %>%
  dplyr::filter(symbol == "STMN2" | symbol == "AARS1" | symbol == "HDGFL2" | symbol == "UNC13A" |
                  symbol == "SYNE1" | symbol == "CYFIP2")
for_plot_dz_0 <- for_plot_dz %>% 
  distinct(symbol, .keep_all = T) %>% 
  mutate(log2FoldChange = 0, padj = 0, source = 0)
for_plot_dz <- rbind(for_plot_dz_0,for_plot_dz) %>% 
  mutate(symbol = factor(symbol, levels = c("AARS1", "HDGFL2", "SYNE1", "CYFIP2", "UNC13A", "STMN2")))

for_plot_dz$source <- as.factor(for_plot_dz$source)
b <- for_plot_dz %>%   
  ggplot() +
  geom_col(aes(x = as.character(source), y = log2FoldChange, fill = symbol), data = big_delta_de_dz_tdp, alpha = 0.3) +
  geom_point(aes(x = source, y = log2FoldChange, color = symbol, group = symbol)) +
  geom_line(aes(x = source, y = log2FoldChange, color = symbol, group = symbol)) +
  scale_fill_manual(values = "black") +
  scale_color_manual(values = c("#A6CEE3", "#1F78B4", "#FF7F00", "#FDBF6F", "#FB9A99", "#E31A1C"))+ # "#B2DF8A" "#33A02C" "#CAB2D6" "#6A3D9A")) +
  scale_y_continuous(limits = c(-5.2,0.5)) +
  labs(x = "Doxycycline concentration", y = "Log2 Fold Change compared to control", color = "", fill = "Gene") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))

#ggsave("~/Desktop/curves_nmd_dz.png", width = 6, height = 4)

#library(patchwork)

a+b
#ggsave("~/Desktop/caio.png", caio, width = 8, height = 4, dpi = 300)

```




## Figure 4 - Features
```{r, echo=FALSE, warning=FALSE, message=FALSE}
get_signif_stars <- function(p_value) {
  if (p_value < 0.001) {
    return("***")
  } else if (p_value < 0.01) {
    return("**")
  } else if (p_value < 0.05) {
    return("*")
  } else {
    return("NS")
  }
}

```

```{r splice_site_strength}
get_significance <- function(df, value, group, category_col, category) {
  test_formula <- as.formula(paste(value, "~", group))
  p_test <- kruskal.test(test_formula, data = df %>% filter(!!sym(category_col) == category))
  #p1_annotation <- paste("p =", format(p1_test$p.value, digits = 3), get_signif_stars(p1_test$p.value))
  return(get_signif_stars(p_test$p.value))
}

print("2A - Splice site strength")
df_5ss_strength <- read.csv(here::here('data', 'maxent/5ss_strength_combined.csv'))
df_3ss_strength <- read.csv(here::here('data', 'maxent/3ss_strength_combined.csv'))
df_ss_strength <- rbind(
  df_5ss_strength  %>%
    mutate(site = "5' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, new_cate, site, ID_exon),
  df_3ss_strength  %>%
    mutate(site = "3' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, new_cate, site, ID_exon)
) %>%
  filter(new_cate %in% c("Early", "Late"))

ss_strength_plot <- df_ss_strength %>%
  ggplot(aes(x=site, y = log2(ss_ratio), fill = new_cate)) +
  geom_boxplot() +
  geom_abline(intercept = 0, slope = 0, color = "black", linetype = "dashed") +
  # geom_signif(
  #  y_position = c(max(df_ss_strength$ss_ratio) * 1.05,max(df_ss_strength$ss_ratio) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
  #  annotation = c(get_significance(df_ss_strength,
  #                                  "ss_ratio", "new_cate", "site", "3' splice site"),
  #                 get_significance(df_ss_strength,
  #                                  "ss_ratio", "new_cate", "site", "5' splice site")), tip_length = 0.02
  # ) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  #ylim(-4.1, 2.5) +
  labs(y = "Log2 Ratio MaxEnt score cryptic site vs annotated site", x = "") +
  theme_classic() +
  theme(text = element_text(size = plot_text_size)) +
  theme(legend.position = "top") +
  guides(fill = "none")
ss_strength_plot
```

```{r conservation_cell_only}
print("2B - Conservation score for CE and non-ce in combined categories")
cons_score_df <- master_table %>%
  filter(new_cate %in% c("Early", "Late")) %>%
  subset(!is.na(phastCons_alt) & !is.na(phastCons_unchanged)) %>%
  pivot_longer(cols = c(phastCons_alt, phastCons_unchanged), names_to = "category", values_to = "cons_score")

cons_score_df_cryptic <- cons_score_df %>% filter(cryptic == "yes")
cons_score_df_not_cryptic <- cons_score_df %>% filter(cryptic == "no")

a <- cons_score_df_cryptic %>%
  ggplot(aes(x = category, y = cons_score, fill = new_cate)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Novel splice site", "Unchanged splice site")) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_alt"),
                   get_significance(cons_score_df_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_unchanged"))
  ) +
  #facet_wrap(facets = vars(cryptic)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

cons_score_plot <- cons_score_df %>%
  ggplot(aes(x = category, y = cons_score, fill = new_cate)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
        facet_wrap(~cryptic,
               labeller = as_labeller(
               c(no = "Alternative splicing", yes = "Cryptic splicing"))) +
  scale_x_discrete(labels = c("Alternatively used splice site", "Unchanged splice site")) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_alt"),
                   get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "new_cate", "category", "phastCons_unchanged"))
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
cons_score_plot
```

```{r exon_in_frame}
result_mrg = fread('data/result_exon_table_merged.csv')
tmp_gr = result_mrg %>% 
    separate(exon_coord,into = c("chr","start","end")) %>% 
    makeGRangesFromDataFrame(,keep.extra.columns = TRUE)

seq_tmp <- getSeq(BSgenome.Hsapiens.UCSC.hg38,tmp_gr)


result_mrg = result_mrg %>% 
    mutate(width = seq_tmp %>% width()) %>% 
    mutate(div_3 = ifelse(width %% 3==0,"In-frame","Out-frame")) 


exon_counts <- result_mrg %>%
    filter(last_exon == FALSE) %>% 
  group_by(new_cate, cryptic, div_3) %>%
  summarise(count = n_distinct(exon_coord)) %>%
  ungroup()

frame_early <- exon_counts %>%
  #filter(cryptic=="yes") %>%
    filter(new_cate == "Early") %>%
    pivot_wider(names_from = div_3, values_from = count, values_fill = 0L) %>%
    select(-new_cate) %>%
    column_to_rownames('cryptic') %>% 
    chisq.test() %>% broom::tidy()

frame_late <- exon_counts %>%
    filter(new_cate == "Late") %>%
    pivot_wider(names_from = div_3, values_from = count, values_fill = 0L) %>%
    select(-new_cate) %>%
    column_to_rownames('cryptic') %>% 
    chisq.test() %>% broom::tidy()

symmetry <- exon_counts %>% 
    filter(new_cate %in% c("Early","Late")) %>%
    ggplot(aes(x = cryptic, y = count, fill = div_3)) +
  geom_bar(stat = "identity", position = "dodge") +
        facet_wrap(~new_cate) +
  geom_signif(data = data.frame(new_cate = c("Early","Late"), div_3=c('In-frame', 'Out-frame')),
              aes(y_position=c(61, 61), xmin=c(0.8, 0.8), xmax=c(2.2, 2.2),
              annotations=c(get_signif_stars(frame_early$p.value), 
                            get_signif_stars(frame_late$p.value))), tip_length=0.02, manual = T) +
  labs(
    #title = "Number of Symmetric and Non-Symmetric Exons by Type",
    x = "",
    y = "Number of exons",
    fill = ""
  ) +
  scale_x_discrete(labels=c("Alternative splicing", "Cryptic splicing"))+
  scale_fill_manual(
    values = plot_inframe_colors,
  ) +
  theme_classic() +
  guides(fill="none")+
  #theme(legend.position = "top") +
  theme(text = element_text(size = plot_text_size), axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
all_exons <- read.csv("data/inclusion_rates.csv")
all_exons <- all_exons %>%
  mutate(
    div_3 = ifelse((exon_stop-exon_start) %% 3==0,"In-frame","Out-frame")
  )

all_exons <- all_exons %>%
  filter(!is.na(incl_rate))%>%
  mutate(incl_rate_group = cut(incl_rate, 
                                breaks = c(0.01, 0.25, 0.85, 0.99, Inf), 
                                labels = c("1-25%", "25-85%", "85-99%", ">99%")))
                                #labels = c("<0.85", "0.85-0.995", ">0.995")))

plot_cont_table <- na.omit(all_exons) %>%
  group_by(incl_rate_group, div_3) %>%
  summarise(count = n(), .groups = 'drop')

cont_table <- plot_cont_table %>%
    pivot_wider(names_from = div_3, values_from = count)

group_pairs <- list(
  c(1,2), 
  c(1,3),  
  c(1,4)   
)

p_values <- sapply(group_pairs, function(pair) {
  chisq.test(as.matrix(cont_table %>% select(`In-frame`, `Out-frame`))[pair,])$p.value
})

ggplot(plot_cont_table, aes(x = incl_rate_group, y = count, fill = div_3)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "",
       x = "Inclusion rate",
       y = "Number of exons",
       fill = "") +
  theme_classic() +
  scale_fill_manual(
    values = plot_inframe_colors,
  ) +
  geom_signif(comparisons = list(c("1-25%", "25-85%"), 
                                 c("1-25%", "85-99%"), 
                                 c("1-25%", ">99%")),
              annotations = sprintf("p = %.1e", p_values),
              y_position = c(max(plot_cont_table$count)*1.1, 
                             max(plot_cont_table$count)*1.2, 
                             max(plot_cont_table$count)*1.3))
```

```{r kmer sh-sy5y, echo=FALSE, warning=FALSE, message=FALSE}
#C. Kmers enrichment analysis in SH-SY5Y
print("2C - Kmers enrichment analysis in SH-SY5Y")

kmer_size <- 6
yg_motifs <- c('TGTGTG', 'GTGTGT', 'GTGTGC', 'CGTGTG', 'TGCGTG', 'TGTGCG')
ya_motifs <- c('ATGTGT', 'TGTGTA', 'TGCATG', 'GTGTAT', 'TGTATG', 'GTATGT')
aa_motifs <- c('AATGAA', 'TTGAAT', 'ATGAAT', 'GTGTGA', 'TGAATG', 'GAATGA', 'GTGAAT', 'GAATGT')
all_motifs <- c(yg_motifs, ya_motifs, aa_motifs)
categories <- c(rep("YG", 6), rep("YA", 6), rep("AA", 8))

count_hexamers <- function(row) {
  if (row[["junc_cat"]] == "exon_skip" || is.na(row[["paste_into_igv_junction_annotated"]])) {
    return(NULL)
  }

  coords <- strsplit(row[["paste_into_igv_junction_annotated"]], "[:-]")[[1]]
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = coords[1], start = as.numeric(coords[2]), end = as.numeric(coords[3]))
  if (row[["strand"]] == "-") {
    seq <- reverseComplement(seq)
  }
  counts <- sapply(all_motifs, function(motif) countPattern(motif, seq, max.mismatch = 0))
  names(counts) <- all_motifs

  return(counts)
}

master_table_filt <- master_table %>%
  filter(new_cate %in% c("Early", "Late")) %>%
  filter(junc_cat != "exon_skip") %>%
  distinct(paste_into_igv_junction_annotated, .keep_all = T)
kmer_table <- apply(master_table_filt, 1, count_hexamers)
all_counts_df <- as.data.frame(t(kmer_table))

all_counts_df$type <- master_table_filt$new_cate
all_counts_df$cryptic <- master_table_filt$cryptic

summary_counts <- all_counts_df %>%
  group_by(type, cryptic) %>%
  summarise(across(all_motifs, sum))

means <- apply(summary_counts[, -c(1,2)], 1, mean)
sds <- apply(summary_counts[, -c(1,2)], 1, sd)
z_scores <- sweep(summary_counts[, -c(1,2)], MARGIN = 1, STATS = means, FUN = "-") / sds
heatmap_data <- as.matrix(z_scores)
rownames(heatmap_data) <- paste(summary_counts$type, ifelse(summary_counts$cryptic == "yes", "Cryptic", "Not cryptic"))

sort_kmers_by_zscore <- function(kmers, z_scores) {
  kmers[order(z_scores[kmers], decreasing = TRUE)]
}

sort_categories_by_zscore <- function(kmers, z_scores) {
  categories[order(z_scores[kmers], decreasing = TRUE)]
}

sorted_all_kmers <- sort_kmers_by_zscore(all_motifs, heatmap_data["Early Cryptic",]+heatmap_data["Late Cryptic",])
sorted_categories <- sort_categories_by_zscore(all_motifs, heatmap_data["Early Cryptic",]+heatmap_data["Late Cryptic",])

annotation <- HeatmapAnnotation(Motif = factor(sorted_categories, levels = c("YG", "YA", "AA")),
                                #show_legend = c(F, F, F),
                                col = list(Motif = c(YG = "palegreen3", YA = "turquoise3", AA = "tan1")),
                                which = "col")
kmer_heatmap <- Heatmap(heatmap_data[c("Early Cryptic","Early Not cryptic","Late Cryptic","Late Not cryptic") , sorted_all_kmers],
        cluster_columns = F,
        cluster_rows = F,
        show_row_names = T,
        show_column_names = T,
        show_row_dend = F,
        bottom_annotation = annotation,
        col = colorRampPalette(brewer.pal(12, "RdPu"))(100),
        show_heatmap_legend = FALSE)
kmer_heatmap

```

```{r ug_dense sh-y5y, echo=FALSE, warning=FALSE, message=FALSE}
#D. Density UG motifs by early-late in SH-SY5Y
print("2D and 2E - Density UG motifs by early-late and top-bottom in SH-SY5Y")
sequences <- sapply(seq_along(master_table$paste_into_igv_junction_annotated), function(i) {
  coord <- master_table$paste_into_igv_junction_annotated[i]
  chr <- sub("^(.*):.*$", "\\1", coord)
  range <- as.numeric(unlist(strsplit(sub(".*:(.*)$", "\\1", coord), "-")))
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = chr, start = range[1], end = range[2])

  if (master_table$strand[i] == "-") {
    seq <- reverseComplement(seq)
  }

  return(seq)
})

master_table$longest_ug_region <- Biostrings::vcountPattern('TGTGTG', DNAStringSet(sequences), max.mismatch = 1)

master_table <- master_table %>%
  mutate(
    start = as.numeric(sub(".*:(\\d+)-.*", "\\1", paste_into_igv_junction_annotated)),
    end = as.numeric(sub(".*-(\\d+)", "\\1", paste_into_igv_junction_annotated)),
    intron_len = end - start,
    ug_density = longest_ug_region / intron_len
  ) %>%
  select(-start, -end)

ug_repeats <- master_table %>%
  filter(new_cate %in% c("Early", "Late")) %>%
  ggplot(aes(x = cryptic, y = ug_density, fill = new_cate)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(new_cate %in% c("Early", "Late")),
                                    "ug_density", "new_cate", "cryptic", "no"),
                   get_significance(master_table %>% filter(new_cate %in% c("Early", "Late")),
                                    "ug_density", "new_cate", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

#E. Density UG motifs by top-bottom in SH-SY5Y
master_table$max_psi_sh <- factor(master_table$max_psi_sh, levels=c("Top", "Bottom"))
master_table %>%
  filter(max_psi_sh %in% c("Top", "Bottom")) %>%
  ggplot(aes(x = cryptic, y = ug_density, fill = max_psi_sh)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) +
  scale_fill_manual(values = plot_psi_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(max_psi_sh %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_sh", "cryptic", "no"),
                   get_significance(master_table %>% filter(max_psi_sh %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_sh", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#F. Distance iCLIP TDP43 binding site SH-SY5Y
print("2F - Distance iCLIP TDP43 binding site SH-SY5Y")
df_iclip <- read.csv(here::here('data', 'all_events.closest_tdp_bind.bed'), sep="\t")

f1 <-df_iclip %>%
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>%
  filter(junc_cat != "exon_skip") %>%
  filter(abs(dist) <= 1000) %>%
  #filter(tdp43_sensitivity_sh!="Early")%>%
  ggplot(aes(x = dist, fill = cryptic)) +
  geom_density(alpha = 0.5, adjust = 0.1) +
  coord_cartesian(xlim=c(-150,150))+
  scale_fill_manual(values = c("seagreen2", "darkorchid")) +
  guides(fill = "none") +
  labs(x = 'Distance to the nearest TDP-binding site', y = 'Density') +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

f1

```

```{r binding_sites_heatmap}
# Calculate total counts for each TDP43 sensitivity group
total_counts <- df_iclip %>%
  group_by(tdp43_sensitivity_sh, cryptic) %>%
  summarize(total = n())

# Calculate percentages
iclip_heatmap_data <- df_iclip %>%
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>%
  filter(abs(dist) <= 250) %>%
  group_by(tdp43_sensitivity_sh, cryptic) %>%
  summarize(count = n(), .groups = "drop_last") %>%
  left_join(total_counts, by = c("tdp43_sensitivity_sh", "cryptic")) %>%
  mutate(percentage = round(count / total * 100, 0)) %>%
  select(-count, -total) %>%
  pivot_wider(
    names_from = cryptic, 
    values_from = percentage, 
    values_fill = 0
  ) %>%
  column_to_rownames(var = "tdp43_sensitivity_sh")

heatmap_plot_df <- iclip_heatmap_data %>%
  rownames_to_column(var = "tdp43_sensitivity_sh") %>%
  pivot_longer(cols = c("yes", "no"), names_to = "Cryptic", values_to = "Percentage")
heatmap_plot_df$tdp43_sensitivity_sh <- factor(heatmap_plot_df$tdp43_sensitivity_sh, levels=c("Late","Early"))
sites_heatmap <- ggplot(heatmap_plot_df, aes(x = Cryptic, y = tdp43_sensitivity_sh, fill = Percentage)) +
  geom_tile() +
  geom_text(aes(label = paste0(Percentage, "%")), vjust = 0.5, hjust = 0.5) +
  scale_fill_gradient(low="saddlebrown", high="palegreen1") +
  labs(title = "", 
       y = "TDP43 sensitivity", 
       x = "Cryptic") +
  scale_x_discrete(labels=c("Alternative splicing", "Cryptic splicing"))+
  guides(fill="none") +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
```

```{r rnamaps_figure}
#G. RNA maps 
early_donors <- readDNAStringSet("data/maxent/donors_early_combined.fa")
late_donors <- readDNAStringSet("data/maxent/donors_late_combined.fa")

g1 <- ggplot() +
  geom_logo(data = as.character(early_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "#772FFF", "T" = "#772FFF")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  ) +
  theme(text = element_text(size = plot_text_size))
g2 <- ggplot() +
  geom_logo(data = as.character(late_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "#772FFF", "T" = "#772FFF")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  ) +
  theme(text = element_text(size = plot_text_size))
g <- g1 / g2
g
```

```{r}
# make figure - gives error so commenitng out for now
#row1 <- cowplot::plot_grid(ss_strength_plot, cons_score_plot, nrow=1,label_size = 16, labels=c("A","B"), rel_widths = c(1,2))
#row2 <- cowplot::plot_grid(symmetry, grid.grabExpr(draw(kmer_heatmap)), nrow=1,label_size = 16, labels=c("C","D"), rel_widths = c(1,2))
#donors_logo <- magick::image_read_pdf(here::here('main_fig_plus_files',"donors_logo.pdf"), density = 300)
#row3 <- cowplot::plot_grid(ug_repeats, g, nrow=1,label_size = 16, labels=c("E","F"), rel_widths = c(1,2))
#row4 <- cowplot::plot_grid(f1, sites_heatmap, nrow=1,label_size = 16, labels=c("G","H", "I"), rel_widths = c(2,1))

#fig2 <- cowplot::plot_grid( row1, row2, row3, row4, nrow=4,  label_size = 16, labels=c("","","",""))
```
