---
title: "Per-CE contributions to the cryptic burden score"
author: "AL"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    df_print: paged
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo    = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width  = 11,
    fig.height = 7,
    dpi        = 150
)
```



# Setup

Run the main burden script. This produces the per-sample burden tables
(`motor_burden`, `frontal_burden`, `cord_burden`, `rimod_burden`), the
junction QC table (`summary_post`), the queried junction data
(`nygc_queried`, `rimod_queried`), and per-cohort metadata
(`motor_meta`, `frontal_meta`, `cord_meta`, `rimod_meta`). It caches the
parquet queries, so the second run is fast.

```{r source-burden}
source("burden_analyses.R")
library(tidyverse)
library(patchwork)
library(glue)
```

# Helper functions

Three helpers do all the per-CE work: one returns the long per-sample
per-CE contribution table (the matrix of `scale_psi` values, unsummed),
one ranks CEs within a disease group, and one wraps the three-panel
"what drives the burden in this cohort" figure.

```{r helpers}
#' Per-sample per-CE contribution to the burden score.
#' Mirrors compute_burden() but stops short of summing across junctions.
compute_ce_contributions <- function(queried_data,
                                     samples_meta,
                                     summary_post,
                                     group_label,
                                     sample_col   = "sample",
                                     fpr_thresh   = 0.1,
                                     tpr_thresh   = 0.1,
                                     min_tdp_path = 5) {

    passing <- summary_post %>%
        filter(fpr_value      <= fpr_thresh,
               tpr_value      >= tpr_thresh,
               value_tdp_path >= min_tdp_path,
               group          == group_label) %>%
        pull(paste_into_igv_junction)

    queried_data %>%
        mutate(paste_into_igv_junction = as.character(glue("{chrom}:{start}-{end}"))) %>%
        filter(sample                  %in% samples_meta[[sample_col]],
               paste_into_igv_junction %in% passing) %>%
        group_by(paste_into_igv_junction) %>%
        mutate(scale_psi = scale(psi) %>% as.vector()) %>%
        ungroup() %>%
        distinct(sample, paste_into_igv_junction, scale_psi) %>%
        left_join(samples_meta, by = setNames(sample_col, "sample"))
}

#' Rank CEs by mean contribution within one disease group.
#'
#' Operates on the z-scored contributions table only. Detection / gene-
#' name annotation are NOT computed here - they come from `summary_post`
#' via `ce_qc_annotation()` and are joined in by `compute_ce_drivers()`.
rank_ce_in_group <- function(contributions, burden_tbl,
                             group_col, group_value,
                             gene_annotation = NULL,
                             sample_col      = "sample") {

    grp_samples <- contributions %>%
        filter(.data[[group_col]] == group_value) %>%
        pull(.data[[sample_col]]) %>% unique()
    if (length(grp_samples) == 0) {
        warning("No samples for ", group_col, " == ", group_value)
        return(tibble())
    }

    mean_burden <- burden_tbl %>%
        filter(.data[[sample_col]] %in% grp_samples) %>%
        pull(cryptic_burden) %>% mean(na.rm = TRUE)

    ranked <- contributions %>%
        filter(.data[[sample_col]] %in% grp_samples) %>%
        group_by(paste_into_igv_junction) %>%
        summarise(mean_scale_psi   = mean(scale_psi,   na.rm = TRUE),
                  median_scale_psi = median(scale_psi, na.rm = TRUE),
                  n_samples        = sum(!is.na(scale_psi)),
                  .groups = "drop") %>%
        mutate(frac_of_mean_burden = mean_scale_psi / mean_burden,
               n_in_group          = length(grp_samples),
               group_value         = group_value) %>%
        arrange(desc(mean_scale_psi))

    if (!is.null(gene_annotation))
        ranked <- ranked %>% left_join(gene_annotation,
                                       by = "paste_into_igv_junction")
    ranked
}

#' Build per-CE annotation for one cohort by pulling gene name and the
#' pre-computed detection metric (`frac_ob_tdp_path`, fraction of TDP+
#' samples in this cohort where the CE was called) directly from
#' `summary_post`. Replaces the previous `compute_ce_penetrance()` helper
#' that recomputed detection from raw PSI - the QC pipeline already did
#' this work and the resulting fractions are exactly what we want.
ce_qc_annotation <- function(summary_post_tbl, qc_group_label) {
    summary_post_tbl %>%
        filter(group == qc_group_label) %>%
        distinct(paste_into_igv_junction, gene_name,
                 frac_ob_tdp_path,  frac_ob_not_path,
                 value_tdp_path,    n_samples_tdp_path,
                 value_not_path,    n_samples_not_path,
                 auc)
}

#' Add a human-readable label column: "GENE (coords)" if gene known.
ce_label <- function(df) {
    df %>%
        mutate(label = ifelse(!is.na(gene_name) & gene_name != "",
                              paste0(gene_name, " (", paste_into_igv_junction, ")"),
                              paste_into_igv_junction))
}

#' Collapse multiple junctions per gene by keeping the row with the
#' highest value in `sort_col`. Junctions without a gene_name are kept
#' as distinct events. Use to dedupe before picking top-N for figures.
top_ce_per_gene <- function(ranked, sort_col = "mean_scale_psi") {
    ranked %>%
        mutate(.gene_key = ifelse(!is.na(gene_name) & gene_name != "",
                                  gene_name, paste_into_igv_junction)) %>%
        group_by(.gene_key) %>%
        slice_max(.data[[sort_col]], n = 1, with_ties = FALSE) %>%
        ungroup() %>%
        select(-.gene_key) %>%
        arrange(desc(.data[[sort_col]]))
}

#' Compute the ranked CE table + cumulative-burden saturation summary
#' for a single cohort. Joins gene name and the pre-computed detection
#' fraction (`frac_ob_tdp_path`) from `summary_post` onto the ranked
#' z-scored contributions. Returns ranked / cumulative / n50 / n80 / n_tot.
compute_ce_drivers <- function(contributions, burden_tbl,
                               group_col, group_value,
                               cohort_label,
                               summary_post_tbl,
                               qc_group_label) {

    ranked <- rank_ce_in_group(contributions, burden_tbl,
                               group_col, group_value)
    qc <- ce_qc_annotation(summary_post_tbl, qc_group_label)
    ranked <- ranked %>% left_join(qc, by = "paste_into_igv_junction")

    cum <- ranked %>%
        arrange(desc(mean_scale_psi)) %>%
        mutate(cum_contrib = cumsum(mean_scale_psi),
               cum_frac    = cum_contrib / sum(mean_scale_psi),
               rank        = row_number(),
               rank_frac   = rank / n())
    n50   <- cum %>% filter(cum_frac >= 0.50) %>% slice_head(n = 1) %>% pull(rank)
    n80   <- cum %>% filter(cum_frac >= 0.80) %>% slice_head(n = 1) %>% pull(rank)
    n_tot <- nrow(cum)
    list(ranked = ranked, cum = cum, cohort = cohort_label,
         n50 = n50, n80 = n80, n_tot = n_tot)
}
```

```{r gene-annotation}
junction_gene <- summary_post %>%
    distinct(paste_into_igv_junction, gene_name)
```

# Per-cohort contribution tables

```{r compute-contributions}
ce_contrib_motor <- compute_ce_contributions(
    nygc_queried,  motor_meta,   summary_post, "Motor Cortex")
ce_contrib_frontal <- compute_ce_contributions(
    nygc_queried,  frontal_meta, summary_post, "FTD frontal/temporal")
ce_contrib_cord  <- compute_ce_contributions(
    nygc_queried,  cord_meta,    summary_post, "Cord")
ce_contrib_rimod <- compute_ce_contributions(
    rimod_queried, rimod_meta,   summary_post, "RiMOD")
```

# Q1: which CEs drive the burden?

The burden score is `sum(scale_psi)` across passing CEs per sample. To
answer "which CEs drive it" within a disease group we just take the mean
of `scale_psi` per CE across samples in that group. We do this for all
four burden cohorts so the answer can be generalised.

```{r drivers-all-cohorts}
driver_specs <- tribble(
    ~contributions,     ~burden,        ~group_col,          ~group,     ~cohort_label,                       ~qc_group,              ~prefix,
    ce_contrib_motor,   motor_burden,   "tdp_proteinopathy", "ALS-TDP",  "NYGC Motor Cortex (ALS-TDP)",       "Motor Cortex",         "nygc_motor_alstdp",
    ce_contrib_frontal, frontal_burden, "tdp_proteinopathy", "FTLD-TDP", "NYGC Frontal/Temporal (FTLD-TDP)",  "FTD frontal/temporal", "nygc_frontal_ftldtdp",
    ce_contrib_cord,    cord_burden,    "tdp_proteinopathy", "ALS-TDP",  "NYGC Spinal Cord (ALS-TDP)",        "Cord",                 "nygc_cord_alstdp",
    ce_contrib_rimod,   rimod_burden,   "tdp_proteinopathy", "FTLD-TDP", "RiMOD Frontal Cortex (FTLD-TDP)",   "RiMOD",                "rimod_ftldtdp"
)

drivers <- pmap(driver_specs, function(contributions, burden, group_col,
                                       group, cohort_label, qc_group, prefix) {
    compute_ce_drivers(contributions    = contributions,
                       burden_tbl       = burden,
                       group_col        = group_col,
                       group_value      = group,
                       cohort_label     = cohort_label,
                       summary_post_tbl = summary_post,
                       qc_group_label   = qc_group)
}) %>% set_names(driver_specs$prefix)

# Persist the ranked tables per cohort
walk2(drivers, names(drivers), function(d, prefix) {
    write_csv(d$ranked,
              file.path(OUTPUT_DIR, paste0(prefix, "_ranked_ce.csv")))
})
```


## Figure 1: top-20 CEs per cohort



One panel per cohort, top 20 CEs by mean contribution, bars coloured by
penetrance (the fraction of samples in that group with at least 2 spliced reads).

```{r fig1-top20, fig.width=16, fig.height=10}
build_top20_panel <- function(d) {
    d$ranked %>%
        top_ce_per_gene() %>%      # collapse to one junction per gene first
        ce_label() %>%
        slice_head(n = 20) %>%
        ggplot(aes(x    = mean_scale_psi,
                   y    = fct_reorder(label, mean_scale_psi),
                   fill = frac_ob_tdp_path)) +
        geom_col() +
        scale_fill_viridis_c(limits = c(0, 1),
                             name   = "% Detected",
                             labels = scales::percent_format(accuracy = 1)) +
        theme_bw(base_size = 11) +
        labs(x = "Mean contribution to cryptic burden",
             y = NULL,
             title = d$cohort)
}

panels_top20    <- map(drivers, build_top20_panel)

fig_top20 <- wrap_plots(panels_top20, ncol = 2) +
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "A",
                    title = "Top-20 CEs by mean contribution to burden",
                    theme = theme(plot.title = element_text(face = "bold")))

ggsave(file.path(OUTPUT_DIR, "fig1_top20_ces_all_cohorts.pdf"),
       fig_top20, width = 16, height = 10)
ggsave(file.path(OUTPUT_DIR, "fig1_top20_ces_all_cohorts.png"),
       fig_top20, width = 16, height = 10, dpi = 200)

fig_top20
```

## Figure 2: cross-cohort overlap of the top-20 drivers

A reviewer-facing direct comparison: of the four cohorts' top-20 gene-
deduped drivers, which genes appear in the top-20 of more than one
cohort? The union of these sets defines the "ever top-20" gene list;
each cohort marks whether (and at what rank) each gene appears in its
own top-20. Genes are ordered by cohort breadth (how many of the four
cohorts they appear in) and then by mean rank.

```{r fig2-overlap}
# Top-20 deduped per cohort, in a long table
top20_long <- imap_dfr(drivers, function(d, prefix) {
    d$ranked %>%
        top_ce_per_gene() %>%
        slice_head(n = 20) %>%
        mutate(rank     = row_number(),
               cohort   = d$cohort,
               gene_key = ifelse(!is.na(gene_name) & gene_name != "",
                                 gene_name, paste_into_igv_junction))
})

# Cohort breadth and mean rank per gene_key, used for ordering
gene_breadth <- top20_long %>%
    group_by(gene_key) %>%
    summarise(n_cohorts = n_distinct(cohort),
              mean_rank = mean(rank),
              best_rank = min(rank),
              .groups   = "drop") %>%
    arrange(desc(n_cohorts), mean_rank)

gene_levels <- gene_breadth %>% pull(gene_key)

cohort_levels <- c("NYGC Motor Cortex (ALS-TDP)",
                   "NYGC Frontal/Temporal (FTLD-TDP)",
                   "NYGC Spinal Cord (ALS-TDP)",
                   "RiMOD Frontal Cortex (FTLD-TDP)")

write_csv(top20_long  %>% select(cohort, rank, gene_key, gene_name,
                                 paste_into_igv_junction, mean_scale_psi,
                                 frac_ob_tdp_path),
          file.path(OUTPUT_DIR, "top20_per_cohort_long.csv"))
write_csv(gene_breadth,
          file.path(OUTPUT_DIR, "top20_gene_breadth.csv"))

message("\nTop-20 driver overlap summary:")
print(gene_breadth %>% count(n_cohorts, name = "n_genes"))
```

```{r fig2-overlap-plot, fig.width=11, fig.height=11}
p_tile <- top20_long %>%
    mutate(gene_key = factor(gene_key, levels = rev(gene_levels)),
           cohort   = factor(cohort,   levels = cohort_levels)) %>%
    ggplot(aes(x = cohort, y = gene_key, fill = rank)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(label  = rank,
                  colour = rank > 10), size = 3) +
    scale_fill_gradient(low = "#D64933", high = "#FAD3CB",
                        limits = c(1, 20), name = "Rank") +
    scale_colour_manual(values = c(`TRUE` = "black", `FALSE` = "white"),
                        guide  = "none") +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    theme(axis.text.x = element_text(angle = 25, hjust = 1),
          panel.grid  = element_blank()) +
    labs(x = NULL, y = NULL,
         title = "Rank of each top-20 driver gene across cohorts",
         subtitle = "Filled cell = gene is in that cohort's top 20; number = rank within that cohort")

p_breadth <- gene_breadth %>%
    mutate(gene_key = factor(gene_key, levels = rev(gene_levels))) %>%
    ggplot(aes(x = n_cohorts, y = gene_key)) +
    geom_col(fill = "grey40") +
    geom_text(aes(label = n_cohorts), hjust = -0.3, size = 3) +
    scale_x_continuous(limits = c(0, 4.6), breaks = 1:4) +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major.y = element_blank()) +
    labs(x = "# cohorts in top-20", y = NULL,
         title = "Cohort breadth")

fig_top20_overlap <- p_tile + p_breadth + plot_layout(widths = c(3.2, 1))

ggsave(file.path(OUTPUT_DIR, "fig2_top20_overlap_across_cohorts.pdf"),
       fig_top20_overlap, width = 11, height = 11)
ggsave(file.path(OUTPUT_DIR, "fig2_top20_overlap_across_cohorts.png"),
       fig_top20_overlap, width = 11, height = 11, dpi = 200)

fig_top20_overlap
```


## Figure 3: top-20 CEs per cohort, ranked by detection

Same structure as Figure 1, but with the axes inverted: x-axis is
detection (`frac_ob_tdp_path` — fraction of TDP+ samples in this cohort
where the CE was called), fill is mean contribution to the burden. Genes
are now ranked by how *widely detected* they are rather than by how much
they contribute. Comparing Figures 1 and 3 shows that the top
contributors and the most widely detected CEs are not necessarily the
same — a CE can be detected in nearly all samples but contribute modestly
(because it sits close to the cohort PSI mean), or contribute heavily
while being called in fewer samples.

```{r fig3-top20-detection, fig.width=16, fig.height=10}
# Shared rank-contribution limits so the viridis colour scale is comparable
# across cohorts (rank 1 = top contributor in that cohort).
max_rank_across_cohorts <- max(map_int(drivers, function(d) nrow(d$ranked)))

build_top20_detection_panel <- function(d, rank_limits) {
    d$ranked %>%
        arrange(desc(mean_scale_psi)) %>%
        mutate(rank_contribution = row_number()) %>%
        top_ce_per_gene(sort_col = "frac_ob_tdp_path") %>%
        ce_label() %>%
        slice_max(frac_ob_tdp_path, n = 20, with_ties = FALSE) %>%
        ggplot(aes(x    = frac_ob_tdp_path,
                   y    = fct_reorder(label, frac_ob_tdp_path),
                   fill = rank_contribution)) +
        geom_col() +
        scale_fill_viridis_c(name      = "Contribution\nrank",
                             direction = -1,
                             limits    = rank_limits) +
        scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                           limits = c(0, 1)) +
        theme_bw(base_size = 11) +
        labs(x = "% TDP+ samples with the CE detected",
             y = NULL,
             title = d$cohort)
}

panels_top20_det <- map(drivers, build_top20_detection_panel,
                        rank_limits = c(1, max_rank_across_cohorts))

fig_top20_det <- wrap_plots(panels_top20_det, ncol = 2) +
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "A",
                    title = "Top-20 CEs by detection in TDP+ samples (per cohort)",
                    theme = theme(plot.title = element_text(face = "bold")))

ggsave(file.path(OUTPUT_DIR, "fig3_top20_ces_by_detection.pdf"),
       fig_top20_det, width = 16, height = 10)
ggsave(file.path(OUTPUT_DIR, "fig3_top20_ces_by_detection.png"),
       fig_top20_det, width = 16, height = 10, dpi = 200)

fig_top20_det
```

## Figure 4: rank-by-contribution vs rank-by-detection, per cohort

A direct head-to-head: within each cohort, every passing CE has a rank
by mean contribution and a separate rank by detection
(`frac_ob_tdp_path`). Plot the two against each other. Points on the
dashed diagonal have identical ranks under both metrics; points off the
diagonal are the CEs where the metrics disagree. Spearman ρ in each
panel quantifies how tightly the two rankings track.

```{r fig4-rank-rank, fig.width=12, fig.height=11}
rank_compare <- imap_dfr(drivers, function(d, prefix) {
    d$ranked %>%
        top_ce_per_gene(sort_col = "mean_scale_psi") %>%
        arrange(desc(mean_scale_psi)) %>%
        mutate(rank_contribution = row_number()) %>%
        arrange(desc(frac_ob_tdp_path)) %>%
        mutate(rank_detection    = row_number()) %>%
        ce_label() %>%
        mutate(cohort     = d$cohort,
               gene_key   = ifelse(!is.na(gene_name) & gene_name != "",
                                   gene_name, paste_into_igv_junction),
               divergence = abs(rank_contribution - rank_detection))
})

rank_cor <- rank_compare %>%
    group_by(cohort) %>%
    summarise(rho   = cor(rank_contribution, rank_detection,
                          method = "spearman"),
              n_ces = n(),
              .groups = "drop") %>%
    mutate(label = sprintf("Spearman ρ = %.2f   (n CEs = %d)", rho, n_ces))

# Per-cohort labels: top-5 by contribution OR top-5 by detection OR
# the 6 largest divergers
rank_compare_labelled <- rank_compare %>%
    group_by(cohort) %>%
    mutate(label_show = ifelse(
        rank_contribution <= 5 |
            rank_detection <= 5 |
            row_number(desc(divergence)) <= 6,
        gene_key, NA_character_)) %>%
    ungroup()

cohort_levels <- c("NYGC Motor Cortex (ALS-TDP)",
                   "NYGC Frontal/Temporal (FTLD-TDP)",
                   "NYGC Spinal Cord (ALS-TDP)",
                   "RiMOD Frontal Cortex (FTLD-TDP)")

p_rank_compare <- rank_compare_labelled %>%
    mutate(cohort = factor(cohort, levels = cohort_levels)) %>%
    ggplot(aes(rank_contribution, rank_detection)) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", colour = "grey50") +
    geom_point(aes(colour = divergence), alpha = 0.75, size = 2) +
    scale_colour_viridis_c(option = "magma", direction = -1,
                           name = "|Δ rank|") +
    geom_text(aes(label = label_show), size = 2.8,
              na.rm = TRUE, check_overlap = TRUE, vjust = -0.6) +
    geom_text(data  = rank_cor %>%
                  mutate(cohort = factor(cohort, levels = cohort_levels)),
              aes(x = Inf, y = -Inf, label = label),
              inherit.aes = FALSE,
              hjust = 1.05, vjust = -1, size = 3.5, fontface = "bold") +
    facet_wrap(~ cohort, ncol = 2, scales = "free") +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    labs(x = "Rank by mean contribution (1 = highest)",
         y = "Rank by % detected (1 = highest)",
         title = "Contribution rank vs detection rank, per cohort",
         subtitle = "Dashed line = identical ranks. Off-diagonal points = CEs where contribution and detection disagree.")

write_csv(rank_compare %>%
              select(cohort, gene_key, gene_name, paste_into_igv_junction,
                     mean_scale_psi, frac_ob_tdp_path,
                     rank_contribution, rank_detection, divergence),
          file.path(OUTPUT_DIR, "rank_contribution_vs_detection.csv"))

ggsave(file.path(OUTPUT_DIR, "fig4_rank_contribution_vs_detection.pdf"),
       p_rank_compare, width = 12, height = 11)
ggsave(file.path(OUTPUT_DIR, "fig4_rank_contribution_vs_detection.png"),
       p_rank_compare, width = 12, height = 11, dpi = 200)

p_rank_compare
```


## Figure 5: cross-cohort overlap of the top-20 by detection

Same tile-and-breadth layout as Figure 2, but the top-20 per cohort is
defined by detection (`frac_ob_tdp_path`) rather than by mean
contribution. Cross-referencing Figures 2 and 5 shows whether the
"widely-detected" set has more cross-cohort overlap than the "high-
contribution" set.

```{r fig5-detection-overlap}
top20_long_det <- imap_dfr(drivers, function(d, prefix) {
    d$ranked %>%
        top_ce_per_gene(sort_col = "frac_ob_tdp_path") %>%
        slice_max(frac_ob_tdp_path, n = 20, with_ties = FALSE) %>%
        mutate(rank     = row_number(),
               cohort   = d$cohort,
               gene_key = ifelse(!is.na(gene_name) & gene_name != "",
                                 gene_name, paste_into_igv_junction))
})

gene_breadth_det <- top20_long_det %>%
    group_by(gene_key) %>%
    summarise(n_cohorts = n_distinct(cohort),
              mean_rank = mean(rank),
              best_rank = min(rank),
              .groups   = "drop") %>%
    arrange(desc(n_cohorts), mean_rank)

gene_levels_det <- gene_breadth_det %>% pull(gene_key)

cohort_levels <- c("NYGC Motor Cortex (ALS-TDP)",
                   "NYGC Frontal/Temporal (FTLD-TDP)",
                   "NYGC Spinal Cord (ALS-TDP)",
                   "RiMOD Frontal Cortex (FTLD-TDP)")

write_csv(top20_long_det %>%
              select(cohort, rank, gene_key, gene_name,
                     paste_into_igv_junction, frac_ob_tdp_path,
                     mean_scale_psi),
          file.path(OUTPUT_DIR, "top20_per_cohort_by_detection_long.csv"))
write_csv(gene_breadth_det,
          file.path(OUTPUT_DIR, "top20_by_detection_gene_breadth.csv"))

message("\nTop-20-by-detection overlap summary:")
print(gene_breadth_det %>% count(n_cohorts, name = "n_genes"))
```

```{r fig5-detection-overlap-plot, fig.width=11, fig.height=11}
p_tile_det <- top20_long_det %>%
    mutate(gene_key = factor(gene_key, levels = rev(gene_levels_det)),
           cohort   = factor(cohort,   levels = cohort_levels)) %>%
    ggplot(aes(x = cohort, y = gene_key, fill = rank)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(label  = rank,
                  colour = rank > 10), size = 3) +
    scale_fill_viridis_c(limits = c(1, 20), direction = -1, name = "Rank") +
    scale_colour_manual(values = c(`TRUE` = "black", `FALSE` = "white"),
                        guide  = "none") +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    theme(axis.text.x = element_text(angle = 25, hjust = 1),
          panel.grid  = element_blank()) +
    labs(x = NULL, y = NULL,
         title = "Rank of each top-20-by-detection gene across cohorts",
         subtitle = "Filled cell = gene is in that cohort's top 20 by % TDP+ detected; number = rank")

p_breadth_det <- gene_breadth_det %>%
    mutate(gene_key = factor(gene_key, levels = rev(gene_levels_det))) %>%
    ggplot(aes(x = n_cohorts, y = gene_key)) +
    geom_col(fill = "grey40") +
    geom_text(aes(label = n_cohorts), hjust = -0.3, size = 3) +
    scale_x_continuous(limits = c(0, 4.6), breaks = 1:4) +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major.y = element_blank()) +
    labs(x = "# cohorts in top-20 by detection", y = NULL,
         title = "Cohort breadth")

fig_top20_det_overlap <- p_tile_det + p_breadth_det +
    plot_layout(widths = c(3.2, 1))

ggsave(file.path(OUTPUT_DIR, "fig5_top20_detection_overlap_across_cohorts.pdf"),
       fig_top20_det_overlap, width = 11, height = 11)
ggsave(file.path(OUTPUT_DIR, "fig5_top20_detection_overlap_across_cohorts.png"),
       fig_top20_det_overlap, width = 11, height = 11, dpi = 200)

fig_top20_det_overlap
```



# Q2: is GRN-FTLD-TDP just C9 with the volume turned up?

For each FTLD cohort, compute the per-CE GRN-minus-C9 delta in mean
contribution. Then ask whether the *direction* and *magnitude* of those
deltas reproduce between NYGC and RiMOD. If GRN samples are broadly
higher than C9 samples across the same CEs in both cohorts, the deltas
will be predominantly positive and the cross-cohort scatter will be
sign-concordant.

## Per-cohort GRN and C9 rankings

```{r grn-c9-rankings}
ranked_frontal_grn <- rank_ce_in_group(
    ce_contrib_frontal, frontal_burden,
    "simplified_mutations", "GRN",     gene_annotation = junction_gene)
ranked_frontal_c9  <- rank_ce_in_group(
    ce_contrib_frontal, frontal_burden,
    "simplified_mutations", "C9ORF72", gene_annotation = junction_gene)
ranked_rimod_grn   <- rank_ce_in_group(
    ce_contrib_rimod, rimod_burden,
    "DiseaseCode", "FTD-GRN",          gene_annotation = junction_gene)
ranked_rimod_c9    <- rank_ce_in_group(
    ce_contrib_rimod, rimod_burden,
    "DiseaseCode", "FTD-C9",           gene_annotation = junction_gene)

# Paired tables (GRN, C9) per cohort
make_paired <- function(grn, c9) {
    grn %>%
        select(paste_into_igv_junction, gene_name,
               mean_grn = mean_scale_psi) %>%
        inner_join(c9 %>% select(paste_into_igv_junction,
                                 mean_c9 = mean_scale_psi),
                   by = "paste_into_igv_junction") %>%
        mutate(delta = mean_grn - mean_c9)
}
paired_nygc  <- make_paired(ranked_frontal_grn, ranked_frontal_c9)
paired_rimod <- make_paired(ranked_rimod_grn,   ranked_rimod_c9)
```

## Cross-cohort reproducibility of the GRN-vs-C9 delta

```{r cross-cohort}
cross_cohort <- paired_nygc %>%
    select(paste_into_igv_junction, gene_name,
           delta_nygc = delta, mean_grn_nygc = mean_grn, mean_c9_nygc = mean_c9) %>%
    inner_join(
        paired_rimod %>% select(paste_into_igv_junction,
                                delta_rimod    = delta,
                                mean_grn_rimod = mean_grn,
                                mean_c9_rimod  = mean_c9),
        by = "paste_into_igv_junction"
    )

n_overlap <- nrow(cross_cohort)

quad <- cross_cohort %>%
    mutate(quadrant = case_when(
        delta_nygc > 0 & delta_rimod > 0 ~ "GRN-high in both",
        delta_nygc < 0 & delta_rimod < 0 ~ "C9-high in both",
        delta_nygc > 0 & delta_rimod < 0 ~ "GRN-high NYGC, C9-high RiMOD",
        delta_nygc < 0 & delta_rimod > 0 ~ "C9-high NYGC, GRN-high RiMOD",
        TRUE ~ "tie"))

quad_counts <- quad %>% count(quadrant)
quad_counts

concordant <- sum(quad$quadrant %in% c("GRN-high in both", "C9-high in both"))
discordant <- sum(quad$quadrant %in% c("GRN-high NYGC, C9-high RiMOD",
                                       "C9-high NYGC, GRN-high RiMOD"))
binom_test <- binom.test(concordant, concordant + discordant, p = 0.5,
                         alternative = "greater")

cross_pearson  <- cor.test(cross_cohort$delta_nygc, cross_cohort$delta_rimod,
                           method = "pearson")
cross_spearman <- suppressWarnings(cor.test(cross_cohort$delta_nygc,
                                            cross_cohort$delta_rimod,
                                            method = "spearman"))

cross_stats <- tibble(
    n_CEs        = n_overlap,
    pearson_r    = cross_pearson$estimate,
    pearson_p    = cross_pearson$p.value,
    spearman_rho = cross_spearman$estimate,
    spearman_p   = cross_spearman$p.value,
    concordant   = concordant,
    discordant   = discordant,
    pct_concord  = round(100 * concordant / (concordant + discordant), 1),
    binom_p      = binom_test$p.value
)
cross_stats
```

```{r reproducible-lists}
reproducible_grn_high <- cross_cohort %>%
    filter(delta_nygc > 0, delta_rimod > 0) %>%
    mutate(combined_delta = delta_nygc + delta_rimod) %>%
    arrange(desc(combined_delta))

reproducible_c9_high <- cross_cohort %>%
    filter(delta_nygc < 0, delta_rimod < 0) %>%
    mutate(combined_delta = delta_nygc + delta_rimod) %>%
    arrange(combined_delta)

```

## Figure 6: cross-cohort reproducibility of GRN-vs-C9 delta

Two-panel figure. Panel A: scatter of per-CE GRN-minus-C9 delta in NYGC
(x) vs RiMOD (y), with sign-concordant CEs in red and discordant CEs in
grey. Panel B: paired bar plot of the top-15 reproducibly-GRN-high CEs
(positive delta in both cohorts), side-by-side per cohort.

```{r fig6-cross-cohort, fig.width=14, fig.height=12}
# Labels: top concordant in each direction + biggest discordant CEs
to_label <- bind_rows(
    reproducible_grn_high %>% slice_head(n = 8),
    reproducible_c9_high  %>% slice_head(n = 8),
    cross_cohort %>%
        filter(sign(delta_nygc) != sign(delta_rimod),
               !is.na(delta_nygc), !is.na(delta_rimod)) %>%
        mutate(disc = abs(delta_nygc - delta_rimod)) %>%
        arrange(desc(disc)) %>% slice_head(n = 5)
) %>% distinct(paste_into_igv_junction) %>% pull(paste_into_igv_junction)

stats_text <- sprintf(
    "Pearson r = %.3f (p = %.2g)\nSpearman ρ = %.3f (p = %.2g)\nConcordant: %d / %d (%.0f%%, binom p = %.2g)",
    cross_pearson$estimate,  cross_pearson$p.value,
    cross_spearman$estimate, cross_spearman$p.value,
    concordant, concordant + discordant,
    100 * concordant / (concordant + discordant),
    binom_test$p.value
)

p_cross <- cross_cohort %>%
    mutate(label_show = gene_name,
           sign_match = sign(delta_nygc) == sign(delta_rimod) &
               delta_nygc != 0 & delta_rimod != 0) %>%
    ggplot(aes(delta_nygc, delta_rimod)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(aes(colour = sign_match), alpha = 0.7) +
    scale_colour_manual(values = c(`TRUE` = "#D64933", `FALSE` = "grey60"),
                        labels = c(`TRUE` = "Same direction",
                                   `FALSE` = "Opposite"),
                        name = NULL) +
    geom_text(aes(label = label_show), size = 3,
              na.rm = TRUE, check_overlap = TRUE, vjust = -0.5) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.2,
             size = 3.5, label = stats_text) +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    labs(x = "GRN - C9 mean contribution (NYGC)",
         y = "GRN - C9 mean contribution (RiMOD)",
         title = sprintf("Cross-cohort reproducibility of GRN-C9 delta (n CEs = %d)",
                         n_overlap))




p_cross
```

## Figure 7: per-sample breadth of the GRN vs C9 burden

For each sample, count the number of CEs whose cohort-wide z-scored PSI
exceeds successive thresholds. This is a per-sample breadth measure
(how many CEs is *this sample* sitting above the cohort baseline for),
not a group-mean measure. Then compare the distribution of those counts
across mutational groups within each FTLD cohort.

```{r fig7-breadth-per-sample, fig.width=12, fig.height=6}
group_colours <- c(GRN = "#D64933", C9 = "#2E6F95")

per_sample_thresholds <- function(contributions, group_col, grn_value,
                                  c9_value, cohort_label) {
    contributions %>%
        filter(.data[[group_col]] %in% c(grn_value, c9_value)) %>%
        group_by(sample, .grp = .data[[group_col]]) %>%
        summarise(
            n_above_0      = sum(scale_psi >= 0,   na.rm = TRUE),
            n_above_half   = sum(scale_psi >= 0.5, na.rm = TRUE),
            n_above_1      = sum(scale_psi >= 1,   na.rm = TRUE),
            n_above_1_half = sum(scale_psi >= 1.5, na.rm = TRUE),
            n_above_2      = sum(scale_psi >= 2,   na.rm = TRUE),
            .groups = "drop"
        ) %>%
        mutate(group  = ifelse(.grp == grn_value, "GRN", "C9"),
               cohort = cohort_label) %>%
        select(-.grp)
}

per_sample_breadth <- bind_rows(
    per_sample_thresholds(ce_contrib_frontal, "simplified_mutations",
                          "GRN", "C9ORF72",
                          "NYGC Frontal/Temporal"),
    per_sample_thresholds(ce_contrib_rimod,   "DiseaseCode",
                          "FTD-GRN", "FTD-C9",
                          "RiMOD Frontal Cortex")
)

breadth_n <- per_sample_breadth %>%
    distinct(cohort, group, sample) %>%
    count(cohort, group) %>%
    pivot_wider(names_from = group, values_from = n) %>%
    mutate(cohort_label = sprintf("%s\nGRN n = %d  |  C9 n = %d",
                                  cohort, GRN, C9))

breadth_long <- per_sample_breadth %>%
    pivot_longer(starts_with("n_above_"),
                 names_to = "threshold", values_to = "n_ces") %>%
    mutate(threshold = recode(threshold,
                              n_above_0      = ">= 0",
                              n_above_half   = ">= 0.5",
                              n_above_1      = ">= 1",
                              n_above_1_half = ">= 1.5",
                              n_above_2      = ">= 2"),
           threshold = factor(threshold,
                              levels = c(">= 0", ">= 0.5",
                                         ">= 1", ">= 1.5", ">= 2"))) %>%
    left_join(breadth_n %>% select(cohort, cohort_label), by = "cohort") %>%
    mutate(cohort_label = factor(cohort_label,
                                 levels = breadth_n$cohort_label))

p_breadth <- breadth_long %>%
    ggplot(aes(x = threshold, y = n_ces, fill = group)) +
    geom_boxplot(position = position_dodge(width = 0.75),
                 width = 0.6, outlier.size = 0.7) +
    facet_wrap(~ cohort_label, ncol = 2) +
    scale_fill_manual(values = group_colours, name = "Mutation") +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    labs(x = "Per-sample cohort-z-scored PSI threshold",
         y = "Number of CEs in sample above threshold",
         title = "Per-sample breadth: CEs above cohort z-score thresholds")

p_breadth
```

