---
title: "Per-CE contributions to the cryptic burden score"
author: "AL"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    df_print: paged
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo    = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width  = 11,
    fig.height = 7,
    dpi        = 150
)
```

# Background


The composite burden score is defined per sample as
\( B_i = \sum_j z_{ij} \) where \(z_{ij} = \text{scale}(\psi)\) is the
junction-wise z-scored PSI of CE *j* across all samples in the queried
cohort (see `compute_burden()` in `burden_analyses.R`). Because the
composite is an additive sum, the **mean** of `scale_psi` for one CE across
a disease group **is** that CE's mean contribution to the burden in that
group, in the same units as the burden itself. Per-CE attribution is
therefore mechanical and exact.

This document addresses two questions:

1. **Q1 - Shape of the score.** Within each cohort, are some CEs driving
   the burden more than others, or is the contribution broadly distributed
   across many CEs?
2. **Q2 - GRN vs C9 dissociation.** In GRN-FTLD-TDP versus C9-FTLD-TDP,
   are the burden-driving CEs the same set (consistent with "GRN is just
   C9 with the volume turned up") or different (consistent with mechanism-
   specific cryptic signatures)? We test this within each FTLD cohort and
   then check whether any differential pattern is reproducible across two
   independent cohorts (NYGC, RiMOD).

# Setup

Run the main burden script. This produces the per-sample burden tables
(`motor_burden`, `frontal_burden`, `cord_burden`, `rimod_burden`), the
junction QC table (`summary_post`), the queried junction data
(`nygc_queried`, `rimod_queried`), and per-cohort metadata
(`motor_meta`, `frontal_meta`, `cord_meta`, `rimod_meta`). It caches the
parquet queries, so the second run is fast.

```{r source-burden}
source("burden_analyses.R")
library(tidyverse)
library(patchwork)
library(glue)
```

# Helper functions

Three helpers do all the per-CE work: one returns the long per-sample
per-CE contribution table (the matrix of `scale_psi` values, unsummed),
one ranks CEs within a disease group, and one wraps the three-panel
"what drives the burden in this cohort" figure.

```{r helpers}
#' Per-sample per-CE contribution to the burden score.
#' Mirrors compute_burden() but stops short of summing across junctions.
compute_ce_contributions <- function(queried_data,
                                     samples_meta,
                                     summary_post,
                                     group_label,
                                     sample_col   = "sample",
                                     fpr_thresh   = 0.1,
                                     tpr_thresh   = 0.1,
                                     min_tdp_path = 5) {

    passing <- summary_post %>%
        filter(fpr_value      <= fpr_thresh,
               tpr_value      >= tpr_thresh,
               value_tdp_path >= min_tdp_path,
               group          == group_label) %>%
        pull(paste_into_igv_junction)

    queried_data %>%
        mutate(paste_into_igv_junction = as.character(glue("{chrom}:{start}-{end}"))) %>%
        filter(sample                  %in% samples_meta[[sample_col]],
               paste_into_igv_junction %in% passing) %>%
        group_by(paste_into_igv_junction) %>%
        mutate(scale_psi = scale(psi) %>% as.vector()) %>%
        ungroup() %>%
        distinct(sample, paste_into_igv_junction, scale_psi) %>%
        left_join(samples_meta, by = setNames(sample_col, "sample"))
}

#' Rank CEs by mean contribution within one disease group.
#'
#' Operates on the z-scored contributions table only. Penetrance is NOT
#' computed here - it comes from raw PSI via `compute_ce_penetrance()`
#' and is joined in by `compute_ce_drivers()`.
rank_ce_in_group <- function(contributions, burden_tbl,
                             group_col, group_value,
                             gene_annotation = NULL,
                             sample_col      = "sample") {

    grp_samples <- contributions %>%
        filter(.data[[group_col]] == group_value) %>%
        pull(.data[[sample_col]]) %>% unique()
    if (length(grp_samples) == 0) {
        warning("No samples for ", group_col, " == ", group_value)
        return(tibble())
    }

    mean_burden <- burden_tbl %>%
        filter(.data[[sample_col]] %in% grp_samples) %>%
        pull(cryptic_burden) %>% mean(na.rm = TRUE)

    ranked <- contributions %>%
        filter(.data[[sample_col]] %in% grp_samples) %>%
        group_by(paste_into_igv_junction) %>%
        summarise(mean_scale_psi   = mean(scale_psi,   na.rm = TRUE),
                  median_scale_psi = median(scale_psi, na.rm = TRUE),
                  n_samples        = sum(!is.na(scale_psi)),
                  .groups = "drop") %>%
        mutate(frac_of_mean_burden = mean_scale_psi / mean_burden,
               n_in_group          = length(grp_samples),
               group_value         = group_value) %>%
        arrange(desc(mean_scale_psi))

    if (!is.null(gene_annotation))
        ranked <- ranked %>% left_join(gene_annotation,
                                       by = "paste_into_igv_junction")
    ranked
}

#' Per-CE detection penetrance within a disease group.
#'
#' Reads raw PSI directly from the queried-data parquet bind. Penetrance
#' is the fraction of group samples where `psi > psi_detection_thresh`.
#' Independent of the burden's z-scoring step.
#'
#' @param queried_data         Bound parquet rows with chrom/start/end/psi/sample
#' @param samples_meta         Metadata with sample_col + group_col
#' @param group_col            Column in samples_meta to filter on
#' @param group_value          Value within group_col
#' @param sample_col           Column mapping samples_meta to queried_data$sample
#' @param psi_detection_thresh PSI threshold for "detected" (default 0)
compute_ce_penetrance <- function(queried_data, samples_meta,
                                  group_col, group_value,
                                  sample_col           = "sample",
                                  psi_detection_thresh = 0) {

    grp_samples <- samples_meta %>%
        filter(.data[[group_col]] == group_value) %>%
        pull(.data[[sample_col]]) %>% unique()
    if (length(grp_samples) == 0) return(tibble())

    queried_data %>%
        mutate(paste_into_igv_junction = as.character(glue("{chrom}:{start}-{end}"))) %>%
        filter(sample %in% grp_samples) %>%
        distinct(sample, paste_into_igv_junction, psi) %>%
        group_by(paste_into_igv_junction) %>%
        summarise(mean_psi          = mean(psi, na.rm = TRUE),
                  median_psi        = median(psi, na.rm = TRUE),
                  penetrance        = mean(psi > psi_detection_thresh, na.rm = TRUE),
                  n_detected        = sum(psi > psi_detection_thresh,  na.rm = TRUE),
                  n_with_psi_value  = sum(!is.na(psi)),
                  .groups = "drop")
}

#' Add a human-readable label column: "GENE (coords)" if gene known.
ce_label <- function(df) {
    df %>%
        mutate(label = ifelse(!is.na(gene_name) & gene_name != "",
                              paste0(gene_name, " (", paste_into_igv_junction, ")"),
                              paste_into_igv_junction))
}

#' Collapse multiple junctions per gene to the highest-contribution one.
#' Junctions without a gene_name are kept as distinct events.
#' Use to dedupe before picking top-N for figures.
top_ce_per_gene <- function(ranked) {
    ranked %>%
        mutate(.gene_key = ifelse(!is.na(gene_name) & gene_name != "",
                                  gene_name, paste_into_igv_junction)) %>%
        group_by(.gene_key) %>%
        slice_max(mean_scale_psi, n = 1, with_ties = FALSE) %>%
        ungroup() %>%
        select(-.gene_key) %>%
        arrange(desc(mean_scale_psi))
}

#' Compute the ranked CE table + cumulative-burden saturation summary
#' for a single cohort. Joins penetrance (from raw PSI) onto the ranked
#' z-scored contributions. Returns ranked / cumulative / n50 / n80 / n_tot.
compute_ce_drivers <- function(contributions, burden_tbl,
                               group_col, group_value,
                               cohort_label,
                               queried_data,
                               samples_meta,
                               junction_gene_tbl,
                               psi_detection_thresh = 0) {

    ranked <- rank_ce_in_group(contributions, burden_tbl,
                               group_col, group_value,
                               gene_annotation = junction_gene_tbl)
    pen    <- compute_ce_penetrance(queried_data, samples_meta,
                                    group_col, group_value,
                                    psi_detection_thresh = psi_detection_thresh)
    ranked <- ranked %>% left_join(pen, by = "paste_into_igv_junction")

    cum <- ranked %>%
        arrange(desc(mean_scale_psi)) %>%
        mutate(cum_contrib = cumsum(mean_scale_psi),
               cum_frac    = cum_contrib / sum(mean_scale_psi),
               rank        = row_number(),
               rank_frac   = rank / n())
    n50   <- cum %>% filter(cum_frac >= 0.50) %>% slice_head(n = 1) %>% pull(rank)
    n80   <- cum %>% filter(cum_frac >= 0.80) %>% slice_head(n = 1) %>% pull(rank)
    n_tot <- nrow(cum)
    list(ranked = ranked, cum = cum, cohort = cohort_label,
         n50 = n50, n80 = n80, n_tot = n_tot)
}
```

```{r gene-annotation}
junction_gene <- summary_post %>%
    distinct(paste_into_igv_junction, gene_name)
```

# Per-cohort contribution tables

```{r compute-contributions}
ce_contrib_motor <- compute_ce_contributions(
    nygc_queried,  motor_meta,   summary_post, "Motor Cortex")
ce_contrib_frontal <- compute_ce_contributions(
    nygc_queried,  frontal_meta, summary_post, "FTD frontal/temporal")
ce_contrib_cord  <- compute_ce_contributions(
    nygc_queried,  cord_meta,    summary_post, "Cord")
ce_contrib_rimod <- compute_ce_contributions(
    rimod_queried, rimod_meta,   summary_post, "RiMOD")
```

# Q1: which CEs drive the burden?

The burden score is `sum(scale_psi)` across passing CEs per sample. To
answer "which CEs drive it" within a disease group we just take the mean
of `scale_psi` per CE across samples in that group. We do this for all
four burden cohorts so the answer can be generalised.

```{r drivers-all-cohorts}
driver_specs <- tribble(
    ~contributions,     ~burden,        ~queried_data, ~samples_meta, ~group_col,          ~group,     ~cohort_label,                       ~prefix,
    ce_contrib_motor,   motor_burden,   nygc_queried,  motor_meta,    "tdp_proteinopathy", "ALS-TDP",  "NYGC Motor Cortex (ALS-TDP)",       "nygc_motor_alstdp",
    ce_contrib_frontal, frontal_burden, nygc_queried,  frontal_meta,  "tdp_proteinopathy", "FTLD-TDP", "NYGC Frontal/Temporal (FTLD-TDP)",  "nygc_frontal_ftldtdp",
    ce_contrib_cord,    cord_burden,    nygc_queried,  cord_meta,     "tdp_proteinopathy", "ALS-TDP",  "NYGC Spinal Cord (ALS-TDP)",        "nygc_cord_alstdp",
    ce_contrib_rimod,   rimod_burden,   rimod_queried, rimod_meta,    "tdp_proteinopathy", "FTLD-TDP", "RiMOD Frontal Cortex (FTLD-TDP)",   "rimod_ftldtdp"
)

drivers <- pmap(driver_specs, function(contributions, burden, queried_data,
                                       samples_meta, group_col, group,
                                       cohort_label, prefix) {
    compute_ce_drivers(contributions       = contributions,
                       burden_tbl          = burden,
                       group_col           = group_col,
                       group_value         = group,
                       cohort_label        = cohort_label,
                       queried_data        = queried_data,
                       samples_meta        = samples_meta,
                       junction_gene_tbl   = junction_gene,
                       psi_detection_thresh = 0)
}) %>% set_names(driver_specs$prefix)

# Persist the ranked tables per cohort
walk2(drivers, names(drivers), function(d, prefix) {
    write_csv(d$ranked,
              file.path(OUTPUT_DIR, paste0(prefix, "_ranked_ce.csv")))
})
```

## Cumulative-burden saturation per cohort

```{r saturation-summary}
driver_summary <- imap_dfr(drivers, function(d, prefix) {
    tibble(cohort  = d$cohort,
           n_total = d$n_tot,
           n50     = d$n50,  pct50 = round(100 * d$n50 / d$n_tot, 1),
           n80     = d$n80,  pct80 = round(100 * d$n80 / d$n_tot, 1))
})
driver_summary
write_csv(driver_summary,
          file.path(OUTPUT_DIR, "ce_drivers_saturation_summary.csv"))
```

## Figure 1: cumulative contribution to burden across cohorts

One curve per cohort, x = rank as a fraction of passing CEs (so cohorts
are directly comparable despite different panel sizes). The dashed
diagonal is the uniform-contribution reference: a CE panel where every CE
contributes equally would trace it exactly. The further a curve sits below
the diagonal, the more concentrated the burden is in a few CEs.

```{r fig1-cumulative, fig.width=10, fig.height=7}
cohort_colours <- c(
    `NYGC Motor Cortex (ALS-TDP)`       = "#5F4BB6",
    `NYGC Frontal/Temporal (FTLD-TDP)`  = "#ACD7E5",
    `NYGC Spinal Cord (ALS-TDP)`        = "#4A5240",
    `RiMOD Frontal Cortex (FTLD-TDP)`   = "#EF7F7F"
)


all_cum <- imap_dfr(drivers, function(d, prefix) {
    d$cum %>% mutate(cohort = d$cohort)
})

cohort_anchors <- imap_dfr(drivers, function(d, prefix) {
    tibble(cohort = d$cohort,
           x      = d$n50 / d$n_tot,
           y      = 0.5,
           label  = sprintf("%s: 50%% at %d/%d (%.0f%%)",
                            d$cohort, d$n50, d$n_tot,
                            100 * d$n50 / d$n_tot))
})

fig_cumulative <- ggplot(all_cum, aes(rank_frac, cum_frac, colour = cohort)) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", colour = "grey50") +
    geom_step(linewidth = 1) +
    geom_hline(yintercept = c(0.5, 0.8),
               linetype = "dotted", colour = "grey70") +
    scale_colour_manual(values = cohort_colours, name = NULL) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                       limits = c(0, 1)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                       limits = c(0, 1)) +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    theme(legend.position = "bottom") +
    labs(x = "Fraction of CEs (rank / total)",
         y = "Cumulative fraction of mean burden",
         title = "Cumulative contribution to cryptic burden, by cohort",
         subtitle = "Dashed line = uniform-contribution reference. Curves further above the diagonal = more concentrated burden.")

ggsave(file.path(OUTPUT_DIR, "fig1_cumulative_contribution_all_cohorts.pdf"),
       fig_cumulative, width = 10, height = 7)
ggsave(file.path(OUTPUT_DIR, "fig1_cumulative_contribution_all_cohorts.png"),
       fig_cumulative, width = 10, height = 7, dpi = 200)

fig_cumulative
```

**Interpretation.** Because CEs are sorted in descending order of
contribution, every curve sits at or above the dashed diagonal; the
diagonal itself would correspond to perfectly uniform contribution (every
CE adds the same amount). The further a curve bows above the diagonal,
the more concentrated the burden is in its top-ranked CEs. Three cortical
cohorts (motor, frontal/temporal, RiMOD) hug the diagonal closely — the
burden is broadly distributed across many CEs and no single CE dominates.
Spinal cord bows noticeably further above the diagonal: the top ~22% of
CEs accounts for 50% of the burden, suggesting a more focal contribution
in cord. Sample size and the smaller passing-CE panel in cord may
contribute to this; the comparison is descriptive, not mechanistically
established.

## Figure 2: top-20 CEs per cohort

Because some events generate multiple junctions per gene (e.g. a single
cryptic exon contributes two flanking junctions), each cohort's top-20 is
first deduplicated by gene: for genes with multiple passing junctions we
keep the one with the highest mean contribution. Unannotated junctions
(no gene_name) are retained as distinct events.

One panel per cohort, top 20 CEs by mean contribution, bars coloured by
penetrance (the fraction of samples in that group with `scale_psi > 0`).

```{r fig2-top20, fig.width=16, fig.height=10}
build_top20_panel <- function(d) {
    d$ranked %>%
        top_ce_per_gene() %>%      # collapse to one junction per gene first
        ce_label() %>%
        slice_head(n = 20) %>%
        ggplot(aes(x    = mean_scale_psi,
                   y    = fct_reorder(label, mean_scale_psi),
                   fill = penetrance)) +
        geom_col() +
        scale_fill_gradient(low = "grey80", high = "#D64933",
                            limits = c(0, 1),
                            name = "Penetrance") +
        theme_bw(base_size = 11) +
        labs(x = "Mean contribution to cryptic burden",
             y = NULL,
             title = d$cohort)
}

panels_top20    <- map(drivers, build_top20_panel)

fig_top20 <- wrap_plots(panels_top20, ncol = 2) +
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "A",
                    title = "Top-20 CEs by mean contribution to burden",
                    theme = theme(plot.title = element_text(face = "bold")))

ggsave(file.path(OUTPUT_DIR, "fig2_top20_ces_all_cohorts.pdf"),
       fig_top20, width = 16, height = 10)
ggsave(file.path(OUTPUT_DIR, "fig2_top20_ces_all_cohorts.png"),
       fig_top20, width = 16, height = 10, dpi = 200)

fig_top20
```

## Figure 3: cross-cohort overlap of the top-20 drivers

A reviewer-facing direct comparison: of the four cohorts' top-20 gene-
deduped drivers, which genes appear in the top-20 of more than one
cohort? The union of these sets defines the "ever top-20" gene list;
each cohort marks whether (and at what rank) each gene appears in its
own top-20. Genes are ordered by cohort breadth (how many of the four
cohorts they appear in) and then by mean rank.

```{r fig3-overlap}
# Top-20 deduped per cohort, in a long table
top20_long <- imap_dfr(drivers, function(d, prefix) {
    d$ranked %>%
        top_ce_per_gene() %>%
        slice_head(n = 20) %>%
        mutate(rank     = row_number(),
               cohort   = d$cohort,
               gene_key = ifelse(!is.na(gene_name) & gene_name != "",
                                 gene_name, paste_into_igv_junction))
})

# Cohort breadth and mean rank per gene_key, used for ordering
gene_breadth <- top20_long %>%
    group_by(gene_key) %>%
    summarise(n_cohorts = n_distinct(cohort),
              mean_rank = mean(rank),
              best_rank = min(rank),
              .groups   = "drop") %>%
    arrange(desc(n_cohorts), mean_rank)

gene_levels <- gene_breadth %>% pull(gene_key)

cohort_levels <- c("NYGC Motor Cortex (ALS-TDP)",
                   "NYGC Frontal/Temporal (FTLD-TDP)",
                   "NYGC Spinal Cord (ALS-TDP)",
                   "RiMOD Frontal Cortex (FTLD-TDP)")

write_csv(top20_long  %>% select(cohort, rank, gene_key, gene_name,
                                 paste_into_igv_junction, mean_scale_psi,
                                 penetrance, mean_psi),
          file.path(OUTPUT_DIR, "top20_per_cohort_long.csv"))
write_csv(gene_breadth,
          file.path(OUTPUT_DIR, "top20_gene_breadth.csv"))

message("\nTop-20 driver overlap summary:")
print(gene_breadth %>% count(n_cohorts, name = "n_genes"))
```

```{r fig3-overlap-plot, fig.width=11, fig.height=11}
p_tile <- top20_long %>%
    mutate(gene_key = factor(gene_key, levels = rev(gene_levels)),
           cohort   = factor(cohort,   levels = cohort_levels)) %>%
    ggplot(aes(x = cohort, y = gene_key, fill = rank)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(label  = rank,
                  colour = rank > 10), size = 3) +
    scale_fill_gradient(low = "#D64933", high = "#FAD3CB",
                        limits = c(1, 20), name = "Rank") +
    scale_colour_manual(values = c(`TRUE` = "black", `FALSE` = "white"),
                        guide  = "none") +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    theme(axis.text.x = element_text(angle = 25, hjust = 1),
          panel.grid  = element_blank()) +
    labs(x = NULL, y = NULL,
         title = "Rank of each top-20 driver gene across cohorts",
         subtitle = "Filled cell = gene is in that cohort's top 20; number = rank within that cohort")

p_breadth <- gene_breadth %>%
    mutate(gene_key = factor(gene_key, levels = rev(gene_levels))) %>%
    ggplot(aes(x = n_cohorts, y = gene_key)) +
    geom_col(fill = "grey40") +
    geom_text(aes(label = n_cohorts), hjust = -0.3, size = 3) +
    scale_x_continuous(limits = c(0, 4.6), breaks = 1:4) +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major.y = element_blank()) +
    labs(x = "# cohorts in top-20", y = NULL,
         title = "Cohort breadth")

fig_top20_overlap <- p_tile + p_breadth + plot_layout(widths = c(3.2, 1))

ggsave(file.path(OUTPUT_DIR, "fig3_top20_overlap_across_cohorts.pdf"),
       fig_top20_overlap, width = 11, height = 11)
ggsave(file.path(OUTPUT_DIR, "fig3_top20_overlap_across_cohorts.png"),
       fig_top20_overlap, width = 11, height = 11, dpi = 200)

fig_top20_overlap
```

**Interpretation.** Genes at the top of the figure (breadth = 4) are
top-20 drivers in every cohort — the canonical, tissue-agnostic CE set
contributing most to the composite burden. Genes appearing in only one
cohort (breadth = 1) are the cohort-specific tail. The number of
universal-vs-specific genes is a direct readout of whether the burden
score is dominated by a shared core or by tissue-/disease-specific
subsets.

# Q2: is GRN-FTLD-TDP just C9 with the volume turned up?

Test the hypothesis "the GRN burden is a uniform amplification of the C9
burden" by comparing per-CE mean contributions in the two groups within
each FTLD cohort. If the hypothesis holds, mean contributions should be
strongly correlated (same CEs drive both), with the GRN means shifted up
relative to C9 (positive intercept and/or slope > 1).

## Per-cohort GRN and C9 rankings

```{r grn-c9-rankings}
ranked_frontal_grn <- rank_ce_in_group(
    ce_contrib_frontal, frontal_burden,
    "simplified_mutations", "GRN",     gene_annotation = junction_gene)
ranked_frontal_c9  <- rank_ce_in_group(
    ce_contrib_frontal, frontal_burden,
    "simplified_mutations", "C9ORF72", gene_annotation = junction_gene)
ranked_rimod_grn   <- rank_ce_in_group(
    ce_contrib_rimod, rimod_burden,
    "DiseaseCode", "FTD-GRN",          gene_annotation = junction_gene)
ranked_rimod_c9    <- rank_ce_in_group(
    ce_contrib_rimod, rimod_burden,
    "DiseaseCode", "FTD-C9",           gene_annotation = junction_gene)

# Paired tables (GRN, C9) per cohort
make_paired <- function(grn, c9) {
    grn %>%
        select(paste_into_igv_junction, gene_name,
               mean_grn = mean_scale_psi) %>%
        inner_join(c9 %>% select(paste_into_igv_junction,
                                 mean_c9 = mean_scale_psi),
                   by = "paste_into_igv_junction") %>%
        mutate(delta = mean_grn - mean_c9)
}
paired_nygc  <- make_paired(ranked_frontal_grn, ranked_frontal_c9)
paired_rimod <- make_paired(ranked_rimod_grn,   ranked_rimod_c9)
```

## Within-cohort GRN vs C9 means

```{r grn-vs-c9-stats}
fit_grn_c9 <- function(paired) {
    fit <- lm(mean_grn ~ mean_c9, data = paired)
    s   <- summary(fit)
    list(fit       = fit,
         intercept = unname(coef(fit)[1]),
         slope     = unname(coef(fit)[2]),
         slope_ci  = confint(fit)["mean_c9", ],
         pearson   = cor.test(paired$mean_grn, paired$mean_c9, method = "pearson"),
         spearman  = suppressWarnings(cor.test(paired$mean_grn, paired$mean_c9,
                                               method = "spearman")))
}

fit_nygc  <- fit_grn_c9(paired_nygc)
fit_rimod <- fit_grn_c9(paired_rimod)

within_cohort_stats <- tibble(
    cohort      = c("NYGC frontal/temporal", "RiMOD frontal cortex"),
    pearson_r   = c(fit_nygc$pearson$estimate,  fit_rimod$pearson$estimate),
    pearson_p   = c(fit_nygc$pearson$p.value,   fit_rimod$pearson$p.value),
    spearman_r  = c(fit_nygc$spearman$estimate, fit_rimod$spearman$estimate),
    spearman_p  = c(fit_nygc$spearman$p.value,  fit_rimod$spearman$p.value),
    intercept   = c(fit_nygc$intercept,         fit_rimod$intercept),
    slope       = c(fit_nygc$slope,             fit_rimod$slope)
)
within_cohort_stats
```

Within each cohort the per-CE means in GRN and C9 are essentially
uncorrelated (NYGC) or weakly anti-correlated (RiMOD). That looks like a
rejection of "uniform volume scaling" at first glance, but the relevant
quantity is the GRN-minus-C9 *delta* per CE, not the GRN-vs-C9 correlation
— see the cross-cohort analysis below.

## Cross-cohort reproducibility of the GRN-vs-C9 delta

```{r cross-cohort}
cross_cohort <- paired_nygc %>%
    select(paste_into_igv_junction, gene_name,
           delta_nygc = delta, mean_grn_nygc = mean_grn, mean_c9_nygc = mean_c9) %>%
    inner_join(
        paired_rimod %>% select(paste_into_igv_junction,
                                delta_rimod    = delta,
                                mean_grn_rimod = mean_grn,
                                mean_c9_rimod  = mean_c9),
        by = "paste_into_igv_junction"
    )

n_overlap <- nrow(cross_cohort)

quad <- cross_cohort %>%
    mutate(quadrant = case_when(
        delta_nygc > 0 & delta_rimod > 0 ~ "GRN-high in both",
        delta_nygc < 0 & delta_rimod < 0 ~ "C9-high in both",
        delta_nygc > 0 & delta_rimod < 0 ~ "GRN-high NYGC, C9-high RiMOD",
        delta_nygc < 0 & delta_rimod > 0 ~ "C9-high NYGC, GRN-high RiMOD",
        TRUE ~ "tie"))

quad_counts <- quad %>% count(quadrant)
quad_counts

concordant <- sum(quad$quadrant %in% c("GRN-high in both", "C9-high in both"))
discordant <- sum(quad$quadrant %in% c("GRN-high NYGC, C9-high RiMOD",
                                       "C9-high NYGC, GRN-high RiMOD"))
binom_test <- binom.test(concordant, concordant + discordant, p = 0.5,
                         alternative = "greater")

cross_pearson  <- cor.test(cross_cohort$delta_nygc, cross_cohort$delta_rimod,
                           method = "pearson")
cross_spearman <- suppressWarnings(cor.test(cross_cohort$delta_nygc,
                                            cross_cohort$delta_rimod,
                                            method = "spearman"))

cross_stats <- tibble(
    n_CEs        = n_overlap,
    pearson_r    = cross_pearson$estimate,
    pearson_p    = cross_pearson$p.value,
    spearman_rho = cross_spearman$estimate,
    spearman_p   = cross_spearman$p.value,
    concordant   = concordant,
    discordant   = discordant,
    pct_concord  = round(100 * concordant / (concordant + discordant), 1),
    binom_p      = binom_test$p.value
)
cross_stats
```

```{r reproducible-lists}
reproducible_grn_high <- cross_cohort %>%
    filter(delta_nygc > 0, delta_rimod > 0) %>%
    mutate(combined_delta = delta_nygc + delta_rimod) %>%
    arrange(desc(combined_delta))

reproducible_c9_high <- cross_cohort %>%
    filter(delta_nygc < 0, delta_rimod < 0) %>%
    mutate(combined_delta = delta_nygc + delta_rimod) %>%
    arrange(combined_delta)

write_csv(cross_cohort,
          file.path(OUTPUT_DIR, "cross_cohort_grn_vs_c9.csv"))
write_csv(reproducible_grn_high,
          file.path(OUTPUT_DIR, "reproducible_grn_high_ces.csv"))
write_csv(reproducible_c9_high,
          file.path(OUTPUT_DIR, "reproducible_c9_high_ces.csv"))
```

## Figure 4: cross-cohort reproducibility of GRN-vs-C9 delta

Two-panel figure. Panel A: scatter of per-CE GRN-minus-C9 delta in NYGC
(x) vs RiMOD (y), with sign-concordant CEs in red and discordant CEs in
grey. Panel B: paired bar plot of the top-15 reproducibly-GRN-high CEs
(positive delta in both cohorts), side-by-side per cohort.

```{r fig4-cross-cohort, fig.width=14, fig.height=12}
# Labels: top concordant in each direction + biggest discordant CEs
to_label <- bind_rows(
    reproducible_grn_high %>% slice_head(n = 8),
    reproducible_c9_high  %>% slice_head(n = 8),
    cross_cohort %>%
        filter(sign(delta_nygc) != sign(delta_rimod),
               !is.na(delta_nygc), !is.na(delta_rimod)) %>%
        mutate(disc = abs(delta_nygc - delta_rimod)) %>%
        arrange(desc(disc)) %>% slice_head(n = 5)
) %>% distinct(paste_into_igv_junction) %>% pull(paste_into_igv_junction)

stats_text <- sprintf(
    "Pearson r = %.3f (p = %.2g)\nSpearman ρ = %.3f (p = %.2g)\nConcordant: %d / %d (%.0f%%, binom p = %.2g)",
    cross_pearson$estimate,  cross_pearson$p.value,
    cross_spearman$estimate, cross_spearman$p.value,
    concordant, concordant + discordant,
    100 * concordant / (concordant + discordant),
    binom_test$p.value
)

p_cross <- cross_cohort %>%
    mutate(label_show = gene_name,
           sign_match = sign(delta_nygc) == sign(delta_rimod) &
               delta_nygc != 0 & delta_rimod != 0) %>%
    ggplot(aes(delta_nygc, delta_rimod)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(aes(colour = sign_match), alpha = 0.7) +
    scale_colour_manual(values = c(`TRUE` = "#D64933", `FALSE` = "grey60"),
                        labels = c(`TRUE` = "Same direction",
                                   `FALSE` = "Opposite"),
                        name = NULL) +
    geom_text(aes(label = label_show), size = 3,
              na.rm = TRUE, check_overlap = TRUE, vjust = -0.5) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.2,
             size = 3.5, label = stats_text) +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    labs(x = "GRN - C9 mean contribution (NYGC)",
         y = "GRN - C9 mean contribution (RiMOD)",
         title = sprintf("Cross-cohort reproducibility of GRN-C9 delta (n CEs = %d)",
                         n_overlap))

p_repro_bar <- reproducible_grn_high %>%
    slice_head(n = 15) %>%
    pivot_longer(c(delta_nygc, delta_rimod),
                 names_to = "cohort", values_to = "delta") %>%
    mutate(cohort = recode(cohort, delta_nygc = "NYGC", delta_rimod = "RiMOD"),
           label  = ifelse(!is.na(gene_name) & gene_name != "",
                           paste0(gene_name, " (", paste_into_igv_junction, ")"),
                           paste_into_igv_junction)) %>%
    ggplot(aes(x = delta,
               y = fct_reorder(label, delta, .fun = max),
               fill = cohort)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.7) +
    scale_fill_manual(values = c(NYGC = "#D64933", RiMOD = "#2E6F95")) +
    theme_bw(base_size = PLOT_TEXT_SIZE) +
    labs(x = "GRN − C9 mean contribution",
         y = NULL, fill = "Cohort",
         title = "Reproducible 'GRN-high vs C9' CEs (top 15 by combined delta)")

fig_cross <- p_cross / p_repro_bar +
    plot_layout(heights = c(1, 1.2)) +
    plot_annotation(tag_levels = "A")

ggsave(file.path(OUTPUT_DIR, "fig4_grn_vs_c9_cross_cohort.pdf"),
       fig_cross, width = 14, height = 12)
ggsave(file.path(OUTPUT_DIR, "fig4_grn_vs_c9_cross_cohort.png"),
       fig_cross, width = 14, height = 12, dpi = 200)

fig_cross
```

