---
title: "ftd_cluster_processing"
author: "Anna-Leigh Brown"
date: "2024-07-29"
output: html_document
---

```{r setup, include=FALSE}
library(data.table)
library(pROC)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
give.n <- function(x){
return(c(y = median(x)*1.6, label = length(x)))
}
```

### Reading in the data


```{r readData}
master_table <- fread(here::here("data/master_table.csv"))
compendium_events = fread('data/compendium_cryptics.csv')
cell_specific = fread("/Users/annaleigh/Downloads/cell_specific_cryptics.csv")
meta_rimod = fread("data/metadata_rimod.csv")

plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#1331a7", "Intermediate" = "steelblue","SH-SY5Y only" = 'grey','SK-N-BE(2) only' = 'darkgrey')

cell_type_df = cell_specific %>% 
    select(junctions_coords,gene_name,matches('dpsi')) %>% 
    melt() %>% 
    separate(variable, into = c(NA,'celltype')) %>% 
    group_by(junctions_coords) %>% 
    mutate(cryptic = value >=0.1) %>% 
    mutate(n_cellType = sum(cryptic,na.rm = TRUE)) %>% 
    ungroup() %>% 
    filter(cryptic == TRUE) %>% 
    group_by(junctions_coords,gene_name) %>% 
    reframe(cells = paste(celltype, collapse = ','),n_cellType) %>% unique()

master_table = master_table %>% 
    mutate(tdp43_sensitivity_be2 = ifelse(tdp43_sensitivity_be2 == '', "SH-SY5Y only",tdp43_sensitivity_be2)) %>% 
    mutate(tdp43_sensitivity_sh = ifelse(tdp43_sensitivity_sh == '', "SK-N-BE(2) only",tdp43_sensitivity_sh)) %>% 
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "SH-SY5Y only" ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_sh == "SK-N-BE(2) only" ~ tdp43_sensitivity_be2,
                                tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_be2 !=  tdp43_sensitivity_sh ~ 'Ambigious'))
    

rimod_junction = arrow::read_parquet('/Users/annaleigh/Desktop/rimod_clustered.parquet')
rimod_clustered = rimod_junction %>% 
    melt(id.vars = c("chr","start","end","strand","cluster",'paste_into_igv_junction')) %>%
    mutate(variable = gsub(".none.output","",variable)) %>% 
    left_join(meta_rimod[,.(sample,disease,genotype)],by = c("variable" = "sample")) %>% 
    as.data.table()
```

## Filter spurious junctions

Leafcutter's clustering script can be a bit aggressive, so we'll going to post-hoc filter out junctions without enough expression in the control brains

```{r filterJunctions, echo=FALSE}
#keep just the cryptic clusters
these_clusters = rimod_clustered %>% 
    filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction | 
               paste_into_igv_junction %in% master_table$paste_into_igv_junction |
               paste_into_igv_junction %in% cell_specific$junctions_coords) %>% 
    pull(cluster) %>% unique()

length(these_clusters)

expressed_junctions = rimod_clustered %>% 
    filter(cluster %in% these_clusters) %>% 
    filter(disease == 'normal') %>% 
    group_by(paste_into_igv_junction) %>% 
    reframe(paste_into_igv_junction,cluster,chr,start,end,strand,avg_count = mean(value),med_count = median(value)) %>% 
    filter(med_count > 5) %>% 
    unique()  

```

Ok now we keep one of either junctions with median expression > 5 in the controls - a cryptic event found in compendium, cell type, or TDP-43 dosage curve

```{r normalizeJunctions, echo = FALSE}

rimod_clustered_long_psi = rimod_clustered %>% 
    filter(cluster %in% these_clusters) %>% 
    filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction | 
           paste_into_igv_junction %in% master_table$paste_into_igv_junction |
           paste_into_igv_junction %in% cell_specific$junctions_coords | 
           paste_into_igv_junction %in% expressed_junctions$paste_into_igv_junction) %>% 
    group_by(cluster,variable) %>% 
    mutate(cluster_count = sum(value)) %>% 
    ungroup() %>% 
    mutate(psi = value / cluster_count) 
```

Now everything has been normalized to a psi metric let's find out if anything is predictive of TDP-43 path vs not path

```{r ROCanalysis}
modeling_ftd_rpm = rimod_clustered_long_psi %>% 
        filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction | 
           paste_into_igv_junction %in% master_table$paste_into_igv_junction |
           paste_into_igv_junction %in% cell_specific$junctions_coords ) %>% #just keeping the events found in compendium, curve, or cell type
    mutate(psi = ifelse(is.na(psi),0,psi)) %>% 
    mutate(psi = ifelse(is.nan(psi),0,psi)) %>% 
    mutate(numeric_prediction = ifelse(genotype %in% c("C9orf72","GRN"), 1, 0)) %>% 
    distinct(variable,paste_into_igv_junction,numeric_prediction,psi,genotype) %>% 
    pivot_wider(values_from = 'psi',
                names_from = 'paste_into_igv_junction',values_fill = 0) %>% 
    as.data.table() 


modeling_ftd_rpm_norm = modeling_ftd_rpm %>% 
  mutate(across(starts_with("chr"), ~ scale(.) %>% as.vector())) %>% 
    select(where(~ !any(is.nan(.))))


library(pROC)


ROCvector = modeling_ftd_rpm_norm %>% colnames()
ROCvector = ROCvector[!(ROCvector %in% c("variable", "numeric_prediction","genotype"))]


my_df = data.table()

for (predictor in ROCvector) {

    pROC_obj <- roc_(data = modeling_ftd_rpm_norm,"numeric_prediction",predictor,quiet = TRUE)
    
    predictor_value <- 0
    fpr_value <- coords(pROC_obj, x = predictor_value, input = "threshold", ret = "specificity", best.method = "closest.topleft", transpose = FALSE)
    # Calculate the FPR
    fpr_value <- 1 - fpr_value
    tpr_value <- coords(pROC_obj, x = predictor_value, input = "threshold", ret = "sensitivity", best.method = "closest.topleft", transpose = FALSE)

    lil_list = list(paste_into_igv_junction = predictor,
                    auc = as.numeric(pROC_obj$auc),
                    avg_fpr = 1 - mean(pROC_obj$specificities),
                    avg_tpr = mean(pROC_obj$sensitivities),
                    fpr_value = as.numeric(fpr_value),
                    tpr_value = as.numeric(tpr_value))
    my_df = rbind(my_df,lil_list)

}

```

Alright we have a df of the True Positive Rate, False Positive Rate and auc. Arguable the FPR is more important than AUC for these questions

```{r filterLowFPR}

my_df %>% 
    left_join(compendium_events %>% distinct(gene_name,paste_into_igv_junction,junc_cat,n_significant_data)) %>% 
    filter(!is.na(n_significant_data)) %>% 
    arrange(avg_fpr) %>% 
    mutate(binned_fpr = cut_interval(avg_fpr,n = 5)) %>% 
    mutate(binned_ds = cut_interval(n_significant_data,n = 3)) %>% 
    arrange(-n_significant_data) %>% 
    mutate(outliers = ifelse(n_significant_data == 1 & avg_fpr < 0.1,gene_name,NA_character_)) %>% 
    # filter(n_significant_data > 3 & FPR > 0.5)
    ggplot(aes(y = avg_fpr, x = binned_ds)) + 
    geom_boxplot() +
    stat_summary(fun.data = give.n, geom = "text", 
                 fun.y = median,
                 position = position_dodge(width = 0.75)) + 
    theme_classic() +
    ggrepel::geom_text_repel(aes(label = outliers),max.overlaps = 100) + 
    geom_hline(yintercept = 0.1, linetype = 'dotted')


my_df %>% 
    left_join(compendium_events %>% distinct(gene_name,paste_into_igv_junction,junc_cat,n_significant_data)) %>% 
    filter(!is.na(n_significant_data)) %>% 
    arrange(avg_fpr) %>% 
    filter(n_significant_data == 1 & avg_fpr < 0.1)
```

