---
title: "ftd_cluster_processing"
author: "Anna-Leigh Brown"
date: "2024-07-29"
output: html_document
---

```{r setup, include=FALSE}
library(data.table)
library(pROC)
library(tidyverse)
library(rstatix)
library(ggpubr)
library(dunn.test)
library(tidy)
knitr::opts_chunk$set(echo = FALSE)
give.n <- function(x){
return(c(y = median(x)*1.6, label = length(x)))
}

junction_plot = function(junction,title = "",metric = 'psi',table = rimod_clustered_long_psi){
    if(metric == 'psi'){
            p = table %>% 
        mutate(path = genotype %in% c('control', 'MAPT')) %>% 
        mutate(genotype = fct_relevel(genotype,'control','MAPT')) %>% 
        as.data.table() %>% 
        filter(paste_into_igv_junction == junction) %>% 
        ggplot(aes(y = psi,x = genotype,fill = path)) + 
        geom_boxplot() + 
        geom_jitter(height = 0) +
        theme_classic() + 
        ylab("Junction PSI") +
        scale_fill_manual(values = c("red",'grey'),name = "TDP-43 proteinopathy") +
        ggtitle(title) + 
        theme(legend.position = 'none')
    
    }else{

        p = table %>% 
        mutate(path = genotype %in% c('control', 'MAPT')) %>% 
        mutate(genotype = fct_relevel(genotype,'control','MAPT')) %>% 
        as.data.table() %>% 
        filter(paste_into_igv_junction == junction) %>% 
        ggplot(aes(y = value,x = genotype,fill = path)) + 
        geom_boxplot() + 
        geom_jitter(height = 0) +
        theme_classic() + 
        ylab("Junction count") +
        scale_fill_manual(values = c("red",'grey'),name = "TDP-43 proteinopathy") +
        ggtitle(title) + 
        theme(legend.position = 'none')
    
    }

    print(p)
    
    return(p)

}
```


### Reading in the data


```{r readData}
master_table <- read.csv(here::here("data/master_table.csv"))
master_table <- master_table %>%
    mutate(tdp43_sensitivity_be2 = ifelse(tdp43_sensitivity_be2 == "","Not found",tdp43_sensitivity_be2)) %>% 
    mutate(tdp43_sensitivity_sh = ifelse(tdp43_sensitivity_sh == "","Not found",tdp43_sensitivity_sh)) %>% 
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "Not found" ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_sh == "Not found" ~ tdp43_sensitivity_be2,
                                tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                                tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late',
                                tdp43_sensitivity_sh == 'Intermediate'~  'Intermediate')) %>% 
        mutate(new_cate = ifelse(is.na(new_cate),"Ambiguous",new_cate)) 

compendium_events = fread('data/compendium_cryptics.csv')
meta_rimod = fread("data/metadata_rimod.csv")
wide_chx = fread('data/wide_chx.csv')
dark_chx = wide_chx %>% filter(chx_effect == 'Dark cryptic')
plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#1331a7", "Intermediate" = "steelblue","SH-SY5Y only" = 'grey','SK-N-BE(2) only' = 'darkgrey')

rimod_clustered = fread("~/Desktop/RIMOD_Lowered_perind.counts.gz")
```

## Rearrange RiMOD FTD

```{r}
rimod_clustered_long_psi = rimod_clustered %>% 
    melt(id.vars = c("chrom")) %>% 
    dt_separate(value,into =  c("junction_count", "cluster_count"), sep = '/') 

rimod_clustered_long_psi = rimod_clustered_long_psi %>% 
    mutate(variable = gsub(".none.output","",variable)) %>% 
    rename(sample = variable) %>% 
    mutate(junction_count = as.numeric(junction_count),cluster_count = as.numeric(cluster_count)) %>% 
    dt_separate(chrom,into =  c("seqname", "start","end",'cluster'), sep = ':') %>% 
    mutate(strand = stringr::str_sub(cluster,-1,-1))

rimod_clustered_long_psi = rimod_clustered_long_psi %>% 
    mutate(paste_into_igv_junction = paste0(seqname,":",start,"-",end))
#keep just the cryptic clusters
these_clusters = rimod_clustered_long_psi %>%
    filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction |
               paste_into_igv_junction %in% master_table$paste_into_igv_junction |
               paste_into_igv_junction %in% dark_chx$paste_into_igv_junction) %>%
    pull(cluster) %>% unique()


rimod_clustered_long_psi %>% 
    filter(cluster %in% these_clusters) %>% 
    mutate(psi = junction_count / cluster_count) %>% 
    left_join(meta_rimod) 
    
#     filter(cluster %in% these_clusters)
```


Now everything has been normalized to a psi metric let's find out if anything is predictive of TDP-43 path vs not path

```{r ROCanalysis}
modeling_ftd_rpm = rimod_clustered_long_psi %>% 
        left_join(meta_rimod) %>% 
        filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction | 
           paste_into_igv_junction %in% master_table$paste_into_igv_junction ) %>% 
        mutate(psi = junction_count / cluster_count) %>% #just keeping the events found in compendium, curve, or cell type
    mutate(psi = ifelse(is.na(psi),0,psi)) %>% 
    mutate(psi = ifelse(is.nan(psi),0,psi)) %>% 
    mutate(numeric_prediction = ifelse(genotype %in% c("C9orf72","GRN"), 1, 0)) %>% 
    distinct(sample,paste_into_igv_junction,numeric_prediction,psi,genotype) %>% 
    pivot_wider(values_from = 'psi',
                names_from = 'paste_into_igv_junction',values_fill = 0) %>% 
    as.data.table() 


modeling_ftd_rpm_norm = modeling_ftd_rpm %>% 
  mutate(across(starts_with("chr"), ~ scale(.) %>% as.vector())) %>% 
    select(where(~ !any(is.nan(.))))


library(pROC)


ROCvector = modeling_ftd_rpm_norm %>% colnames()
ROCvector = ROCvector[!(ROCvector %in% c("sample", "numeric_prediction","genotype"))]


my_df = data.table()

for (predictor in ROCvector) {

    pROC_obj <- roc_(data = modeling_ftd_rpm_norm,"numeric_prediction",predictor,quiet = TRUE)
    
    predictor_value <- 0
    fpr_value <- coords(pROC_obj, x = predictor_value, input = "threshold", ret = "specificity", best.method = "closest.topleft", transpose = FALSE)
    # Calculate the FPR
    fpr_value <- 1 - fpr_value
    tpr_value <- coords(pROC_obj, x = predictor_value, input = "threshold", ret = "sensitivity", best.method = "closest.topleft", transpose = FALSE)

    lil_list = list(paste_into_igv_junction = predictor,
                    auc = as.numeric(pROC_obj$auc),
                    avg_fpr = 1 - mean(pROC_obj$specificities[-c(1, length(pROC_obj$sensitivities))]),
                    avg_tpr = mean(pROC_obj$sensitivities[-c(1, length(pROC_obj$sensitivities))]),
                    fpr_value = as.numeric(fpr_value),
                    tpr_value = as.numeric(tpr_value))
    my_df = rbind(my_df,lil_list)

}

```

Alright we have a df of the True Positive Rate, False Positive Rate and auc. Arguable the FPR is more important than AUC for these questions

```{r filterLowFPR}

my_df %>%   
    left_join(compendium_events %>% distinct(gene_name,paste_into_igv_junction,junc_cat,n_significant_data)) %>% 
    filter(!is.na(n_significant_data)) %>% 
    arrange(avg_fpr) %>% 
    mutate(binned_fpr = cut_interval(fpr_value,n = 5)) %>% 
    mutate(binned_ds = cut_interval(n_significant_data,n = 3)) %>% 
    arrange(-n_significant_data) %>% 
    mutate(outliers = case_when(n_significant_data == 1 & fpr_value < 0.1 & tpr_value > 0.4 ~ gene_name,
                                n_significant_data >=7 & fpr_value > 0.2 ~ gene_name, TRUE ~ NA_character_)) %>%
    # filter(n_significant_data > 3 & FPR > 0.5)
    ggplot(aes(y =fpr_value, x = binned_ds)) + 
    geom_boxplot() +
    stat_summary(fun.data = give.n, geom = "text", 
                 fun.y = median,
                 position = position_dodge(width = 0.75)) + 
    theme_classic() +
    ggrepel::geom_text_repel(aes(label = outliers),max.overlaps = 100) + 
    geom_hline(yintercept = 0.1, linetype = 'dotted') + 
    ylab("False positive rate in RiMOD FTD dataset (TDP-43 path vs not)") + 
    xlab("N datasets observed")



my_df %>% 
    left_join(compendium_events %>% distinct(gene_name,paste_into_igv_junction,junc_cat,n_significant_data)) %>% 
    filter(!is.na(n_significant_data)) %>% 
    arrange(avg_fpr) %>% 
    filter(n_significant_data == 1 & avg_fpr < 0.1)
```


```{r fpr late early}
my_df %>%  
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    filter(cryptic == 'yes') %>% 
    arrange(avg_fpr) %>% 
    mutate(binned_fpr = cut_interval(fpr_value,n = 5)) %>% 
    filter(new_cate != "Ambiguous") %>% 
    # filter(n_significant_data > 3 & FPR > 0.5)
    ggplot(aes(y =fpr_value, x = new_cate)) + 
    geom_boxplot() +
    stat_summary(fun.data = give.n, geom = "text", 
                 fun.y = median,
                 position = position_dodge(width = 0.75)) + 
    theme_classic() +
    # ggrepel::geom_text_repel(aes(label = outliers),max.overlaps = 100) + 
    geom_hline(yintercept = 0.1, linetype = 'dotted') + 
    ylab("False positive rate in RiMOD FTD dataset (TDP-43 path vs not)") + 
    xlab("TDP-43 sensitivity") 
```



```{r features_fpr}
my_df %>%   
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    arrange(avg_fpr) %>% 
    filter(fpr_value < 0.1) %>% 
    filter(new_cate != "Ambiguous") %>% 
    # filter(n_significant_data > 3 & FPR > 0.5)
    ggplot(aes(y =width, x = new_cate)) + 
    ggpubr::stat_compare_means() +
    geom_boxplot() +
    stat_summary(fun.data = give.n, geom = "text", 
                 fun.y = median,
                 position = position_dodge(width = 0.75)) + 
    theme_classic() +
    # ggrepel::geom_text_repel(aes(label = outliers),max.overlaps = 100) + 
    ylab("Width of annotated intron") + 
    xlab("TDP-43 sensitivity") +
    scale_y_log10() 

my_df %>%  
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    arrange(avg_fpr) %>% 
    filter(fpr_value <= 0.1) %>% 
    filter(new_cate != "Ambiguous") %>% 
    # filter(n_significant_data > 3 & FPR > 0.5)
    mutate(longest_ug_region = ifelse(is.na(longest_ug_region),0,longest_ug_region)) %>% 
    ggplot(aes(y =longest_ug_region, x = as.factor(new_cate))) + 
    ggpubr::stat_compare_means() +
    geom_boxplot() +
    stat_summary(fun.data = give.n, geom = "text", 
                 fun.y = median,
                 position = position_dodge(width = 0.75)) + 
    theme_classic() +
    # ggrepel::geom_text_repel(aes(label = outliers),max.overlaps = 100) + 
    ylab("Longest UG Region") + 
    xlab("TDP-43 sensitivity") +
    scale_y_log10() +
    ggtitle("FPR < 0.1 events")


my_df %>%   
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    arrange(avg_fpr) %>% 
    filter(fpr_value <= 0.1) %>% 
    filter(new_cate != "Intermediate") %>% 
    # filter(new_cate != "Ambiguous") %>% 
    # filter(n_significant_data > 3 & FPR > 0.5)
    # mutate(dist_nearest_ug_region = ifelse(is.na(dist_nearest_ug_region),300,dist_nearest_ug_region)) %>% 
    mutate(ug_density = longest_ug_region / width) %>% 
    ggplot(aes(y =dist_nearest_ug_region, x = as.factor(new_cate))) + 
    ggpubr::stat_compare_means() +
    geom_boxplot() +
    stat_summary(fun.data = give.n, geom = "text", 
                 fun.y = median,
                 position = position_dodge(width = 0.75)) + 
    theme_classic() +
    # ggrepel::geom_text_repel(aes(label = outliers),max.overlaps = 100) + 
    ylab("Nearest UG Region") + 
    xlab("TDP-43 sensitivity") +
    ggtitle("FPR < 0.1 events")


my_df %>%  
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    arrange(avg_fpr) %>% 
    # filter(fpr_value <= 0.1) %>% 
    filter(new_cate != "Ambiguous") %>% 
    mutate(ug_density = longest_ug_region / width) %>% 
    ggplot(aes(x = tpr_value,y = dist_nearest_ug_region,color = new_cate)) + 
    geom_point()

```

```{r }
my_df %>%   
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "" ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_sh == "" ~ tdp43_sensitivity_be2,
                            tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_be2 == 'Early' | tdp43_sensitivity_sh == 'Early' ~ 'Early',
                            tdp43_sensitivity_be2 == 'Late' | tdp43_sensitivity_sh == 'Late' ~ 'Late'))  %>%
    arrange(avg_fpr) %>% 
    filter(fpr_value <= 0.1) %>% 
    # filter(new_cate != "Ambiguous") %>% 
    # filter(n_significant_data > 3 & FPR > 0.5)
    # mutate(dist_nearest_ug_region = ifelse(is.na(dist_nearest_ug_region),300,dist_nearest_ug_region)) %>% 
    mutate(ug_density = longest_ug_region / width) %>% 
    ggplot(aes(y =dist_nearest_ug_region, x = as.factor(max_psi_sh))) + 
    ggpubr::stat_compare_means() +
    geom_boxplot() +
    stat_summary(fun.data = give.n, geom = "text", 
                 fun.y = median,
                 position = position_dodge(width = 0.75)) + 
    theme_classic() +
    # ggrepel::geom_text_repel(aes(label = outliers),max.overlaps = 100) + 
    ylab("Nearest UG Region") + 
    xlab("TDP-43 sensitivity") +
    ggtitle("FPR < 0.1 events")

```




```{r}
bind_table = master_table %>% 
    distinct(paste_into_igv_junction,gene,strand) %>% 
    dplyr::rename(gene_name = gene) %>% 
    rbind((compendium_events %>% distinct(gene_name,paste_into_igv_junction,strand))) %>% 
    dplyr::rename(junctions_coords = paste_into_igv_junction) %>% 
    unique()  %>% 
    left_join(annotables::grch38 %>% select(symbol,strand),by = c("gene_name" = "symbol")) %>% 
    mutate(strand_final = ifelse(is.na(strand.x),strand.y,strand.x)) %>% 
    mutate(strand_final = ifelse(strand_final == 1, "+","-")) %>% 
    select(-strand.x,-strand.y) %>% 
    dplyr::rename(strand = strand_final) %>% 
    unique()

```




Let's try a similar approach with the RiMOD data that we did with Liu - wilcox test etc

```{r wilcox_test}
wilcox_test_res = rimod_clustered_long_psi %>% 
    filter(paste_into_igv_junction %in% bind_table$junctions_coords) %>% 
    mutate(condition = ifelse(genotype %in% c("C9orf72","GRN"),"TDP-43 proteinopathy","non-TDP-43-proteinopathy")) %>% 
    group_by(paste_into_igv_junction) %>% 
    nest() |> 
    mutate(wicox_test_result = map(data, ~ broom::tidy(wilcox.test(psi ~ condition, data = .x))),
           mean_path = map_dbl(data, ~ mean(.x$psi[.x$condition == "TDP-43 proteinopathy"], na.rm = TRUE)),
           mean_nonpath = map_dbl(data, ~ mean(.x$psi[.x$condition == "non-TDP-43-proteinopathy"], na.rm = TRUE)),
           ratio = log2(mean_path / mean_nonpath)) |> 
    unnest() %>% 
    distinct(paste_into_igv_junction,p.value,ratio,mean_path,mean_nonpath,ratio) %>% 
    left_join(bind_table,by = c("paste_into_igv_junction" = "junctions_coords")) %>% 
    arrange(ratio) %>% 
    mutate(adj_pvalue = p.adjust(p.value))

fwrite(wilcox_test_res, 'data/wilcox_test_res_rimod.csv')
```


```{r rimod_contingency}
# these_ones_rimod = wilcox_test_res %>% 
#     arrange(-ratio) %>% 
#     left_join(master_table %>% select(paste_into_igv_junction,new_cate,cryptic)) %>% 
#     filter(!is.na(new_cate)) %>% 
#     filter(cryptic == 'yes') %>% 
#     filter(adj_pvalue < 0.01 & ratio > 0) %>% 
#     pull(paste_into_igv_junction)

these_ones_rimod = my_df %>% filter(fpr_value <= 0.1 & tpr_value >= 0.1) %>% pull(paste_into_igv_junction)
contingency_table_rimod <- master_table %>%
    filter(cryptic == 'yes') %>% 
    filter(new_cate != "Ambiguous") %>% 
    select(paste_into_igv_junction, new_cate) %>%
    filter(!is.na(new_cate)) %>%
    mutate(sig_diff_pos_neg = paste_into_igv_junction %in% these_ones_rimod) %>%
    group_by(new_cate, sig_diff_pos_neg) %>%
    summarise(n = n_distinct(paste_into_igv_junction), .groups = "drop") %>%
    pivot_wider(names_from = sig_diff_pos_neg, values_from = n, values_fill = 0) %>% 
    column_to_rownames("new_cate") 

pairwise_results_rimod = rstatix::pairwise_fisher_test(contingency_table_rimod) 

comparisons <- list(
    c("Early", "Intermediate"),
    c("Early", "Late"),
    c("Intermediate", "Late")
)
    

contingency_table_rimod %>%
    rownames_to_column('new_cate') %>%
    melt() %>%
    group_by(new_cate) %>%
    mutate(prop = value / sum(value)) %>% 
    ggplot(aes(x = new_cate, y = value, fill = variable)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(
        aes(label = scales::percent(prop, accuracy = 1)),
        position = position_stack(0.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = "FPR < 0.1 C9orf72+GRN\nvs Control+MAPT") +
    scale_fill_manual(values = c("#4E6E58", "#FF5E5B")) + 
    theme(legend.position = 'bottom') +
    ggsignif::geom_signif(
        comparisons = comparisons,
        annotations = scales::pvalue(pairwise_results_rimod$p.adj, accuracy = 0.001),
        y_position = c(340, 320, 300),  # Adjusted y-positions above the bars
        textsize = 3,
        tip_length = 0
    )

```


```{r rimod_psi_plot, fig.dim = c(8, 6)}
plot_data_rimod = rimod_clustered_long_psi %>% 
    mutate(psi = junction_count / cluster_count) %>% #just keeping the events found in compendium, curve, or cell type
    mutate(psi = ifelse(is.na(psi),0,psi)) %>% 
    mutate(psi = ifelse(is.nan(psi),0,psi)) %>% 
    filter(paste_into_igv_junction %in% master_table$paste_into_igv_junction) %>% 
    mutate(selective = ifelse(paste_into_igv_junction %in% these_ones_rimod,'Upregulated in C9orf72+GRN\nvs Control+MAPT', 
                              'N.S.' )) %>% 
    left_join(master_table %>% select(new_cate,paste_into_igv_junction,gene,cryptic)) %>% 
    unique() %>% 
    filter(!is.na(new_cate)) %>% 
    mutate(psi = ifelse(is.nan(psi),0,psi))

plot_data_rimod %>% 
    left_join(meta_rimod) %>% 
    filter(genotype %in% c("GRN","C9orf72")) %>% 
    filter(selective != "N.S." & cryptic == 'yes') %>% 
    ggplot(aes(x = genotype, y = psi,fill = new_cate)) + 
    geom_boxplot(outliers = FALSE) +
    theme_classic() + 
    scale_fill_manual(values = plot_sensitivity_colors) + 
    stat_compare_means(hide.ns = TRUE,
                       textsize = 3,
                       tip.length = 0) +
    theme(legend.position = "none") +
    facet_wrap(~selective) + 
    ylab("PSI")



# dunn_results <- dunn.test(subset(plot_data_rimod,condition == "TDP-43 negative" & selective != "N.S.")$psi,
#                           subset(plot_data_rimod,condition == "TDP-43 negative" & selective != "N.S.")$new_cate,
#                           kw = TRUE, label = TRUE)
# 
# plot_data_rimod %>% 
#     filter(cluster_count >=10) %>% 
#     filter(selective != "N.S.") %>% 
#     ggplot(aes(x = genotype, y = psi,fill = new_cate)) + 
#     geom_boxplot() +
#     theme_classic() + 
#     facet_wrap(~gene,scales = 'free_y') +
#     scale_fill_manual(values = plot_sensitivity_colors) + 
#     theme(legend.position = "none") 
#     facet_wrap(~gene,scales = 'free_y')
#     ggsignif::geom_signif(
#         data = subset(plot_data,condition == "TDP-43 negative"),
#         comparisons = comparisons,
#         annotations = scales::pvalue(dunn_results$P.adjusted, accuracy = 0.001),
#         y_position = c(1, 1.1, 1.2),  # Adjusted y-positions above the bars
#         textsize = 3,
#         tip_length = 0
#     ) + 
#     ylab("PSI")

```


```{r clustering_by_expression}
cell_type_fraction = fread('/Users/annaleigh/Desktop/rimod_ftd_rnaseq_cell_type_fractions.txt') 
more_meta = fread("/Users/annaleigh/Desktop/rimod_ftd_dataset_table_v3.txt")
ptest = plot_data_rimod %>% 
    filter(genotype %in% c("C9orf72","GRN")) %>% 
    filter(cryptic == 'yes' & selective != "N.S.") %>% 
    select(variable,psi,paste_into_igv_junction) %>% 
    pivot_wider(names_from = variable,
                values_from = psi) %>% 
    column_to_rownames('paste_into_igv_junction') %>% 
    prcomp()

ptest$rotation %>% as.data.frame() %>% 
    rownames_to_column("RimodID") %>% 
    left_join(more_meta) %>% 
    ggplot(aes(x = PC1, y = PC2,color = Sex)) + 
    geom_point()

ptest$rotation %>% as.data.frame() %>% 
    rownames_to_column("RimodID") %>% 
    left_join(more_meta) %>% 
    ggplot(aes(x = PC1, y = PC2,color = DiseaseCode)) + 
    geom_point()

ptest$rotation %>% as.data.frame() %>% 
    rownames_to_column("RimodID") %>% 
    left_join(more_meta) %>% 
    ggplot(aes(x = PC1, y = PC2,color = RIN)) + 
    geom_point()

pc_scores <- ptest$rotation %>% 
    as.data.frame() %>%
    rownames_to_column("RimodID") # Add row names as a column for identification

kmeans_result <- kmeans(pc_scores[, 2:3], centers = 2) # 2 clusters based on PC1 and PC2


clustered_data <- pc_scores %>%
    mutate(cluster = as.factor(kmeans_result$cluster))

clustered_data %>% 
    left_join(more_meta) %>% 
    ggplot(aes(x = PC1, y = PC2,color = as.factor(cluster))) + 
    geom_point()

clustered_data %>% 
    left_join(more_meta) %>% 
    ggplot(aes(x = cluster,y = as.numeric(Age))) + 
    geom_boxplot() + 
    stat_compare_means()
clustered_data %>% 
    left_join(more_meta) %>% 
    ggplot(aes(x = cluster,y = as.numeric(PH))) + 
    geom_boxplot() +
    stat_compare_means()

clustered_data %>% 
    left_join(more_meta) %>% 
    ggplot(aes(x = cluster,y = as.numeric(PMD))) + 
    geom_boxplot() +
    stat_compare_means()

clustered_data %>% 
    left_join(more_meta) %>% 
    ggplot(aes(x = cluster,y = as.numeric(RIN))) + 
    geom_boxplot() +
    stat_compare_means()

clustered_data %>% 
    left_join(cell_type_fraction,by = c("RimodID" = "sample_id")) %>% 
    select(cluster:ExNeurons) %>% 
    melt() %>% 
    ggplot(aes(x = cluster,y = value)) + 
    geom_boxplot() +
    stat_compare_means() + 
    facet_wrap(~variable)

clustered_data %>% 
    left_join(plot_data_rimod,by =c ('RimodID' = 'variable')) %>% 
    filter(cryptic == 'yes' & selective != "N.S.") %>% 
    ggplot(aes(x = new_cate,y = psi,fill = cluster.x)) + 
    geom_boxplot()

clustered_data %>% 
    left_join(plot_data_rimod,by =c ('RimodID' = 'variable')) %>% 
    filter(cryptic == 'yes' & selective != "N.S.") %>% 
    ggplot(aes(x = gene,y = psi,fill = cluster.x)) + 
    geom_violin() + 
    facet_wrap(~gene,scales = 'free') + 
    stat_compare_means()

clustered_data %>% 
    left_join(plot_data_rimod,by =c ('RimodID' = 'variable')) %>% 
    filter(cryptic == 'yes' & selective != "N.S.") %>% 
    filter(gene == 'NYNRIN') %>% 
    select(cluster.)

```


```{r}
full_rimod_psi = rimod_clustered %>% 
    filter(cluster %in% these_clusters) %>% 
    group_by(cluster,variable) %>% 
    mutate(cluster_count = sum(value)) %>% 
    ungroup() %>% 
    mutate(psi = value / cluster_count) %>% 
    mutate(psi = ifelse(is.na(psi),0,psi)) %>% 
    mutate(psi = ifelse(is.nan(psi),0,psi)) 

wilcox_test_res_tau = full_rimod_psi %>% 
    mutate(condition = ifelse(genotype %in% c("MAPT"),"TAU","not tau")) %>% 
    group_by(paste_into_igv_junction) %>% 
    nest() |> 
    mutate(wicox_test_result = map(data, ~ broom::tidy(wilcox.test(psi ~ condition, data = .x))),
           mean_tau = map_dbl(data, ~ mean(.x$psi[.x$condition == "TAU"], na.rm = TRUE)),
           mean_not_tau = map_dbl(data, ~ mean(.x$psi[.x$condition == "not tau"], na.rm = TRUE)),
           ratio = log2(mean_tau / mean_not_tau)) |> 
    unnest() %>% 
    distinct(paste_into_igv_junction,p.value,ratio,mean_tau,mean_not_tau,ratio) %>% 
    left_join(bind_table) %>% 
    arrange(-ratio) %>% 
    mutate(adj_pvalue = p.adjust(p.value))

wilcox_test_res_tau %>% 
    filter(adj_pvalue < 0.01) %>% 
    as.data.table() %>% 
    filter(ratio >1)

```




Cell-type fraction correlation?
```{r}
cell_type_fraction = fread('/Users/annaleigh/Desktop/rimod_ftd_rnaseq_cell_type_fractions.txt') 
rimod_clustered_long_psi %>% 
    filter(paste_into_igv_junction %in% master_table$paste_into_igv_junction) %>% 
    left_join(master_table %>% select(paste_into_igv_junction,new_cate)) %>%
    filter(new_cate %in% c("Early","Intermediate","Late")) %>% 
    filter(genotype %in% c("C9orf72",'GRN')) %>% 
    unique() %>% 
    left_join(cell_type_fraction,by = c("variable" = 'sample_id')) %>% 
    select(-(chr:cluster),-cluster_count,-value,-new_cate) %>% 
    filter(paste_into_igv_junction %in% paste_into_igv_junction)
    pivot_wider(names_from = 'paste_into_igv_junction',
                values_from = 'psi',
                values_fill = 0) %>% 
    cor_mat(-variable,-disease,-genotype)
    

cor.mat %>%
    cor_reorder() %>%
    pull_lower_triangle() %>%
    cor_plot(label = TRUE)

```

