---
title: "supplementary_fig"
author: "MZ; AM; ALB"
date: "2024-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
###load packages/libraries/scripts
library(UpSetR)
library(plyranges)
library(tidyverse)
library(ggrepel)
library(patchwork)
library(ggpubr)
library(ggsignif)
library(ggseqlogo)
library(Biostrings)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ComplexHeatmap)
library(RColorBrewer)
library(data.table)
library(ggstats)
library(ggrepel)
source(here::here("scripts/odds_ratio_plots.R"))
source(here::here('scripts/trial_nmd_plots.R'))


plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#1331a7", "Intermediate" = "steelblue")
plot_psi_colors <- c("Top" = "#388f78", "Bottom" = "#de9e4a", "Intermediate" = "steelblue")
plot_text_size <- 10
```


```{r upset_plot_compendium}
compendium_events = fread('data/compendium_cryptics.csv')

compendium_events = compendium_events %>% 
    mutate(plot_comparison = case_when(comparison == "Ferguson-HeLA" ~ 'HeLa',
                                       comparison == "Appocher-SKNBE2" ~ 'SK-N-BE(2) - 1',
                                       comparison == 'Brown-i3Cortical' ~ "i3 Cortical - 1",
                                       comparison == "Brown-SKNBE2" ~ 'SK-N-BE(2) - 2',
                                       comparison == 'Brown-SHSY5Y' ~ "SH-SY5Y - 2",
                                       comparison == 'Melamed-SHSY5Y'  ~ "SH-SY5Y - 1",
                                       comparison == 'Seddighi-i3Cortical' ~ "i3 Cortical - 2",
                                       comparison == 'Humphrey-i3Cortical' ~ "i3 Cortical - 3",
                                       comparison == 'Vargas-i3LMN' ~ "i3 Lower motor neuron",
                                       comparison == 'Klim-MotorNeuron' ~ "iPSC Motor neuron"))
# Create a binary presence matrix
binary_matrix <- compendium_events %>%
    distinct(paste_into_igv_junction,comparison) %>% 
    mutate(present = 1) %>%
    pivot_wider(names_from = comparison, values_from = present, values_fill = 0) 

# Set junctions as rownames
binary_matrix = binary_matrix %>% 
    column_to_rownames('paste_into_igv_junction') %>% 
    as.data.frame()

upset(binary_matrix, 
      nsets = 10,
      order.by = "freq", 
      main.bar.color = "steelblue", 
      sets.bar.color = "grey30")
```



```{r tdp_correlation_n_cryptics}
library(openxlsx)
library(dplyr)

# Load all sheets
xlsx_file <- "/Users/annaleigh/Desktop/deseq2_outputs_compendium.xlsx"
sheet_names <- getSheetNames(xlsx_file)
dfs <- purrr::map(sheet_names, ~read.xlsx(xlsx_file,.x))

# Extract TARDBP rows and log2FoldChange values
tardbp_results <- purrr::map2_dfr(dfs, sheet_names, ~{
  filtered <- filter(.x, gene_name == "TARDBP")
  if (nrow(filtered) > 0) {
    data.frame(
      comparison = .y,
      log2FoldChange = filtered$log2FoldChange
    )
  } else {
    NULL
  }
})

# View result
print(tardbp_results)
compendium_events %>% 
    group_by(comparison) %>% 
    summarize(n_junction = n_distinct(paste_into_igv_junction)) %>% 
    left_join(tardbp_results) %>% 
    separate(comparison,into = c(NA,'celltype'),remove = FALSE) %>% 
    ggplot(aes(x = log2FoldChange,y = n_junction)) + 
    geom_point(size = 3,aes(color = celltype)) +
    theme_bw() + 
    ylab("N cryptic junctions") + 
    xlab("TARDBP KD level") +
    stat_cor()
```

# Sample numbers of the post-mortem datasets

```{r}
meta_rimod = fread("data/metadata_rimod.csv")
meta_nygc = fread('data/metadta_nygc.csv')
library(dplyr)
library(knitr)

# Create the three tables
table1 <- meta_nygc %>%  
    filter(rna_tissue_source_simplified %in% c("Cortex_Frontal","Cortex_Temporal")) %>% 
    filter(tdp_proteinopathy %in% c("Control","FTLD-TAU","FTLD-FUS","FTLD-TDP")) %>%
    distinct(sample_id_rna, external_subject_id, tdp_proteinopathy, rna_tissue_source_simplified) %>%
    group_by(tdp_proteinopathy, rna_tissue_source_simplified) %>% 
    summarize(n_sample = n_distinct(sample_id_rna), 
              n_patients = n_distinct(external_subject_id),
              .groups = 'drop')

table2 <- meta_nygc %>% 
    filter(grepl("Cortex_Motor", rna_tissue_source_simplified)) %>% 
    filter(tdp_proteinopathy %in% c("Control","ALS-TDP","ALS-non-TDP")) %>% 
    distinct(sample_id_rna, external_subject_id, tdp_proteinopathy, rna_tissue_source_simplified) %>%
    group_by(tdp_proteinopathy, rna_tissue_source_simplified) %>% 
    summarize(n_sample = n_distinct(sample_id_rna), 
              n_patients = n_distinct(external_subject_id),
              .groups = 'drop')

table3 <- meta_nygc %>% 
    filter(grepl("Cord", rna_tissue_source_simplified)) %>% 
    distinct(sample_id_rna, external_subject_id, tdp_proteinopathy) %>%
    group_by(tdp_proteinopathy) %>% 
    summarize(n_sample = n_distinct(sample_id_rna), 
              n_patients = n_distinct(external_subject_id),
              .groups = 'drop') %>% 
    mutate(rna_tissue_source_simplified = 'Cord_Spinal')

# Combine the tables
combined_table <- bind_rows(table1, table2, table3) %>%
  dplyr::rename(`Proteinopathy` = tdp_proteinopathy,
         `Tissue Source` = rna_tissue_source_simplified,
         `Samples (n)` = n_sample,
         `Patients (n)` = n_patients)

# Display with kable

meta_rimod %>% 
    mutate(tdp_proteinopathy = case_when(genotype %in% c("C9orf72","GRN")~ "FTLD-TDP",
                                         genotype == 'control' ~ "Control",
                                         genotype == 'MAPT' ~ "FTLD-TAU")) %>% 
    distinct(sample, tdp_proteinopathy) %>%
    group_by(tdp_proteinopathy) %>% 
        summarize(n_sample = n_distinct(sample), 
              n_patients = n_distinct(sample),
              .groups = 'drop')
    

meta_nygc %>%  
    filter(grepl("Cord", rna_tissue_source_simplified)|rna_tissue_source_simplified %in% c("Cortex_Frontal","Cortex_Temporal",'Cortex_Motor')) %>% 
    filter(tdp_proteinopathy %in% c("Control","FTLD-TAU","FTLD-FUS","FTLD-TDP",c("Control","ALS-TDP","ALS-non-TDP"))) %>% 
    distinct(rna_tissue_source_simplified,tdp_proteinopathy,external_subject_id,sample_id_rna) %>% 
    mutate(case_control = ifelse(tdp_proteinopathy == 'Control', 'control', 'case')) %>% 
    group_by(case_control) %>% 
            summarize(n_sample = n_distinct(sample_id_rna), 
              n_patients = n_distinct(external_subject_id))

```


# Sequencing features of post-mortem datasets

```{r sequencing_features_rimod_vs_nygc}
source("scripts/box_plot_helper.R")
meta_rimod = fread("data/metadata_rimod.csv")
meta_nygc = fread('data/metadta_nygc.csv')
lib_r = fread("data/library_stats_rimod.csv")
library_nygc = fread('data/nygc_library_depth.csv')

rimod_stats = meta_rimod %>% 
    mutate(post_mortem_interval_in_hours = dminutes(PMD) / dhours(1)) %>% 
    mutate(tdp_proteinopathy = case_when(DiseaseCode == 'control' ~ "Control",
                                         DiseaseCode == 'FTD-MAPT' ~ "FTLD-TAU",
                                         DiseaseCode %in% c('FTD-C9','FTD-GRN') ~ "FTLD-TDP")) %>% 
    dplyr::rename(rin = RIN,age_at_death = Age) %>% 
    mutate(dataset = "RiMOD FTD") %>% 
    left_join(lib_r %>% select(Sample,RNAseq_STAR_uniquely_mapped),by = c("sample" = "Sample")) %>% 
    dplyr::rename(library_size = RNAseq_STAR_uniquely_mapped) %>% 
    distinct(age_at_death,tdp_proteinopathy,post_mortem_interval_in_hours,rin,dataset,library_size) 



nygc_stats_frontal = meta_nygc %>% 
    filter(rna_tissue_source_simplified %in% c("Cortex_Frontal","Cortex_Temporal")) %>% 
    filter(tdp_proteinopathy %in% c("Control","FTLD-TAU","FTLD-FUS","FTLD-TDP")) %>%  
    distinct(sample,age_at_death,tdp_proteinopathy,post_mortem_interval_in_hours,rin) %>% 
    mutate(post_mortem_interval_in_hours = as.numeric(post_mortem_interval_in_hours)) %>% 
    mutate(age_at_death = ifelse(age_at_death == '90 or Older', 99,age_at_death)) %>% 
    mutate(age_at_death = as.numeric(age_at_death)) %>% 
    mutate(dataset = "NYGC Frontal") %>% 
    mutate(sample = gsub(".final","",sample)) %>% 
    left_join(library_nygc, by = c("sample" = "V1")) %>% 
    dplyr::rename(library_size = V2)

nygc_stats_als_motor = meta_nygc %>% 
    filter(grepl("Cortex_Motor",rna_tissue_source_simplified)) %>% 
    filter(tdp_proteinopathy %in% c("Control","ALS-TDP","ALS-non-TDP")) %>% 
    distinct(sample,age_at_death,tdp_proteinopathy,post_mortem_interval_in_hours,rin) %>% 
    mutate(post_mortem_interval_in_hours = as.numeric(post_mortem_interval_in_hours)) %>% 
    mutate(age_at_death = ifelse(age_at_death == '90 or Older', 99,age_at_death)) %>% 
    mutate(age_at_death = as.numeric(age_at_death)) %>% 
    mutate(dataset = "NYGC Motor") %>% 
    mutate(sample = gsub(".final","",sample)) %>% 
    left_join(library_nygc, by = c("sample" = "V1")) %>% 
    dplyr::rename(library_size = V2)


nygc_stats_cord = meta_nygc %>% 
    filter(grepl("Cord",rna_tissue_source_simplified)) %>% 
    filter(tdp_proteinopathy %in% c("Control","ALS-TDP","ALS-non-TDP")) %>% 
    distinct(sample,age_at_death,tdp_proteinopathy,post_mortem_interval_in_hours,rin) %>% 
    mutate(post_mortem_interval_in_hours = as.numeric(post_mortem_interval_in_hours)) %>% 
    mutate(age_at_death = ifelse(age_at_death == '90 or Older', 99,age_at_death)) %>% 
    mutate(age_at_death = as.numeric(age_at_death)) %>% 
    mutate(dataset = "NYGC Cord") %>% 
    mutate(sample = gsub(".final","",sample)) %>% 
    left_join(library_nygc, by = c("sample" = "V1")) %>% 
    dplyr::rename(library_size = V2)



combined_data <- nygc_stats_als_motor %>%
    select(-sample) %>%
    rbind(nygc_stats_cord %>% select(-sample)) %>%
    rbind(nygc_stats_frontal %>% select(-sample)) %>%
    rbind(rimod_stats) %>% 
    mutate(dataset = case_match(dataset,
                                "NYGC Cord" ~ "ALS\nNYGC\nSpinal\ncord",
                                "NYGC Frontal" ~ "FTLD\nNYGC\nFrontal/temporal\ncortex",
                                "NYGC Motor" ~ "ALS\nNYGC\nMotor\ncortex",
                                "RiMOD FTD" ~ "FTLD\nRiMOD\nFrontal\ncortex"))
combined_data$rin = as.numeric(combined_data$rin)

age = plot_significant_comparisons(combined_data,column_name =  "age_at_death", ylab_chosen =  "Age at death")
age = age +     theme(text = element_text(size = 18),axis.text.x = element_text(size = 10),plot.subtitle  = element_text(size = 10))
rin = plot_significant_comparisons(combined_data,column_name =  "rin", ylab_chosen =  "RIN")
rin = rin +   theme(text = element_text(size = 18),axis.text.x = element_text(size = 10),plot.subtitle  = element_text(size = 10))

pmi = plot_significant_comparisons(combined_data,column_name =  "post_mortem_interval_in_hours", ylab_chosen =  "Postmortem interval (h)")
pmi = pmi +     theme(text = element_text(size = 18),axis.text.x = element_text(size = 10),plot.subtitle  = element_text(size = 10))
size_of_lib = plot_significant_comparisons(combined_data,column_name =  "library_size", ylab_chosen =  "Library depth")
size_of_lib = size_of_lib  +  theme(text = element_text(size = 18),
                                    axis.text.x = element_text(size = 10),plot.subtitle  = element_text(size = 10)) +
        scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))  
ggarrange(plotlist = list(age,rin,pmi,size_of_lib),ncol = 2,nrow = 2)

```



# Does baseline gene expression drive detection rate?

```{r nygc_gene_expression_detection_rate}
id_to_ens = fread("data/gencode.v42_geneid_genename.tsv")

summary_post <- fread("data/cryptic_detection_summary_long_version.csv")

nygc_metadata = fread("/Users/annaleigh/Desktop/nygc_metadata_done.csv") %>% 
    dplyr::rename(sample = V4) %>% 
    select(-bam_path,-bam_name,-id) %>% 
        mutate(group = case_when(grepl("Cord",rna_tissue_source_simplified) ~ 'Cord',
                             grepl("Cortex_Motor",rna_tissue_source_simplified) & 
                                 tdp_proteinopathy %in% c("Control","ALS-TDP","ALS-non-TDP") ~ "Motor Cortex",
                             rna_tissue_source_simplified %in% c("Cortex_Frontal","Cortex_Temporal") &
                                 tdp_proteinopathy %in% c("Control","FTLD-TAU","FTLD-FUS","FTLD-TDP") ~ "FTD frontal/temporal")) %>% 
    filter(rna_tissue_source_simplified %in% c("Cortex_Frontal","Cortex_Temporal","Cortex_Motor") |
               grepl("Cord",rna_tissue_source_simplified)) 


cryptic_genes = summary_post %>% 
    left_join(id_to_ens,by = c("gene_name")) 

tpm_cal = fread('data/cryptic_genes_nygc_tpm.csv.gz')


    

norm_expresion = tpm_cal %>% 
    filter(Gene_Id %in% cryptic_genes$gene_id) %>% 
    filter(Sample %in% nygc_metadata$sample_id_rna) %>% 
    left_join(nygc_metadata %>% select(sample_id_rna,group,tdp_proteinopathy,
                                       rna_tissue_source_simplified),by = c("Sample" = 'sample_id_rna')) %>% 
    filter(!is.na(group)) %>% 
    # filter(tdp_proteinopathy != "Control") %>% 
    group_by(Gene_Id,group,tdp_proteinopathy) %>% 
    summarize(mean_expr = mean(TPM)) 


plot_nygc = norm_expresion %>% 
    left_join(cryptic_genes %>% filter(fpr_value <= 0.1) %>% 
                  select(gene_id,gene_name,group,frac_ob_tdp_path),
              by = c("Gene_Id" = "gene_id", "group")) %>% 
    filter(!is.na(frac_ob_tdp_path)) %>% 
    unique()

```


```{r rimod_gene}
meta_rimod = fread("data/metadata_rimod.csv")

rimod_vst = fread("data/rimod_ftd_tpm.csv")
rimod_norm = rimod_vst %>% 
    separate(Sample, into = 'sample') %>% 
        left_join(meta_rimod) %>% 
    mutate(tdp_proteinopathy = case_when(genotype %in% c("C9orf72","GRN") ~ 'RiMOD FTD TDP',
                                         genotype == 'control' ~ "Control",
                                         genotype == 'MAPT' ~ 'RiMOD FTD non-TDP')) %>% 
    filter(Gene_Id %in% cryptic_genes$gene_id) %>% 
    group_by(Gene_Id,tdp_proteinopathy) %>% 
    summarize(mean_expr = mean(TPM)) %>% 
    mutate(group = "RiMOD") %>% 
        left_join(cryptic_genes %>% filter(fpr_value <= 0.1) %>% 
                  select(gene_id,gene_name,group,frac_ob_tdp_path),
              by = c("Gene_Id" = "gene_id", "group")) %>% 
    filter(!is.na(frac_ob_tdp_path)) %>% 
    unique() 


rbind(rimod_norm,plot_nygc) %>% 
    filter(tdp_proteinopathy == 'Control') %>% 
    ggplot(aes(x = mean_expr,
               y = frac_ob_tdp_path)) + 
    geom_point() + 
    facet_wrap(~group,nrow = 2,scales = 'free') + 
    scale_x_continuous(trans = scales::pseudo_log_trans()) +
    geom_text_repel(aes(label = gene_name)) +
    stat_cor() + 
    ylab("Fraction TDP-43 proteinopathy samples detected") +
    scale_y_continuous(labels = scales::percent_format()) +
    xlab("Mean TPM in control tissue") + 
    theme_bw() +
    theme(text = element_text(size = 18))



rbind(rimod_norm,plot_nygc) %>% 
    filter(tdp_proteinopathy %in% c("RiMOD FTD TDP","FTLD-TDP","ALS-TDP")) %>% 
    ggplot(aes(x = mean_expr,
               y = frac_ob_tdp_path)) + 
    geom_point(alpha = 0.7) + 
    facet_wrap(~group,nrow = 2,scales = 'free') + 
    scale_x_continuous(trans = scales::pseudo_log_trans()) +
    geom_text_repel(aes(label = gene_name)) +
    stat_cor() + 
    ylab("Fraction TDP-43 proteinopathy samples detected") +
    scale_y_continuous(labels = scales::percent_format()) +
    xlab("Mean TPM in TDP-43 path tissue") + 
    theme_bw() +
    theme(text = element_text(size = 18))


```


# Are cryptics which are selective only in cortex/cord due to gene expression?

```{r cord only vs cortex}
source('scripts/rank_plot.R')

not_in_the_cord = cryptic_genes %>%
  filter(fpr_value <= 0.1 & tpr_value >= 0.1) %>%
  filter(value_tdp_path >= 5) %>%
  filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction) %>%
  distinct(group, gene_name) %>%
  mutate(observed = TRUE) %>%
  pivot_wider(
    names_from = 'group',
    values_from = 'observed',
    values_fill = FALSE
  ) %>%
  unique() %>%
  mutate(total_count = rowSums(across(where(is.logical)))) %>% 
    filter(Cord == FALSE & total_count == 3) %>% 
    left_join(cryptic_genes %>% distinct(gene_name,gene_id)) 


only_in_the_cord = cryptic_genes %>%
  filter(fpr_value <= 0.1 & tpr_value >= 0.1) %>%
  filter(value_tdp_path >= 5) %>%
  filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction) %>%
  distinct(group, gene_name) %>%
  mutate(observed = TRUE) %>%
  pivot_wider(
    names_from = 'group',
    values_from = 'observed',
    values_fill = FALSE
  ) %>%
  unique() %>%
  mutate(total_count = rowSums(across(where(is.logical)))) %>% 
    filter(Cord == TRUE & total_count == 1) %>% 
    left_join(cryptic_genes %>% distinct(gene_name,gene_id)) 
    
only_in_the_motor_cortex = cryptic_genes %>%
  filter(fpr_value <= 0.1 & tpr_value >= 0.1) %>%
  filter(value_tdp_path >= 5) %>%
  filter(paste_into_igv_junction %in% compendium_events$paste_into_igv_junction) %>%
  distinct(group, gene_name) %>%
  mutate(observed = TRUE) %>%
  pivot_wider(
    names_from = 'group',
    values_from = 'observed',
    values_fill = FALSE
  ) %>%
  unique() %>%
  mutate(total_count = rowSums(across(where(is.logical)))) %>% 
    filter(`Motor Cortex` == TRUE & total_count == 1) %>% 
    left_join(cryptic_genes %>% distinct(gene_name,gene_id)) 

my_df = norm_expresion %>% 
    filter(tdp_proteinopathy == 'Control' & group %in% c("Cord","Motor Cortex")) %>% 
    mutate(only_in_cord = Gene_Id %in% only_in_the_cord$gene_id) %>% 
    mutate(only_in_motr = Gene_Id %in% only_in_the_motor_cortex$gene_id) %>% 
    left_join(cryptic_genes %>% distinct(gene_name,gene_id),by = c("Gene_Id" = 'gene_id')) %>% 
    mutate(log10_mean = log10(mean_expr)) %>% 
    select(-mean_expr) %>% 
    pivot_wider(names_from = 'group',
                values_from = 'log10_mean')
    

motor_only_cord_expression = create_rank_plot(
  data = my_df,
  expression_col = "Cord",
  gene_name_col = "gene_name",
  cord_only_col = "only_in_motr"
)
motor_only_cord_expression + 
    ggtitle('Gene expression of motor cortex specific cryptic in spinal cord') + 
    theme(plot.title = element_text(size = 15))
cord_only_motor_expression = create_rank_plot(
  data = my_df,
  expression_col = "Motor Cortex",
  gene_name_col = "gene_name",
  cord_only_col = "only_in_cord"
)
cord_only_motor_expression + 
    ggtitle('Gene expression of spinal cord specific cryptic in motor cortex') +
    theme(plot.title = element_text(size = 15))

```

# Are the bulk identified junctions also present in FACs sorted nuclei?

```{r}
data_summary = fread('data/liu_overlap.csv')
ggplot(data_summary, aes(x = group, y = proportion, fill = direction)) +
  geom_bar(stat = "identity") +
      geom_text(aes(label = count), 
            position = position_stack(vjust = 0.5), 
            color = "white", fontweight = "bold",size = 8) +
  # geom_text(aes(label = paste0(round(percentage, 1), "%")), 
  #           position = position_stack(vjust = 0.5), 
  #           color = "white", fontweight = "bold",size = 8) +
  labs(title = str_wrap("Cryptic junctions in bulk RNA-seq upregulated in FTLD TDP-43-negative neuronal nuclei",width = 55),
       x = element_blank(), y = "Proportion", fill = "Direction") +
  theme_minimal() +
scale_fill_manual(values = c("#BF5B17","#386CB0")) + 
  scale_y_continuous(labels = scales::percent) + 
    theme(legend.position = 'bottom') +
    theme(text = element_text(size = 18),plot.title = element_text(size = 15))


```

# How does library depth vary in compendium datasets?

```{r uniquely_mapped_compendium}
two = readxl::read_excel('data/compendium_library_depth.xlsx', sheet = 'Compendium')
tardbp_results = fread('data/tardbp_kd_level_compendium.csv')
compendium_events = fread('data/compendium_cryptics.csv')
n_junc_per = compendium_events %>% 
    group_by(comparison) %>% 
    summarise(n_cryptic_junc = n_distinct(paste_into_igv_junction))



for_lm = two %>% 
    group_by(dataset) %>% 
    summarize(avg_depth = mean(uniquely_mapped)) %>% 
    left_join(n_junc_per, by = c("dataset" = 'comparison')) %>% 
    left_join(tardbp_results,by = c("dataset" = 'comparison')) %>% 
    separate(dataset,remove = FALSE, into = c(NA,"cell_type"))


lm(n_cryptic_junc ~ log2FoldChange*avg_depth,data = for_lm) %>% 
    broom::tidy()
```

# Is library depth stable throughout the 2 dose-dependent curves

```{r uniquely_mapped_curves}
one = readxl::read_excel('data/compendium_library_depth.xlsx', sheet = 'Curves')
    
sushi_curve = plot_significant_comparisons(one %>% filter(comparison == 'SH-SY5Y curve'),column_name =  "uniquely_mapped", ylab_chosen =  "Uniquely mapped reads") 
sushi_curve = sushi_curve + geom_jitter(width = 0.2) + 
    scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6),
                       limits = c(30*10^6,80*10^6))  +
        theme(text = element_text(size = 18))


sk_curvecurve = plot_significant_comparisons(one %>% filter(comparison == 'SK-N-BE(2) curve'),column_name =  "uniquely_mapped", ylab_chosen =  "Uniquely mapped reads")
sk_curvecurve = sk_curvecurve + geom_jitter(width = 0.2) + 
        scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6),
                       limits = c(30*10^6,80*10^6))  +
    theme(text = element_text(size = 18))
ggarrange(plotlist = list(sushi_curve,sk_curvecurve))
```





```{r, echo=FALSE, warning=FALSE, message=FALSE}
###top50_upf1 - load this table!!
results_table_upf1vsctrl <- read.csv(here::here("data/results_table_upf1_vs_ctrl.csv"))
results_table_UPF1_ctrl <- read.csv(here::here("data/results_table_UPF1_ctrl.csv"))

results_table_upf1vsctrl_filtered <- results_table_upf1vsctrl %>%
  filter(log2FoldChange > 1 | log2FoldChange < -1) %>%
  filter(!is.na(symbol))

results_table_UPF1_ctrl_filtered <- results_table_UPF1_ctrl %>%
  filter(symbol %in% results_table_upf1vsctrl_filtered$symbol) %>%
  filter(!is.na(symbol))

results_table_upf1vsctrl_filtered_2 <- results_table_upf1vsctrl_filtered %>%
  filter(symbol %in% results_table_UPF1_ctrl_filtered$symbol)

ggplot(data = results_table_upf1vsctrl_filtered_2, aes(x = reorder(symbol, log2FoldChange), y = 0, fill = log2FoldChange)) +
  geom_tile() +
  geom_tile(data = results_table_UPF1_ctrl_filtered, aes(x = symbol, y = -1, fill = log2FoldChange)) +
  theme_classic() +
  colorspace::scale_fill_continuous_diverging() +
  theme(axis.text.x = element_text(angle = 90))
```


# TARDBP RNA-seq levels in 2 curves
```{r, echo=FALSE, warning=FALSE, message=FALSE}
### load data needed
#S1A
normal_counts_sh <- read.csv(here::here('data', 'normed_counts_sh.csv'))
normal_counts_sk <- read.csv(here::here('data', 'normed_counts_sk.csv'))

master_table <- read.csv(here::here("data/master_table.csv"))

print("S3A - levels from DeSeq in SH and SK")
b <- normal_counts_sh[,c(2:19,21)] %>% 
  dplyr::filter(symbol == "TARDBP") %>%
  pivot_longer(c(1:18)) %>%
  mutate(group = gsub("^[^_]+_|_[^_]+$", "", name)) %>%
  ggplot(aes(x = group, y = value)) +
  geom_boxplot(alpha = 0.1, show.legend = F) +
  geom_point(position = position_dodge2(width = 1), show.legend = F) +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 Counts per million') +
  theme_classic() +
  theme(text = element_text(size = 10))
d <- normal_counts_sk[,c(2:25,27)] %>% 
  dplyr::filter(symbol == "TARDBP") %>%
  pivot_longer(c(1:24)) %>%
  mutate(group = gsub("^[^_]+_|_[^_]+$", "", name)) %>%
  ggplot(aes(x = group, y = value)) +
  geom_boxplot(alpha = 0.1, show.legend = F) +
  geom_point(position = position_dodge2(width = 1), show.legend = F) +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 Counts per million') +
  scale_x_discrete(labels = c("0", "0.02", "0.03", "0.04", "0.05", "0.075", "0.1", "1")) +
  theme_classic() +
  theme(text = element_text(size = 10))
b + d
```

# Are late / early genes more likely to be DE?
```{r de_late_early_sh}
diff_expr_sh <- read.csv(here::here('data', 'diff_expr_sh.csv'))
diff_expr_sk <- read.csv(here::here('data', 'diff_expr_sk.csv'))
master_table = as.data.table(master_table)
fold_cut = 0
diff_plt_sh = diff_expr_sh %>% 
    left_join(unique(master_table[junction_for_clustering == 'yes',.(gene,last_exon,cryptic,tdp43_sensitivity_sh)]),by = c("symbol" = 'gene')) %>% 
    filter(cryptic == 'yes' | is.na(last_exon)) %>% 
    filter(symbol != "RNA5-8SN2") %>% 
    mutate(fc_direction = ifelse(log2FoldChange > 0, "up","down")) %>%
    mutate(sig_de = ifelse(padj < 0.05 & abs(log2FoldChange) >= fold_cut, "sig", "non_sig")) %>% 
    filter(!is.na(sig_de)) %>% 
    mutate(sig_de = glue::glue("{sig_de}\n{fc_direction}")) %>% 
    distinct(symbol,last_exon,cryptic,tdp43_sensitivity_sh,fc_direction,sig_de,source) %>% 
    mutate(tdp43_sensitivity_sh = ifelse(is.na(tdp43_sensitivity_sh),'Non-splicing',tdp43_sensitivity_sh)) %>% 
    mutate(tdp43_sensitivity_sh = ifelse(tdp43_sensitivity_sh == '','Non-splicing',tdp43_sensitivity_sh)) %>% 
    mutate(tdp43_sensitivity_sh = ifelse(tdp43_sensitivity_sh == 'Not found','Non-splicing',tdp43_sensitivity_sh)) 

sh_d_gene = diff_plt_sh %>% 
    group_by(source,tdp43_sensitivity_sh) %>% 
    summarize(denom = n_distinct(symbol))

diff_plt_sh %>% 
    group_by(tdp43_sensitivity_sh, fc_direction, sig_de, source) %>% 
    summarize(n_genes = n_distinct(symbol)) %>% 
    ungroup() %>% 
    left_join(sh_d_gene) %>% 
    mutate(prop = n_genes / denom) %>% 
    filter(!grepl("non_sig", sig_de)) %>% 
    ggplot(aes(x = tdp43_sensitivity_sh, 
               y = prop, 
               fill = sig_de)) + 
    geom_col(position = position_dodge(preserve = "single"), color = 'black') + 
    geom_text(aes(label = n_genes), 
              position = position_dodge(width = 0.9), 
              vjust = -0.5, 
              size = 4) + # Adding white text on top of the bars
    facet_wrap(~source) + 
    theme_classic() + 
    ylab("Proportion of genes") + 
    xlab(element_blank()) + 
    theme(legend.position = 'bottom') + 
    scale_fill_manual(values = c("#2B4162", "#721817"))

```

```{r de_late_early_be2}

diff_plt_be2 = diff_expr_sk %>% 
    left_join(unique(master_table[junction_for_clustering == 'yes',.(gene,last_exon,cryptic,tdp43_sensitivity_be2)]),by = c("symbol" = 'gene')) %>% 
    filter(symbol != "RNA5-8SN2") %>% 
    mutate(fc_direction = ifelse(log2FoldChange > 0, "up","down")) %>%
    mutate(sig_de = ifelse(padj < 0.05 & abs(log2FoldChange) >= fold_cut, "sig", "non_sig")) %>% 
    filter(!is.na(sig_de)) %>% 
    mutate(sig_de = glue::glue("{sig_de}\n{fc_direction}")) %>% 
    distinct(symbol,last_exon,cryptic,tdp43_sensitivity_be2,fc_direction,sig_de,source) %>% 
    mutate(tdp43_sensitivity_be2 = ifelse(is.na(tdp43_sensitivity_be2),'non-splicing',tdp43_sensitivity_be2)) %>% 
    mutate(tdp43_sensitivity_be2 = ifelse(tdp43_sensitivity_be2 == '','non-splicing',tdp43_sensitivity_be2)) %>% 
    mutate(tdp43_sensitivity_be2 = ifelse(tdp43_sensitivity_be2 == 'Not found','non-splicing',tdp43_sensitivity_be2)) 

be2_d_gene = diff_plt_be2 %>% 
    group_by(source,tdp43_sensitivity_be2) %>% 
    summarize(denom = n_distinct(symbol))

diff_plt_be2 %>% 
    group_by(tdp43_sensitivity_be2, fc_direction, sig_de, source) %>% 
    summarize(n_genes = n_distinct(symbol)) %>% 
    ungroup() %>% 
    left_join(be2_d_gene) %>% 
    mutate(prop = n_genes / denom) %>% 
    filter(!grepl("non_sig", sig_de)) %>% 
    ggplot(aes(x = tdp43_sensitivity_be2, 
               y = prop, 
               fill = sig_de)) + 
    geom_col(position = position_dodge(preserve = "single"), color = 'black') + 
    geom_text(aes(label = n_genes), 
              position = position_dodge(width = 0.9), 
              vjust = -0.5, 
              size = 4) + # Adding white text on top of the bars
    facet_wrap(~source) + 
    theme_classic() + 
    ylab("Proportion of genes") + 
    xlab(element_blank()) + 
    theme(legend.position = 'bottom') + 
    scale_fill_manual(values = c("#2B4162", "#721817"))

```

# Poison exon expression in NMD

```{r}
library(ggridges)
chx_control = fread('data/chx_control.csv')
upf1_control = fread('data/upf1_control.csv')
smg1_control_cyto = fread("data/smg1_control_cyto.csv")
smg1_control_nuc = fread("data/smg1_control_nuc.csv")
smg1_control_bulk = fread("data/smg1_control_bulk.csv")

# https://github.com/HudsonAlpha/poison_exon_variant_analysis

poison_intron = rtracklayer::import('data/poison_exon_containing_introns.bed')
poison_casette = rtracklayer::import('data/poison_exon_cassettes.bed')

library(GenomicRanges)

result <- subtract(poison_intron, poison_casette)
result = unlist(result)





data_cyto <- result %>% 
  as.data.table() %>% 
  mutate(end = end + 1) %>% 
  select(-strand) %>% 
  left_join(smg1_control_cyto %>% distinct(seqnames, start, end, strand,
                                           gene_name, mean_dpsi_per_lsv_junction,
                                           probability_changing)) %>% 
  filter(!is.na(gene_name)) %>% 
  mutate(dataset = "SH-SY5Y\nSMG1-11j inhibitor\nCytoplasm")


data_nuc <- result %>% 
    as.data.table() %>% 
  mutate(end = end + 1) %>% 
  select(-strand) %>% 
  left_join(smg1_control_nuc %>% distinct(seqnames, start, end, strand, 
                                          gene_name, mean_dpsi_per_lsv_junction, 
                                          probability_changing)) %>% 
  filter(!is.na(gene_name)) %>% 
  mutate(dataset = "SH-SY5Y\nSMG1-11j inhibitor\nNucleus")

data_smg1_bulk <- result %>% 
    as.data.table() %>% 
  mutate(end = end + 1) %>% 
  select(-strand) %>% 
  left_join(smg1_control_bulk %>% distinct(seqnames, start, end, strand, 
                                          gene_name, mean_dpsi_per_lsv_junction, 
                                          probability_changing)) %>% 
  filter(!is.na(gene_name)) %>% 
  mutate(dataset = "SH-SY5Y\nSMG1-11j inhibitor\nPseudo-bulk")

data_upf1 <- result %>% 
    as.data.table() %>% 
  mutate(end = end + 1) %>% 
  select(-strand) %>% 
  left_join(upf1_control %>% distinct(seqnames, start, end, strand, 
                                          gene_name, mean_dpsi_per_lsv_junction, 
                                          probability_changing)) %>% 
  filter(!is.na(gene_name)) %>% 
  mutate(dataset = "iPSC Cortical Neuron\nUPF1 KD")

data_chx <- result %>% 
    as.data.table() %>% 
  mutate(end = end + 1) %>% 
  select(-strand) %>% 
  left_join(chx_control %>% distinct(seqnames, start, end, strand, 
                                          gene_name, mean_dpsi_per_lsv_junction, 
                                          probability_changing)) %>% 
  filter(!is.na(gene_name)) %>% 
  mutate(dataset = "SH-SY5Y\nCyclohexmide")

all_data_sets_nmdi <- rbind(data_cyto, data_nuc, data_smg1_bulk,data_upf1,data_chx)


all_data_sets_nmdi %>% 
    filter(probability_changing > 0.9) %>% 
    mutate(dataset = fct_relevel(dataset,"SH-SY5Y\nSMG1-11j inhibitor\nCytoplasm","SH-SY5Y\nSMG1-11j inhibitor\nNucleus",'SH-SY5Y\nSMG1-11j inhibitor\nPseudo-bulk')) %>% 
    ggplot(aes(x = mean_dpsi_per_lsv_junction)) + 
    geom_histogram() + 
    facet_wrap(~dataset) +
    theme_bw() + 
    ylab("N significant poison exon junctions") +
    theme(text = element_text(size = 20),
          axis.text.x = element_text(size = 14)) +
    theme(legend.position = 'none') +
    xlab('Delta PSI') + 
    scale_x_continuous(labels = scales::percent) + 
    geom_vline(xintercept = 0, linetype = 'dotted')



all_data_sets_nmdi %>% 
    mutate(sig = probability_changing > 0.9) %>% 
    ggplot(aes(x = mean_dpsi_per_lsv_junction, 
              color = dataset,
              y = -log10(1 - probability_changing),
              alpha = sig)) +  # Map alpha to sig
    geom_point() +
    scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.5), guide = "none") +
    theme_bw() + 
    geom_hline(yintercept = -log10(0.1)) +
    geom_vline(xintercept = 0,linetype = 'dotted')  +
    xlab("Delta PSI NMDi vs Control")


```

# UPF1 and TDP-43 levels in NMD experiments

```{r}
library(openxlsx)
library(dplyr)

# Load all sheets
xlsx_file <- "/Users/annaleigh/Desktop/NMD_deseq_chx_upf1.xlsx"
sheet_names <- getSheetNames(xlsx_file)
sheet_names = sheet_names[grepl("upf1|ctrl",sheet_names)]
dfs <- purrr::map(sheet_names, ~read.xlsx(xlsx_file,.x))

# Extract TARDBP rows and log2FoldChange values
deseq_nmd_results <- purrr::map2_dfr(dfs, sheet_names, ~{
  filtered <- filter(.x, symbol %in% c("TARDBP","UPF1"))
  if (nrow(filtered) > 0) {
    data.frame(
      comparison = .y,
      log2FoldChange = filtered$log2FoldChange,
      gene_name  = filtered$symbol,
    padj  = filtered$padj

    )
  } else {
    NULL
  }
})

deseq_nmd_results %>% 
    mutate(comparison = case_match(comparison, "upf1_vs_ctrl" ~ "UPF1 KD\nvs\nControl",
                                   "UPF1ctrl_UPF1TDP43" ~ "UPF1 KD\nvs\nUPF1 KD + TDP-43 KD",
                                   "ctrlTDP43_UPF1TDP43" ~ "TDP-43 KD\nvs\nUPF1 KD + TDP-43 KD",
                                   "ctrlctrl_ctrlTDP43" ~ "TDP-43 KD\nvs\nControl")) %>% 
  mutate(sig = padj < 0.01) %>% 
  ggplot(aes(
    x = comparison,
    y = log2FoldChange,
    fill = gene_name,
    alpha = sig,          # map alpha to sig
    color = sig           # map outline color to sig
  )) + 
  geom_col(position = 'dodge2', linewidth = 1) + # linewidth controls outline thickness
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.7)) +
  scale_color_manual(values = c(`TRUE` = "black", `FALSE` = NA)) +
  guides(alpha = "none", color = "none") +
    theme_bw() +
    theme(text = element_text(size = 20)) +
    xlab(element_blank()) +
    ylab("Log2 Fold Change") + 
    geom_hline(yintercept = 0,linetype = 'dotted') +
    guides(fill = guide_legend(title = "Gene name")) +
    theme(legend.position = 'bottom')

```


# Spaghetti plot for APA events in SH-SY5Y
```{r,echo=FALSE,warning=FALSE,message=FALSE}


shsy5y_papa = fread("data/shsy5y_papa_curve.csv")

sh_plot = shsy5y_papa |>
    select(le_id,mean_ppau_contrast,experiment,gene_name,timing) |>
    complete(le_id,experiment,fill = list(mean_ppau_contrast = 0))  |>
    dplyr::group_by(le_id) %>%
    fill(timing, .direction = "downup") %>%
    ungroup() |>
    mutate(gene_name = ifelse(experiment == 'zanovello_shsy5y_curve_0075',
                              gene_name,NA_character_)) %>%
    rbind(shsy5y_papa %>% distinct(le_id,gene_name,timing) %>% mutate(experiment = '_fakezero') %>% mutate(mean_ppau_contrast = 0)) %>% 
    ggplot(aes(x = experiment,
               y = mean_ppau_contrast,
               group = le_id,color = timing)) +
    geom_point() +
    geom_line() +
    geom_text_repel(aes(label = gene_name),show.legend = FALSE) +
    theme_bw() +
    geom_hline(yintercept = 0.1,linetype = 'dotted') +
    scale_color_manual(values = c("#E376BF","#4A6EDD")) +
    scale_x_discrete(labels=c("Control","+", "++", "+++","++++","+++++")) +
    facet_wrap(~timing) +
    ylab("Delta % polyA usage") +
    theme(text = element_text(size = 20)) +
    xlab(element_blank()) +
    theme(legend.position = 'none') +
    scale_y_continuous(labels = scales::percent_format()) 


    

```

# Spaghetti plot for APA events in BE(2)
```{r}
sknbe2_papa = fread("data/sknbe2_papa_curve.csv")

be2_plot = sknbe2_papa %>% 
    select(le_id,mean_ppau_contrast,experiment,gene_name,timing) |> 
    complete(le_id,experiment,fill = list(mean_ppau_contrast = 0))  |> 
    dplyr::group_by(le_id) %>%
    fill(timing, .direction = "downup") %>%
    ungroup() |>
    rbind(sknbe2_papa %>% distinct(le_id,gene_name,timing) %>% mutate(experiment = '_fakezero') %>% mutate(mean_ppau_contrast = 0)) %>% 
    mutate(gene_name = ifelse(experiment == 'zanovello_skndz_curve_1',
                              gene_name,NA_character_)) |>

    ggplot(aes(x = experiment,
               y = mean_ppau_contrast,
               group = le_id,color = timing)) +
    geom_point() +
    geom_line() +
    geom_text_repel(aes(label = gene_name),show.legend = FALSE) +
    theme_bw() +
    geom_hline(yintercept = 0.1,linetype = 'dotted') +
    scale_color_manual(values = c("#E376BF","#4A6EDD")) +
    scale_x_discrete(labels=c("Control","+", "++", "+++","++++","+++++","+++\n+++","+++\n+++++")) +
    ylab("Delta % polyA usage") +
    scale_y_continuous(labels = scales::percent_format()) +
    facet_wrap(~timing) +
        theme(text = element_text(size = 20)) +
    xlab(element_blank()) +
        theme(legend.position = 'none') +
    scale_y_continuous(labels = scales::percent_format()) 



ggarrange(plotlist = list(sh_plot,be2_plot),nrow = 2)

```


# UPF1 KD & SMG1 KD NMD-insentive events are more likely to be detected, they are not more likely to be selective

```{r, echo=FALSE, warning=FALSE, message=FALSE}
summary_post = fread("data/summary_metrics_post_mortem.csv")

smg1_summary_rest <- generate_summary_table(
    summary_post, 
    pseudo_bulk_plot_slope$filtered_table,
    category_column = "color_gene_name",
    baseline_category = "non-NMD rescued"
)

fisher_result_smg1 = run_fisher_test(smg1_summary_rest)
smg1_plots = make_final_plot(smg1_summary_rest,fisher_result_smg1)
smg1_plots$detection_plot + ggtitle('SMG1i junctions')
smg1_plots$selectivity_plot + ggtitle('SMG1i junctions')

upf1_summary_rest <- generate_summary_table(
    summary_post, 
    upf1_slope_plot_table$filtered_table,
    category_column = "color_gene_name",
    baseline_category = "non-NMD rescued"
)

fisher_result_upf1 = run_fisher_test(upf1_summary_rest)
upf1_plots = make_final_plot(upf1_summary_rest,fisher_result_upf1)
upf1_plots$detection_plot + ggtitle('UPF1 junctions')
upf1_plots$selectivity_plot + ggtitle('UPF1 junctions')



```

# NMD association with detection frequency across KDs - UPF1 + SMG1i

```{r}
# Step 2: Get unique junctions with their category classification
df = compendium_events %>%
    distinct(n_significant_data,paste_into_igv_junction) %>% 
    mutate(group = case_when(n_significant_data <= 3 ~ "1-3",
                          n_significant_data >= 4 & n_significant_data <= 6 ~ "4-6",
                          TRUE ~ "7-10")) %>% 
    left_join(wide_upf1) %>% 
    group_by(group) %>% 
    filter(!is.na(color_gene_name)) %>% 
    mutate(n_junc = n_distinct(paste_into_igv_junction)) %>% 
    distinct(color_gene_name,group,paste_into_igv_junction,n_junc) %>% 
    group_by(color_gene_name,group) %>% 
    summarize(n_bucket = n_distinct(paste_into_igv_junction)) %>% 
    ungroup() %>% 
    unique()
xtabs(n_bucket ~ color_gene_name + group,data = df) %>% 
    prop.table(margin = 2)

```


```


### Supplementary figure 3 - NMD
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# reading in the master table again for debug, can remove later
master_table <- read.csv(here::here("data/master_table.csv"))
master_table <- master_table %>%
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "" ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_sh == "" ~ tdp43_sensitivity_be2,
                                tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                                tdp43_sensitivity_be2 != 'Late' & tdp43_sensitivity_sh == 'Early' ~ 'Early',
                                tdp43_sensitivity_be2 != 'Early' & tdp43_sensitivity_sh == 'Late' ~ 'Late',
                                tdp43_sensitivity_sh == 'Intermediate'~  'Intermediate')) %>% 
    mutate(new_cate = ifelse(is.na(new_cate),"Ambiguous",new_cate))

big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))
big_data_chx = big_data_chx %>% 
    separate(paste_into_igv_junction,remove = FALSE,into = c("chr",'start','end'),
             convert = TRUE)


wide_chx = big_data_chx |> 
    select(paste_into_igv_junction,.id,mean_psi_per_lsv_junction,color_gene_name) |> 
    pivot_wider(names_from = '.id',values_from = 'mean_psi_per_lsv_junction',
                values_fn = max) %>% 
    unique() %>% 
    mutate(normal_cryptic_chx = Control_Control < 0.05 &  Control_TDP43KD > 0.1) |> 
    mutate(chx_cryptic = Control_Control < 0.05 &  Cycloheximide_TDP43KD > 0.1) |> 
    mutate(chx_effect = case_when((Cycloheximide_TDP43KD - Control_TDP43KD) > 0.05 ~ "NMD rescued",
                                  Control_TDP43KD > 0.94 ~ "PSI > 95% at baseline",
                                  TRUE ~ "non-NMD rescued")) %>% 
    left_join(big_data_chx %>% distinct(paste_into_igv_junction,gene_name,gene_id)) %>% 
    mutate(ensgene =  gsub("\\..*","",gene_id))


chx_dark = wide_chx |> 
    filter(normal_cryptic_chx == FALSE & chx_cryptic == TRUE & chx_effect == "NMD rescued") |> 
    pull(paste_into_igv_junction) |> unique()

wide_chx = wide_chx %>% 
    mutate(chx_effect = ifelse(paste_into_igv_junction %in% chx_dark, "Dark cryptic",chx_effect))
```

```{r LateEarlyBE2}
sk_nmd_sens = wide_chx %>% 
    distinct(paste_into_igv_junction,chx_effect) %>% 
    left_join(master_table) %>% 
    filter(!is.na(tdp43_sensitivity_be2)) %>% 
    filter(tdp43_sensitivity_be2 != "") %>% 
    filter(junction_for_clustering == 'yes') %>% 
    mutate(chx_effect = ifelse(chx_effect == "Dark cryptic","NMD rescued",chx_effect)) %>% 
    group_by(chx_effect,tdp43_sensitivity_be2) %>% 
    reframe(n = n_distinct(paste_into_igv_junction))

ggplot(sk_nmd_sens) +
    aes(x = tdp43_sensitivity_be2, fill = chx_effect, weight = n,by = as.factor(tdp43_sensitivity_be2)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'Cycloheximide treatment') +
    scale_fill_manual(values = c("#3288BD","#D53E4F","#BFCC94")) + 
    theme(legend.position = 'bottom') 


sk_nmd_sens %>% 
    filter(chx_effect != "PSI > 95% at baseline") %>% 
    pivot_wider(names_from = 'chx_effect',values_from = 'n') %>% 
    column_to_rownames("tdp43_sensitivity_be2") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()
```

```{r NMDSensNewCate}
new_cate_nmd_sens = wide_chx %>% 
    distinct(paste_into_igv_junction,chx_effect) %>% 
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    filter(new_cate != "") %>% 
    filter(junction_for_clustering == 'yes') %>% 
    mutate(chx_effect = ifelse(chx_effect == "Dark cryptic","NMD rescued",chx_effect)) %>% 
    group_by(chx_effect,new_cate) %>% 
    reframe(n = n_distinct(paste_into_igv_junction))

ggplot(new_cate_nmd_sens) +
    aes(x = new_cate, fill = chx_effect, weight = n, by = as.factor(new_cate)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'Cycloheximide treatment') +
    scale_fill_manual(values = c("#3288BD","#D53E4F","#BFCC94")) + 
    theme(legend.position = 'bottom') 

new_cate_nmd_sens %>% 
    filter(chx_effect != "PSI > 95% at baseline") %>% 
    pivot_wider(names_from = 'chx_effect',values_from = 'n') %>% 
    column_to_rownames("new_cate") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()
```



### Supplementary figure 4 - features SH-SY5Y

A. Conservation score for CE in SK-N-BE(2)
B. Conservation score for non-CE in SK-N-BE(2)
C. Kmers enrichment analysis in SK-N-BE(2)
D. Density UG motifs by early-late in SK-N-BE(2)
E. Density UG motifs by top-bottom in  SK-N-BE(2)
F. Distance iCLIP TDP43 binding site SK-N-BE(2)
G. RNA maps based on figure F
H. 5'ss CE vs normal
I. 3'ss CE vs normal

```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("2A and 2B")

get_signif_stars <- function(p_value) {
  if (p_value < 0.001) {
    return("***")
  } else if (p_value < 0.01) {
    return("**")
  } else if (p_value < 0.05) {
    return("*")
  } else {
    return("NS")  # not significant
  }
}
  
get_significance <- function(df, value, group, category_col, category) {
  test_formula <- as.formula(paste(value, "~", group))
  p_test <- kruskal.test(test_formula, data = df %>% filter(!!sym(category_col) == category))
  return(get_signif_stars(p_test$p.value))
}

cons_score_df <- master_table %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>% 
  subset(!is.na(phastCons_alt) & !is.na(phastCons_unchanged)) %>%
  pivot_longer(cols = c(phastCons_alt, phastCons_unchanged), names_to = "category", values_to = "cons_score") 

cons_score_df_cryptic <- cons_score_df %>% filter(cryptic == "yes") 
cons_score_df_not_cryptic <- cons_score_df %>% filter(cryptic == "no") 
a <- cons_score_df_cryptic %>% 
  ggplot(aes(x = category, y = cons_score, fill = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Novel splice site", "Unchanged splice site")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_alt"), 
                   get_significance(cons_score_df_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_unchanged")) 
  ) +
  #facet_wrap(facets = vars(cryptic)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

b <- cons_score_df_not_cryptic %>% 
  ggplot(aes(x = category, y = cons_score, fill = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Alternatively used splice site", "Unchanged splice site")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_alt"), 
                   get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_unchanged"))
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
a + b
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#C. Kmers enrichment analysis in  SK-N-BE(2)
print("2C")

kmer_size <- 6
yg_motifs <- c('GTGTGT', 'TGTGTG', 'CGTGTG', 'TGCGTG', 'TGTGCG', 'GTGTGC')
ya_motifs <- c('ATGTGT', 'GTATGT', 'GTGTAT', 'TGTGTA', 'TGTATG', 'TGCATG')
aa_motifs <- c('GTGTGA', 'AATGAA', 'GAATGA', 'TGAATG', 'ATGAAT', 'GTGAAT', 'GAATGT', 'TTGAAT')
all_motifs <- c(yg_motifs, ya_motifs, aa_motifs)

count_hexamers <- function(row) {
  if (row[["junc_cat"]] == "exon_skip" || is.na(row[["paste_into_igv_junction_annotated"]])) {
    return(NULL)
  }
  
  coords <- strsplit(row[["paste_into_igv_junction_annotated"]], "[:-]")[[1]]
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = coords[1], start = as.numeric(coords[2]), end = as.numeric(coords[3]))
  if (row[["strand"]] == "-") {
    seq <- reverseComplement(seq)
  }
  counts <- sapply(all_motifs, function(motif) countPattern(motif, seq, max.mismatch = 0))
  names(counts) <- all_motifs
  
  return(counts)
}
master_table_filt <- master_table %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>% filter(junc_cat != "exon_skip")
kmer_table <- apply(master_table_filt, 1, count_hexamers)
all_counts_df <- as.data.frame(t(kmer_table))

all_counts_df$type <- master_table_filt$tdp43_sensitivity_be2
all_counts_df$cryptic <- master_table_filt$cryptic

summary_counts <- all_counts_df %>%
  group_by(type, cryptic) %>%
  summarise(across(all_motifs, sum))

means <- apply(summary_counts[, -c(1,2)], 1, mean)
sds <- apply(summary_counts[, -c(1,2)], 1, sd)
z_scores <- sweep(summary_counts[, -c(1,2)], MARGIN = 1, STATS = means, FUN = "-") / sds
heatmap_data <- as.matrix(z_scores)

sort_kmers_by_zscore <- function(kmers, z_scores) {
  kmers[order(z_scores[kmers], decreasing = TRUE)]
}

sorted_yg <- sort_kmers_by_zscore(yg_motifs, heatmap_data[1,])
sorted_ya <- sort_kmers_by_zscore(ya_motifs, heatmap_data[1,])
sorted_aa <- sort_kmers_by_zscore(aa_motifs, heatmap_data[1,])

sorted_all_kmers <- c(sorted_yg, sorted_ya, sorted_aa)
categories <- c(rep("YG", 6), rep("YA", 6), rep("AA", 8))

annotation <- HeatmapAnnotation(Motif = factor(categories, levels = c("YG", "YA", "AA")),
                                show_legend = c(F, F, F),
                                col = list(Motif = c(YG = "purple", YA = "orange", AA = "darkgreen")),
                                which = "col")

rownames(heatmap_data) <- paste(summary_counts$type, ifelse(summary_counts$cryptic == "yes", "Cryptic", "Not cryptic"))
c <- Heatmap(heatmap_data[c("Early Cryptic","Early Not cryptic","Late Cryptic","Late Not cryptic") , sorted_all_kmers], 
        cluster_columns = F,
        cluster_rows = F,
        show_row_names = T,
        show_column_names = T,
        show_row_dend = F,
        bottom_annotation = annotation,
        col = colorRampPalette(brewer.pal(3, "YlGnBu"))(100),
        show_heatmap_legend = FALSE)
c
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#D. Density UG motifs by early-late in  SK-N-BE(2)
print("2D and 2E")
sequences <- sapply(seq_along(master_table$paste_into_igv_junction_annotated), function(i) {
  coord <- master_table$paste_into_igv_junction_annotated[i]
  chr <- sub("^(.*):.*$", "\\1", coord)
  range <- as.numeric(unlist(strsplit(sub(".*:(.*)$", "\\1", coord), "-")))
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = chr, start = range[1], end = range[2])

  if (master_table$strand[i] == "-") {
    seq <- reverseComplement(seq)
  }
  
  return(seq)
})

master_table$longest_ug_region <- Biostrings::vcountPattern('TGTGTG', DNAStringSet(sequences), max.mismatch = 1)

master_table <- master_table %>% 
  mutate(
    start = as.numeric(sub(".*:(\\d+)-.*", "\\1", paste_into_igv_junction_annotated)),
    end = as.numeric(sub(".*-(\\d+)", "\\1", paste_into_igv_junction_annotated)),
    intron_len = end - start,
    ug_density = longest_ug_region / intron_len
  ) %>%
  select(-start, -end)

d <- master_table %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>% 
  ggplot(aes(x = cryptic, y = ug_density, fill = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(tdp43_sensitivity_be2 %in% c("Early", "Late")),
                                    "ug_density", "tdp43_sensitivity_be2", "cryptic", "no"), 
                   get_significance(master_table %>% filter(tdp43_sensitivity_be2 %in% c("Early", "Late")),
                                    "ug_density", "tdp43_sensitivity_be2", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

#E. Density UG motifs by top-bottom in  SK-N-BE(2)
master_table$max_psi_be2 <- factor(master_table$max_psi_be2, levels=c("Top", "Bottom"))
e <- master_table %>% 
  filter(max_psi_be2 %in% c("Top", "Bottom")) %>%
  ggplot(aes(x = cryptic, y = ug_density, fill = max_psi_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) + 
  scale_fill_manual(values = plot_psi_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(max_psi_be2 %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_be2", "cryptic", "no"), 
                   get_significance(master_table %>% filter(max_psi_be2 %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_be2", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
d + e

#F. Distance iCLIP TDP43 binding site  SK-N-BE(2)
print("2F")
df_iclip <- read.csv(here::here('data', 'all_events.closest_tdp_bind.bed'), sep="\t")
df_iclip$tdp_len <- df_iclip$end - df_iclip$start + 1

# f <- df_iclip %>% 
#   filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>%
#   filter(abs(dist) <= 1000) %>%
#   ggplot(aes(x = dist, fill = tdp43_sensitivity_be2)) +
#   geom_density(alpha = 0.5, adjust = 1) +  
#   facet_wrap(~cryptic, 
#              ncol = 2,
#              labeller = labeller(cryptic=c("yes" = "Cryptic", "no" = "Not cryptic"))) +
#   scale_fill_manual(values = plot_sensitivity_colors) +
#   guides(fill = "none") +
#   labs(x = 'Distance to the nearest TDP-binding site', y = 'Density') +  
#   theme_classic() +
#   theme(text = element_text(size = plot_text_size))
# 
# f

#G. RNA maps based on figure F
early_donors <- readDNAStringSet("data/maxent/donors_early_be2.fa")
late_donors <- readDNAStringSet("data/maxent/donors_late_be2.fa")

g1 <- ggplot() + 
  geom_logo(data = as.character(early_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g2 <- ggplot() + 
  geom_logo(data = as.character(late_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g <- g1 / g2
g

#H. 5'ss CE vs normal
df_5ss_strength <- read.csv(here::here('data', 'maxent/5ss_strength_sh.csv'))
h <- df_5ss_strength %>%
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>% 
  ggplot(aes(x = ss_score_cryptic, y =ss_score_ann, color = tdp43_sensitivity_sh)) +
  geom_point(size = 2, alpha = 0.8) +  
  scale_color_manual(values = plot_sensitivity_colors) +
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed") + 
  coord_cartesian(xlim = c(-30, 15), ylim = c(-30, 15)) + 
  labs(x = "MaxEnt score (5'ss cryptic site)", y = "MaxEnt score (5'ss annotated site)") +
  theme_minimal() +
  theme(legend.position = "top") +
  guides(color = "none")

#I. 3'ss CE vs normal
df_3ss_strength <- read.csv(here::here('data', 'maxent/3ss_strength_sh.csv'))
i <- df_3ss_strength %>%
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>% 
  ggplot(aes(x = ss_score_cryptic, y = ss_score_ann, color = tdp43_sensitivity_sh)) +
  geom_point(size = 2, alpha = 0.8) +  
  scale_color_manual(values = plot_sensitivity_colors) +
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed") + 
  coord_cartesian(xlim = c(-30, 15), ylim = c(-30, 15)) + 
  labs(x = "MaxEnt score (3'ss cryptic site)", y = "MaxEnt score (3'ss annotated site)") +
  theme_minimal() +
  theme(legend.position = "top") +
  guides(color = "none")
h + i
```
