---
title: "supplementary_fig"
author: "MZ"
date: "2024-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
###load packages/libraries/scripts
library(tidyverse)
library(ggrepel)
library(patchwork)

source(here::here('trial_nmd_plots.R'))
#check you're in right folder
here::here()
```

### Supplementary figure 1 - sensitivity of CEs to TDP43 LoF is driven by their cis features
```{r, echo=FALSE, warning=FALSE, message=FALSE}
### load data needed
#S1A
normal_counts_sh <- read.csv(here::here('data', 'normed_counts_sh.csv'))
normal_counts_sk <- read.csv(here::here('data', 'normed_counts_sk.csv'))

#S1?
master_table <- read.csv(here::here("data/master_table.csv"))

#S1F #S1G
#apa_sh <- read.csv(here::here())
#apa_sk <- read.csv(here::here())

print("S1A")
b <- normal_counts_sh[,c(2:19,21)] %>% 
  dplyr::filter(symbol == "TARDBP") %>%
  pivot_longer(c(1:18)) %>%
  mutate(group = gsub("^[^_]+_|_[^_]+$", "", name)) %>%
  ggplot(aes(x = group, y = value)) +
  geom_boxplot(alpha = 0.1, show.legend = F) +
  geom_point(position = position_dodge2(width = 1), show.legend = F) +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 Counts per million') +
  theme_classic() +
  theme(text = element_text(size = 10))
d <- normal_counts_sk[,c(2:25,27)] %>% 
  dplyr::filter(symbol == "TARDBP") %>%
  pivot_longer(c(1:24)) %>%
  mutate(group = gsub("^[^_]+_|_[^_]+$", "", name)) %>%
  ggplot(aes(x = group, y = value)) +
  geom_boxplot(alpha = 0.1, show.legend = F) +
  geom_point(position = position_dodge2(width = 1), show.legend = F) +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 Counts per million') +
  scale_x_discrete(labels = c("0", "0.02", "0.03", "0.04", "0.05", "0.075", "0.1", "1")) +
  theme_classic() +
  theme(text = element_text(size = 10))
b + d

print("S1B")

print("S1C")

print("S1D")
g <- master_table %>% 
    filter(tdp43_sensitivity_sh != "") %>% 
    ggplot(aes(x = UMAP_1_sh, y = UMAP_2_sh, color = as.character(tdp43_sensitivity_sh))) +
    geom_point(show.legend = F) +
    xlab("UMAP 1") +
    ylab("UMAP 2") +
    labs(color = "TDP43 sensitivity") +
    scale_color_brewer(palette = "Set1") +
    theme_bw()
h <- master_table %>% 
    filter(tdp43_sensitivity_be2 != "") %>% 
    ggplot(aes(x = UMAP_1_be2, y = UMAP_2_be2, color = as.character(tdp43_sensitivity_be2))) +
    geom_point() +
    xlab("UMAP 1") +
    ylab("UMAP 2") +
    labs(color = "TDP43 sensitivity") +
    scale_color_brewer(palette = "Set1") +
    theme_bw()
g+h

print("S1E")

print("S1F")
#apa_sh
print("S1G")
#apa_sk
```

A. levels from DeSeq in SH and SK
B. Concordance between SH and SK (scatterplot) - to do!!
C. UMAP clustering of events in SH and BE2 together - to do!!
D. UMAP clustering of events in SH and in BE2 - separated
E. differences in differential splicing are not due to changes in gene expression
E. Spaghetti plot for APA events in SH-SY5Y
F. Spaghetti plot for APA events in BE(2)

### Supplementary figure 2 - features 
```{r}

```

A. Conservation score for CE in SK-N-BE(2)
B. Conservation score for non-CE in SK-N-BE(2)
C. Kmers enrichment analysis in SK-N-BE(2)
D. Density UG motifs by early-late in SK-N-BE(2)
E. Density UG motifs by top-bottom in SH-SY5Y
F. Distance iCLIP TDP43 binding site SK-N-BE(2)
G. RNA maps based on figure F
H. 5'ss CE vs normal
I. 3'ss CE vs normal

### Supplementary figure 3 - NMD
```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("S3A")
gene_list_slope_upf1 <- c("STMN2", "UNC13A", "AARS1", "HDGFL2", "CYFIP2", "SYNE1")
input_list <- c("ctrl_ctrl",
                "ctrl_UPF1", 
                "TDP43_ctrl", 
                "TDP43_UPF1")
big_data_upf1 <- read.csv(here::here('data','nmd_or_not_upf1.csv'))
a <- slope_plot_nmd(big_data_upf1, 
               gene_list_slope_upf1)

print("S3B")

```

A. Slope plot UPF1
B. Correlation CHX UPF1 - groups of rescued / non rescued
C. TDP-43 inhibition doesn't impair NMD (top 50 UPF1)
D. TDP-43 inhibition doesn't impair NMD (known poison exons)


