---
title: "supplementary_fig"
author: "MZ"
date: "2024-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
###load packages/libraries/scripts
library(tidyverse)
library(ggrepel)
library(patchwork)
<<<<<<< Updated upstream
=======
library(ggpubr)
>>>>>>> Stashed changes
library(ggsignif)
library(ggseqlogo)
library(Biostrings)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ComplexHeatmap)
library(RColorBrewer)

source(here::here('trial_nmd_plots.R'))
#check you're in right folder
here::here()

plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#1331a7", "Intermediate" = "steelblue")
plot_psi_colors <- c("Top" = "#388f78", "Bottom" = "#de9e4a", "Intermediate" = "steelblue")
plot_text_size <- 10
```

### Supplementary figure 1 - sensitivity of CEs to TDP43 LoF is driven by their cis features
```{r, echo=FALSE, warning=FALSE, message=FALSE}
### load data needed
#S1A
normal_counts_sh <- read.csv(here::here('data', 'normed_counts_sh.csv'))
normal_counts_sk <- read.csv(here::here('data', 'normed_counts_sk.csv'))

#S1?
master_table <- read.csv(here::here("data/master_table.csv"))

#S1F #S1G
#apa_sh <- read.csv(here::here())
#apa_sk <- read.csv(here::here())

print("S1A - levels from DeSeq in SH and SK")
b <- normal_counts_sh[,c(2:19,21)] %>% 
  dplyr::filter(symbol == "TARDBP") %>%
  pivot_longer(c(1:18)) %>%
  mutate(group = gsub("^[^_]+_|_[^_]+$", "", name)) %>%
  ggplot(aes(x = group, y = value)) +
  geom_boxplot(alpha = 0.1, show.legend = F) +
  geom_point(position = position_dodge2(width = 1), show.legend = F) +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 Counts per million') +
  theme_classic() +
  theme(text = element_text(size = 10))
d <- normal_counts_sk[,c(2:25,27)] %>% 
  dplyr::filter(symbol == "TARDBP") %>%
  pivot_longer(c(1:24)) %>%
  mutate(group = gsub("^[^_]+_|_[^_]+$", "", name)) %>%
  ggplot(aes(x = group, y = value)) +
  geom_boxplot(alpha = 0.1, show.legend = F) +
  geom_point(position = position_dodge2(width = 1), show.legend = F) +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 Counts per million') +
  scale_x_discrete(labels = c("0", "0.02", "0.03", "0.04", "0.05", "0.075", "0.1", "1")) +
  theme_classic() +
  theme(text = element_text(size = 10))
b + d

print("S1B - Concordance between SH and SK scatterplot")
#scatterplot from upgrade
master_table %>% 
    ggplot(aes(x = dox075, y = dox1)) +
    geom_point() +
    geom_smooth(method = 'lm', se = F) +
    stat_cor() +
    xlab("Max PSI in SH-SY5Y") +
    ylab("Max PSI in SK-N-BE(2)") +
    theme_classic()

<<<<<<< Updated upstream
print("S1C")

print("S1D")
=======
print("S1C - UMAP clustering of events in SH and BE2 together")
master_table %>% 
    filter(tdp43_sensitivity_sh != "") %>% 
    ggplot(aes(x = UMAP_1_together, y = UMAP_2_together, color = as.character(cluster_together))) +
    geom_point(show.legend = F) +
    xlab("UMAP 1 together") +
    ylab("UMAP 2 together") +
    labs(color = "cluster") +
    scale_color_brewer(palette = "Set1") +
    theme_bw()


print("S1D - UMAP clustering of events in SH and in BE2 separated")
>>>>>>> Stashed changes
g <- master_table %>% 
    filter(tdp43_sensitivity_be2 != "") %>% 
    ggplot(aes(x = UMAP_1_sh, y = UMAP_2_sh, color = as.character(tdp43_sensitivity_be2))) +
    geom_point(show.legend = F) +
    xlab("UMAP 1") +
    ylab("UMAP 2") +
    labs(color = "TDP43 sensitivity") +
    scale_color_manual(values = plot_sensitivity_colors) +
    theme_bw()
h <- master_table %>% 
    filter(tdp43_sensitivity_be2 != "") %>% 
    ggplot(aes(x = UMAP_1_be2, y = UMAP_2_be2, color = as.character(tdp43_sensitivity_be2))) +
    geom_point() +
    xlab("UMAP 1") +
    ylab("UMAP 2") +
    labs(color = "TDP43 sensitivity") +
    scale_color_manual(values = plot_sensitivity_colors) +
    theme_bw()
g+h

print("S1E - differences in differential splicing are not due to changes in gene expression")

print("S1F")
#apa_sh
print("S1G")
#apa_sk
```

A. levels from DeSeq in SH and SK
B. Concordance between SH and SK (scatterplot) - to do!!
C. UMAP clustering of events in SH and BE2 together
D. UMAP clustering of events in SH and in BE2 - separated
E. differences in differential splicing are not due to changes in gene expression??
F. Spaghetti plot for APA events in SH-SY5Y
G. Spaghetti plot for APA events in BE(2)

### Supplementary figure 2 - features 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("2A and 2B")

get_signif_stars <- function(p_value) {
  if (p_value < 0.001) {
    return("***")
  } else if (p_value < 0.01) {
    return("**")
  } else if (p_value < 0.05) {
    return("*")
  } else {
    return("NS")  # not significant
  }
}
  
get_significance <- function(df, value, group, category_col, category) {
  test_formula <- as.formula(paste(value, "~", group))
  p_test <- kruskal.test(test_formula, data = df %>% filter(!!sym(category_col) == category))
  return(get_signif_stars(p_test$p.value))
}

cons_score_df <- master_table %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>% 
  subset(!is.na(phastCons_alt) & !is.na(phastCons_unchanged)) %>%
  pivot_longer(cols = c(phastCons_alt, phastCons_unchanged), names_to = "category", values_to = "cons_score") 

cons_score_df_cryptic <- cons_score_df %>% filter(cryptic == "yes") 
cons_score_df_not_cryptic <- cons_score_df %>% filter(cryptic == "no") 
a <- cons_score_df_cryptic %>% 
  ggplot(aes(x = category, y = cons_score, fill = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Novel splice site", "Unchanged splice site")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_alt"), 
                   get_significance(cons_score_df_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_unchanged")) 
  ) +
  #facet_wrap(facets = vars(cryptic)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

b <- cons_score_df_not_cryptic %>% 
  ggplot(aes(x = category, y = cons_score, fill = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Alternatively used splice site", "Unchanged splice site")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_alt"), 
                   get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_unchanged"))
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
a + b
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#C. Kmers enrichment analysis in  SK-N-BE(2)
print("2C")

kmer_size <- 6
yg_motifs <- c('GTGTGT', 'TGTGTG', 'CGTGTG', 'TGCGTG', 'TGTGCG', 'GTGTGC')
ya_motifs <- c('ATGTGT', 'GTATGT', 'GTGTAT', 'TGTGTA', 'TGTATG', 'TGCATG')
aa_motifs <- c('GTGTGA', 'AATGAA', 'GAATGA', 'TGAATG', 'ATGAAT', 'GTGAAT', 'GAATGT', 'TTGAAT')
all_motifs <- c(yg_motifs, ya_motifs, aa_motifs)

count_hexamers <- function(row) {
  if (row[["junc_cat"]] == "exon_skip" || is.na(row[["paste_into_igv_junction_annotated"]])) {
    return(NULL)
  }
  
  coords <- strsplit(row[["paste_into_igv_junction_annotated"]], "[:-]")[[1]]
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = coords[1], start = as.numeric(coords[2]), end = as.numeric(coords[3]))
  if (row[["strand"]] == "-") {
    seq <- reverseComplement(seq)
  }
  counts <- sapply(all_motifs, function(motif) countPattern(motif, seq, max.mismatch = 0))
  names(counts) <- all_motifs
  
  return(counts)
}
master_table_filt <- master_table %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>% filter(junc_cat != "exon_skip")
kmer_table <- apply(master_table_filt, 1, count_hexamers)
all_counts_df <- as.data.frame(t(kmer_table))

all_counts_df$type <- master_table_filt$tdp43_sensitivity_be2
all_counts_df$cryptic <- master_table_filt$cryptic

summary_counts <- all_counts_df %>%
  group_by(type, cryptic) %>%
  summarise(across(all_motifs, sum))

means <- apply(summary_counts[, -c(1,2)], 1, mean)
sds <- apply(summary_counts[, -c(1,2)], 1, sd)
z_scores <- sweep(summary_counts[, -c(1,2)], MARGIN = 1, STATS = means, FUN = "-") / sds
heatmap_data <- as.matrix(z_scores)

sort_kmers_by_zscore <- function(kmers, z_scores) {
  kmers[order(z_scores[kmers], decreasing = TRUE)]
}

sorted_yg <- sort_kmers_by_zscore(yg_motifs, heatmap_data[1,])
sorted_ya <- sort_kmers_by_zscore(ya_motifs, heatmap_data[1,])
sorted_aa <- sort_kmers_by_zscore(aa_motifs, heatmap_data[1,])

sorted_all_kmers <- c(sorted_yg, sorted_ya, sorted_aa)
categories <- c(rep("YG", 6), rep("YA", 6), rep("AA", 8))

annotation <- HeatmapAnnotation(Motif = factor(categories, levels = c("YG", "YA", "AA")),
                                show_legend = c(F, F, F),
                                col = list(Motif = c(YG = "purple", YA = "orange", AA = "darkgreen")),
                                which = "col")

rownames(heatmap_data) <- paste(summary_counts$type, ifelse(summary_counts$cryptic == "yes", "Cryptic", "Not cryptic"))
c <- Heatmap(heatmap_data[c("Early Cryptic","Early Not cryptic","Late Cryptic","Late Not cryptic") , sorted_all_kmers], 
        cluster_columns = F,
        cluster_rows = F,
        show_row_names = T,
        show_column_names = T,
        show_row_dend = F,
        bottom_annotation = annotation,
        col = colorRampPalette(brewer.pal(3, "YlGnBu"))(100),
        show_heatmap_legend = FALSE)
c
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#D. Density UG motifs by early-late in  SK-N-BE(2)
print("2D and 2E")
sequences <- sapply(seq_along(master_table$paste_into_igv_junction_annotated), function(i) {
  coord <- master_table$paste_into_igv_junction_annotated[i]
  chr <- sub("^(.*):.*$", "\\1", coord)
  range <- as.numeric(unlist(strsplit(sub(".*:(.*)$", "\\1", coord), "-")))
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = chr, start = range[1], end = range[2])

  if (master_table$strand[i] == "-") {
    seq <- reverseComplement(seq)
  }
  
  return(seq)
})

master_table$longest_ug_region <- Biostrings::vcountPattern('TGTGTG', DNAStringSet(sequences), max.mismatch = 1)

master_table <- master_table %>% 
  mutate(
    start = as.numeric(sub(".*:(\\d+)-.*", "\\1", paste_into_igv_junction_annotated)),
    end = as.numeric(sub(".*-(\\d+)", "\\1", paste_into_igv_junction_annotated)),
    intron_len = end - start,
    ug_density = longest_ug_region / intron_len
  ) %>%
  select(-start, -end)

d <- master_table %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>% 
  ggplot(aes(x = cryptic, y = ug_density, fill = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(tdp43_sensitivity_be2 %in% c("Early", "Late")),
                                    "ug_density", "tdp43_sensitivity_be2", "cryptic", "no"), 
                   get_significance(master_table %>% filter(tdp43_sensitivity_be2 %in% c("Early", "Late")),
                                    "ug_density", "tdp43_sensitivity_be2", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

#E. Density UG motifs by top-bottom in  SK-N-BE(2)
master_table$max_psi_be2 <- factor(master_table$max_psi_be2, levels=c("Top", "Bottom"))
e <- master_table %>% 
  filter(max_psi_be2 %in% c("Top", "Bottom")) %>%
  ggplot(aes(x = cryptic, y = ug_density, fill = max_psi_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) + 
  scale_fill_manual(values = plot_psi_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(max_psi_be2 %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_be2", "cryptic", "no"), 
                   get_significance(master_table %>% filter(max_psi_be2 %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_be2", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
d + e

#F. Distance iCLIP TDP43 binding site  SK-N-BE(2)
print("2F")
df_iclip <- read.csv(here::here('data', 'all_events.closest_tdp_bind.bed'), sep="\t")
df_iclip$tdp_len <- df_iclip$end - df_iclip$start + 1

f <- df_iclip %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>%
  filter(abs(dist) <= 1000) %>%
  ggplot(aes(x = dist, fill = tdp43_sensitivity_be2)) +
  geom_density(alpha = 0.5, adjust = 1) +  
  facet_wrap(~cryptic, 
             ncol = 2,
             labeller = labeller(cryptic=c("yes" = "Cryptic", "no" = "Not cryptic"))) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  labs(x = 'Distance to the nearest TDP-binding site', y = 'Density') +  
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

f

#G. RNA maps based on figure F
early_donors <- readDNAStringSet("data/maxent/donors_early_be2.fa")
late_donors <- readDNAStringSet("data/maxent/donors_late_be2.fa")

g1 <- ggplot() + 
  geom_logo(data = as.character(early_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g2 <- ggplot() + 
  geom_logo(data = as.character(late_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g <- g1 / g2
g

#H. 5'ss CE vs normal
df_5ss_strength <- read.csv(here::here('data', 'maxent/5ss_strength.csv'))
h <- df_5ss_strength %>%
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>% 
  ggplot(aes(x = X5_ss_score_cryptic, y =X5_ss_score_ann, color = tdp43_sensitivity_sh)) +
  geom_point(size = 2, alpha = 0.8) +  
  scale_color_manual(values = plot_sensitivity_colors) +
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed") + 
  coord_cartesian(xlim = c(-30, 15), ylim = c(-30, 15)) + 
  labs(x = "MaxEnt score (5'ss cryptic site)", y = "MaxEnt score (5'ss annotated site)") +
  theme_minimal() +
  theme(legend.position = "top") +
  guides(color = "none")

#I. 3'ss CE vs normal
df_3ss_strength <- read.csv(here::here('data', 'maxent/3ss_strength.csv'))
i <- df_3ss_strength %>%
  filter(tdp43_sensitivity_sh %in% c("Early", "Late")) %>% 
  ggplot(aes(x = X3_ss_score_cryptic, y =X3_ss_score_ann, color = tdp43_sensitivity_sh)) +
  geom_point(size = 2, alpha = 0.8) +  
  scale_color_manual(values = plot_sensitivity_colors) +
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed") + 
  coord_cartesian(xlim = c(-30, 15), ylim = c(-30, 15)) + 
  labs(x = "MaxEnt score (3'ss cryptic site)", y = "MaxEnt score (3'ss annotated site)") +
  theme_minimal() +
  theme(legend.position = "top") +
  guides(color = "none")
h + i
```

A. Conservation score for CE in SK-N-BE(2)

B. Conservation score for non-CE in SK-N-BE(2)

C. Kmers enrichment analysis in SK-N-BE(2)

D. Density UG motifs by early-late in SK-N-BE(2)
<<<<<<< Updated upstream
E. Density UG motifs by top-bottom in  SK-N-BE(2)
=======
<<<<<<< HEAD

E. Density UG motifs by top-bottom in SH-SY5Y

=======
E. Density UG motifs by top-bottom in  SK-N-BE(2)
>>>>>>> 0b9c0e2228595ac6a483dcb2fe50b70533dbc2c3
>>>>>>> Stashed changes
F. Distance iCLIP TDP43 binding site SK-N-BE(2)

G. RNA maps based on figure F

H. 5'ss CE vs normal

I. 3'ss CE vs normal

### Supplementary figure 3 - NMD
```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("S3A")
gene_list_slope_upf1 <- c("STMN2", "UNC13A", "AARS1", "HDGFL2", "CYFIP2", "SYNE1")
input_list <- c("ctrl_ctrl",
                "ctrl_UPF1", 
                "TDP43_ctrl", 
                "TDP43_UPF1")
big_data_upf1 <- read.csv(here::here('data','nmd_or_not_upf1.csv'))
a <- slope_plot_nmd(big_data_upf1, 
               gene_list_slope_upf1)

print("S3B")

```

A. Slope plot UPF1

B. Correlation CHX UPF1 - groups of rescued / non rescued

C. TDP-43 inhibition doesn't impair NMD (top 50 UPF1)

D. TDP-43 inhibition doesn't impair NMD (known poison exons)
