---
title: "supplementary_fig"
author: "MZ"
date: "2024-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
###load packages/libraries/scripts
library(data.table)
library(tidyverse)
library(ggrepel)
library(patchwork)
library(ggpubr)
library(ggstats)
library(ggsignif)
library(ggseqlogo)
library(Biostrings)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ComplexHeatmap)
library(RColorBrewer)

source(here::here('trial_nmd_plots.R'))
#check you're in right folder
here::here()

plot_sensitivity_colors <- c("Early" = "#de53b0", "Late" = "#1331a7", "Intermediate" = "steelblue")
plot_psi_colors <- c("Top" = "#388f78", "Bottom" = "#de9e4a", "Intermediate" = "steelblue")
plot_text_size <- 10
```

### Supplementary figure 1 - sensitivity of CEs to TDP43 LoF is driven by their cis features

A. levels from DeSeq in SH and SK
B. Concordance between SH and SK (scatterplot) - to do!!
C. UMAP clustering of events in SH and BE2 together
D. UMAP clustering of events in SH and in BE2 - separated
E. differences in differential splicing are not due to changes in gene expression??

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### load data needed
#S1A
normal_counts_sh <- read.csv(here::here('data', 'normed_counts_sh.csv'))
normal_counts_sk <- read.csv(here::here('data', 'normed_counts_sk.csv'))

#S1?
master_table <- read.csv(here::here("data/master_table.csv"))

#S1F #S1G
#apa_sh <- read.csv(here::here())
#apa_sk <- read.csv(here::here())

print("S1A - levels from DeSeq in SH and SK")
b <- normal_counts_sh[,c(2:19,21)] %>% 
  dplyr::filter(symbol == "TARDBP") %>%
  pivot_longer(c(1:18)) %>%
  mutate(group = gsub("^[^_]+_|_[^_]+$", "", name)) %>%
  ggplot(aes(x = group, y = value)) +
  geom_boxplot(alpha = 0.1, show.legend = F) +
  geom_point(position = position_dodge2(width = 1), show.legend = F) +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 Counts per million') +
  theme_classic() +
  theme(text = element_text(size = 10))
d <- normal_counts_sk[,c(2:25,27)] %>% 
  dplyr::filter(symbol == "TARDBP") %>%
  pivot_longer(c(1:24)) %>%
  mutate(group = gsub("^[^_]+_|_[^_]+$", "", name)) %>%
  ggplot(aes(x = group, y = value)) +
  geom_boxplot(alpha = 0.1, show.legend = F) +
  geom_point(position = position_dodge2(width = 1), show.legend = F) +
  xlab(expression(paste('Doxycycline concentration (', mu,'g/mL)'))) +
  ylab('TDP-43 Counts per million') +
  scale_x_discrete(labels = c("0", "0.02", "0.03", "0.04", "0.05", "0.075", "0.1", "1")) +
  theme_classic() +
  theme(text = element_text(size = 10))
b + d

print("S1B - Concordance between SH and SK scatterplot")
#scatterplot from upgrade
master_table %>% 
    ggplot(aes(x = dox075, y = dox1)) +
    geom_point() +
    geom_smooth(method = 'lm', se = F) +
    stat_cor() +
    xlab("Max PSI in SH-SY5Y") +
    ylab("Max PSI in SK-N-BE(2)") +
    theme_classic()

print("S1C")

print("S1D")

print("S1C - UMAP clustering of events in SH and BE2 together")
master_table %>% 
    filter(tdp43_sensitivity_sh != "") %>% 
    ggplot(aes(x = UMAP_1_together, y = UMAP_2_together, color = as.character(cluster_together))) +
    geom_point(show.legend = F) +
    xlab("UMAP 1 together") +
    ylab("UMAP 2 together") +
    labs(color = "cluster") +
    scale_color_brewer(palette = "Set1") +
    theme_bw()


print("S1D - UMAP clustering of events in SH and in BE2 separated")

g <- master_table %>% 
    filter(tdp43_sensitivity_be2 != "") %>% 
    ggplot(aes(x = UMAP_1_sh, y = UMAP_2_sh, color = as.character(tdp43_sensitivity_sh))) +
    geom_point(show.legend = F) +
    xlab("UMAP 1") +
    ylab("UMAP 2") +
    labs(color = "TDP43 sensitivity") +
    scale_color_manual(values = plot_sensitivity_colors) +
    theme_bw()
h <- master_table %>% 
    filter(tdp43_sensitivity_be2 != "") %>% 
    ggplot(aes(x = UMAP_1_be2, y = UMAP_2_be2, color = as.character(tdp43_sensitivity_be2))) +
    geom_point() +
    xlab("UMAP 1") +
    ylab("UMAP 2") +
    labs(color = "TDP43 sensitivity") +
    scale_color_manual(values = plot_sensitivity_colors) +
    theme_bw()
g+h

print("S1E - differences in differential splicing are not due to changes in gene expression")

print("S1F")
#apa_sh
print("S1G")
#apa_sk
```

#### F. Spaghetti plot for APA events in SH-SY5Y
```{r,echo=FALSE,warning=FALSE,message=FALSE}
shsy5y_papa = fread("data/shsy5y_papa_curve.csv")

shsy5y_papa |> 
    select(le_id,mean_ppau_contrast,experiment,gene_name,timing) |> 
    complete(le_id,experiment,fill = list(mean_ppau_contrast = 0))  |> 
    dplyr::group_by(le_id) %>%
    fill(timing, .direction = "downup") %>%
    ungroup() |> 
    mutate(gene_name = ifelse(experiment == 'zanovello_shsy5y_curve_0075',
                              gene_name,NA_character_)) %>% 
    ggplot(aes(x = experiment,
               y = mean_ppau_contrast,
               group = le_id,color = timing)) + 
    geom_point() + 
    geom_line() + 
    geom_text_repel(aes(label = gene_name),show.legend = FALSE) +
    theme_bw() + 
    geom_hline(yintercept = 0.1,linetype = 'dotted') + 
    scale_color_manual(values = c("#E376BF","#4A6EDD")) +
    scale_x_discrete(labels=c("+", "++", "+++","++++","+++++")) 
    
    
```

#### G. Spaghetti plot for APA events in BE(2)

```{r}
shsy5y_papa = fread("data/sknbe2_papa_curve.csv")

    select(le_id,mean_ppau_contrast,experiment,gene_name,timing) |> 
    complete(le_id,experiment,fill = list(mean_ppau_contrast = 0))  |> 
    dplyr::group_by(le_id) %>%
    fill(timing, .direction = "downup") %>%
    ungroup() |> 
    mutate(gene_name = ifelse(experiment == 'zanovello_skndz_curve_1',
                              gene_name,NA_character_)) |> 
    ggplot(aes(x = experiment,
               y = mean_ppau_contrast,
               group = le_id,color = timing)) + 
    geom_point() + 
    geom_line() + 
    geom_text_repel(aes(label = gene_name),show.legend = FALSE) +
    theme_bw() + 
    geom_hline(yintercept = 0.1,linetype = 'dotted') + 
    scale_color_manual(values = c("#E376BF","#4A6EDD")) +
    scale_x_discrete(labels=c("+", "++", "+++","++++","+++++","++++++","++++++")) +
    ylab("% polyA usage") + 
    scale_y_continuous(labels = scales::percent_format())

```



### Supplementary figure 2 - features 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("2A and 2B")

get_signif_stars <- function(p_value) {
  if (p_value < 0.001) {
    return("***")
  } else if (p_value < 0.01) {
    return("**")
  } else if (p_value < 0.05) {
    return("*")
  } else {
    return("NS")  # not significant
  }
}
  
get_significance <- function(df, value, group, category_col, category) {
  test_formula <- as.formula(paste(value, "~", group))
  p_test <- kruskal.test(test_formula, data = df %>% filter(!!sym(category_col) == category))
  return(get_signif_stars(p_test$p.value))
}

cons_score_df <- master_table %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>% 
  subset(!is.na(phastCons_alt) & !is.na(phastCons_unchanged)) %>%
  pivot_longer(cols = c(phastCons_alt, phastCons_unchanged), names_to = "category", values_to = "cons_score") 

cons_score_df_cryptic <- cons_score_df %>% filter(cryptic == "yes") 
cons_score_df_not_cryptic <- cons_score_df %>% filter(cryptic == "no") 
a <- cons_score_df_cryptic %>% 
  ggplot(aes(x = category, y = cons_score, fill = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Novel splice site", "Unchanged splice site")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_alt"), 
                   get_significance(cons_score_df_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_unchanged")) 
  ) +
  #facet_wrap(facets = vars(cryptic)) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

b <- cons_score_df_not_cryptic %>% 
  ggplot(aes(x = category, y = cons_score, fill = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('PhastCons score') +
  scale_x_discrete(labels = c("Alternatively used splice site", "Unchanged splice site")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(1.1, 1.1), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2), tip_length = 0.02,
    annotation = c(get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_alt"), 
                   get_significance(cons_score_df_not_cryptic,
                                    "cons_score", "tdp43_sensitivity_be2", "category", "phastCons_unchanged"))
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
a + b
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#C. Kmers enrichment analysis in  SK-N-BE(2)
print("2C")

kmer_size <- 6
yg_motifs <- c('TGTGTG', 'GTGTGT', 'GTGTGC', 'CGTGTG', 'TGCGTG', 'TGTGCG')
ya_motifs <- c('ATGTGT', 'TGTGTA', 'TGCATG', 'GTGTAT', 'TGTATG', 'GTATGT')
aa_motifs <- c('AATGAA', 'TTGAAT', 'ATGAAT', 'GTGTGA', 'TGAATG', 'GAATGA', 'GTGAAT', 'GAATGT')
all_motifs <- c(yg_motifs, ya_motifs, aa_motifs)
categories <- c(rep("YG", 6), rep("YA", 6), rep("AA", 8))

count_hexamers <- function(row) {
  if (row[["junc_cat"]] == "exon_skip" || is.na(row[["paste_into_igv_junction_annotated"]])) {
    return(NULL)
  }
  
  coords <- strsplit(row[["paste_into_igv_junction_annotated"]], "[:-]")[[1]]
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = coords[1], start = as.numeric(coords[2]), end = as.numeric(coords[3]))
  if (row[["strand"]] == "-") {
    seq <- reverseComplement(seq)
  }
  counts <- sapply(all_motifs, function(motif) countPattern(motif, seq, max.mismatch = 0))
  names(counts) <- all_motifs
  
  return(counts)
}

master_table_filt <- master_table %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>% 
  filter(junc_cat != "exon_skip") %>%
  distinct(paste_into_igv_junction_annotated, .keep_all = T)
kmer_table <- apply(master_table_filt, 1, count_hexamers)
all_counts_df <- as.data.frame(t(kmer_table))

all_counts_df$type <- master_table_filt$tdp43_sensitivity_be2
all_counts_df$cryptic <- master_table_filt$cryptic

summary_counts <- all_counts_df %>%
  group_by(type, cryptic) %>%
  summarise(across(all_motifs, sum))

means <- apply(summary_counts[, -c(1,2)], 1, mean)
sds <- apply(summary_counts[, -c(1,2)], 1, sd)
z_scores <- sweep(summary_counts[, -c(1,2)], MARGIN = 1, STATS = means, FUN = "-") / sds
heatmap_data <- as.matrix(z_scores)
rownames(heatmap_data) <- paste(summary_counts$type, ifelse(summary_counts$cryptic == "yes", "Cryptic", "Not cryptic"))

sort_kmers_by_zscore <- function(kmers, z_scores) {
  kmers[order(z_scores[kmers], decreasing = TRUE)]
}

sort_categories_by_zscore <- function(kmers, z_scores) {
  categories[order(z_scores[kmers], decreasing = TRUE)]
}

sorted_all_kmers <- sort_kmers_by_zscore(all_motifs, heatmap_data["Early Cryptic",]+heatmap_data["Late Cryptic",])
sorted_categories <- sort_categories_by_zscore(all_motifs, heatmap_data["Early Cryptic",]+heatmap_data["Late Cryptic",])

annotation <- HeatmapAnnotation(Motif = factor(sorted_categories, levels = c("YG", "YA", "AA")),
                                show_legend = c(F, F, F),
                                col = list(Motif = c(YG = "palegreen3", YA = "turquoise3", AA = "tan1")),
                                which = "col")

c <- Heatmap(heatmap_data[c("Early Cryptic","Early Not cryptic","Late Cryptic","Late Not cryptic") , sorted_all_kmers],
        cluster_columns = F,
        cluster_rows = F,
        show_row_names = T,
        show_column_names = T,
        show_row_dend = F,
        bottom_annotation = annotation,
        col = colorRampPalette(brewer.pal(12, "RdPu"))(100),
        show_heatmap_legend = FALSE)
c
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#D. Density UG motifs by early-late in  SK-N-BE(2)
print("2D and 2E")
sequences <- sapply(seq_along(master_table$paste_into_igv_junction_annotated), function(i) {
  coord <- master_table$paste_into_igv_junction_annotated[i]
  chr <- sub("^(.*):.*$", "\\1", coord)
  range <- as.numeric(unlist(strsplit(sub(".*:(.*)$", "\\1", coord), "-")))
  seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, names = chr, start = range[1], end = range[2])

  if (master_table$strand[i] == "-") {
    seq <- reverseComplement(seq)
  }
  
  return(seq)
})

master_table$longest_ug_region <- Biostrings::vcountPattern('TGTGTG', DNAStringSet(sequences), max.mismatch = 1)

master_table <- master_table %>% 
  mutate(
    start = as.numeric(sub(".*:(\\d+)-.*", "\\1", paste_into_igv_junction_annotated)),
    end = as.numeric(sub(".*-(\\d+)", "\\1", paste_into_igv_junction_annotated)),
    intron_len = end - start,
    ug_density = longest_ug_region / intron_len
  ) %>%
  select(-start, -end)

d <- master_table %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>% 
  ggplot(aes(x = cryptic, y = ug_density, fill = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) + 
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(tdp43_sensitivity_be2 %in% c("Early", "Late")),
                                    "ug_density", "tdp43_sensitivity_be2", "cryptic", "no"), 
                   get_significance(master_table %>% filter(tdp43_sensitivity_be2 %in% c("Early", "Late")),
                                    "ug_density", "tdp43_sensitivity_be2", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

#E. Density UG motifs by top-bottom in  SK-N-BE(2)
master_table$max_psi_be2 <- factor(master_table$max_psi_be2, levels=c("Top", "Bottom"))
e <- master_table %>% 
  filter(max_psi_be2 %in% c("Top", "Bottom")) %>%
  ggplot(aes(x = cryptic, y = ug_density, fill = max_psi_be2)) +
  geom_boxplot() +
  xlab("") +
  ylab('UG-repeat density') +
  scale_x_discrete(labels = c("Not cryptic", "Cryptic")) + 
  scale_fill_manual(values = plot_psi_colors) +
  guides(fill = "none") +
  geom_signif(
    y_position = c(max(master_table$ug_density) * 1.05,max(master_table$ug_density) * 1.05), xmin = c(0.8, 1.8), xmax = c(1.2, 2.2),
    annotation = c(get_significance(master_table %>% filter(max_psi_be2 %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_be2", "cryptic", "no"), 
                   get_significance(master_table %>% filter(max_psi_be2 %in% c("Top", "Bottom")),
                                    "ug_density", "max_psi_be2", "cryptic", "yes")), tip_length = 0.02
  ) +
  theme_classic() +
  theme(text = element_text(size = plot_text_size))
d + e

#F. Distance iCLIP TDP43 binding site  SK-N-BE(2)
print("2F")
df_iclip <- read.csv(here::here('data', 'all_events.closest_tdp_bind.bed'), sep="\t")
df_iclip$tdp_len <- df_iclip$end - df_iclip$start + 1

f <- df_iclip %>% 
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late")) %>%
  filter(abs(dist) <= 1000) %>%
  ggplot(aes(x = dist, fill = tdp43_sensitivity_be2)) +
  geom_density(alpha = 0.5, adjust = 1) +  
  facet_wrap(~cryptic, 
             ncol = 2,
             labeller = labeller(cryptic=c("yes" = "Cryptic", "no" = "Not cryptic"))) +
  scale_fill_manual(values = plot_sensitivity_colors) +
  guides(fill = "none") +
  labs(x = 'Distance to the nearest TDP-binding site', y = 'Density') +  
  theme_classic() +
  theme(text = element_text(size = plot_text_size))

f

#G. RNA maps based on figure F
early_donors <- readDNAStringSet("data/maxent/donors_early_be2.fa")
late_donors <- readDNAStringSet("data/maxent/donors_late_be2.fa")

g1 <- ggplot() + 
  geom_logo(data = as.character(early_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g2 <- ggplot() + 
  geom_logo(data = as.character(late_donors), method = "probability")+
  scale_fill_manual(values = c("A" = "black", "C" = "black", "G" = "palegreen2", "T" = "palegreen2")) +
  labs( x = "Position", y = "Probability") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")  # Adjust plot margins
  )
g <- g1 / g2
g

#H. MaxEnt CE vs normal SH-SY5Y
df_5ss_strength <- read.csv(here::here('data', 'maxent/5ss_strength_sh.csv'))
df_3ss_strength <- read.csv(here::here('data', 'maxent/3ss_strength_sh.csv'))
df_ss_strength <- rbind(
  df_5ss_strength  %>%
    mutate(site = "5' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, tdp43_sensitivity_sh, site, ID_exon),
  df_3ss_strength  %>%
    mutate(site = "3' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, tdp43_sensitivity_sh, site, ID_exon)
) %>%
  filter(tdp43_sensitivity_sh %in% c("Early", "Late"))
h <- df_ss_strength %>% 
  ggplot(aes(x=site, y = ss_ratio, color = tdp43_sensitivity_sh)) +
  geom_boxplot() +
  geom_abline(intercept = 1, slope = 0, color = "gray", linetype = "dashed") + 
  scale_color_manual(values = plot_sensitivity_colors) +
  ylim(-4.1, 2.5) +
  labs(y = "MaxEnt score cryptic site vs annotated site", x = "") +
  theme_minimal() +
  theme(legend.position = "top") +
  guides(color = "none")

#I. MaxEnt CE vs normal SK-N-BE(2)
df_5ss_strength <- read.csv(here::here('data', 'maxent/5ss_strength_be2.csv'))
df_3ss_strength <- read.csv(here::here('data', 'maxent/3ss_strength_be2.csv'))
df_ss_strength <- rbind(
  df_5ss_strength  %>%
    mutate(site = "5' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, tdp43_sensitivity_be2, site, ID_exon),
  df_3ss_strength  %>%
    mutate(site = "3' splice site", ss_ratio = ss_score_cryptic/ss_score_ann) %>%
    select(ss_score_ann, ss_score_cryptic, ss_ratio, tdp43_sensitivity_be2, site, ID_exon)
) %>%
  filter(tdp43_sensitivity_be2 %in% c("Early", "Late"))
i <- df_ss_strength %>% 
  ggplot(aes(x=site, y = ss_ratio, color = tdp43_sensitivity_be2)) +
  geom_boxplot() +
  geom_abline(intercept = 1, slope = 0, color = "gray", linetype = "dashed") + 
  scale_color_manual(values = plot_sensitivity_colors) +
  ylim(-4.1, 2.5) +
  labs(y = "MaxEnt score cryptic site vs annotated site", x = "") +
  theme_minimal() +
  theme(legend.position = "top") +
  guides(color = "none")
h + i
```

A. Conservation score for CE in SK-N-BE(2)

B. Conservation score for non-CE in SK-N-BE(2)

C. Kmers enrichment analysis in SK-N-BE(2)

D. Density UG motifs by early-late in SK-N-BE(2)

E. Density UG motifs by top-bottom in  SK-N-BE(2)

F. Distance iCLIP TDP43 binding site SK-N-BE(2)

G. RNA maps based on figure F

H. MaxEnt CE vs normal SH-SY5Y

I. MaxEnt CE vs normal SK-N-BE(2)

### Supplementary figure 3 - NMD

A. Slope plot UPF1


```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("S3A")
gene_list_slope_upf1 <- c("STMN2", "UNC13A", "AARS1", "HDGFL2", "CYFIP2", "SYNE1")
input_list <- c("ctrl_ctrl",
                "ctrl_UPF1", 
                "TDP43_ctrl", 
                "TDP43_UPF1")
big_data_upf1 <- read.csv(here::here('data','nmd_or_not_upf1.csv'))
a <- slope_plot_nmd(big_data_upf1, 
               gene_list_slope_upf1)

print("S3B")

```


B. late/early UPF1

```{r LateEarlyNMD, echo=FALSE, warning=FALSE, message=FALSE}
# reading in the master table again for debug, can remove later
master_table <- read.csv(here::here("data/master_table.csv"))


sh_nmd_sens_upf1 = big_data_upf1 %>% 
    distinct(paste_into_igv_junction,color_gene_name) %>% 
    left_join(master_table) %>% 
    filter(!is.na(tdp43_sensitivity_sh)) %>% 
    filter(tdp43_sensitivity_sh != "") %>% 
    group_by(color_gene_name,tdp43_sensitivity_sh) %>% 
    reframe(n = n_distinct(paste_into_igv_junction)) 

ggplot(sh_nmd_sens_upf1) +
    aes(x = tdp43_sensitivity_sh, fill = color_gene_name, weight = n,by = as.factor(tdp43_sensitivity_sh)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'UPF1 KD') +
    scale_fill_manual(values = c("#3288BD","#D53E4F")) + 
    theme(legend.position = 'bottom')

sh_nmd_sens_upf1 %>% 
    pivot_wider(names_from = 'color_gene_name',values_from = 'n') %>% 
    column_to_rownames("tdp43_sensitivity_sh") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()

sk_nmd_sens_upf1 = big_data_upf1 %>% 
    distinct(paste_into_igv_junction,color_gene_name) %>% 
    left_join(master_table) %>% 
    filter(!is.na(tdp43_sensitivity_be2)) %>% 
    filter(tdp43_sensitivity_be2 != "") %>% 
    group_by(color_gene_name,tdp43_sensitivity_be2) %>% 
    reframe(n = n_distinct(paste_into_igv_junction)) 

sk_nmd_sens_upf1 = big_data_upf1 %>% 
    distinct(paste_into_igv_junction,color_gene_name) %>% 
    left_join(master_table) %>% 
    filter(!is.na(new_cate)) %>% 
    filter(new_cate != "") %>% 
    group_by(color_gene_name,new_cate) %>% 
    reframe(n = n_distinct(paste_into_igv_junction)) 

ggplot(sk_nmd_sens_upf1) +
    aes(x = new_cate, fill = color_gene_name, weight = n,by = as.factor(new_cate)) +
    geom_bar() +
    geom_text(
        aes(label = scales::percent(after_stat(prop), accuracy = 1)),
        stat = "prop",
        position = position_stack(.5)
    ) +
    theme_classic() + 
    labs(y = 'N junctions',
         x = element_blank(),
         fill = 'UPF1 KD ') +
    scale_fill_manual(values = c("#3288BD","#D53E4F")) + 
    theme(legend.position = 'bottom')

sk_nmd_sens_upf1 %>% 
    pivot_wider(names_from = 'color_gene_name',values_from = 'n') %>% 
    column_to_rownames("new_cate") %>% 
    as.matrix() %>% 
    rstatix::pairwise_fisher_test()
```

C. Correlation CHX UPF1 - groups of rescued / non rescued

D. TDP-43 inhibition doesn't impair NMD (top 50 UPF1)

E. TDP-43 inhibition doesn't impair NMD (known poison exons)

### Supplementary figure 4 - in vivo

### Supplementary figure 4 - early late is consistent across compendium

```{r earlyLateCompendiumSH, echo=FALSE, warning=FALSE, message=FALSE}
compendium_events = fread('data/compendium_cryptics.csv')
master_table %>% 
    left_join(compendium_events %>% select(paste_into_igv_junction,n_significant_data)) %>% 
    unique() %>% 
    select(tdp43_sensitivity_sh,n_significant_data,paste_into_igv_junction,new_cate) %>% 
    mutate(n_significant_data = ifelse(is.na(n_significant_data),0,n_significant_data)) %>% 
    # filter(tdp43_sensitivity_sh != "") %>% 
    ggplot(aes(new_cate, y = n_significant_data)) + 
    geom_boxplot() + 
    labs(y = 'N datasets cryptic detected',
         x = element_blank()) + 
    theme_classic() + 
    stat_compare_means()

compendium_events %>% 
    left_join(master_table,by = c("paste_into_igv_junction")) %>% 
    unique()  %>% 
    filter(!is.na(new_cate)) %>% 
    mutate(new_cate = case_when(tdp43_sensitivity_be2 == "" ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_sh == "" ~ tdp43_sensitivity_be2,
                            tdp43_sensitivity_sh == tdp43_sensitivity_be2 ~ tdp43_sensitivity_sh,
                            tdp43_sensitivity_be2 == 'Early' | tdp43_sensitivity_sh == 'Early' ~ 'Early',
                            tdp43_sensitivity_be2 == 'Late' | tdp43_sensitivity_sh == 'Late' ~ 'Late'))  %>% 
    mutate(n_significant_data = ifelse(is.na(n_significant_data),0,n_significant_data)) %>% 
        select(tdp43_sensitivity_sh,n_significant_data,paste_into_igv_junction,new_cate)  %>% 
    # filter(tdp43_sensitivity_sh != "") %>% 
    ggplot(aes(new_cate, y = n_significant_data)) + 
    geom_boxplot() + 
    labs(y = 'N datasets cryptic detected',
         x = element_blank()) + 
    theme_classic() + 
    stat_compare_means()
```

```{r earlyLateCompendiumSK, echo=FALSE, warning=FALSE, message=FALSE}
master_table %>% 
    left_join(compendium_events %>% select(paste_into_igv_junction,n_significant_data)) %>% 
    unique() %>% 
    select(tdp43_sensitivity_be2,n_significant_data,paste_into_igv_junction) %>% 
    mutate(n_significant_data = ifelse(is.na(n_significant_data),0,n_significant_data)) %>% 
    filter(tdp43_sensitivity_be2 != "") %>% 
    ggplot(aes(tdp43_sensitivity_be2, y = n_significant_data)) + 
    geom_boxplot() + 
    labs(y = 'N datasets cryptic detected',
         x = element_blank()) + 
    theme_classic() + 
    stat_compare_means()
```


```{r NMDcompendiumCHX, echo=FALSE, warning=FALSE, message=FALSE}
compendium_events = fread('data/compendium_cryptics.csv')
big_data_chx <- read.csv(here::here('data','nmd_or_not.csv'))

compendium_events %>% 
    left_join(big_data_chx %>% distinct(paste_into_igv_junction,color_gene_name)) %>% 
    mutate(color_gene_name = ifelse(color_gene_name == 'non-NMD rescued', "NMD insensitive","NMD sensitive")) %>% 
    filter(!is.na(color_gene_name)) %>% 
    distinct(paste_into_igv_junction,color_gene_name,n_significant_data) %>% 
    ggplot(aes(color = color_gene_name, x = n_significant_data)) + 
    stat_ecdf()  +
    labs(x = 'N datasets cryptic detected') + 
    theme_classic()

compendium_events %>% 
    left_join(big_data_chx %>% distinct(paste_into_igv_junction,color_gene_name)) %>%
    filter(!is.na(color_gene_name)) %>%
    mutate(color_gene_name = ifelse(color_gene_name == 'non-NMD rescued', "NMD insensitive","NMD sensitive")) %>% 
    distinct(paste_into_igv_junction,color_gene_name,n_significant_data,gene_name) %>%
    mutate(plot_name = ifelse(color_gene_name == "NMD sensitive" & n_significant_data > 5,gene_name,NA_character_)) %>% 
    ggplot(aes(color_gene_name, y = n_significant_data)) +
    geom_boxplot() +
    labs(y = 'N datasets cryptic detected',
         x = element_blank()) +
    theme_classic() +
    stat_compare_means() + 
    geom_text_repel(aes(label = plot_name))


```

```{r NMDcompendiumUPF1, echo=FALSE, warning=FALSE, message=FALSE}
compendium_events = fread('data/compendium_cryptics.csv')
big_data_upf1 <- read.csv(here::here('data','nmd_or_not_upf1.csv'))


compendium_events %>% 
    left_join(big_data_upf1 %>% distinct(paste_into_igv_junction,color_gene_name)) %>% 
    filter(!is.na(color_gene_name)) %>% 
    distinct(paste_into_igv_junction,color_gene_name,n_significant_data) %>% 
    ggplot(aes(color = color_gene_name, x = n_significant_data)) + 
    stat_ecdf()  +
    labs(x = 'N datasets cryptic detected') + 
    theme_classic()

compendium_events %>% 
    left_join(big_data_upf1 %>% distinct(paste_into_igv_junction,color_gene_name)) %>%
    filter(!is.na(color_gene_name)) %>%
    mutate(color_gene_name = ifelse(color_gene_name == 'non-NMD rescued', "NMD insensitive","NMD sensitive")) %>% 
    mutate(plot_name = ifelse(color_gene_name == "NMD sensitive" & n_significant_data > 5,gene_name,NA_character_)) %>% 
    distinct(paste_into_igv_junction,color_gene_name,n_significant_data,plot_name) %>%
    ggplot(aes(color_gene_name, y = n_significant_data)) +
    geom_boxplot() +
    labs(y = 'N datasets cryptic detected',
         x = element_blank()) +
    theme_classic() +
    stat_compare_means()+
    geom_text_repel(aes(label = plot_name))


```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
print("S4A")

print("S4B")

print("S4C")
maxc_type %>%
  filter(group != "H2O") %>%
    filter(tdp43_sensitivity_sh != "") %>% 
  ggplot(aes(x = sample_name_ari, y = gene_name, fill = fraction_cryptic)) +
  geom_tile() +
  #scale_fill_viridis_d() +
  scale_fill_gradient(low = "white", high = "black") +
  facet_grid(rows = vars(tdp43_sensitivity_sh), cols = vars(group), 
             scales = "free", space = "free") +
  ylab("") + xlab("Sample name") + labs(fill = "CE fraction") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))

```

